### 2021-04-15 tidymodels_analysis.Rmd
r markdown report of results

---
output: html_document
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, comment = "")

library(tidyverse)
library(nortest)
library(FSA)
library(tidymodels)
library(factoextra)
library(plotly)
library(usemodels)
library(ranger)
library(vip)

```

## Dysregulation of hemoglobin and oxygenation in cerebral malaria {.tabset}

```{r data}

# read in data
df_hbox <- read.csv("Data/98 first visit records with cerebral oxygenation stats_trimmed.csv") %>% as_tibble()

# get rid of "Other"
df_hbox <- df_hbox %>% 
  filter(Status != "Other") %>% 
  mutate(Status = factor(Status, levels = c("HV", "UM", "CM")))

# identify duplicated rows (TM0003, TM2001)
df_hbox <- df_hbox %>% filter(duplicated(Subject.ID..NIAID.) == FALSE)

# Convert age to numeric
df_hbox <- df_hbox %>% 
  mutate(Age = gsub(" Years, ", "_", Age)) %>% 
  mutate(Age = gsub(" Months ", "", Age)) %>% 
  separate(Age, into = c("Years", "Months"), sep = "_") %>% 
  mutate(Years = as.numeric(Years)) %>% 
  mutate(Months = as.numeric(Months)/12) %>% 
  mutate(Age = Years + Months) %>% 
  select(-c(Years, Months))

# Convert sex to factor
df_hbox <- df_hbox %>% mutate(Sex = as.factor(Sex))

# Trim for variables of interest (VOI)
df_hbox_trim <- df_hbox %>% 
  dplyr::select(c(Subject.ID..NIAID., Status, Glucose, Hematocrit, Lactate, Temperature,
                  Arginine.umol.L, Haptoglobin..mg.dl., Hemoglobin..uM., Whole.Blood.Nitrite, 
                  cerebral.hbdeox, cerebral.hbdeox.alpha, cerebral.hbox, cerebral.hbox.alpha,
                  cerebral.sat, cerebral.sat.alpha, cerebral.thc, cerebral.thc.alpha, cerebral.thc.norm.std)) %>%

  dplyr::rename("subject_id" = "Subject.ID..NIAID.", "Arginine (umol/L)" = "Arginine.umol.L", "Haptoglobin (mg/dL)" = "Haptoglobin..mg.dl.", 
                "Hemoglobin (uM)" = "Hemoglobin..uM.", "WB_Nitrite" = "Whole.Blood.Nitrite", "DeoxyHb" = "cerebral.hbdeox", 
                "DeoxyHb_alpha" = "cerebral.hbdeox.alpha", "OxyHb" = "cerebral.hbox", "OxyHb_alpha" = "cerebral.hbox.alpha", 
                "O2sat" = "cerebral.sat", "O2sat_alpha" = "cerebral.sat.alpha", "THC" = "cerebral.thc", 
                "THC_alpha" = "cerebral.thc.alpha", "THC_sd" = "cerebral.thc.norm.std")

# impute missing data with median by group
f_impute <- function(x, na.rm = TRUE) (replace(x, is.na(x), median(x, na.rm = na.rm)))

df_hbox_impute <- df_hbox_trim %>% 
  group_by(Status) %>% 
  mutate_at(3:ncol(.), f_impute) %>% 
  ungroup()

# pivot long
df_hbox_long <- df_hbox_trim %>% pivot_longer(cols = 3:ncol(.), names_to = "measurement", values_to = "value")

```

### Exploratory analysis

#### Histogram to explore normality of data
```{r plot1, fig.width = 15, fig.height = 8}

df_hbox_long %>%

  ggplot(aes(x = value)) +
  geom_histogram(bins = 20) +
  facet_grid(Status ~ measurement, scales = "free") +

  theme(strip.text.y = element_text(size = 15),
        strip.text.x = element_text(size = 11),
        axis.title.x = element_blank(),
        axis.title.y = element_text(size = 15),
        axis.text = element_text(size = 12),
        axis.text.x = element_text(angle = 45, hjust = 0.9))

```

#### Plots of all the variables by status
```{r plot2, fig.width = 15, fig.height = 15}

df_hbox_long %>%

  ggplot(aes(x = Status, y = value)) +
  geom_violin(aes(color = Status)) +
  geom_jitter(position = position_jitter(0.2), shape = 1) +
  stat_summary(fun = "median", geom = "crossbar", aes(color = Status), size = 0.2, width = 0.5) +
  facet_wrap(~measurement, scales = "free_y") +

  theme_bw() +
  theme(legend.position = "none")  +

  theme(strip.text = element_text(size = 15),
        axis.title = element_text(size = 15),
        axis.title.y = element_blank(),
        axis.text = element_text(size = 15))

```

### PCA

#### Can we cluster patients based on available variables?

```{r pca_data}

# convert tibble to df
df_hbox_impute_df <- df_hbox_impute %>% select(-c(subject_id, Status)) %>% as.data.frame(df_hbox_impute) 
rownames(df_hbox_impute_df) <- df_hbox_impute$subject_id

# compute PCA
my_pca <- prcomp(df_hbox_impute_df, scale = TRUE)

# define groups
groups <- as.factor(df_hbox_impute$Status)

# Eigenvalues
pca_eig_val <- get_eigenvalue(my_pca)

# Results for Variables
pca_res_var <- get_pca_var(my_pca)

# Results for individuals
pca_res_ind <- get_pca_ind(my_pca)

```

#### 2D PCA
```{r pca_2d, fig.width = 10, fig.height = 10}

p_pca <- fviz_pca_ind(my_pca,
                      col.ind = groups, # color by groups
                      palette = c("#00AFBB",  "#FC4E07", "#7CAE00"),
                      addEllipses = TRUE, # Concentration ellipses
                      ellipse.type = "confidence",
                      legend.title = "Groups",
                      repel = TRUE
)

p_pca$layers[[4]] <- NULL # get rid of text labels

p_pca

```

#### Biplot
```{r pca_biplot, fig.width = 12, fig.height = 12}

# Biplot of individuals and variables
p_biplot <- fviz_pca_biplot(my_pca, repel = TRUE,
                            col.var = "black", # Variables color
                            col.ind = groups  # Individuals color
)

p_biplot$layers[[3]] <- NULL
p_biplot$layers[[2]] <- NULL

p_biplot

```

#### Determine biggest contributors to each PC
```{r pca_contrib}

df_pca_contrib <- pca_res_var$contrib %>% 
  as_tibble(rownames = "variable") %>% 
  arrange(-Dim.1, -Dim.2)

df_contrib <- tibble(Dim.1 = paste0(df_pca_contrib$variable, 
                                    sep = ": ", 
                                    round(df_pca_contrib$Dim.1, 2)),
                     Dim.2 = paste0(arrange(df_pca_contrib, -Dim.2)$variable, 
                                    sep = ": ", 
                                    round(arrange(df_pca_contrib, -Dim.2)$Dim.2, 2)),
                     Dim.3 = paste0(arrange(df_pca_contrib, -Dim.3)$variable, 
                                    sep = ": ", 
                                    round(arrange(df_pca_contrib, -Dim.3)$Dim.3, 2)))

df_contrib %>% as.data.frame()

```

#### 3D PCA

```{r pca_3d}

df_pca <- pca_res_ind$coord %>% 
  as_tibble(rownames = "subject") %>% 
  mutate(Status = df_hbox_impute$Status)

plot_ly(x = df_pca$Dim.1, y = df_pca$Dim.2, z = df_pca$Dim.3, 
        type = "scatter3d", mode = "markers", color = df_pca$Status)

```

### Random forest predictive modeling: all data

#### Build, tune, and finalize model
- Random forest model
- bootstrap resampling (w/ replacement)
- BoxCox transform all predictors, remove variables with zero variance, kNN imputation
- tune for optimal ROC AUC and accuracy

```{r model1}

# Data
hbox_df <- df_hbox_trim %>% select(-subject_id)

# Build a model

  # split data set
  set.seed(123)
  hbox_split <- initial_split(hbox_df, strata = Status)
  hbox_train <- training(hbox_split)
  hbox_test <- testing(hbox_split)

  # resamples (not a lot of data, need to use cross validation)
  set.seed(234)
  # vfold_cv(hbox_train, strata = Status) # test too small
  hbox_folds <- bootstraps(hbox_train, strata = Status)

  # Scaffolding for setting up common types of models
  # use_ranger(Status ~ ., data = hbox_df)

  # Copy usemodels code...

  # Create recipe
  ranger_recipe <- 
    recipe(formula = Status ~ ., data = hbox_df) %>%

    # BoxCox transformation to normality
    step_BoxCox(all_predictors(), -all_outcomes()) %>%

    # remove variables with non-zero variance
    step_nzv(all_predictors(), -all_outcomes()) %>%

    # impute missing data using K-nearest neighbors
    step_knnimpute(all_predictors(), -all_outcomes())

  # Model specifications (set for tuning)
  ranger_spec <- 
    rand_forest(mtry = tune(), min_n = tune(), trees = 1000) %>% 
    set_mode("classification") %>% 
    set_engine("ranger")

  # Create workflow
  ranger_workflow <- 
    workflow() %>% 
    add_recipe(ranger_recipe) %>% 
    add_model(ranger_spec)

# Tune model

  # Initial tune on bootstraps
  set.seed(84374)
  ranger_tune <-
    tune_grid(ranger_workflow, 
              resamples = hbox_folds, 
              grid = 11)

# Finalize model

  # finalize workflow
  final_rf <- ranger_workflow %>% 
    finalize_workflow(select_best(ranger_tune, metric = "accuracy"))

  final_rs_res <- final_rf %>% 
    fit_resamples(resamples = hbox_folds,
                  metrics = metric_set(roc_auc, accuracy, sens, spec),
                  control = control_resamples(save_pred = TRUE))

  hbox_fit <- last_fit(final_rf, hbox_split) # fitting to training data, evaluating on testing

```

#### Results

```{r model1_res}

# computed on test set
collect_metrics(hbox_fit) %>% select(.metric, .estimate) %>% as.data.frame()

# collect predictions on test set
df_pred <- collect_predictions(hbox_fit) %>%
  select(c(Status, .pred_HV, .pred_UM, .pred_CM, .pred_class))

df_pred %>% as_data_frame()

df_pred %>% filter(Status != .pred_class) %>% as.data.frame() # only got one wrong

```

```{r model1_plot_res}

conf_mat_resampled(final_rs_res) %>%

  ggplot(aes(x = Truth, y = Freq, fill = Prediction)) + 
  geom_col(position = "dodge") +
  ggtitle("Confusion matrix") +
  theme_bw()

```

#### Determine variable importance
```{r model1_imp}

# new specification for ranger model
imp_spec <- ranger_spec %>% 
  finalize_model(select_best(ranger_tune, metric = "accuracy")) %>% 
  set_engine("ranger", importance = "permutation")

# plot importance 
workflow() %>% 
  add_recipe(ranger_recipe) %>% 
  add_model(imp_spec) %>% 
  fit(hbox_train) %>% 
  pull_workflow_fit() %>% 
  vip(aesthetics = list(alpha = 0.8, fill = "midnightblue")) +

  theme_bw() +
  ggtitle("Variable importance")

```

### Random forest predictive modeling: NIRS data only

#### Build, tune, and finalize model
Same code, but built model only using variables measured using NIRS

```{r model2}

# Data
hbox_df2 <- df_hbox_impute %>% 
  select(c(Status, DeoxyHb, DeoxyHb_alpha, OxyHb, OxyHb_alpha, O2sat, O2sat_alpha, THC, THC_alpha, THC_sd))

# Build a model

  # split data set
  set.seed(456)
                hbox_split2 <- initial_split(hbox_df2, strata = Status)
                hbox_train2 <- training(hbox_split2)
                hbox_test2 <- testing(hbox_split2)

  # resamples (not a lot of data, need to use cross validation)
  set.seed(789)
  hbox_folds2 <- bootstraps(hbox_train2, strata = Status)

  # Scaffolding for setting up common types of models
  # use_ranger(Status ~ ., data = hbox_df2)

  # Copy usemodels code...

  # Create recipe
  ranger_recipe2 <- 
    recipe(formula = Status ~ ., data = hbox_df2) %>%

    # BoxCox transformation to normality
    step_BoxCox(all_predictors(), -all_outcomes()) %>%

    # remove variables with non-zero variance
    step_nzv(all_predictors(), -all_outcomes()) %>%

    # impute missing data using K-nearest neighbors
    step_knnimpute(all_predictors(), -all_outcomes())

  # Model specifications (set for tuning)
  ranger_spec2 <- 
    rand_forest(mtry = tune(), min_n = tune(), trees = 1000) %>% 
    set_mode("classification") %>% 
    set_engine("ranger")

  # Create workflow
  ranger_workflow2 <- 
    workflow() %>% 
    add_recipe(ranger_recipe2) %>% 
    add_model(ranger_spec2)

# Tune model

  # Initial tune on bootstraps
  set.seed(101112)
  ranger_tune2 <-
    tune_grid(ranger_workflow2, 
              resamples = hbox_folds2, 
              grid = 11)

# Finalize model

  # finalize workflow
  final_rf2 <- ranger_workflow2 %>% 
    finalize_workflow(select_best(ranger_tune2, metric = "accuracy"))

  final_rs_res2 <- final_rf2 %>% 
    fit_resamples(resamples = hbox_folds2,
                  metrics = metric_set(roc_auc, accuracy, sens, spec),
                  control = control_resamples(save_pred = TRUE))

  hbox_fit2 <- last_fit(final_rf2, hbox_split2) # fitting to training data, evaluating on testing

```

#### Results

```{r model2_res}

# computed on test set
collect_metrics(final_rs_res2) %>% select(.metric, mean, std_err, n) %>% as.data.frame()

# collect predictions on test set
df_pred2 <- collect_predictions(hbox_fit2) %>%
  select(c(Status, .pred_HV, .pred_UM, .pred_CM, .pred_class))

df_pred2 %>% as_data_frame()

df_pred2 %>% filter(Status != .pred_class) %>% as.data.frame() # only got one wrong

```

```{r model2_plot_res}

conf_mat_resampled(final_rs_res2) %>%

  ggplot(aes(x = Truth, y = Freq, fill = Prediction)) + 
  geom_col(position = "dodge") +
  ggtitle("Confusion matrix") +
  theme_bw()

```

#### Determine variable importance
```{r model2_imp}

# new specification for ranger model
imp_spec2 <- ranger_spec2 %>% 
  finalize_model(select_best(ranger_tune2, metric = "accuracy")) %>% 
  set_engine("ranger", importance = "permutation")

# plot importance 
workflow() %>% 
  add_recipe(ranger_recipe2) %>% 
  add_model(imp_spec2) %>% 
  fit(hbox_train2) %>% 
  pull_workflow_fit() %>% 
  vip(aesthetics = list(alpha = 0.8, fill = "midnightblue")) +

  theme_bw() +
  ggtitle("Variable importance")

```

### 2021-04-15 Scripts/tidymodels_analysis.R

@@ -102,8 +102,6 @@
 
 # Use PCA to explore if these variables cluster Status --------------------
 
-     
-
         # convert tibble to df
         df_hbox_impute_df <- df_hbox_impute %>% select(-c(subject_id, Status)) %>% as.data.frame(df_hbox_impute) 
         rownames(df_hbox_impute_df) <- df_hbox_impute$subject_id

### 2021-05-04 delete.Rmd
updated analysis - variable correlations and one-way ANOVAs

---
title: "format"
author: "Rachel Smith"
date: "5/4/2021"
output: html_document
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, comment = "")

library(tidyverse)
library(nortest)
library(FSA)
library(tidymodels)
library(factoextra)
library(ggpubr)
library(ggrepel)
library(grid)
library(pheatmap)
library(plotly)
library(corrr)
library(reshape)
library(RColorBrewer)
library(usemodels)
library(ranger)
library(vip)
library(MASS)

```

# Dysregulation of hemoglobin and oxygenation in cerebral malaria {.tabset}

```{r data}

# read in data
df_hbox <- read.csv("Data/98 first visit records with cerebral oxygenation stats_trimmed.csv") %>% as_tibble()

# get rid of "Other"
df_hbox <- df_hbox %>% 
  filter(Status != "Other") %>% 
  mutate(Status = factor(Status, levels = c("HV", "UM", "CM")))

# identify duplicated rows (TM0003, TM2001)
df_hbox <- df_hbox %>% filter(duplicated(Subject.ID..NIAID.) == FALSE)

# Convert age to numeric
df_hbox <- df_hbox %>% 
  mutate(Age = gsub(" Years, ", "_", Age)) %>% 
  mutate(Age = gsub(" Months ", "", Age)) %>% 
  separate(Age, into = c("Years", "Months"), sep = "_") %>% 
  mutate(Years = as.numeric(Years)) %>% 
  mutate(Months = as.numeric(Months)/12) %>% 
  mutate(Age = Years + Months) %>% 
  dplyr::select(-c(Years, Months))

# Convert sex to factor
df_hbox <- df_hbox %>% mutate(Sex = as.factor(Sex))

# Trim for variables of interest (VOI)
df_hbox_trim <- df_hbox %>% 
  dplyr::select(c(Subject.ID..NIAID., Status, Glucose, Hematocrit, Lactate, Temperature,
                  Arginine.umol.L, Haptoglobin..mg.dl., Hemoglobin..uM., Whole.Blood.Nitrite, 
                  cerebral.hbdeox, cerebral.hbdeox.alpha, cerebral.hbox, cerebral.hbox.alpha,
                  cerebral.sat, cerebral.sat.alpha, cerebral.thc, cerebral.thc.alpha, cerebral.thc.norm.std)) %>%

  dplyr::rename("subject_id" = "Subject.ID..NIAID.", "Arginine (umol/L)" = "Arginine.umol.L", "Haptoglobin (mg/dL)" = "Haptoglobin..mg.dl.", 
                "Hemoglobin (uM)" = "Hemoglobin..uM.", "WB_Nitrite" = "Whole.Blood.Nitrite", "DeoxyHb" = "cerebral.hbdeox", 
                "DeoxyHb_alpha" = "cerebral.hbdeox.alpha", "OxyHb" = "cerebral.hbox", "OxyHb_alpha" = "cerebral.hbox.alpha", 
                "O2sat" = "cerebral.sat", "O2sat_alpha" = "cerebral.sat.alpha", "THC" = "cerebral.thc", 
                "THC_alpha" = "cerebral.thc.alpha", "THC_sd" = "cerebral.thc.norm.std")

# impute missing data with median by group
f_impute <- function(x, na.rm = TRUE) (replace(x, is.na(x), median(x, na.rm = na.rm)))

df_hbox_impute <- df_hbox_trim %>% 
  group_by(Status) %>% 
  mutate_at(3:ncol(.), f_impute) %>% 
  ungroup()

# pivot long
df_hbox_long <- df_hbox_trim %>% pivot_longer(cols = 3:ncol(.), names_to = "measurement", values_to = "value")

```

## One way ANOVAs {.tabset}

### Influence healthy vs malaria

#### Arginine
```{r arginine_show, echo = TRUE, eval = FALSE}

# create model for BoxCox transformation
arg_model <- glm(`Arginine (umol/L)` ~ Status, data = df_hbox_impute)
boxcox(arg_model, lambda = seq(-0.5, 1, by = 0.1), plotit = TRUE)

# BoxCox transform
df_arg <- df_hbox_impute %>% mutate(arg_transform = (`Arginine (umol/L)`^-0.25 - 1)/-0.25)

# model on transformed data
glm(arg_transform ~ Status, data = df_arg) %>% summary()

```

```{r arg_model}

glm(`Arginine (umol/L)` ~ Status, data = df_hbox_impute) %>% summary()

df_hbox_long %>% filter(measurement == "Arginine (umol/L)") %>%

  ggplot(aes(x = Status, y = value)) +
  geom_violin(aes(color = Status)) +
  geom_jitter(position = position_jitter(0.2), shape = 1) +
  stat_summary(fun = "median", geom = "crossbar", aes(color = Status), size = 0.2, width = 0.5) +

  theme_bw() +
  theme(legend.position = "none")  +
  ggtitle("Arginine (umol/L)") +

  theme(strip.text = element_text(size = 15),
        axis.title = element_text(size = 15),
        axis.title.y = element_blank(),
        axis.text = element_text(size = 15))

```

#### Temperature
```{r temp}

glm(Temperature ~ Status, data = df_hbox_impute) %>% summary()
aov(Temperature ~ Status, data = df_hbox_impute) %>% TukeyHSD() %>% 
  .$Status %>% 
  as.data.frame() %>% 
  rownames_to_column("Status") %>% 
  dplyr::select(Status, `p adj`)

```

#### Hematocrit
```{r hct}

glm(Hematocrit ~ Status, data = df_hbox_impute) %>% summary()
aov(Temperature ~ Status, data = df_hbox_impute) %>% TukeyHSD() %>% 
  .$Status %>% 
  as.data.frame() %>% 
  rownames_to_column("Status") %>% 
  dplyr::select(Status, `p adj`)

```

### Influence CM vs UM

#### O2sat_alpha
```{r O2sat_alpha}

glm(O2sat_alpha ~ Status, data = df_hbox_impute) %>% summary()
aov(O2sat_alpha ~ Status, data = df_hbox_impute) %>% TukeyHSD() %>% 
  .$Status %>% 
  as.data.frame() %>% 
  rownames_to_column("Status") %>% 
  dplyr::select(Status, `p adj`)

```

#### THC_alpha
```{r THC_alpha}

glm(THC_alpha ~ Status, data = df_hbox_impute) %>% summary()
aov(THC_alpha ~ Status, data = df_hbox_impute) %>% TukeyHSD() %>% 
  .$Status %>% 
  as.data.frame() %>% 
  rownames_to_column("Status") %>% 
  dplyr::select(Status, `p adj`)

```

#### THC
```{r THC}

glm(THC ~ Status, data = df_hbox_impute) %>% summary()
aov(THC ~ Status, data = df_hbox_impute) %>% TukeyHSD() %>% 
  .$Status %>% 
  as.data.frame() %>% 
  rownames_to_column("Status") %>% 
  dplyr::select(Status, `p adj`)

```

#### OxyHb
```{r OxyHb}

glm(OxyHb ~ Status, data = df_hbox_impute) %>% summary()
aov(OxyHb ~ Status, data = df_hbox_impute) %>% TukeyHSD() %>% 
  .$Status %>% 
  as.data.frame() %>% 
  rownames_to_column("Status") %>% 
  dplyr::select(Status, `p adj`)

```

#### Lactate
```{r Lactate}

glm(Lactate ~ Status, data = df_hbox_impute) %>% summary()
aov(Lactate ~ Status, data = df_hbox_impute) %>% TukeyHSD() %>% 
  .$Status %>% 
  as.data.frame() %>% 
  rownames_to_column("Status") %>% 
  dplyr::select(Status, `p adj`)

```

#### ?OxyHb_alpha
```{r OxyHb_alpha}

glm(OxyHb_alpha ~ Status, data = df_hbox_impute) %>% summary()
aov(OxyHb_alpha ~ Status, data = df_hbox_impute) %>% TukeyHSD() %>% 
  .$Status %>% 
  as.data.frame() %>% 
  rownames_to_column("Status") %>% 
  dplyr::select(Status, `p adj`)

```

### 2021-05-04 tidymodels_analysis.Rmd
updated analysis - variable correlations and one-way ANOVAs

@@
 library(factoextra)
+library(ggpubr)
+library(ggrepel)
+library(grid)
+library(pheatmap)
 library(plotly)
+library(corrr)
+library(reshape)
+library(RColorBrewer)
 library(usemodels)
@@
 library(vip)
+library(MASS)
 
@@
 
-## Dysregulation of hemoglobin and oxygenation in cerebral malaria {.tabset}
+# Dysregulation of hemoglobin and oxygenation in cerebral malaria {.tabset}
 
@@
   mutate(Age = Years + Months) %>% 
-  select(-c(Years, Months))
+  dplyr::select(-c(Years, Months))
 
@@
 
-### Exploratory analysis
+## Exploratory analysis
 
@@
 
-### PCA
+## PCA
 
@@
 # convert tibble to df
-df_hbox_impute_df <- df_hbox_impute %>% select(-c(subject_id, Status)) %>% as.data.frame(df_hbox_impute) 
+df_hbox_impute_df <- df_hbox_impute %>% dplyr::select(-c(subject_id, Status)) %>% as.data.frame(df_hbox_impute) 
 rownames(df_hbox_impute_df) <- df_hbox_impute$subject_id
@@
 
-p_pca <- fviz_pca_ind(my_pca,
-                      col.ind = groups, # color by groups
-                      palette = c("#00AFBB",  "#FC4E07", "#7CAE00"),
-                      addEllipses = TRUE, # Concentration ellipses
-                      ellipse.type = "confidence",
-                      legend.title = "Groups",
-                      repel = TRUE
-)
+df_pnts <- my_pca$x %>% as_tibble() %>% mutate(group = groups)
 
-p_pca$layers[[4]] <- NULL # get rid of text labels
+df_pnts %>%
 
-p_pca
+  ggplot(aes(x = PC1, y = PC2, color = group, fill = group)) +
+  geom_point(size = 3) +
+  stat_conf_ellipse(level = 0.95, npoint = 100, bary = TRUE, alpha = 0.1, geom = "polygon") +
+  stat_mean(aes(shape = group), na.rm = FALSE) +
+  geom_vline(xintercept = 0, lty = 2, color = "black") +
+  geom_hline(yintercept = 0, lty = 2, color = "black") +
+
+  xlab(paste0("PC1: ", round(pca_eig_val$variance.percent[1], 2), "% of variance")) +
+  ylab(paste0("PC1: ", round(pca_eig_val$variance.percent[2], 2), "% of variance")) +
+
+  theme_bw()
 
@@
 
-# Biplot of individuals and variables
-p_biplot <- fviz_pca_biplot(my_pca, repel = TRUE,
-                            col.var = "black", # Variables color
-                            col.ind = groups  # Individuals color
-)
+df_vars <- my_pca$rotation %>% as_tibble() %>% mutate(variable = rownames(my_pca$rotation))
 
-p_biplot$layers[[3]] <- NULL
-p_biplot$layers[[2]] <- NULL
+# plot
 
-p_biplot
+  # plot patients
+  ggplot() +
+  geom_point(data = df_pnts, mapping = aes(x = PC1, y = PC2, color = group), 
+             size = 3) +
+  stat_conf_ellipse(data = df_pnts, mapping = aes(x = PC1, y = PC2, color = group, fill = group),
+                    level = 0.95, npoint = 100, bary = TRUE, alpha = 0.1, geom = "polygon") +
+  stat_mean(data = df_pnts, mapping = aes(x = PC1, y = PC2, color = group, shape = group),
+            na.rm = FALSE) +
+
+  # plot variables
+  geom_point(data = df_vars, aes(x = PC1*10, y = PC2*10)) +
+  geom_text_repel(data = df_vars, aes(x = PC1*10, y = PC2*10, label = variable, size = 10), show.legend = FALSE) +
+  geom_segment(data = df_vars, aes(x = 0, y = 0, xend = PC1*10, yend = PC2*10), arrow = arrow()) +
+
+  # draw lines
+  geom_vline(xintercept = 0, lty = 2, color = "black") +
+  geom_hline(yintercept = 0, lty = 2, color = "black") +
+
+  # design
+  xlab(paste0("PC1: ", round(pca_eig_val$variance.percent[1], 2), "% of variance")) +
+  ylab(paste0("PC1: ", round(pca_eig_val$variance.percent[2], 2), "% of variance")) +
+
+  theme_bw()
 
@@
   as_tibble(rownames = "subject") %>% 
-  mutate(Status = df_hbox_impute$Status)
+  mutate(Status = factor(df_hbox_impute$Status, levels = c("UM", "HV", "CM")))
 
@@
 
-### Random forest predictive modeling: all data
+## Variable correlations
+
+```{r correlation, fig.heigh = 10, fig.width = 10}
+
+m_cor <- df_hbox_impute %>% dplyr::select(-c(subject_id, Status)) %>% cor()
+
+melt(m_cor) %>% 
+  ggplot(aes(x = X1, y = X2, fill = value, label = round(value,2))) + 
+  geom_tile() +
+  scale_fill_gradient2(low = "#4575b4", high = "#d73027") +
+  geom_text(size = 3) +
+
+  xlab("") +
+  ylab("") +
+  theme(axis.text.x = element_text(angle = 45, hjust = 0.9))
+
+```
+
+## Random forest predictive modeling {.tabset}
+
+### All variables
 
@@
 # Data
-hbox_df <- df_hbox_trim %>% select(-subject_id)
+hbox_df <- df_hbox_trim %>% dplyr::select(-subject_id)
 
@@
 # computed on test set
-collect_metrics(hbox_fit) %>% select(.metric, .estimate) %>% as.data.frame()
+collect_metrics(hbox_fit) %>% dplyr::select(.metric, .estimate) %>% as.data.frame()
 
@@
 df_pred <- collect_predictions(hbox_fit) %>%
-  select(c(Status, .pred_HV, .pred_UM, .pred_CM, .pred_class))
+  dplyr::select(c(Status, .pred_HV, .pred_UM, .pred_CM, .pred_class))
 
-df_pred %>% as_data_frame()
+df_pred %>% as.data.frame()
 
@@
 
-  ggplot(aes(x = Truth, y = Freq, fill = Prediction)) + 
+  ggplot(aes(x = Truth, y = Freq, fill = Prediction, label = Freq)) + 
   geom_col(position = "dodge") +
+  geom_text(position = position_dodge(width = 0.9)) +
   ggtitle("Confusion matrix") +
@@
 
-### Random forest predictive modeling: NIRS data only
+### NIRS variables only
 
@@
 hbox_df2 <- df_hbox_impute %>% 
-  select(c(Status, DeoxyHb, DeoxyHb_alpha, OxyHb, OxyHb_alpha, O2sat, O2sat_alpha, THC, THC_alpha, THC_sd))
+  dplyr::select(c(Status, DeoxyHb, DeoxyHb_alpha, OxyHb, OxyHb_alpha, O2sat, O2sat_alpha, THC, THC_alpha, THC_sd))
 
@@
 # computed on test set
-collect_metrics(final_rs_res2) %>% select(.metric, mean, std_err, n) %>% as.data.frame()
+collect_metrics(final_rs_res2) %>% dplyr::select(.metric, mean, std_err, n) %>% as.data.frame()
 
@@
 df_pred2 <- collect_predictions(hbox_fit2) %>%
-  select(c(Status, .pred_HV, .pred_UM, .pred_CM, .pred_class))
+  dplyr::select(c(Status, .pred_HV, .pred_UM, .pred_CM, .pred_class))
 
-df_pred2 %>% as_data_frame()
+df_pred2 %>% as.data.frame()
 
@@
 
-  ggplot(aes(x = Truth, y = Freq, fill = Prediction)) + 
+  ggplot(aes(x = Truth, y = Freq, fill = Prediction, label = Freq)) + 
   geom_col(position = "dodge") +
+  geom_text(position = position_dodge(width = 1)) +
   ggtitle("Confusion matrix") +
@@
 ```
+
+### Other variables
+
+#### Build, tune, and finalize model
+Same code, but built model only using other variables (Arginine, lactate, hematocrit, temperature, whole blood nitrite, haptoglobin, glucose)
+
+```{r model3}
+
+# Data
+hbox_df3 <- df_hbox_impute %>% dplyr::select(-c(subject_id, DeoxyHb, DeoxyHb_alpha, OxyHb, OxyHb_alpha, O2sat, O2sat_alpha, THC, THC_alpha, THC_sd))
+
+# Build a model
+
+  # split data set
+  set.seed(1234)
+  hbox_split3 <- initial_split(hbox_df3, strata = Status)
+  hbox_train3 <- training(hbox_split3)
+  hbox_test3 <- testing(hbox_split3)
+
+  # resamples (not a lot of data, need to use cross validation)
+  set.seed(2345)
+  hbox_folds3 <- bootstraps(hbox_train3, strata = Status)
+
+  # Copy usemodels code...
+
+  # Create recipe
+  ranger_recipe3 <- 
+    recipe(formula = Status ~ ., data = hbox_df3) %>%
+
+    # BoxCox transformation to normality
+    step_BoxCox(all_predictors(), -all_outcomes()) %>%
+
+    # remove variables with non-zero variance
+    step_nzv(all_predictors(), -all_outcomes()) %>%
+
+    # impute missing data using K-nearest neighbors
+    step_knnimpute(all_predictors(), -all_outcomes())
+
+  # Model specifications (set for tuning)
+  ranger_spec3 <- 
+    rand_forest(mtry = tune(), min_n = tune(), trees = 1000) %>% 
+    set_mode("classification") %>% 
+    set_engine("ranger")
+
+  # Create workflow
+  ranger_workflow3 <- 
+    workflow() %>% 
+    add_recipe(ranger_recipe3) %>% 
+    add_model(ranger_spec3)
+
+# Tune model
+
+  # Initial tune on bootstraps
+  set.seed(8434)
+  ranger_tune3 <-
+    tune_grid(ranger_workflow3, 
+              resamples = hbox_folds3, 
+              grid = 11)
+
+# Finalize model
+
+  # finalize workflow
+  final_rf3 <- ranger_workflow3 %>% 
+    finalize_workflow(select_best(ranger_tune3, metric = "accuracy"))
+
+  final_rs_res3 <- final_rf3 %>% 
+    fit_resamples(resamples = hbox_folds3,
+                  metrics = metric_set(roc_auc, accuracy, sens, spec),
+                  control = control_resamples(save_pred = TRUE))
+
+  hbox_fit3 <- last_fit(final_rf3, hbox_split3) # fitting to training data, evaluating on testing
+
+```
+
+#### Results
+
+```{r model3_res}
+
+# computed on test set
+collect_metrics(final_rs_res3) %>% dplyr::select(.metric, mean, std_err, n) %>% as.data.frame()
+
+# collect predictions on test set
+df_pred3 <- collect_predictions(hbox_fit3) %>%
+  dplyr::select(c(Status, .pred_HV, .pred_UM, .pred_CM, .pred_class))
+
+df_pred3 %>% as.data.frame()
+
+df_pred3 %>% filter(Status != .pred_class) %>% as.data.frame()
+
+```
+
+```{r model3_plot_res}
+
+conf_mat_resampled(final_rs_res3) %>%
+
+  ggplot(aes(x = Truth, y = Freq, fill = Prediction, label = Freq)) + 
+  geom_col(position = "dodge") +
+  geom_text(position = position_dodge(width = 1)) +
+  ggtitle("Confusion matrix") +
+  theme_bw()
+
+```
+
+#### Determine variable importance
+```{r model3_imp}
+
+# new specification for ranger model
+imp_spec3 <- ranger_spec3 %>% 
+  finalize_model(select_best(ranger_tune3, metric = "accuracy")) %>% 
+  set_engine("ranger", importance = "permutation")
+
+# plot importance 
+workflow() %>% 
+  add_recipe(ranger_recipe3) %>% 
+  add_model(imp_spec3) %>% 
+  fit(hbox_train3) %>% 
+  pull_workflow_fit() %>% 
+  vip(aesthetics = list(alpha = 0.8, fill = "midnightblue")) +
+
+  theme_bw() +
+  ggtitle("Variable importance")
+
+```
+
+## One way ANOVAs {.tabset}
+
+### Influence healthy vs malaria
+
+#### Arginine
+```{r arginine_show, echo = TRUE, eval = FALSE}
+
+# create model for BoxCox transformation
+arg_model <- glm(`Arginine (umol/L)` ~ Status, data = df_hbox_impute)
+boxcox(arg_model, lambda = seq(-0.5, 1, by = 0.1), plotit = TRUE)
+
+# BoxCox transform
+df_arg <- df_hbox_impute %>% mutate(arg_transform = (`Arginine (umol/L)`^-0.25 - 1)/-0.25)
+
+# model on transformed data
+glm(arg_transform ~ Status, data = df_arg) %>% summary()
+
+```
+
+```{r arg_model}
+
+glm(`Arginine (umol/L)` ~ Status, data = df_hbox_impute) %>% summary()
+
+aov(`Arginine (umol/L)` ~ Status, data = df_hbox_impute) %>% TukeyHSD() %>% 
+  .$Status %>% 
+  as.data.frame() %>% 
+  rownames_to_column("Status") %>% 
+  dplyr::select(Status, `p adj`)
+
+```
+
+```{r arg_plot}
+
+df_hbox_long %>% filter(measurement == "Arginine (umol/L)") %>%
+
+  ggplot(aes(x = Status, y = value)) +
+  geom_violin(aes(color = Status)) +
+  geom_jitter(position = position_jitter(0.2), shape = 1) +
+  stat_summary(fun = "median", geom = "crossbar", aes(color = Status), size = 0.2, width = 0.5) +
+
+  theme_bw() +
+  theme(legend.position = "none")  +
+  ggtitle("Arginine (umol/L)") +
+
+  theme(strip.text = element_text(size = 15),
+        axis.title = element_text(size = 15),
+        axis.title.y = element_blank(),
+        axis.text = element_text(size = 15))
+
+```
+
+#### Temperature
+```{r temp}
+
+glm(Temperature ~ Status, data = df_hbox_impute) %>% summary()
+
+aov(Temperature ~ Status, data = df_hbox_impute) %>% TukeyHSD() %>% 
+  .$Status %>% 
+  as.data.frame() %>% 
+  rownames_to_column("Status") %>% 
+  dplyr::select(Status, `p adj`)
+
+```
+
+```{r temp_plot}
+
+df_hbox_long %>% filter(measurement == "Temperature") %>%
+
+  ggplot(aes(x = Status, y = value)) +
+  geom_violin(aes(color = Status)) +
+  geom_jitter(position = position_jitter(0.2), shape = 1) +
+  stat_summary(fun = "median", geom = "crossbar", aes(color = Status), size = 0.2, width = 0.5) +
+
+  theme_bw() +
+  theme(legend.position = "none")  +
+  ggtitle("Temperature") +
+
+  theme(strip.text = element_text(size = 15),
+        axis.title = element_text(size = 15),
+        axis.title.y = element_blank(),
+        axis.text = element_text(size = 15))
+
+```
+
+#### Hematocrit
+```{r hct}
+
+glm(Hematocrit ~ Status, data = df_hbox_impute) %>% summary()
+
+aov(Temperature ~ Status, data = df_hbox_impute) %>% TukeyHSD() %>% 
+  .$Status %>% 
+  as.data.frame() %>% 
+  rownames_to_column("Status") %>% 
+  dplyr::select(Status, `p adj`)
+
+```
+
+```{r hct_plot}
+
+df_hbox_long %>% filter(measurement == "Hematocrit") %>%
+
+  ggplot(aes(x = Status, y = value)) +
+  geom_violin(aes(color = Status)) +
+  geom_jitter(position = position_jitter(0.2), shape = 1) +
+  stat_summary(fun = "median", geom = "crossbar", aes(color = Status), size = 0.2, width = 0.5) +
+
+  theme_bw() +
+  theme(legend.position = "none")  +
+  ggtitle("Hematocrit") +
+
+  theme(strip.text = element_text(size = 15),
+        axis.title = element_text(size = 15),
+        axis.title.y = element_blank(),
+        axis.text = element_text(size = 15))
+
+```
+
+### Influence CM vs UM
+
+#### O2sat_alpha
+```{r O2sat_alpha}
+
+glm(O2sat_alpha ~ Status, data = df_hbox_impute) %>% summary()
+
+aov(O2sat_alpha ~ Status, data = df_hbox_impute) %>% TukeyHSD() %>% 
+  .$Status %>% 
+  as.data.frame() %>% 
+  rownames_to_column("Status") %>% 
+  dplyr::select(Status, `p adj`)
+
+```
+
+```{r O2sat_alpha_plot}
+
+df_hbox_long %>% filter(measurement == "O2sat_alpha") %>%
+
+  ggplot(aes(x = Status, y = value)) +
+  geom_violin(aes(color = Status)) +
+  geom_jitter(position = position_jitter(0.2), shape = 1) +
+  stat_summary(fun = "median", geom = "crossbar", aes(color = Status), size = 0.2, width = 0.5) +
+
+  theme_bw() +
+  theme(legend.position = "none")  +
+  ggtitle("O2sat_alpha") +
+
+  theme(strip.text = element_text(size = 15),
+        axis.title = element_text(size = 15),
+        axis.title.y = element_blank(),
+        axis.text = element_text(size = 15))
+
+```
+
+#### THC_alpha
+```{r THC_alpha}
+
+glm(THC_alpha ~ Status, data = df_hbox_impute) %>% summary()
+
+aov(THC_alpha ~ Status, data = df_hbox_impute) %>% TukeyHSD() %>% 
+  .$Status %>% 
+  as.data.frame() %>% 
+  rownames_to_column("Status") %>% 
+  dplyr::select(Status, `p adj`)
+
+```
+
+```{r THC_alpha_plot}
+
+df_hbox_long %>% filter(measurement == "THC_alpha") %>%
+
+  ggplot(aes(x = Status, y = value)) +
+  geom_violin(aes(color = Status)) +
+  geom_jitter(position = position_jitter(0.2), shape = 1) +
+  stat_summary(fun = "median", geom = "crossbar", aes(color = Status), size = 0.2, width = 0.5) +
+
+  theme_bw() +
+  theme(legend.position = "none")  +
+  ggtitle("THC_alpha") +
+
+  theme(strip.text = element_text(size = 15),
+        axis.title = element_text(size = 15),
+        axis.title.y = element_blank(),
+        axis.text = element_text(size = 15))
+
+```
+
+#### THC
+```{r THC}
+
+glm(THC ~ Status, data = df_hbox_impute) %>% summary()
+
+aov(THC ~ Status, data = df_hbox_impute) %>% TukeyHSD() %>% 
+  .$Status %>% 
+  as.data.frame() %>% 
+  rownames_to_column("Status") %>% 
+  dplyr::select(Status, `p adj`)
+
+```
+
+```{r THC_plot}
+
+df_hbox_long %>% filter(measurement == "THC") %>%
+
+  ggplot(aes(x = Status, y = value)) +
+  geom_violin(aes(color = Status)) +
+  geom_jitter(position = position_jitter(0.2), shape = 1) +
+  stat_summary(fun = "median", geom = "crossbar", aes(color = Status), size = 0.2, width = 0.5) +
+
+  theme_bw() +
+  theme(legend.position = "none")  +
+  ggtitle("THC") +
+
+  theme(strip.text = element_text(size = 15),
+        axis.title = element_text(size = 15),
+        axis.title.y = element_blank(),
+        axis.text = element_text(size = 15))
+
+```
+
+#### OxyHb
+```{r OxyHb}
+
+glm(OxyHb ~ Status, data = df_hbox_impute) %>% summary()
+
+aov(OxyHb ~ Status, data = df_hbox_impute) %>% TukeyHSD() %>% 
+  .$Status %>% 
+  as.data.frame() %>% 
+  rownames_to_column("Status") %>% 
+  dplyr::select(Status, `p adj`)
+
+```
+
+```{r OxyHb_plot}
+
+df_hbox_long %>% filter(measurement == "OxyHb") %>%
+
+  ggplot(aes(x = Status, y = value)) +
+  geom_violin(aes(color = Status)) +
+  geom_jitter(position = position_jitter(0.2), shape = 1) +
+  stat_summary(fun = "median", geom = "crossbar", aes(color = Status), size = 0.2, width = 0.5) +
+
+  theme_bw() +
+  theme(legend.position = "none")  +
+  ggtitle("OxyHb") +
+
+  theme(strip.text = element_text(size = 15),
+        axis.title = element_text(size = 15),
+        axis.title.y = element_blank(),
+        axis.text = element_text(size = 15))
+
+```
+
+#### Lactate
+```{r Lactate}
+
+glm(Lactate ~ Status, data = df_hbox_impute) %>% summary()
+
+aov(Lactate ~ Status, data = df_hbox_impute) %>% TukeyHSD() %>% 
+  .$Status %>% 
+  as.data.frame() %>% 
+  rownames_to_column("Status") %>% 
+  dplyr::select(Status, `p adj`)
+
+```
+
+```{r Lactate_plot}
+
+df_hbox_long %>% filter(measurement == "Lactate") %>%
+
+  ggplot(aes(x = Status, y = value)) +
+  geom_violin(aes(color = Status)) +
+  geom_jitter(position = position_jitter(0.2), shape = 1) +
+  stat_summary(fun = "median", geom = "crossbar", aes(color = Status), size = 0.2, width = 0.5) +
+
+  theme_bw() +
+  theme(legend.position = "none")  +
+  ggtitle("Lactate") +
+
+  theme(strip.text = element_text(size = 15),
+        axis.title = element_text(size = 15),
+        axis.title.y = element_blank(),
+        axis.text = element_text(size = 15))
+
+```
+
+#### ?OxyHb_alpha
+```{r OxyHb_alpha}
+
+glm(OxyHb_alpha ~ Status, data = df_hbox_impute) %>% summary()
+
+aov(OxyHb_alpha ~ Status, data = df_hbox_impute) %>% TukeyHSD() %>% 
+  .$Status %>% 
+  as.data.frame() %>% 
+  rownames_to_column("Status") %>% 
+  dplyr::select(Status, `p adj`)
+
+```
+
+```{r OxyHb_alpha_plot}
+
+df_hbox_long %>% filter(measurement == "OxyHb_alpha") %>%
+
+  ggplot(aes(x = Status, y = value)) +
+  geom_violin(aes(color = Status)) +
+  geom_jitter(position = position_jitter(0.2), shape = 1) +
+  stat_summary(fun = "median", geom = "crossbar", aes(color = Status), size = 0.2, width = 0.5) +
+
+  theme_bw() +
+  theme(legend.position = "none")  +
+  ggtitle("OxyHb_alpha") +
+
+  theme(strip.text = element_text(size = 15),
+        axis.title = element_text(size = 15),
+        axis.title.y = element_blank(),
+        axis.text = element_text(size = 15))
+
+```

### 2021-05-04 Scripts/tidymodels_analysis.R

@@ -293,7 +293,7 @@
                 
                 
         # Data
-        hbox_df2 <- df_hbox_impute %>% select(-c(subject_id, Glucose, Hematocrit, Lactate, Temperature, Arginine, Haptoglobin, Hemoglobin, WB_Nitrite))
+        hbox_df2 <- df_hbox_impute %>% select(c(Status, DeoxyHb, DeoxyHb_alpha, OxyHb, OxyHb_alpha, O2sat, O2sat_alpha, THC, THC_alpha, THC_sd))
                 
         # Build a model
                 
@@ -384,16 +384,103 @@
                         pull_workflow_fit() %>% 
                         vip(aesthetics = list(alpha = 0.8, fill = "midnightblue"))
                 
-                
-                
-                
-                
-                
-                
-                
+
 
                 
+# Same procedure but only with other data --------------------------------------------------------------
                 
+                
+        # Data
+        hbox_df3 <- df_hbox_impute %>% select(-c(subject_id, DeoxyHb, DeoxyHb_alpha, OxyHb, OxyHb_alpha, O2sat, O2sat_alpha, THC, THC_alpha, THC_sd))
+                
+        # Build a model
+                
+                # split data set
+                set.seed(1234)
+                hbox_split3 <- initial_split(hbox_df3, strata = Status)
+                hbox_train3 <- training(hbox_split3)
+                hbox_test3 <- testing(hbox_split3)  
+                
+                # resamples (not a lot of data, need to use cross validation)
+                set.seed(2345)
+                vfold_cv(hbox_train3, strata = Status) # test too small
+                hbox_folds3 <- bootstraps(hbox_train3, strata = Status)
+                
+                # Scaffolding for setting up common types of models
+                use_ranger(Status ~ ., data = hbox_df3)
+                
+                # Copy usemodels code...
+                
+                # Create recipe
+                ranger_recipe3 <- 
+                        recipe(formula = Status ~ ., data = hbox_df3) %>% 
+                        
+                        # BoxCox transformation to normality
+                        step_BoxCox(all_predictors(), -all_outcomes()) %>% 
+                        
+                        # remove variables with non-zero variance
+                        step_nzv(all_predictors(), -all_outcomes()) %>% 
+                        
+                        # impute missing data using K-nearest neighbors
+                        step_knnimpute(all_predictors(), -all_outcomes()) 
+                
+                # Model specifications (set for tuning)
+                ranger_spec3 <- 
[…]

### 2021-05-05 log_reg_analysis.Rmd
logistic regression model

---
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = FALSE, comment = "", message = FALSE, warning = FALSE, fig.height = 8, fig.width = 10)

```

# Cerebral oxygenation in cerebral malaria: logistic regression analysis {.tabset}

```{r data}

library(tidyverse)

# read in data
df_hbox <- read.csv("Data/98 first visit records with cerebral oxygenation stats_trimmed.csv") %>% as_tibble()

# get rid of "Other"
df_hbox <- df_hbox %>% 
  filter(Status != "Other") %>% 
  mutate(Status = factor(Status, levels = c("HV", "UM", "CM")))

# identify duplicated rows (TM0003, TM2001)
df_hbox <- df_hbox %>% filter(duplicated(Subject.ID..NIAID.) == FALSE)

# Convert age to numeric
df_hbox <- df_hbox %>% 
  mutate(Age = gsub(" Years, ", "_", Age)) %>% 
  mutate(Age = gsub(" Months ", "", Age)) %>% 
  separate(Age, into = c("Years", "Months"), sep = "_") %>% 
  mutate(Years = as.numeric(Years)) %>% 
  mutate(Months = as.numeric(Months)/12) %>% 
  mutate(Age = Years + Months) %>% 
  dplyr::select(-c(Years, Months))

# Convert sex to factor
df_hbox <- df_hbox %>% mutate(Sex = as.factor(Sex))

# Trim for variables of interest (VOI)
df_hbox_trim <- df_hbox %>% 
  dplyr::select(c(Subject.ID..NIAID., Status, Glucose, Hematocrit, Lactate, Temperature,
                  Arginine.umol.L, Haptoglobin..mg.dl., Hemoglobin..uM., Whole.Blood.Nitrite, 
                  cerebral.hbdeox, cerebral.hbdeox.alpha, cerebral.hbox, cerebral.hbox.alpha,
                  cerebral.sat, cerebral.sat.alpha, cerebral.thc, cerebral.thc.alpha, cerebral.thc.norm.std)) %>%

  dplyr::rename("subject_id" = "Subject.ID..NIAID.", "Arginine (umol/L)" = "Arginine.umol.L", "Haptoglobin (mg/dL)" = "Haptoglobin..mg.dl.", 
                "Hemoglobin (uM)" = "Hemoglobin..uM.", "WB_Nitrite" = "Whole.Blood.Nitrite", "DeoxyHb" = "cerebral.hbdeox", 
                "DeoxyHb_alpha" = "cerebral.hbdeox.alpha", "OxyHb" = "cerebral.hbox", "OxyHb_alpha" = "cerebral.hbox.alpha", 
                "O2sat" = "cerebral.sat", "O2sat_alpha" = "cerebral.sat.alpha", "THC" = "cerebral.thc", 
                "THC_alpha" = "cerebral.thc.alpha", "THC_sd" = "cerebral.thc.norm.std")

# impute missing data with median by group
f_impute <- function(x, na.rm = TRUE) (replace(x, is.na(x), median(x, na.rm = na.rm)))

df_hbox_impute <- df_hbox_trim %>% 
  group_by(Status) %>% 
  mutate_at(3:ncol(.), f_impute) %>% 
  ungroup()

# pivot long
df_hbox_long <- df_hbox_trim %>% pivot_longer(cols = 3:ncol(.), names_to = "measurement", values_to = "value")

```

## Exploratory analysis

```{r voi}

# filter to just UM and CM
df_hbox_log <- df_hbox_impute %>% filter(Status != "HV") %>% select(-subject_id)

# select only VOI
df_hbox_log_voi <- df_hbox_log %>% select(1, 10:18)

# convert CM vs UM to 1 and 0
df_hbox_voi_01 <- df_hbox_log_voi %>% mutate(Status = ifelse(Status == "CM", 1, 0) %>% as.factor())

```

```{r voi_plot}

df_hbox_log_voi %>% pivot_longer(2:10, names_to = "variable", values_to = "value") %>%

  ggplot(aes(x = Status, y = value)) +
  geom_violin(aes(color = Status)) +
  geom_jitter(position = position_jitter(0.2), shape = 1) +
  stat_summary(fun = "median", geom = "crossbar", aes(color = Status), size = 0.2, width = 0.5) +
  facet_wrap(~variable, scales = "free_y") +

  theme_bw() +
  theme(legend.position = "none")  +

  theme(strip.text = element_text(size = 15),
        axis.title = element_text(size = 15),
        axis.title.y = element_blank(),
        axis.text = element_text(size = 15))
```

```{r histogram, fig.height = 6}

# histogram

df_hbox_log_voi %>% pivot_longer(2:10, names_to = "variable", values_to = "value") %>%

  ggplot(aes(x = value)) +
  geom_histogram(bins = 20) +
  facet_grid(Status ~ variable, scales = "free") +

  theme(strip.text.y = element_text(size = 15),
        strip.text.x = element_text(size = 11),
        axis.title.x = element_blank(),
        axis.title.y = element_text(size = 15),
        axis.text = element_text(size = 12),
        axis.text.x = element_text(angle = 45, hjust = 0.9))

```

## Variable correlations
```{r correlations, fig.show = "hold", out.width = "50%", fig.height = 12, fig.width = 15}

# calculate spearman correlations in each group

f_calc_cor <- function(df, status) {

  df_hbox_impute %>%
    filter(Status == status) %>% 
    dplyr::select(-c(subject_id, Status)) %>% 
    cor(method = "spearman")

}

m_cor_hv <- df_hbox_impute %>% f_calc_cor(status = "HV")

m_cor_um <- df_hbox_impute %>% f_calc_cor(status = "UM")

m_cor_cm <- df_hbox_impute %>% f_calc_cor(status = "CM")

# correlation matrix

  library(reshape)
  library(ggpubr)

  # write plot function
  f_plot_cor <- function(mat) {

    melt(mat) %>%

      ggplot(aes(x = X1, y = X2, fill = value, label = round(value, 2))) + 
      geom_tile() +
      scale_fill_gradient2(low = "#4575b4", high = "#d73027") +
      geom_text(size = 6) +

      xlab("") +
      ylab("") +
      theme(axis.text.x = element_text(angle = 45, hjust = 0.9),
            axis.text = element_text(size = 25),
            legend.text = element_text(size = 40),
            legend.title = element_text(size = 50),
            legend.key.size = unit(2, 'cm'),
            plot.title = element_text(size = 30)
            )

  }

  # extract legend
  p_leg <- m_cor_hv %>% f_plot_cor() %>% get_legend()

  # create plots
  m_cor_hv %>% f_plot_cor() + ggtitle("HV variable correlations") + theme(legend.position = "none")

  m_cor_um %>% f_plot_cor() + ggtitle("UM variable correlations") + theme(legend.position = "none")

  m_cor_cm %>% f_plot_cor() + ggtitle("CM variable correlations") + theme(legend.position = "none")

  as_ggplot(p_leg)

```

## Logistic regression {.tabset}

### Explore data
```{r log_plot}

df_hbox_voi_01 %>% 
  pivot_longer(2:10, names_to = "variable", values_to = "value") %>% 
  mutate(Status = as.numeric(as.character(Status))) %>%

  ggplot(aes(x = value, y = Status)) +
  geom_point() +
  geom_smooth(method = "glm", method.args = list(family = "binomial")) +
  facet_wrap(~variable, scales = "free_x") +

  ylab("Status: 1 = CM, 0 = UM") +
  xlab("") +
  theme_bw()

```

### Model 1

#### Build

- logistic regression model
- variables: DeoxyHb, DeoxyHb_alpha, OxyHb, OxyHb_alpha, O2sat, O2sat_alpha, THC, THC_alpha, THC_sd
- bootstrap x100
- BoxCox transformation

```{r split_boot}

library(tidymodels)
library(rsample)

# split data
set.seed(123)
hbox_split <- initial_split(df_hbox_voi_01, strata = Status)
hbox_train <- training(hbox_split)
hbox_test <- training(hbox_split)

# resample
set.seed(234)
hbox_boot <- bootstraps(hbox_train, times = 100)

```

```{r model_spec}

# recipe
hbox_rec <- recipe(Status ~ ., data = hbox_train) %>% 
  step_BoxCox(all_predictors()) %>% 
  step_nzv(all_predictors())

# workflow
hbox_wf <- workflow() %>% 
  add_recipe(hbox_rec)

# logistic regression model spec
glm_spec <- logistic_reg() %>% 
  set_engine("glm")

```

```{r train_model}

glm_res <- hbox_wf %>% 
  add_model(glm_spec) %>%
  fit_resamples(
    resamples = hbox_boot,
    metrics = metric_set(roc_auc, accuracy, sensitivity, specificity),
    control = tune::control_resamples(save_pred = TRUE)
  )

```

#### Evaluate

Training data

```{r eval_training}

# collect metrics
collect_metrics(glm_res) %>% 
  dplyr::select(.metric, mean, n, std_err) %>% 
  as.data.frame()

# confusion matrix on training data
glm_res %>% 
  conf_mat_resampled() %>% 
  mutate(Prediction = ifelse(Prediction == 1, "CM", "UM")) %>% 
  mutate(Truth = ifelse(Truth == 1, "CM", "UM")) %>%

  # plot
  ggplot(aes(x = Truth, y = Prediction, fill = Freq, label = Freq)) +
  geom_tile() +
  geom_text(size = 6) +
  scale_fill_gradient2(low = "#4575b4", high = "#d73027") +

  ggtitle("Confusion matrix: training data") +
  theme_bw()

```

```{r roc}

glm_res %>% 
  collect_predictions() %>% 
  group_by(id) %>% 
  roc_curve(Status, .pred_0) %>%

  ggplot(aes(1 - specificity, sensitivity, color = id)) +
  geom_abline(lty = 2, color = "gray60", size = 1.5) +
  geom_path(show.legend = FALSE, alpha = 0.6, size = 1.2) +
  coord_equal() +
  ggtitle("ROC") +
  theme_bw()

```

Testing data
```{r test}

# fit to test data
hbox_final <- hbox_wf %>% 
  add_model(glm_spec) %>% 
  last_fit(hbox_split)

# compute metrics on testing data
collect_metrics(hbox_final) %>% 
  dplyr::select(-c(.estimator, .config)) %>% 
  as.data.frame()

# predictions
collect_predictions(hbox_final) %>% 
  count(.pred_class, Status) %>% 
  dplyr::rename("Prediction" = ".pred_class", "Truth" = "Status", "Freq" = "n") %>%  
  mutate(Prediction = ifelse(Prediction == 1, "CM", "UM")) %>% 
  mutate(Truth = ifelse(Truth == 1, "CM", "UM")) %>%

  # plot
  ggplot(aes(x = Truth, y = Prediction, fill = Freq, label = Freq)) +
  geom_tile() +
  geom_text(size = 6) +
  scale_fill_gradient2(low = "#4575b4", high = "#d73027") +

  ggtitle("Confusion matrix: testing data") +
  theme_bw()

```

Coefficients
```{r coefficients}

hbox_final$.workflow[[1]] %>% 
  tidy() %>% 
  as.data.frame()

```

Odds ratios
```{r OR}

# OR
hbox_final$.workflow[[1]] %>% 
  tidy(exponentiate = TRUE) %>% 
  arrange(desc(estimate)) %>% 
  as.data.frame()

```

```{r plot_voi}

df_hbox_log_voi %>%

  ggplot(aes(x = THC_alpha, y = THC_sd, color = Status, fill = Status, size = O2sat_alpha)) +
  geom_point(alpha = 0.7) +
  stat_conf_ellipse(level = 0.95, npoint = 100, bary = TRUE, alpha = 0.1, geom = "polygon") +
  stat_mean(shape = 1, na.rm = FALSE) +
  ggtitle("VOI") +
  theme_bw()

```

### Model 2

#### Build

- logistic regression model
- variables: THC_alpha, THC_sd (O2sat_alpha was NS when included in model, p = 0.42)
- bootstrap x100
- BoxCox transformation

```{r split_boot2}

library(tidymodels)
library(rsample)

# select only significant variables
df_hbox_model2 <- df_hbox_voi_01 %>% dplyr::select(c(Status, THC_alpha, THC_sd))

# split data
set.seed(456)
hbox_split2 <- initial_split(df_hbox_model2, strata = Status)
hbox_train2 <- training(hbox_split2)
hbox_test2 <- training(hbox_split2)

# resample
set.seed(789)
hbox_boot2 <- bootstraps(hbox_train2, times = 100)

```

```{r model_spec2}

# recipe
hbox_rec2 <- recipe(Status ~ ., data = hbox_train2) %>% 
  step_BoxCox(all_predictors()) %>% 
  step_nzv(all_predictors())

# workflow
hbox_wf2 <- workflow() %>% 
  add_recipe(hbox_rec2)

# logistic regression model spec
glm_spec2 <- logistic_reg() %>% 
  set_engine("glm")

```

```{r train_model2}

glm_res2 <- hbox_wf2 %>% 
  add_model(glm_spec2) %>%
  fit_resamples(
    resamples = hbox_boot2,
    metrics = metric_set(roc_auc, accuracy, sensitivity, specificity),
    control = tune::control_resamples(save_pred = TRUE)
  )

```

#### Evaluate

Training data

```{r eval_training2}

# collect metrics
collect_metrics(glm_res2) %>% 
  dplyr::select(.metric, mean, n, std_err) %>% 
  as.data.frame()

# confusion matrix on training data
glm_res2 %>% 
  conf_mat_resampled() %>% 
  mutate(Prediction = ifelse(Prediction == 1, "CM", "UM")) %>% 
  mutate(Truth = ifelse(Truth == 1, "CM", "UM")) %>%

  # plot
  ggplot(aes(x = Truth, y = Prediction, fill = Freq, label = Freq)) +
  geom_tile() +
  geom_text(size = 6) +
  scale_fill_gradient2(low = "#4575b4", high = "#d73027") +

  ggtitle("Confusion matrix: training data") +
  theme_bw()

```

```{r roc2}

glm_res2 %>% 
  collect_predictions() %>% 
  group_by(id) %>% 
  roc_curve(Status, .pred_0) %>%

  ggplot(aes(1 - specificity, sensitivity, color = id)) +
  geom_abline(lty = 2, color = "gray60", size = 1.5) +
  geom_path(show.legend = FALSE, alpha = 0.6, size = 1.2) +
  coord_equal() +
  ggtitle("ROC") +
  theme_bw()

```

Testing data
```{r test2}

# fit to test data
hbox_final2 <- hbox_wf2 %>% 
  add_model(glm_spec2) %>% 
  last_fit(hbox_split2)

# compute metrics on testing data
collect_metrics(hbox_final2) %>% 
  dplyr::select(-c(.estimator, .config)) %>% 
  as.data.frame()

# predictions
collect_predictions(hbox_final2) %>% 
  count(.pred_class, Status) %>% 
  dplyr::rename("Prediction" = ".pred_class", "Truth" = "Status", "Freq" = "n") %>%  
  mutate(Prediction = ifelse(Prediction == 1, "CM", "UM")) %>% 
  mutate(Truth = ifelse(Truth == 1, "CM", "UM")) %>%

  # plot
  ggplot(aes(x = Truth, y = Prediction, fill = Freq, label = Freq)) +
  geom_tile() +
  geom_text(size = 6) +
  scale_fill_gradient2(low = "#4575b4", high = "#d73027") +

  ggtitle("Confusion matrix: testing data") +
  theme_bw()

```

Coefficients
```{r coefficients2}

hbox_final2$.workflow[[1]] %>% 
  tidy() %>% 
  as.data.frame()

```

Odds ratios
```{r OR2}

# OR
hbox_final2$.workflow[[1]] %>% 
  tidy(exponentiate = TRUE) %>% 
  arrange(desc(estimate)) %>% 
  as.data.frame()

```

```{r plot_voi2}

df_hbox_model2 %>%

  ggplot(aes(x = THC_alpha, y = THC_sd, color = Status, fill = Status)) +
  geom_point(alpha = 0.7) +
  stat_conf_ellipse(level = 0.95, npoint = 100, bary = TRUE, alpha = 0.1, geom = "polygon") +
  stat_mean(shape = 1, na.rm = FALSE) +
  ggtitle("VOI") +
  theme_bw()

```

### 2021-05-06 log_reg_analysis.Rmd
updated logistic regression analysis - no split (include all data), bootstrap x1000

@@
 knitr::opts_chunk$set(echo = FALSE, comment = "", message = FALSE, warning = FALSE, fig.height = 8, fig.width = 10)
+
+load("glm_res.Rdata")
 
@@
         axis.text = element_text(size = 15))
+
 ```
@@
 
-  df_hbox_impute %>%
+  df %>%
     filter(Status == status) %>% 
-    dplyr::select(-c(subject_id, Status)) %>% 
+    dplyr::select(-Status) %>% 
     cor(method = "spearman")
@@
 
-m_cor_hv <- df_hbox_impute %>% f_calc_cor(status = "HV")
+m_cor_hv <- df_hbox_impute %>% dplyr::select(c(2, 11:19)) %>% f_calc_cor(status = "HV")
 
-m_cor_um <- df_hbox_impute %>% f_calc_cor(status = "UM")
+m_cor_um <- df_hbox_impute %>% dplyr::select(c(2, 11:19)) %>% f_calc_cor(status = "UM")
 
-m_cor_cm <- df_hbox_impute %>% f_calc_cor(status = "CM")
+m_cor_cm <- df_hbox_impute %>% dplyr::select(c(2, 11:19)) %>% f_calc_cor(status = "CM")
 
@@
       scale_fill_gradient2(low = "#4575b4", high = "#d73027") +
-      geom_text(size = 6) +
+      geom_text(size = 8) +
 
@@
       theme(axis.text.x = element_text(angle = 45, hjust = 0.9),
-            axis.text = element_text(size = 25),
+            axis.text = element_text(size = 35),
             legend.text = element_text(size = 40),
@@
 - variables: DeoxyHb, DeoxyHb_alpha, OxyHb, OxyHb_alpha, O2sat, O2sat_alpha, THC, THC_alpha, THC_sd
-- bootstrap x100
+- bootstrap x1000
 - BoxCox transformation
@@
 
-# split data
-set.seed(123)
-hbox_split <- initial_split(df_hbox_voi_01, strata = Status)
-hbox_train <- training(hbox_split)
-hbox_test <- training(hbox_split)
-
 # resample
 set.seed(234)
-hbox_boot <- bootstraps(hbox_train, times = 100)
+hbox_boot <- bootstraps(df_hbox_voi_01, times = 1000)
 
@@
 # recipe
-hbox_rec <- recipe(Status ~ ., data = hbox_train) %>% 
+hbox_rec <- recipe(Status ~ ., data = df_hbox_voi_01) %>% 
   step_BoxCox(all_predictors()) %>% 
@@
 
-```{r train_model}
+```{r train_model, eval = FALSE}
 
@@
 
+```{r best}
+
+# select best model to be used for evaluation
+glm_best <- glm_res %>% select_best("roc_auc")
+
+# finalize model
+hbox_final <- hbox_wf %>% 
+  add_model(glm_spec) %>% 
+  finalize_workflow(glm_best) %>%
+  fit(df_hbox_voi_01) %>%
+  pull_workflow_fit()
+
+```
+
 #### Evaluate
-
-Training data
 
@@
 
-  ggtitle("Confusion matrix: training data") +
+  ggtitle("Confusion matrix") +
   theme_bw()
@@
 
-```{r roc}
+```{r roc, eval = FALSE}
 
@@
 
-Testing data
-```{r test}
-
-# fit to test data
-hbox_final <- hbox_wf %>% 
-  add_model(glm_spec) %>% 
-  last_fit(hbox_split)
-
-# compute metrics on testing data
-collect_metrics(hbox_final) %>% 
-  dplyr::select(-c(.estimator, .config)) %>% 
-  as.data.frame()
-
-# predictions
-collect_predictions(hbox_final) %>% 
-  count(.pred_class, Status) %>% 
-  dplyr::rename("Prediction" = ".pred_class", "Truth" = "Status", "Freq" = "n") %>%  
-  mutate(Prediction = ifelse(Prediction == 1, "CM", "UM")) %>% 
-  mutate(Truth = ifelse(Truth == 1, "CM", "UM")) %>%
-
-  # plot
-  ggplot(aes(x = Truth, y = Prediction, fill = Freq, label = Freq)) +
-  geom_tile() +
-  geom_text(size = 6) +
-  scale_fill_gradient2(low = "#4575b4", high = "#d73027") +
-
-  ggtitle("Confusion matrix: testing data") +
-  theme_bw()
-
-```
-
 Coefficients
@@
 
-hbox_final$.workflow[[1]] %>% 
+hbox_final %>% 
   tidy() %>% 
@@
 # OR
-hbox_final$.workflow[[1]] %>% 
+hbox_final %>% 
   tidy(exponentiate = TRUE) %>% 
@@
   as.data.frame()
-
-```
-
-```{r plot_voi}
-
-df_hbox_log_voi %>%
-
-  ggplot(aes(x = THC_alpha, y = THC_sd, color = Status, fill = Status, size = O2sat_alpha)) +
-  geom_point(alpha = 0.7) +
-  stat_conf_ellipse(level = 0.95, npoint = 100, bary = TRUE, alpha = 0.1, geom = "polygon") +
-  stat_mean(shape = 1, na.rm = FALSE) +
-  ggtitle("VOI") +
-  theme_bw()
 
@@
 - logistic regression model
-- variables: THC_alpha, THC_sd (O2sat_alpha was NS when included in model, p = 0.42)
-- bootstrap x100
+- variables: THC_alpha, THC_sd
+- bootstrap x1000
 - BoxCox transformation
@@
 
-# split data
-set.seed(456)
-hbox_split2 <- initial_split(df_hbox_model2, strata = Status)
-hbox_train2 <- training(hbox_split2)
-hbox_test2 <- training(hbox_split2)
-
 # resample
 set.seed(789)
-hbox_boot2 <- bootstraps(hbox_train2, times = 100)
+hbox_boot2 <- bootstraps(df_hbox_model2, times = 1000)
 
@@
 # recipe
-hbox_rec2 <- recipe(Status ~ ., data = hbox_train2) %>% 
+hbox_rec2 <- recipe(Status ~ ., data = df_hbox_model2) %>% 
   step_BoxCox(all_predictors()) %>% 
@@
 
-```{r train_model2}
+```{r train_model2, eval = FALSE}
 
@@
     metrics = metric_set(roc_auc, accuracy, sensitivity, specificity),
-    control = tune::control_resamples(save_pred = TRUE)
+    control = tune::control_resamples(save_pred = TRUE, verbose = TRUE)
   )
+
+save(glm_res, glm_res2, file = "glm_res.Rdata")
+
+```
+
+```{r best2}
+
+# select best model to be used for evaluation
+glm_best2 <- glm_res2 %>% select_best("roc_auc")
+
+# finalize model
+hbox_final2 <- hbox_wf2 %>% 
+  add_model(glm_spec2) %>% 
+  finalize_workflow(glm_best2) %>%
+  fit(df_hbox_model2) %>%
+  pull_workflow_fit()
 
@@
 #### Evaluate
-
-Training data
 
@@
 
-```{r roc2}
+```{r roc2, eval = FALSE}
 
@@
 
-Testing data
-```{r test2}
-
-# fit to test data
-hbox_final2 <- hbox_wf2 %>% 
-  add_model(glm_spec2) %>% 
-  last_fit(hbox_split2)
-
-# compute metrics on testing data
-collect_metrics(hbox_final2) %>% 
-  dplyr::select(-c(.estimator, .config)) %>% 
-  as.data.frame()
-
-# predictions
-collect_predictions(hbox_final2) %>% 
-  count(.pred_class, Status) %>% 
-  dplyr::rename("Prediction" = ".pred_class", "Truth" = "Status", "Freq" = "n") %>%  
-  mutate(Prediction = ifelse(Prediction == 1, "CM", "UM")) %>% 
-  mutate(Truth = ifelse(Truth == 1, "CM", "UM")) %>%
-
-  # plot
-  ggplot(aes(x = Truth, y = Prediction, fill = Freq, label = Freq)) +
-  geom_tile() +
-  geom_text(size = 6) +
-  scale_fill_gradient2(low = "#4575b4", high = "#d73027") +
-
-  ggtitle("Confusion matrix: testing data") +
-  theme_bw()
-
-```
-
 Coefficients
@@
 
-hbox_final2$.workflow[[1]] %>% 
+hbox_final2 %>% 
   tidy() %>% 
@@
 
-# OR
-hbox_final2$.workflow[[1]] %>% 
+hbox_final2 %>% 
   tidy(exponentiate = TRUE) %>% 
@@
 
-```{r plot_voi2}
+## THC_alpha & THC_sd
 
-df_hbox_model2 %>%
+```{r violin_plot}
 
-  ggplot(aes(x = THC_alpha, y = THC_sd, color = Status, fill = Status)) +
-  geom_point(alpha = 0.7) +
-  stat_conf_ellipse(level = 0.95, npoint = 100, bary = TRUE, alpha = 0.1, geom = "polygon") +
-  stat_mean(shape = 1, na.rm = FALSE) +
-  ggtitle("VOI") +
+df_hbox_log_voi %>% 
+  pivot_longer(2:10, names_to = "variable", values_to = "value") %>% 
+  filter(variable %in% c("THC_alpha", "THC_sd")) %>%
+
+  ggplot(aes(x = Status, y = value)) +
+  geom_violin(aes(color = Status)) +
+  geom_jitter(position = position_jitter(0.2), shape = 1) +
+  stat_summary(fun = "median", geom = "crossbar", aes(color = Status), size = 0.2, width = 0.5) +
+  facet_wrap(~variable, scales = "free_y") +
+
+  ggtitle("Violin plot") +
+  theme_bw() +
+  theme(legend.position = "none")  +
+
+  theme(strip.text = element_text(size = 15),
+        axis.title = element_text(size = 15),
+        axis.title.y = element_blank(),
+        axis.text = element_text(size = 15))
+
+```
+
+```{r box_plot}
+
+# calculate quantiles of THC_alpha scores
+Q1 <- quantile(df_hbox_log_voi$THC_alpha, 0.25)
+Q2 <- quantile(df_hbox_log_voi$THC_alpha, 0.50)
+Q3 <- quantile(df_hbox_log_voi$THC_alpha, 0.75)
+
+df_hbox_log_voi %>% 
+  dplyr::select(c("Status", "THC_alpha", "THC_sd")) %>%
+
+  # assign quantiles to THC_alpha scores
+  mutate(THCa_quartile = case_when(
+
+    THC_alpha < Q1 ~ 1,
+    Q1 < THC_alpha & THC_alpha < Q2 ~ 2,
+    Q2 < THC_alpha & THC_alpha <= Q3 ~ 3,
+    Q3 < THC_alpha ~ 4
+
+  ) %>% as.factor()) %>%
+
+  # plot
+  ggplot(aes(x = Status, y = THC_sd)) +
+  geom_boxplot(show.legend = FALSE, outlier.shape = NA) +
+  geom_point(aes(color = THCa_quartile), position = position_jitter(0.2), size = 3) +
+  stat_summary(aes(color = THCa_quartile), fun = "median", geom = "crossbar", size = 0.3, width = 0.5) +
+  scale_color_manual(values = c("#F8766D", "#7CAE00", "#00BFC4", "#C77CFF", "black", "black")) +
+
+  ggtitle("Relationship between THC_sd and THC_alpha") +
   theme_bw()
@@
 ```
+
+```{r scatter_plot}
+
+cm_model <- lm(THC_sd ~ THC_alpha, data = df_hbox_log_voi %>% filter(Status == "CM"))
+um_model <- lm(THC_sd ~ THC_alpha, data = df_hbox_log_voi %>% filter(Status == "UM"))
+
+df_hbox_log_voi %>%
+
+  ggplot(aes(x = THC_alpha, y = THC_sd, color = Status, fill = Status)) +
+  geom_point(size = 3, alpha = 0.7) +
+  geom_smooth(method = "lm", se = FALSE) +
+
+  geom_text(label = paste0("y = ", 
+                           round(cm_model$coefficients[[1]], 2), 
+                           " + ", 
+                           round(cm_model$coefficients[[2]], 2), 
+                           "x ; R^2 = ", 
+                           round(summary(cm_model)$adj.r.squared, 2)),
+            x = 1.3, y = 0.55,
+            color = "#00BFC4",
+            size = 6) +
+
+    geom_text(label = paste0("y = ", 
+                           round(um_model$coefficients[[1]], 2), 
+                           " + ", 
+                           round(um_model$coefficients[[2]], 2), 
+                           "x ; R^2 = ", 
+                           round(summary(um_model)$adj.r.squared, 2)),
+            x = 1.3, y = 0.6,
+            color = "#F8766D",
+            size = 6) +
+
+  ggtitle("Scatter plot") +
+  theme_bw()
+
+```

### 2021-05-07 log_reg_analysis.Rmd
add PCA

@@
 
-## Variable correlations
+## PCA & variable correlations
+```{r pca}
+
+library(factoextra)
+
+# convert tibble to df
+df_hbox_voi_df <- df_hbox_log_voi %>% dplyr::select(-Status) %>% as.data.frame() 
+rownames(df_hbox_voi_df) <- df_hbox_impute %>% dplyr::filter(Status != "HV") %>% .$subject_id
+
+# compute PCA
+my_pca <- prcomp(df_hbox_voi_df, scale = TRUE)
+
+# define groups
+groups <- as.factor(df_hbox_log_voi$Status)
+
+# Eigenvalues
+pca_eig_val <- get_eigenvalue(my_pca)
+
+# two data frames for plotting
+df_vars <- my_pca$rotation %>% as_tibble() %>% mutate(variable = rownames(my_pca$rotation))
+df_pnts <- my_pca$x %>% as_tibble() %>% mutate(group = groups)
+
+# plot
+
+library(ggpubr)
+library(ggrepel)
+
+  # plot patients
+  ggplot() +
+  geom_point(data = df_pnts, mapping = aes(x = PC1, y = PC2, color = group), 
+             size = 3) +
+  stat_conf_ellipse(data = df_pnts, mapping = aes(x = PC1, y = PC2, color = group, fill = group),
+                    level = 0.95, npoint = 100, bary = TRUE, alpha = 0.1, geom = "polygon") +
+  stat_mean(data = df_pnts, mapping = aes(x = PC1, y = PC2, color = group, shape = group),
+            na.rm = FALSE) +
+
+  # plot variables
+  geom_point(data = df_vars, aes(x = PC1*10, y = PC2*10)) +
+  geom_text_repel(data = df_vars, aes(x = PC1*10, y = PC2*10, label = variable, size = 10), show.legend = FALSE) +
+  geom_segment(data = df_vars, aes(x = 0, y = 0, xend = PC1*10, yend = PC2*10), arrow = arrow()) +
+
+  # draw lines
+  geom_vline(xintercept = 0, lty = 2, color = "black") +
+  geom_hline(yintercept = 0, lty = 2, color = "black") +
+
+  # design
+  xlab(paste0("PC1: ", round(pca_eig_val$variance.percent[1], 2), "% of variance")) +
+  ylab(paste0("PC1: ", round(pca_eig_val$variance.percent[2], 2), "% of variance")) +
+
+  ggtitle("PCA") +
+  theme_bw()
+
+```
+
 ```{r correlations, fig.show = "hold", out.width = "50%", fig.height = 12, fig.width = 15}

### 2021-05-11 figures.Rmd
manuscript figures draft 1

---
title: "Dysregulation of Cerebral Hemoglobin Content in Malawian Children with Malaria"
author: "Rachel Smith"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include = FALSE}

knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, comment = "", fig.height = 8, fig.width = 10)

library(tidyverse)
library(janitor)
library(factoextra)
library(ggrepel)
library(ggpubr)
library(tidymodels)
library(vip)
library(scales)
library(reshape)

load("model1.Rdata")
load("model2.Rdata")

```

```{r data}

# read in data
df_hbox <- read.csv("Data/98 first visit records with cerebral oxygenation stats_trimmed.csv") %>% as_tibble()

# get rid of "Other"
df_hbox <- df_hbox %>% 
  filter(Status != "Other") %>% 
  mutate(Status = factor(Status, levels = c("HV", "UM", "CM")))

# identify duplicated rows (TM0003, TM2001)
df_hbox <- df_hbox %>% filter(duplicated(Subject.ID..NIAID.) == FALSE)

# Convert age to numeric
df_hbox <- df_hbox %>% 
  mutate(Age = gsub(" Years, ", "_", Age)) %>% 
  mutate(Age = gsub(" Months ", "", Age)) %>% 
  separate(Age, into = c("Years", "Months"), sep = "_") %>% 
  mutate(Years = as.numeric(Years)) %>% 
  mutate(Months = as.numeric(Months)/12) %>% 
  mutate(Age = Years + Months) %>% 
  dplyr::select(-c(Years, Months))

# Convert sex to factor
df_hbox <- df_hbox %>% mutate(Sex = as.factor(Sex))

# Trim for variables of interest (VOI)
df_hbox_trim <- df_hbox %>% 
  dplyr::select(c(Subject.ID..NIAID., Status, Glucose, Hematocrit, Lactate, Temperature,
                  Arginine.umol.L, Haptoglobin..mg.dl., Hemoglobin..uM., Whole.Blood.Nitrite, 
                  cerebral.hbdeox, cerebral.hbdeox.alpha, cerebral.hbox, cerebral.hbox.alpha,
                  cerebral.sat, cerebral.sat.alpha, cerebral.thc, cerebral.thc.alpha, cerebral.thc.norm.std)) %>%

  dplyr::rename("subject_id" = "Subject.ID..NIAID.", "Arginine (umol/L)" = "Arginine.umol.L", "Haptoglobin (mg/dL)" = "Haptoglobin..mg.dl.", 
                "Hemoglobin (uM)" = "Hemoglobin..uM.", "WB_Nitrite" = "Whole.Blood.Nitrite", "DeoxyHb" = "cerebral.hbdeox", 
                "DeoxyHb_alpha" = "cerebral.hbdeox.alpha", "OxyHb" = "cerebral.hbox", "OxyHb_alpha" = "cerebral.hbox.alpha", 
                "O2sat" = "cerebral.sat", "O2sat_alpha" = "cerebral.sat.alpha", "THC" = "cerebral.thc", 
                "THC_alpha" = "cerebral.thc.alpha", "THC_sd" = "cerebral.thc.norm.std")

# impute missing data with median by group
f_impute <- function(x, na.rm = TRUE) (replace(x, is.na(x), median(x, na.rm = na.rm)))

df_hbox_impute <- df_hbox_trim %>% 
  group_by(Status) %>% 
  mutate_at(3:ncol(.), f_impute) %>% 
  ungroup()

# pivot long
df_hbox_long <- df_hbox_trim %>% pivot_longer(cols = 3:ncol(.), names_to = "measurement", values_to = "value")

```

### Table 1: Summary of hematological data by patient group

```{r table1}

# number of subjects 
df_status <- df_hbox_impute %>% 
  count(Status) %>% 
  pivot_wider(1:2, names_from = "Status", values_from = "n") %>% 
  mutate(variable = "Number of subjects") %>% 
  dplyr::select(variable, everything()) %>% 
  mutate_if(is.numeric, as.character)

# Sex
df_sex <- df_hbox %>% 
  group_by(Status) %>% 
  count(Sex) %>% 
  filter(Sex != "") %>% 
  pivot_wider(names_from = "Sex", values_from = "n") %>% 
  mutate(perc_fem = round(Female/(Female + Male)*100, 2)) %>%

  pivot_wider(c(1,4), names_from = "Status", values_from = "perc_fem") %>% 
  mutate(variable = "Sex (% female)") %>% 
  dplyr::select(variable, everything()) %>% 
  mutate_if(is.numeric, as.character)

# Age, hematocrit, and lactate
df_hbox %>% 
  group_by(Status) %>% 
  summarise(`Age (years)` = median(Age, na.rm = TRUE), age_range = range(Age, na.rm = TRUE), 
            `Hematocrit (%)` = median(Hematocrit, na.rm = TRUE), hem_range = range(Hematocrit, na.rm = TRUE),
            `Lactate (mmol/L)` = median(Lactate, na.rm = TRUE), lactate_range = range(Lactate, na.rm = TRUE),
            `Arginine (umol/L)` = median(Arginine.umol.L, na.rm = TRUE), arg_range = range(Arginine.umol.L, na.rm = TRUE),
            `Temperature (°C)` = median(Temperature, na.rm = TRUE), temp_range = range(Temperature, na.rm = TRUE)) %>% 
  mutate_if(is.numeric, round, 1) %>%

  mutate(`Age (years)` = paste0(`Age (years)`, " (", age_range[1], ", ", age_range[2], ")"),
         `Hematocrit (%)` = paste0(`Hematocrit (%)`, " (", hem_range[1], ", ", hem_range[2], ")"),
         `Lactate (mmol/L)` = paste0(`Lactate (mmol/L)`, " (", lactate_range[1], ", ", lactate_range[2], ")"),
         `Arginine (umol/L)` = paste0(`Arginine (umol/L)`, " (", arg_range[1], ", ", arg_range[2], ")"),
         `Temperature (°C)` = paste0(`Temperature (°C)`, " (", temp_range[1], ", ", temp_range[2], ")")) %>% 
  dplyr::select(-c(age_range, hem_range, lactate_range, arg_range, temp_range)) %>% 
  unique() %>% 
  t() %>% 
  row_to_names(row_number = 1) %>% 
  as_tibble(rownames = "variable") %>% 
  add_row(df_status, .before = 1) %>% 
  add_row(df_sex, .before = 2) %>%

  dplyr::rename(" " = "variable",
                "Healthy volunteers (HV)" = "HV",
                "Uncomplicated malaria patients (UM)" = "UM",
                "Cerebral malaria patients (CM)" = "CM") %>% 
  knitr::kable(align = rep('c', 4))

```
**Table 1.** A summary of subject information and hematological data. Group patient data presented as median (range). Note: sex not recorded for 3 CM patients.

### Figure 1: Visual summary of NIRS variables

```{r fig1}

df_hbox_voi <- df_hbox_impute %>% 
  dplyr::select(c(Status, DeoxyHb, DeoxyHb_alpha, OxyHb, OxyHb_alpha, O2sat, O2sat_alpha, THC, THC_alpha, THC_sd)) %>% 
  dplyr::filter(DeoxyHb_alpha < 3) %>% 
  dplyr::rename("THC sd" = "THC_sd", "THCα" = "THC_alpha", "Deoxy Hb" = "DeoxyHb", "Oxy Hb" = "OxyHb", 
                "O2 sat" = "O2sat", "O2 sat α" = "O2sat_alpha", "Deoxy Hb α" = "DeoxyHb_alpha", "Oxy Hb α" = "OxyHb_alpha")

df_hbox_voi_long <- df_hbox_voi %>% 
  pivot_longer(2:10, names_to = "variable", values_to = "value") %>%
  mutate(variable = factor(variable, levels = c("Deoxy Hb", "Oxy Hb", "O2 sat", "THC", "THC sd", 
                                                "Deoxy Hb α", "Oxy Hb α", "O2 sat α", "THCα")))

my_col <- c("CM" = "#00BFC4", "HV" = "#7CAE00", "UM" = "#F8766D")

df_hbox_voi_long %>%

  ggplot(aes(x = Status, y = value)) +
  geom_violin(aes(color = Status)) +
  geom_jitter(position = position_jitter(0.2), shape = 1) +
  stat_summary(fun = "median", geom = "crossbar", aes(color = Status), size = 0.2, width = 0.5) +
  facet_wrap(~variable, scales = "free_y") +

  scale_color_manual(values = my_col) +
  theme_bw() +
  theme(legend.position = "none")  +

  theme(strip.text = element_text(size = 15),
        axis.title = element_text(size = 15),
        axis.title.y = element_blank(),
        axis.text = element_text(size = 15))

```
**Figure 1.** Differences in variables measured by Oxiplex and associated alpha values among patient groups. Note: removed one outlier from DeoxyHb alpha (3.52)

### Figure 2: Random forest modeling using all outcomes and predictors

#### A. PCA
```{r fig2a}

# convert tibble to df
df_hbox_impute_df <- df_hbox_impute %>% 
  dplyr::select(-c(subject_id, Status)) %>% 
  dplyr::rename("Whole blood nitrite" = "WB_Nitrite", "THC sd" = "THC_sd", "THCα" = "THC_alpha", 
                "Deoxy Hb" = "DeoxyHb", "Oxy Hb" = "OxyHb", "O2 sat" = "O2sat", "O2 sat α" = "O2sat_alpha",
                "Deoxy Hb α" = "DeoxyHb_alpha", "Oxy Hb α" = "OxyHb_alpha",
                "Hemoglobin" = "Hemoglobin (uM)", "Haptoglobin" = "Haptoglobin (mg/dL)", "Arginine" = "Arginine (umol/L)") %>% 
  as.data.frame(df_hbox_impute) 
rownames(df_hbox_impute_df) <- df_hbox_impute$subject_id

# compute PCA
my_pca1 <- prcomp(df_hbox_impute_df, scale = TRUE)

# define groups
groups1 <- as.factor(df_hbox_impute$Status)

# Eigenvalues
pca_eig_val1 <- get_eigenvalue(my_pca1)

# extract variables
df_vars1 <- my_pca1$rotation %>% as_tibble() %>% mutate(variable = rownames(my_pca1$rotation))

# extract patients
df_pnts1 <- my_pca1$x %>% as_tibble() %>% mutate(group = groups1)

# plot

  # plot patients
  ggplot() +
  geom_point(data = df_pnts1, 
             mapping = aes(x = PC1, y = PC2, color = group), 
             size = 3) +
  stat_conf_ellipse(data = df_pnts1, 
                    mapping = aes(x = PC1, y = PC2, color = group, fill = group),
                    level = 0.95, npoint = 100, bary = TRUE, alpha = 0.1, geom = "polygon") +
  stat_mean(data = df_pnts1, 
            mapping = aes(x = PC1, y = PC2, color = group, shape = group),
            na.rm = FALSE) +

  # plot variables
  geom_point(data = df_vars1, aes(x = PC1*10, y = PC2*10)) +
  geom_text_repel(data = df_vars1, aes(x = PC1*10, y = PC2*10, label = variable, size = 10), show.legend = FALSE) +
  geom_segment(data = df_vars1, aes(x = 0, y = 0, xend = PC1*10, yend = PC2*10), arrow = arrow()) +

  # draw lines
  geom_vline(xintercept = 0, lty = 2, color = "black") +
  geom_hline(yintercept = 0, lty = 2, color = "black") +

  # design
  scale_color_manual(values = my_col) +
  scale_fill_manual(values = my_col) +
  xlab(paste0("PC1: ", round(pca_eig_val1$variance.percent[1], 2), "% of variance")) +
  ylab(paste0("PC2: ", round(pca_eig_val1$variance.percent[2], 2), "% of variance")) +

  theme_bw() +
  theme(axis.text = element_text(size = 20),
        axis.title = element_text(size = 25),
        legend.text = element_text(size = 15),
        legend.title = element_text(size = 20))

```

#### B. Confusion matrix
```{r fig2b_prep, eval = FALSE}

# Data
df_hbox_model1 <- df_hbox_trim %>% 
  dplyr::select(-subject_id) %>% 
  dplyr::rename("Whole blood nitrite" = "WB_Nitrite", "THC sd" = "THC_sd", "THCα" = "THC_alpha", 
                "Deoxy Hb" = "DeoxyHb", "Oxy Hb" = "OxyHb", "O2 sat" = "O2sat", "O2 sat α" = "O2sat_alpha",
                "Deoxy Hb α" = "DeoxyHb_alpha", "Oxy Hb α" = "OxyHb_alpha",
                "Hemoglobin" = "Hemoglobin (uM)", "Haptoglobin" = "Haptoglobin (mg/dL)", "Arginine" = "Arginine (umol/L)")

# Build a model

  # resamples (not a lot of data, need to use cross validation)
  set.seed(123)
  hbox_boot1 <- bootstraps(df_hbox_model1, strata = Status)

  # Create recipe
  ranger_recipe <- 
    recipe(formula = Status ~ ., data = df_hbox_model1) %>%

    # BoxCox transformation to normality
    step_BoxCox(all_predictors(), -all_outcomes()) %>%

    # remove variables with non-zero variance
    step_nzv(all_predictors(), -all_outcomes()) %>%

    # impute missing data using K-nearest neighbors
    step_knnimpute(all_predictors(), -all_outcomes())

  # Random forest model specifications (set for tuning)
  ranger_spec <- 
    rand_forest(mtry = tune(), min_n = tune(), trees = 1000) %>% 
    set_mode("classification") %>% 
    set_engine("ranger")

  # Create workflow
  ranger_workflow <- 
    workflow() %>% 
    add_recipe(ranger_recipe) %>% 
    add_model(ranger_spec)

# Tune model

  # Initial tune on bootstraps
  set.seed(84374)
  ranger_tune <-
    tune_grid(ranger_workflow, 
              resamples = hbox_boot1, 
              grid = 11)

# Finalize model

  final_rf <- ranger_workflow %>% 
    finalize_workflow(select_best(ranger_tune, metric = "accuracy"))

  final_rs_res <- final_rf %>% 
    fit_resamples(resamples = hbox_boot1,
                  metrics = metric_set(roc_auc, accuracy, sens, spec),
                  control = control_resamples(save_pred = TRUE))

# save objects so model doesn't rerun every time you knit
  save(df_hbox_model1, hbox_boot1, ranger_recipe, ranger_spec, ranger_tune, final_rs_res, file = "model1.Rdata")

```

```{r fig2b_plot}

final_rs_res %>% 
  conf_mat_resampled() %>%

  # plot
  ggplot(aes(x = Truth, y = Prediction, fill = Freq, label = Freq)) +
  geom_tile() +
  geom_text(size = 8) +
  scale_fill_gradient2(low = "#4575b4", high = "#d73027") +

  theme_bw() +
  theme(axis.title = element_text(size = 25),
        axis.text = element_text(size = 20),
        legend.title = element_text(size = 20),
        legend.text = element_text(size = 15))

```

#### C. Variable importance
```{r fig2c}

# new specification for ranger model
imp_spec <- ranger_spec %>% 
  finalize_model(select_best(ranger_tune, metric = "accuracy")) %>% 
  set_engine("ranger", importance = "permutation")

# plot importance 
workflow() %>% 
  add_recipe(ranger_recipe) %>% 
  add_model(imp_spec) %>% 
  fit(df_hbox_model1) %>% 
  pull_workflow_fit() %>% 
  vip(aesthetics = list(alpha = 0.8, fill = "midnightblue")) +

  theme_bw() +
  theme(axis.text = element_text(size = 20),
        axis.title = element_text(size = 25))

```

**Figure 2.** Random forest modeling demonstrates a high degree of discriminatory power between healthy patients and patients with malaria (uncomplicated and cerebral) using primarily on hematological parameters. **(A)** Principal component analysis (PCA) of study patients using hematological and NIRS data. Variables are plotted as vectors in the direction and magnitude of the PC to which they contribute. Color indicates patient status, ellipses show median and 95% confidence interval on the cluster. **(B)** Confusion matrix representing the results of random forest modeling averaged across 25 bootstrap resamples. Freq indicates the average number of times that the event (shown by the x- and y-axes) occurred. **(C)** Relative contribution of the variables to the discriminatory power of the model (showing top 10 only).

### Figure 3: Logistic regression on CM vs UM

#### A. PCA

```{r fig3a}
# select variables of interest
df_hbox_log_voi <- df_hbox_impute %>% 
  filter(Status != "HV") %>% 
  dplyr::select(2, 11:19)

# convert tibble to df
df_hbox_voi_df <- df_hbox_log_voi %>% 
  dplyr::select(-Status) %>% 
  dplyr::rename("THC sd" = "THC_sd", "THCα" = "THC_alpha", "Deoxy Hb" = "DeoxyHb", "Oxy Hb" = "OxyHb", 
                "O2 sat" = "O2sat", "O2 sat α" = "O2sat_alpha", "Deoxy Hb α" = "DeoxyHb_alpha", "Oxy Hb α" = "OxyHb_alpha") %>% 
  as.data.frame() 
rownames(df_hbox_voi_df) <- df_hbox_impute %>% dplyr::filter(Status != "HV") %>% .$subject_id

# compute PCA
my_pca2 <- prcomp(df_hbox_voi_df, scale = TRUE)

# define groups
groups2 <- as.factor(df_hbox_log_voi$Status)

# Eigenvalues
pca_eig_val2 <- get_eigenvalue(my_pca2)

# two data frames for plotting
df_vars2 <- my_pca2$rotation %>% as_tibble() %>% mutate(variable = rownames(my_pca2$rotation))
df_pnts2 <- my_pca2$x %>% as_tibble() %>% mutate(group = groups2)

# plot

  # plot patients
  ggplot() +
  geom_point(data = df_pnts2, mapping = aes(x = PC1, y = PC2, color = group), 
             size = 3) +
  stat_conf_ellipse(data = df_pnts2, mapping = aes(x = PC1, y = PC2, color = group, fill = group),
                    level = 0.95, npoint = 100, bary = TRUE, alpha = 0.1, geom = "polygon") +
  stat_mean(data = df_pnts2, mapping = aes(x = PC1, y = PC2, color = group, shape = group),
            na.rm = FALSE) +

  # plot variables
  geom_point(data = df_vars2, aes(x = PC1*10, y = PC2*10)) +
  geom_text_repel(data = df_vars2, aes(x = PC1*10, y = PC2*10, label = variable, size = 10), show.legend = FALSE) +
  geom_segment(data = df_vars2, aes(x = 0, y = 0, xend = PC1*10, yend = PC2*10), arrow = arrow()) +

  # draw lines
  geom_vline(xintercept = 0, lty = 2, color = "black") +
  geom_hline(yintercept = 0, lty = 2, color = "black") +

  # design
  xlab(paste0("PC1: ", round(pca_eig_val2$variance.percent[1], 2), "% of variance")) +
  ylab(paste0("PC2: ", round(pca_eig_val2$variance.percent[2], 2), "% of variance")) +

  theme_bw() +
  theme(axis.text = element_text(size = 20),
        axis.title = element_text(size = 25),
        legend.text = element_text(size = 15),
        legend.title = element_text(size = 20))

```

#### B. Confusion matrix

```{r fig3b_prep, eval = FALSE}

# convert to 0s and 1s
df_hbox_voi_01 <- df_hbox_log_voi %>% mutate(Status = ifelse(Status == "CM", 1, 0) %>% as.factor())

# select only significant variables
df_hbox_model2 <- df_hbox_voi_01 %>% 
  dplyr::select(c(Status, THC_alpha, THC_sd)) %>% 
  dplyr::rename("THCα" = "THC_alpha", "THC sd" = "THC_sd") %>% 
  mutate(`1/10 * THCα` = THCα/10) %>% 
  dplyr::select(-THCα)

# resample
set.seed(789)
hbox_boot2 <- bootstraps(df_hbox_model2, times = 1000)

# recipe
hbox_rec2 <- recipe(Status ~ ., data = df_hbox_model2) %>% 
  step_BoxCox(all_predictors()) %>% 
  step_nzv(all_predictors())

# workflow
hbox_wf2 <- workflow() %>% 
  add_recipe(hbox_rec2)

# logistic regression model spec
glm_spec2 <- logistic_reg() %>% 
  set_engine("glm")

glm_res2 <- hbox_wf2 %>% 
  add_model(glm_spec2) %>%
  fit_resamples(
    resamples = hbox_boot2,
    metrics = metric_set(roc_auc, accuracy, sensitivity, specificity),
    control = tune::control_resamples(save_pred = TRUE, verbose = TRUE)
  )

# select best model to be used for evaluation
glm_best2 <- glm_res2 %>% select_best("roc_auc")

# finalize model
hbox_final2 <- hbox_wf2 %>% 
  add_model(glm_spec2) %>% 
  finalize_workflow(glm_best2) %>%
  fit(df_hbox_model2) %>%
  pull_workflow_fit()

# save objects so model doesn't rerun 
save(glm_res2, hbox_final2, file = "model2.Rdata")

```

```{r fig3b_plot}

glm_res2 %>% 
  conf_mat_resampled() %>% 
  mutate(Prediction = ifelse(Prediction == 1, "CM", "UM")) %>% 
  mutate(Truth = ifelse(Truth == 1, "CM", "UM")) %>%

  # plot
  ggplot(aes(x = Truth, y = Prediction, fill = Freq, label = Freq)) +
  geom_tile() +
  geom_text(size = 6) +
  scale_fill_gradient2(low = "#4575b4", high = "#d73027") +

  theme_bw() +
  theme(axis.title = element_text(size = 25),
        axis.text = element_text(size = 20),
        legend.title = element_text(size = 20),
        legend.text = element_text(size = 15))

```

**Figure 3** Logistic regression model discriminating CM vs UM patients using only THCα and THC sd as predictors. **(A)** PCA of patients with CM and UM as determined by NIRS variables. Color indicates patient status, ellipses show median and 95% confidence interval on the cluster. **(B)** Confusion matrix representing the results of the logistic regression model averaged across 1000 bootstrap resamples. Freq indicates the average number of times that the event (shown by the x- and y-axes) occurred.

### Table 2. Logistic regression model evaluation

#### A. Model metrics
```{r table2a}

collect_metrics(glm_res2) %>% 
  dplyr::select(.metric, mean, std_err) %>% 
  mutate(mean = mean*100, std_err = std_err*100) %>% 
  mutate(Metric = c("Accuracy", "ROC AUC", "Sensitivity", "Specificity")) %>% 
  dplyr::select(c("Metric", "mean", "std_err")) %>%

  dplyr::rename("Mean across resamples" = "mean", "Standard error" = "std_err") %>% 
  mutate_if(is.numeric, round, 2) %>% 
  as.data.frame() %>%

  knitr::kable(align = rep('c', 4))

```

#### B. Coefficients
```{r table2b}

hbox_final2 %>% 
  tidy() %>% 
  mutate(p.value = as.character(p.value)) %>% 
  mutate_if(is.numeric, round, 2) %>% 
  mutate(p.value = as.numeric(p.value)) %>% 
  mutate(p.value = scientific(p.value, digits = 3)) %>% 
  dplyr::rename("Term" = "term", "Coefficient" = "estimate", "Standard error" = "std.error", 
                "Statistic" = "statistic", "p-value" = "p.value") %>%

  as.data.frame() %>%
  knitr::kable(align = rep('c', 5))

```

**Table 2.** Evaluation of the logistic regression model across 1000 bootstrap resamples.

### Figure 4: Relationship between THCα and THC sd

```{r fig4}

cm_model <- lm(THC_sd ~ THC_alpha, data = df_hbox_log_voi %>% filter(Status == "CM"))
um_model <- lm(THC_sd ~ THC_alpha, data = df_hbox_log_voi %>% filter(Status == "UM"))

df_hbox_log_voi %>% dplyr::rename("THC sd" = "THC_sd", "THCα" = "THC_alpha") %>%

  ggplot(aes(x = THCα, y = `THC sd`, color = Status, fill = Status)) +
  geom_point(size = 3, alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE) +

  geom_text(label = paste0("y = ", 
                           round(cm_model$coefficients[[1]], 2), 
                           " + ", 
                           round(cm_model$coefficients[[2]], 2), 
                           "x ; R^2 = ", 
                           round(summary(cm_model)$adj.r.squared, 2)),
            x = 1.3, y = 0.55,
            color = "#00BFC4",
            size = 6) +

    geom_text(label = paste0("y = ", 
                           round(um_model$coefficients[[1]], 2), 
                           " + ", 
                           round(um_model$coefficients[[2]], 2), 
                           "x ; R^2 = ", 
                           round(summary(um_model)$adj.r.squared, 2)),
            x = 1.3, y = 0.6,
            color = "#F8766D",
            size = 6) +

  theme_bw() +
  theme(axis.text = element_text(size = 20),
        axis.title = element_text(size = 25),
        legend.text = element_text(size = 15),
        legend.title = element_text(size = 20))

```

**Figure 4.** Scatter plot demonstrating the relationship between THCα and THC sd.

### Supplementary figure 1: Variable correlations 
```{r fig_s1, fig.show = "hold", out.width = "50%", fig.height = 12, fig.width = 15}

# calculate spearman correlations in each group

f_calc_cor <- function(df, status) {

  df %>%
    filter(Status == status) %>% 
    dplyr::select(-Status) %>% 
    cor(method = "spearman")

}

df_hbox_cor <- df_hbox_impute %>% 
  dplyr::rename("THC sd" = "THC_sd", "THCα" = "THC_alpha", "Deoxy Hb" = "DeoxyHb", "Oxy Hb" = "OxyHb", 
                "O2 sat" = "O2sat", "O2 sat α" = "O2sat_alpha", "Deoxy Hb α" = "DeoxyHb_alpha", "Oxy Hb α" = "OxyHb_alpha") %>% 
  dplyr::select(Status, `Deoxy Hb`, `Oxy Hb`, `O2 sat`, THC, `THC sd`, `Deoxy Hb α`, `Oxy Hb α`, `O2 sat α`, THCα)

m_cor_hv <- df_hbox_cor %>% f_calc_cor(status = "HV")

m_cor_um <- df_hbox_cor %>% f_calc_cor(status = "UM")

m_cor_cm <- df_hbox_cor %>% f_calc_cor(status = "CM")

# correlation matrix

# write plot function
f_plot_cor <- function(mat) {

  melt(mat) %>%

    mutate_if(is.factor, factor, levels = c("Deoxy Hb", "Oxy Hb", "O2 sat", "THC", "THC sd", "Deoxy Hb α", "Oxy Hb α", "O2 sat α", "THCα")) %>%

    ggplot(aes(x = X1, y = X2, fill = value, label = round(value, 2))) + 
    geom_tile() +
    scale_fill_gradient2(low = "#4575b4", high = "#d73027") +
    geom_text(size = 9) +

    xlab("") +
    ylab("") +
    theme(axis.text.x = element_text(angle = 45, hjust = 0.9),
          axis.text = element_text(size = 35),
          legend.text = element_text(size = 40),
          legend.title = element_text(size = 50),
          legend.key.size = unit(2, 'cm'),
          plot.title = element_text(size = 30)
    )

}

  # extract legend
  p_leg <- m_cor_hv %>% f_plot_cor() %>% get_legend()

  # create plots
  m_cor_hv %>% f_plot_cor() + ggtitle("HV variable correlations") + theme(legend.position = "none")

  m_cor_um %>% f_plot_cor() + ggtitle("UM variable correlations") + theme(legend.position = "none")

  m_cor_cm %>% f_plot_cor() + ggtitle("CM variable correlations") + theme(legend.position = "none")

  as_ggplot(p_leg)

```

**Figure S1.** Shown are the spearman correlations among variables measured by Oxiplex and their associated alpha values. Red indicates a positive correlation; blue indicates a negative correlation.

### Supplementary figure 2: Logistic regression curves of NIRS variables
```{r fig_s2}

df_hbox_voi_long %>% mutate(Status = as.numeric(ifelse(Status == "CM", 1, 0))) %>%

  ggplot(aes(x = value, y = Status)) +
  geom_point() +
  geom_smooth(method = "glm", method.args = list(family = "binomial")) +
  facet_wrap(~variable, scales = "free_x") +

  ylab("Status: 1 = CM, 0 = UM") +
  xlab("") +

  theme_bw() +
  theme(axis.text = element_text(size = 20),
        axis.title = element_text(size = 25),
        legend.text = element_text(size = 15),
        legend.title = element_text(size = 20),
        strip.text = element_text(size = 20))

```

**Figure S2.** Plotted are the individual logistic regression curves of individual variables tested for their discriminatory power between UM and CM. On the y-axis, 1 represents CM while 0 represents UM. A steeper curve indicates higher disciminatory power

### 2021-05-12 Scripts/raw_data.R
read in raw data, remove blips, bandpass filter

@@ -1,0 +1,749 @@
+
+
+
+# read in data -------------------------------------------------------------------------
+
+  library(tidyverse)
+  library(purrr)
+  
+
+  # read in
+  my_files <- list.files(path = "Data/Raw data/", pattern = "*.txt", full.names = TRUE, recursive = FALSE)
+  
+  l_raw_data <- 1:length(my_files) %>% 
+    purrr::map(~read.delim2(file = my_files[.x], header = FALSE, sep = "\t") %>% as_tibble()) 
+  
+  # name with Status
+  l_names <- 1:length(my_files) %>% 
+    purrr::map(~substr(my_files[.x], start = 22, stop = 23)) 
+  
+  names(l_raw_data) <- l_names
+
+
+  
+# format data -------------------------------------------------------------
+
+
+  library(data.table)
+  library(zoo)
+  
+  # format to just THC in workable form
+  l_thc <- 1:length(l_raw_data) %>% 
+    
+    purrr::map(~mutate(l_raw_data[[.x]], Status = l_names[[.x]])) %>% 
+    purrr::map(~filter(.x, !grepl("Patient|Data|[R|r]aw", V1))) %>% 
+    purrr::map(~dplyr::select(.x, c(ncol(.x), 2,4))) %>% 
+    purrr::map(~dplyr::rename(.x, "Time" = "V2", "THC" = "V4")) %>% 
+    purrr::map(~mutate(.x, Time = as.numeric(Time))) %>% 
+    purrr::map(~mutate(.x, THC = as.numeric(THC)))
+  
+  # renumber patients
+  l_thc <- 1:length(l_thc) %>% 
+    purrr::map(~mutate(l_thc[[.x]], number = c(1:length(l_thc))[.x])) %>% 
+    purrr::map(~dplyr::select(.x, c(number, everything())))
+  
+  # condense list to df
+  df_thc <- rbindlist(l_thc) %>% 
+    as_tibble() 
+  
+  # make sure status is HV, UM, or CM
+  df_thc$Status[df_thc$Status == "HC"] <- "HV"
[…]

### 2021-05-13 figures.Rmd
file rearrange, models 3 & 4

@@
 
-final_rs_res %>% 
-  conf_mat_resampled() %>%
+final_res_conf_mat <- final_rs_res %>% conf_mat_resampled() 
+final_res_conf_mat %>% 
+  group_by(Truth) %>% 
+  summarise(total_true = sum(Freq)) %>% 
+  right_join(final_res_conf_mat) %>% 
+  mutate(`Percent occurence` = round(Freq/total_true*100, 2)) %>%
 
   # plot
-  ggplot(aes(x = Truth, y = Prediction, fill = Freq, label = Freq)) +
+  ggplot(aes(x = Truth, y = Prediction, fill = `Percent occurence`, label = `Percent occurence`)) +
   geom_tile() +
@@
 
-glm_res2 %>% 
+final_res_conf_mat2 <- glm_res2 %>% 
   conf_mat_resampled() %>% 
   mutate(Prediction = ifelse(Prediction == 1, "CM", "UM")) %>% 
-  mutate(Truth = ifelse(Truth == 1, "CM", "UM")) %>%
+  mutate(Truth = ifelse(Truth == 1, "CM", "UM"))
+
+final_res_conf_mat2 %>% 
+  group_by(Truth) %>% 
+  summarise(total_true = sum(Freq)) %>% 
+  right_join(final_res_conf_mat2) %>% 
+  mutate(`Percent occurence` = round(Freq/total_true*100, 2)) %>%
 
   # plot
-  ggplot(aes(x = Truth, y = Prediction, fill = Freq, label = Freq)) +
+  ggplot(aes(x = Truth, y = Prediction, fill = `Percent occurence`, label = `Percent occurence`)) +
   geom_tile() +

### 2021-05-13 Scripts/additional_models.R

@@ -1,0 +1,154 @@
+
+
+# model 3: THCa only -----------------------------------------------------------------
+
+
+  # select only THC_alpha
+  df_hbox_model3 <- df_hbox_voi_01 %>% 
+    dplyr::select(c(Status, THC_alpha)) %>% 
+    dplyr::rename("THCα" = "THC_alpha") %>% 
+    mutate(`1/10 * THCα` = THCα/10) %>% 
+    dplyr::select(-THCα)
+  
+  # resample
+  set.seed(8910)
+  hbox_boot3 <- bootstraps(df_hbox_model3, times = 1000)
+  
+  # recipe
+  hbox_rec3 <- recipe(Status ~ ., data = df_hbox_model3) %>% 
+    step_BoxCox(all_predictors()) %>% 
+    step_nzv(all_predictors())
+  
+  # workflow
+  hbox_wf3 <- workflow() %>% 
+    add_recipe(hbox_rec3)
+  
+  # logistic regression model spec
+  glm_spec3 <- logistic_reg() %>% 
+    set_engine("glm")
+  
+  glm_res3 <- hbox_wf3 %>% 
+    add_model(glm_spec3) %>%
+    fit_resamples(
+      resamples = hbox_boot3,
+      metrics = metric_set(roc_auc, accuracy, sensitivity, specificity),
+      control = tune::control_resamples(save_pred = TRUE, verbose = TRUE)
+    )
+  
+  # select best model to be used for evaluation
+  glm_best3 <- glm_res3 %>% select_best("roc_auc")
+  
+  # finalize model
+  hbox_final3 <- hbox_wf3 %>% 
+    add_model(glm_spec3) %>% 
+    finalize_workflow(glm_best3) %>%
+    fit(df_hbox_model3) %>%
+    pull_workflow_fit()
+  
+  ## Evaluate
+  
+  collect_metrics(glm_res3) %>% 
+    dplyr::select(.metric, mean, std_err) %>% 
+    mutate(mean = mean*100, std_err = std_err*100) %>% 
+    mutate(Metric = c("Accuracy", "ROC AUC", "Sensitivity", "Specificity")) %>% 
+    dplyr::select(c("Metric", "mean", "std_err")) %>% 
+    
+    dplyr::rename("Mean across resamples" = "mean", "Standard error" = "std_err") %>% 
+    mutate_if(is.numeric, round, 2) 
+  
+  hbox_final3 %>% 
+    tidy() %>% 
[…]

### 2021-05-13 Scripts/raw_data_processing.R

@@ -730,18 +730,32 @@
   
   df_thc_filt_pass1 <- df_thc_filt_pass %>% dplyr::filter(number == 1)
   
-  DFA(df_thc_filt_pass1$band_pass)
+  dfa::.DFA(df_thc_filt_pass1$band_pass)
   
-
-  library(nonlinearTseries)
-  
-  
+  file <- NYA2008
+  dfa.res <- DFA::DFA(NYA2008)
   
   
+library(MFDFA)  
+  a<-0.9
+  N<-1024
+  tsx<-MFsim(N,a)
+  scale=10:100
+  q<--10:10
+  m<-1
+  b<-MFDFA(tsx, scale, m, q) 
   
+  dev.new()
+  par(mai=rep(1, 4))
+  plot(q, b$Hq, col=1, axes= F, ylab=expression('h'[q]), pch=16, cex.lab=1.8,
+       cex.axis=1.8, main="Hurst exponent",
+       ylim=c(min(b$Hq),max(b$Hq)))
+  grid(col="midnightblue")
+  axis(1)
+  axis(2)
   
-  
-  
+  alpha <- b$spec %>% as_tibble()
+  alpha %>% ggplot(aes(x = 1:20, y = hq)) + geom_point()
   
   
   

### 2021-05-13 figures.Rmd
add variables to table 1

@@
   dplyr::select(c(Subject.ID..NIAID., Status, Glucose, Hematocrit, Lactate, Temperature,
-                  Arginine.umol.L, Haptoglobin..mg.dl., Hemoglobin..uM., Whole.Blood.Nitrite, 
+                  Arginine.umol.L, Haptoglobin..mg.dl., Hemoglobin..uM., 
+                  BP.Diastolic, BP.Systolic, O2.Sat, RR, HR,
+
                   cerebral.hbdeox, cerebral.hbdeox.alpha, cerebral.hbox, cerebral.hbox.alpha,
-                  cerebral.sat, cerebral.sat.alpha, cerebral.thc, cerebral.thc.alpha, cerebral.thc.norm.std)) %>%
+                  cerebral.sat, cerebral.sat.alpha, cerebral.thc, cerebral.thc.alpha)) %>%
 
-  dplyr::rename("subject_id" = "Subject.ID..NIAID.", "Arginine (umol/L)" = "Arginine.umol.L", "Haptoglobin (mg/dL)" = "Haptoglobin..mg.dl.", 
-                "Hemoglobin (uM)" = "Hemoglobin..uM.", "WB_Nitrite" = "Whole.Blood.Nitrite", "DeoxyHb" = "cerebral.hbdeox", 
-                "DeoxyHb_alpha" = "cerebral.hbdeox.alpha", "OxyHb" = "cerebral.hbox", "OxyHb_alpha" = "cerebral.hbox.alpha", 
-                "O2sat" = "cerebral.sat", "O2sat_alpha" = "cerebral.sat.alpha", "THC" = "cerebral.thc", 
-                "THC_alpha" = "cerebral.thc.alpha", "THC_sd" = "cerebral.thc.norm.std")
+  dplyr::rename("subject_id" = "Subject.ID..NIAID.", 
+                "Arginine (umol/L)" = "Arginine.umol.L", "Haptoglobin (mg/dL)" = "Haptoglobin..mg.dl.", 
+                "Hemoglobin (uM)" = "Hemoglobin..uM.", "Lactate (mmol/L)" = "Lactate", "Hematocrit (%)" = "Hematocrit",
+                "Systolic BP" = "BP.Systolic", "Diastolic BP" = "BP.Diastolic", "O2 sat" = "O2.Sat", 
+                "Respiration rate" = "RR", "Heart rate" = "HR",
+
+                "Deoxy Hb" = "cerebral.hbdeox", "Deoxy Hb α" = "cerebral.hbdeox.alpha", 
+                "Oxy Hb" = "cerebral.hbox", "Oxy Hb α" = "cerebral.hbox.alpha",  
+                "Cerebral O2 sat" = "cerebral.sat", "Cerebral O2 sat α" = "cerebral.sat.alpha", 
+                "THC" = "cerebral.thc", "THCα" = "cerebral.thc.alpha")
+
+# Remove patients labeled as "UM" that instead meet the criteria for severe malaria
+
+  # remove UM patient with lactate > 5 
+  df_hbox_trim <- df_hbox_trim %>% dplyr::filter(!(Status == "UM" & `Lactate (mmol/L)` > 5))
+
+  # All UM patients have hematocrit > 15%
+  # df_hbox_trim %>% dplyr::filter(Status == "UM"  & `Hematocrit (%)` <= 15)
 
@@
 # pivot long
-df_hbox_long <- df_hbox_trim %>% pivot_longer(cols = 3:ncol(.), names_to = "measurement", values_to = "value")
+df_hbox_long <- df_hbox_trim %>% pivot_longer(cols = 3:ncol(.), names_to = "variable", values_to = "value")
 
@@
 df_hbox %>%
+
   group_by(Status) %>% 
   summarise(`Age (years)` = median(Age, na.rm = TRUE), age_range = range(Age, na.rm = TRUE),
+
+            `Temperature (°C)` = median(Temperature, na.rm = TRUE), temp_range = range(Temperature, na.rm = TRUE),
+            `Heart rate` = median(HR, na.rm = TRUE), hr_range = range(HR, na.rm = TRUE),
+            `Systolic BP` = median(BP.Systolic, na.rm = TRUE), sys_range = range(BP.Systolic, na.rm = TRUE),
+            `Diastolic BP` = median(BP.Diastolic, na.rm = TRUE), dia_range = range(BP.Diastolic, na.rm = TRUE),
+            `Respiration rate` = median(RR, na.rm = TRUE), rr_range = range(RR, na.rm = TRUE),
+            `O~2~ saturation` = median(O2.Sat, na.rm = TRUE), o2_range = range(O2.Sat, na.rm = TRUE),
+
             `Hematocrit (%)` = median(Hematocrit, na.rm = TRUE), hem_range = range(Hematocrit, na.rm = TRUE),
             `Lactate (mmol/L)` = median(Lactate, na.rm = TRUE), lactate_range = range(Lactate, na.rm = TRUE),
-            `Arginine (umol/L)` = median(Arginine.umol.L, na.rm = TRUE), arg_range = range(Arginine.umol.L, na.rm = TRUE),
-            `Temperature (°C)` = median(Temperature, na.rm = TRUE), temp_range = range(Temperature, na.rm = TRUE)) %>% 
+            `Arginine (μmol/L)` = median(Arginine.umol.L, na.rm = TRUE), arg_range = range(Arginine.umol.L, na.rm = TRUE)) %>% 
   mutate_if(is.numeric, round, 1) %>%
@@
   mutate(`Age (years)` = paste0(`Age (years)`, " (", age_range[1], ", ", age_range[2], ")"),
+
+         `Temperature (°C)` = paste0(`Temperature (°C)`, " (", temp_range[1], ", ", temp_range[2], ")"),
+         `Heart rate` = paste0(`Heart rate`, " (", hr_range[1], ", ", hr_range[2], ")"),
+         `Systolic BP` = paste0(`Systolic BP`, " (", sys_range[1], ", ", sys_range[2], ")"),
+         `Diastolic BP` = paste0(`Diastolic BP`, " (", dia_range[1], ", ", dia_range[2], ")"),
+         `Respiration rate` = paste0(`Respiration rate`, " (", rr_range[1], ", ", rr_range[2], ")"),
+         `O~2~ saturation` = paste0(`O~2~ saturation`, " (", o2_range[1], ", ", o2_range[2], ")"),
+
          `Hematocrit (%)` = paste0(`Hematocrit (%)`, " (", hem_range[1], ", ", hem_range[2], ")"),
          `Lactate (mmol/L)` = paste0(`Lactate (mmol/L)`, " (", lactate_range[1], ", ", lactate_range[2], ")"),
-         `Arginine (umol/L)` = paste0(`Arginine (umol/L)`, " (", arg_range[1], ", ", arg_range[2], ")"),
-         `Temperature (°C)` = paste0(`Temperature (°C)`, " (", temp_range[1], ", ", temp_range[2], ")")) %>% 
-  dplyr::select(-c(age_range, hem_range, lactate_range, arg_range, temp_range)) %>% 
+         `Arginine (μmol/L)` = paste0(`Arginine (μmol/L)`, " (", arg_range[1], ", ", arg_range[2], ")")) %>% 
+  dplyr::select(-c(age_range, temp_range, hr_range, sys_range, dia_range, rr_range, o2_range, hem_range, lactate_range, arg_range)) %>%
+
   unique() %>% 
@@
                 "Healthy volunteers (HV)" = "HV",
-                "Uncomplicated malaria patients (UM)" = "UM",
-                "Cerebral malaria patients (CM)" = "CM") %>% 
+                "Uncomplicated malaria (UM) patients" = "UM",
+                "Cerebral malaria (CM) patients" = "CM") %>% 
   knitr::kable(align = rep('c', 4))
@@
 **Figure 1.** Differences in variables measured by Oxiplex and associated alpha values among patient groups. Note: removed one outlier from DeoxyHb alpha (3.52)
-
-### Figure 2: Random forest modeling using all outcomes and predictors
-
-#### A. PCA
-```{r fig2a}
-
-# convert tibble to df
-df_hbox_impute_df <- df_hbox_impute %>% 
-  dplyr::select(-c(subject_id, Status)) %>% 
-  dplyr::rename("Whole blood nitrite" = "WB_Nitrite", "THC sd" = "THC_sd", "THCα" = "THC_alpha", 
-                "Deoxy Hb" = "DeoxyHb", "Oxy Hb" = "OxyHb", "O2 sat" = "O2sat", "O2 sat α" = "O2sat_alpha",
-                "Deoxy Hb α" = "DeoxyHb_alpha", "Oxy Hb α" = "OxyHb_alpha",
-                "Hemoglobin" = "Hemoglobin (uM)", "Haptoglobin" = "Haptoglobin (mg/dL)", "Arginine" = "Arginine (umol/L)") %>% 
-  as.data.frame(df_hbox_impute) 
-rownames(df_hbox_impute_df) <- df_hbox_impute$subject_id
-
-# compute PCA
-my_pca1 <- prcomp(df_hbox_impute_df, scale = TRUE)
-
-# define groups
-groups1 <- as.factor(df_hbox_impute$Status)
-
-# Eigenvalues
-pca_eig_val1 <- get_eigenvalue(my_pca1)
-
-# extract variables
-df_vars1 <- my_pca1$rotation %>% as_tibble() %>% mutate(variable = rownames(my_pca1$rotation))
-
-# extract patients
-df_pnts1 <- my_pca1$x %>% as_tibble() %>% mutate(group = groups1)
-
-# plot
-
-  # plot patients
-  ggplot() +
-  geom_point(data = df_pnts1, 
-             mapping = aes(x = PC1, y = PC2, color = group), 
-             size = 3) +
-  stat_conf_ellipse(data = df_pnts1, 
-                    mapping = aes(x = PC1, y = PC2, color = group, fill = group),
-                    level = 0.95, npoint = 100, bary = TRUE, alpha = 0.1, geom = "polygon") +
-  stat_mean(data = df_pnts1, 
-            mapping = aes(x = PC1, y = PC2, color = group, shape = group),
-            na.rm = FALSE) +
-
-  # plot variables
-  geom_point(data = df_vars1, aes(x = PC1*10, y = PC2*10)) +
-  geom_text_repel(data = df_vars1, aes(x = PC1*10, y = PC2*10, label = variable, size = 10), show.legend = FALSE) +
-  geom_segment(data = df_vars1, aes(x = 0, y = 0, xend = PC1*10, yend = PC2*10), arrow = arrow()) +
-
-  # draw lines
-  geom_vline(xintercept = 0, lty = 2, color = "black") +
-  geom_hline(yintercept = 0, lty = 2, color = "black") +
-
-  # design
-  scale_color_manual(values = my_col) +
-  scale_fill_manual(values = my_col) +
-  xlab(paste0("PC1: ", round(pca_eig_val1$variance.percent[1], 2), "% of variance")) +
-  ylab(paste0("PC2: ", round(pca_eig_val1$variance.percent[2], 2), "% of variance")) +
-
-  theme_bw() +
-  theme(axis.text = element_text(size = 20),
-        axis.title = element_text(size = 25),
-        legend.text = element_text(size = 15),
-        legend.title = element_text(size = 20))
-
-```
-
-#### B. Confusion matrix
-```{r fig2b_prep, eval = FALSE}
-
-# Data
-df_hbox_model1 <- df_hbox_trim %>% 
-  dplyr::select(-subject_id) %>% 
-  dplyr::rename("Whole blood nitrite" = "WB_Nitrite", "THC sd" = "THC_sd", "THCα" = "THC_alpha", 
-                "Deoxy Hb" = "DeoxyHb", "Oxy Hb" = "OxyHb", "O2 sat" = "O2sat", "O2 sat α" = "O2sat_alpha",
-                "Deoxy Hb α" = "DeoxyHb_alpha", "Oxy Hb α" = "OxyHb_alpha",
-                "Hemoglobin" = "Hemoglobin (uM)", "Haptoglobin" = "Haptoglobin (mg/dL)", "Arginine" = "Arginine (umol/L)")
-
-# Build a model
-
-  # resamples (not a lot of data, need to use cross validation)
-  set.seed(123)
-  hbox_boot1 <- bootstraps(df_hbox_model1, strata = Status)
-
-  # Create recipe
-  ranger_recipe <- 
-    recipe(formula = Status ~ ., data = df_hbox_model1) %>%
-
-    # BoxCox transformation to normality
-    step_BoxCox(all_predictors(), -all_outcomes()) %>%
-
-    # remove variables with non-zero variance
-    step_nzv(all_predictors(), -all_outcomes()) %>%
-
-    # impute missing data using K-nearest neighbors
-    step_knnimpute(all_predictors(), -all_outcomes())
-
-  # Random forest model specifications (set for tuning)
-  ranger_spec <- 
-    rand_forest(mtry = tune(), min_n = tune(), trees = 1000) %>% 
-    set_mode("classification") %>% 
-    set_engine("ranger")
-
-  # Create workflow
-  ranger_workflow <- 
-    workflow() %>% 
-    add_recipe(ranger_recipe) %>% 
-    add_model(ranger_spec)
-
-# Tune model
-
-  # Initial tune on bootstraps
-  set.seed(84374)
-  ranger_tune <-
-    tune_grid(ranger_workflow, 
-              resamples = hbox_boot1, 
-              grid = 11)
-
-# Finalize model
-
-  final_rf <- ranger_workflow %>% 
-    finalize_workflow(select_best(ranger_tune, metric = "accuracy"))
-
-  final_rs_res <- final_rf %>% 
-    fit_resamples(resamples = hbox_boot1,
-                  metrics = metric_set(roc_auc, accuracy, sens, spec),
-                  control = control_resamples(save_pred = TRUE))
-
-# save objects so model doesn't rerun every time you knit
-  save(df_hbox_model1, hbox_boot1, ranger_recipe, ranger_spec, ranger_tune, final_rs_res, file = "model1.Rdata")
-
-```
-
-```{r fig2b_plot}
-
-final_res_conf_mat <- final_rs_res %>% conf_mat_resampled() 
-final_res_conf_mat %>% 
-  group_by(Truth) %>% 
-  summarise(total_true = sum(Freq)) %>% 
-  right_join(final_res_conf_mat) %>% 
-  mutate(`Percent occurence` = round(Freq/total_true*100, 2)) %>%
-
-  # plot
-  ggplot(aes(x = Truth, y = Prediction, fill = `Percent occurence`, label = `Percent occurence`)) +
-  geom_tile() +
-  geom_text(size = 8) +
-  scale_fill_gradient2(low = "#4575b4", high = "#d73027") +
-
-  theme_bw() +
-  theme(axis.title = element_text(size = 25),
-        axis.text = element_text(size = 20),
-        legend.title = element_text(size = 20),
-        legend.text = element_text(size = 15))
-
-```
-
-#### C. Variable importance
-```{r fig2c}
-
-# new specification for ranger model
-imp_spec <- ranger_spec %>% 
-  finalize_model(select_best(ranger_tune, metric = "accuracy")) %>% 
-  set_engine("ranger", importance = "permutation")
-
-# plot importance 
-workflow() %>% 
-  add_recipe(ranger_recipe) %>% 
-  add_model(imp_spec) %>% 
-  fit(df_hbox_model1) %>% 
-  pull_workflow_fit() %>% 
-  vip(aesthetics = list(alpha = 0.8, fill = "midnightblue")) +
-
-  theme_bw() +
-  theme(axis.text = element_text(size = 20),
-        axis.title = element_text(size = 25))
-
-```
-
-**Figure 2.** Random forest modeling demonstrates a high degree of discriminatory power between healthy patients and patients with malaria (uncomplicated and cerebral) using primarily on hematological parameters.
-**(A)** Principal component analysis (PCA) of study patients using hematological and NIRS data.
-Variables are plotted as vectors in the direction and magnitude of the PC to which they contribute.
-Color indicates patient status, ellipses show median and 95% confidence interval on the cluster.
-**(B)** Confusion matrix representing the results of random forest modeling averaged across 25 bootstrap resamples.
-Freq indicates the average number of times that the event (shown by the x- and y-axes) occurred.
-**(C)** Relative contribution of the variables to the discriminatory power of the model (showing top 10 only).
 
@@
 
-### Figure 4: Relationship between THCα and THC sd
-
-```{r fig4}
-
-cm_model <- lm(THC_sd ~ THC_alpha, data = df_hbox_log_voi %>% filter(Status == "CM"))
-um_model <- lm(THC_sd ~ THC_alpha, data = df_hbox_log_voi %>% filter(Status == "UM"))
-
-df_hbox_log_voi %>% dplyr::rename("THC sd" = "THC_sd", "THCα" = "THC_alpha") %>%
-
-  ggplot(aes(x = THCα, y = `THC sd`, color = Status, fill = Status)) +
-  geom_point(size = 3, alpha = 0.7) +
-  geom_smooth(method = "lm", se = FALSE) +
-
-  geom_text(label = paste0("y = ", 
-                           round(cm_model$coefficients[[1]], 2), 
-                           " + ", 
-                           round(cm_model$coefficients[[2]], 2), 
-                           "x ; R^2 = ", 
-                           round(summary(cm_model)$adj.r.squared, 2)),
-            x = 1.3, y = 0.55,
-            color = "#00BFC4",
-            size = 6) +
-
-    geom_text(label = paste0("y = ", 
-                           round(um_model$coefficients[[1]], 2), 
-                           " + ", 
-                           round(um_model$coefficients[[2]], 2), 
-                           "x ; R^2 = ", 
-                           round(summary(um_model)$adj.r.squared, 2)),
-            x = 1.3, y = 0.6,
-            color = "#F8766D",
-            size = 6) +
-
-  theme_bw() +
-  theme(axis.text = element_text(size = 20),
-        axis.title = element_text(size = 25),
-        legend.text = element_text(size = 15),
-        legend.title = element_text(size = 20))
-
-```
-
-**Figure 4.** Scatter plot demonstrating the relationship between THCα and THC sd.
-
 ### Supplementary figure 1: Variable correlations 

### 2021-05-13 figures.Rmd
fig 1 rearrange and add KW p-values

@@
 library(janitor)
+library(FSA)
 library(factoextra)
@@
 df_hbox_voi <- df_hbox_impute %>% 
-  dplyr::select(c(Status, DeoxyHb, DeoxyHb_alpha, OxyHb, OxyHb_alpha, O2sat, O2sat_alpha, THC, THC_alpha, THC_sd)) %>% 
-  dplyr::filter(DeoxyHb_alpha < 3) %>% 
-  dplyr::rename("THC sd" = "THC_sd", "THCα" = "THC_alpha", "Deoxy Hb" = "DeoxyHb", "Oxy Hb" = "OxyHb", 
-                "O2 sat" = "O2sat", "O2 sat α" = "O2sat_alpha", "Deoxy Hb α" = "DeoxyHb_alpha", "Oxy Hb α" = "OxyHb_alpha")
+  dplyr::select(c(Status, `Deoxy Hb`, `Deoxy Hb α`, `Oxy Hb`, `Oxy Hb α`, `Cerebral O2 sat`, `Cerebral O2 sat α`, THC, `THCα`)) %>% 
+  dplyr::filter(`Deoxy Hb α` < 3)
 
 df_hbox_voi_long <- df_hbox_voi %>% 
-  pivot_longer(2:10, names_to = "variable", values_to = "value") %>%
-  mutate(variable = factor(variable, levels = c("Deoxy Hb", "Oxy Hb", "O2 sat", "THC", "THC sd", 
-                                                "Deoxy Hb α", "Oxy Hb α", "O2 sat α", "THCα")))
+  pivot_longer(2:9, names_to = "variable", values_to = "value") %>%
+  mutate(variable = factor(variable, levels = c("Oxy Hb", "Oxy Hb α",   
+                                                 "Deoxy Hb", "Deoxy Hb α", 
+                                                "Cerebral O2 sat", "Cerebral O2 sat α", 
+                                                "THC", "THCα")))
 
@@
   stat_summary(fun = "median", geom = "crossbar", aes(color = Status), size = 0.2, width = 0.5) +
-  facet_wrap(~variable, scales = "free_y") +
+  facet_wrap(~variable, scales = "free_y", nrow = 4, ncol = 2) +
 
@@
 ```
-**Figure 1.** Differences in variables measured by Oxiplex and associated alpha values among patient groups. Note: removed one outlier from DeoxyHb alpha (3.52)
+
+```{r fig1.pvals, eval = FALSE}
+
+# THC
+kruskal.test(THC ~ Status, data = df_hbox_voi)
+dunnTest(THC ~ Status, data = df_hbox_voi, method = "bh")
+
+# Cerebral O2 sat
+kruskal.test(`Cerebral O2 sat` ~ Status, data = df_hbox_voi)
+
+```
+
+**Figure 1.** Differences in variables measured by Oxiplex and associated α values among patient groups.
+THC levels were different between all three groups (Kruskal-Wallis adjusted p-value = 3.25x10^-6^; Dunn Test post-hoc adjusted p-values: CM vs HV = 1.66x10^-6^, CM vs UM = .032, HV vs UM = .033).
+Levels of cerebral O~2~ saturation did not differ among groups (Kruskal-Wallis adjusted p-value = .07).
+Note: removed one outlier from Deoxy Hb α (3.52)
 

### 2021-05-13 Scripts/additional_models.R
create models using cerebral o2 sat and THC + corresponding alpha values

@@ -136,19 +136,85 @@
   
   
   
+# model 5: Cerebral O2 sat + cerebral O2 sat a + THC + THCa --------------------------------------------------------------
+  
+  
+  # select  THC_alpha & THC
+  `df_hbox_model5` <- df_hbox_voi_01 %>% 
+    mutate(`1/10 * THCα` = THCα/10) %>% 
+    dplyr::select(-THCα)
+  
+  # resample
+  set.seed(1012)
+  hbox_boot5 <- bootstraps(df_hbox_model5, times = 100)
+  
+  # recipe
+  hbox_rec5 <- recipe(Status ~ ., data = df_hbox_model5) %>% 
+    step_BoxCox(all_predictors()) %>% 
+    step_nzv(all_predictors())
+  
+  # workflow
+  hbox_wf5 <- workflow() %>% 
+    add_recipe(hbox_rec5)
+  
+  # logistic regression model spec
+  glm_spec5 <- logistic_reg() %>% 
+    set_engine("glm")
+  
+  glm_res5 <- hbox_wf5 %>% 
+    add_model(glm_spec5) %>%
+    fit_resamples(
+      resamples = hbox_boot5,
+      metrics = metric_set(roc_auc, accuracy, sensitivity, specificity),
+      control = tune::control_resamples(save_pred = TRUE, verbose = TRUE)
+    )
+  
+  # select best model to be used for evaluation
+  glm_best5 <- glm_res5 %>% select_best("roc_auc")
+  
+  # finalize model
+  hbox_final5 <- hbox_wf5 %>% 
+    add_model(glm_spec5) %>% 
+    finalize_workflow(glm_best5) %>%
+    fit(df_hbox_model5) %>%
+    pull_workflow_fit()
+  
+  ## Evaluate
+  
+  collect_metrics(glm_res5) %>% 
+    dplyr::select(.metric, mean, std_err) %>% 
+    mutate(mean = mean*100, std_err = std_err*100) %>% 
+    mutate(Metric = c("Accuracy", "ROC AUC", "Sensitivity", "Specificity")) %>% 
+    dplyr::select(c("Metric", "mean", "std_err")) %>% 
+    
+    dplyr::rename("Mean across resamples" = "mean", "Standard error" = "std_err") %>% 
+    mutate_if(is.numeric, round, 2) # model performs worse
+  
+  hbox_final5 %>% 
+    tidy() %>% 
+    mutate(p.value = as.character(p.value)) %>% 
[…]

### 2021-05-13 figures.Rmd
code for Figures_draft2

@@
 
-load("model1.Rdata")
-load("model2.Rdata")
+load("glm_model.Rdata")
 
@@
 
-```{r fig1}
+```{r fig1, fig.height = 12}
 
 df_hbox_voi <- df_hbox_impute %>% 
-  dplyr::select(c(Status, `Deoxy Hb`, `Deoxy Hb α`, `Oxy Hb`, `Oxy Hb α`, `Cerebral O2 sat`, `Cerebral O2 sat α`, THC, `THCα`)) %>% 
+  dplyr::select(c(Status, `Oxy Hb`, `Oxy Hb α`, `Deoxy Hb`, `Deoxy Hb α`, `Cerebral O2 sat`, `Cerebral O2 sat α`, THC, `THCα`)) %>% 
   dplyr::filter(`Deoxy Hb α` < 3)
@@
 Levels of cerebral O~2~ saturation did not differ among groups (Kruskal-Wallis adjusted p-value = .07).
-Note: removed one outlier from Deoxy Hb α (3.52)
+Note: removed one outlier from Deoxy Hb α (3.52).
 
-### Figure 3: Logistic regression on CM vs UM
+### Figure 2: Logistic regression on CM vs UM
 
@@
 
-```{r fig3a}
-# select variables of interest
-df_hbox_log_voi <- df_hbox_impute %>% 
-  filter(Status != "HV") %>% 
-  dplyr::select(2, 11:19)
+```{r fig2a}
+# remove healthy volunteers, Oxy Hb, and Deoxy Hb from analysis
+df_hbox_log_voi <- df_hbox_voi %>% 
+  dplyr::filter(Status != "HV") %>% 
+  dplyr::select(-(2:5))
 
@@
   dplyr::select(-Status) %>% 
-  dplyr::rename("THC sd" = "THC_sd", "THCα" = "THC_alpha", "Deoxy Hb" = "DeoxyHb", "Oxy Hb" = "OxyHb", 
-                "O2 sat" = "O2sat", "O2 sat α" = "O2sat_alpha", "Deoxy Hb α" = "DeoxyHb_alpha", "Oxy Hb α" = "OxyHb_alpha") %>% 
   as.data.frame() 
@@
 # compute PCA
-my_pca2 <- prcomp(df_hbox_voi_df, scale = TRUE)
+my_pca <- prcomp(df_hbox_voi_df, scale = TRUE)
 
 # define groups
-groups2 <- as.factor(df_hbox_log_voi$Status)
+groups <- as.factor(df_hbox_log_voi$Status)
 
 # Eigenvalues
-pca_eig_val2 <- get_eigenvalue(my_pca2)
+pca_eig_val <- get_eigenvalue(my_pca)
 
 # two data frames for plotting
-df_vars2 <- my_pca2$rotation %>% as_tibble() %>% mutate(variable = rownames(my_pca2$rotation))
-df_pnts2 <- my_pca2$x %>% as_tibble() %>% mutate(group = groups2)
+df_vars <- my_pca$rotation %>% as_tibble() %>% mutate(variable = rownames(my_pca$rotation))
+df_pnts <- my_pca$x %>% as_tibble() %>% mutate(group = groups)
 
@@
   ggplot() +
-  geom_point(data = df_pnts2, mapping = aes(x = PC1, y = PC2, color = group), 
-             size = 3) +
-  stat_conf_ellipse(data = df_pnts2, mapping = aes(x = PC1, y = PC2, color = group, fill = group),
+  geom_point(data = df_pnts, mapping = aes(x = PC1, y = PC2, color = group), size = 3) +
+  stat_conf_ellipse(data = df_pnts, mapping = aes(x = PC1, y = PC2, color = group, fill = group),
                     level = 0.95, npoint = 100, bary = TRUE, alpha = 0.1, geom = "polygon") +
-  stat_mean(data = df_pnts2, mapping = aes(x = PC1, y = PC2, color = group, shape = group),
+  stat_mean(data = df_pnts, mapping = aes(x = PC1, y = PC2, color = group, shape = group),
             na.rm = FALSE) +
@@
   # plot variables
-  geom_point(data = df_vars2, aes(x = PC1*10, y = PC2*10)) +
-  geom_text_repel(data = df_vars2, aes(x = PC1*10, y = PC2*10, label = variable, size = 10), show.legend = FALSE) +
-  geom_segment(data = df_vars2, aes(x = 0, y = 0, xend = PC1*10, yend = PC2*10), arrow = arrow()) +
+  geom_point(data = df_vars, aes(x = PC1*10, y = PC2*10)) +
+  geom_text_repel(data = df_vars, aes(x = PC1*10, y = PC2*10, label = variable, size = 10), show.legend = FALSE) +
+  geom_segment(data = df_vars, aes(x = 0, y = 0, xend = PC1*10, yend = PC2*10), arrow = arrow()) +
 
@@
   # design
-  xlab(paste0("PC1: ", round(pca_eig_val2$variance.percent[1], 2), "% of variance")) +
-  ylab(paste0("PC2: ", round(pca_eig_val2$variance.percent[2], 2), "% of variance")) +
+  xlab(paste0("PC1: ", round(pca_eig_val$variance.percent[1], 2), "% of variance")) +
+  ylab(paste0("PC2: ", round(pca_eig_val$variance.percent[2], 2), "% of variance")) +
 
@@
 
-```{r fig3b_prep, eval = FALSE}
+```{r fig2b_prep, eval = FALSE}
 
@@
 # select only significant variables
-df_hbox_model2 <- df_hbox_voi_01 %>% 
-  dplyr::select(c(Status, THC_alpha, THC_sd)) %>% 
-  dplyr::rename("THCα" = "THC_alpha", "THC sd" = "THC_sd") %>% 
+df_hbox_model <- df_hbox_voi_01 %>% 
+  dplyr::select(c(Status, THCα)) %>% 
   mutate(`1/10 * THCα` = THCα/10) %>% 
@@
 # resample
-set.seed(789)
-hbox_boot2 <- bootstraps(df_hbox_model2, times = 1000)
+set.seed(123)
+hbox_boot <- bootstraps(df_hbox_model, times = 1000)
 
 # recipe
-hbox_rec2 <- recipe(Status ~ ., data = df_hbox_model2) %>% 
+hbox_rec <- recipe(Status ~ ., data = df_hbox_model) %>% 
   step_BoxCox(all_predictors()) %>% 
@@
 
-# workflow
-hbox_wf2 <- workflow() %>% 
-  add_recipe(hbox_rec2)
+# create workflow
+hbox_wf <- workflow() %>% add_recipe(hbox_rec)
 
-# logistic regression model spec
-glm_spec2 <- logistic_reg() %>% 
-  set_engine("glm")
+# logistic regression model specifications
+glm_spec <- logistic_reg() %>% set_engine("glm")
 
-glm_res2 <- hbox_wf2 %>% 
-  add_model(glm_spec2) %>%
+# create model on 1000 different resamples
+glm_res <- hbox_wf %>% 
+  add_model(glm_spec) %>%
   fit_resamples(
-    resamples = hbox_boot2,
+    resamples = hbox_boot,
     metrics = metric_set(roc_auc, accuracy, sensitivity, specificity),
@@
 # select best model to be used for evaluation
-glm_best2 <- glm_res2 %>% select_best("roc_auc")
+glm_best <- glm_res %>% select_best("roc_auc")
 
 # finalize model
-hbox_final2 <- hbox_wf2 %>% 
-  add_model(glm_spec2) %>% 
-  finalize_workflow(glm_best2) %>%
-  fit(df_hbox_model2) %>%
+hbox_final <- hbox_wf %>% 
+  add_model(glm_spec) %>% 
+  finalize_workflow(glm_best) %>%
+  fit(df_hbox_model) %>%
   pull_workflow_fit()
@@
 # save objects so model doesn't rerun 
-save(glm_res2, hbox_final2, file = "model2.Rdata")
+save(glm_res, hbox_final, file = "glm_model.Rdata")
 
@@
 
-```{r fig3b_plot}
+```{r fig2b_plot}
 
-final_res_conf_mat2 <- glm_res2 %>% 
+final_res_conf_mat <- glm_res %>% 
   conf_mat_resampled() %>% 
@@
 
-final_res_conf_mat2 %>% 
+final_res_conf_mat %>% 
   group_by(Truth) %>% 
   summarise(total_true = sum(Freq)) %>% 
-  right_join(final_res_conf_mat2) %>% 
+  right_join(final_res_conf_mat) %>% 
   mutate(`Percent occurence` = round(Freq/total_true*100, 2)) %>%
@@
 
-**Figure 3** Logistic regression model discriminating CM vs UM patients using only THCα and THC sd as predictors.
+**Figure 2** Logistic regression model discriminating CM vs UM patients using only THCα as the predictor.
 **(A)** PCA of patients with CM and UM as determined by NIRS variables.
@@
 **(B)** Confusion matrix representing the results of the logistic regression model averaged across 1000 bootstrap resamples.
-Freq indicates the average number of times that the event (shown by the x- and y-axes) occurred.
+Percent occurrence indicates percentage times that the event (shown by the x- and y-axes) occurred.
 
@@
 
-collect_metrics(glm_res2) %>% 
+collect_metrics(glm_res) %>% 
   dplyr::select(.metric, mean, std_err) %>% 
@@
 
-hbox_final2 %>% 
+hbox_final %>% 
   tidy() %>% 
@@
 
-### Supplementary figure 1: Variable correlations 
+### Supplementary figure 1: Spearman correlations of NIRS variables 
 ```{r fig_s1, fig.show = "hold", out.width = "50%", fig.height = 12, fig.width = 15}
@@
 
-df_hbox_cor <- df_hbox_impute %>% 
-  dplyr::rename("THC sd" = "THC_sd", "THCα" = "THC_alpha", "Deoxy Hb" = "DeoxyHb", "Oxy Hb" = "OxyHb", 
-                "O2 sat" = "O2sat", "O2 sat α" = "O2sat_alpha", "Deoxy Hb α" = "DeoxyHb_alpha", "Oxy Hb α" = "OxyHb_alpha") %>% 
-  dplyr::select(Status, `Deoxy Hb`, `Oxy Hb`, `O2 sat`, THC, `THC sd`, `Deoxy Hb α`, `Oxy Hb α`, `O2 sat α`, THCα)
+m_cor_hv <- df_hbox_voi %>% f_calc_cor(status = "HV")
 
-m_cor_hv <- df_hbox_cor %>% f_calc_cor(status = "HV")
+m_cor_um <- df_hbox_voi %>% f_calc_cor(status = "UM")
 
-m_cor_um <- df_hbox_cor %>% f_calc_cor(status = "UM")
-
-m_cor_cm <- df_hbox_cor %>% f_calc_cor(status = "CM")
+m_cor_cm <- df_hbox_voi %>% f_calc_cor(status = "CM")
 
@@
 
-    mutate_if(is.factor, factor, levels = c("Deoxy Hb", "Oxy Hb", "O2 sat", "THC", "THC sd", "Deoxy Hb α", "Oxy Hb α", "O2 sat α", "THCα")) %>%
+    mutate_if(is.factor, factor, levels = c("Oxy Hb", "Deoxy Hb", "Cerebral O2 sat", "THC", "Oxy Hb α", "Deoxy Hb α", "Cerebral O2 sat α", "THCα")) %>%
 
@@
     ylab("") +
+    theme_bw() +
     theme(axis.text.x = element_text(angle = 45, hjust = 0.9),
@@
 ### Supplementary figure 2: Logistic regression curves of NIRS variables
-```{r fig_s2}
+```{r fig_s2, fig.height = 12}
 
@@
   geom_smooth(method = "glm", method.args = list(family = "binomial")) +
-  facet_wrap(~variable, scales = "free_x") +
+  facet_wrap(~variable, scales = "free_x", nrow = 4, ncol = 2) +
 
@@
 On the y-axis, 1 represents CM while 0 represents UM.
-A steeper curve indicates higher disciminatory power
+A steeper curve indicates higher discriminatory power

### 2021-05-13 figures.Rmd
remove 5.8 lactate from UM and lactate data for HV in table 1

@@
 # Age, hematocrit, and lactate
-df_hbox %>%
+df_clinical_res <- df_hbox %>%
+
+  dplyr::filter(Subject.ID..NIAID. != "TM1018") %>%
 
@@
                 "Uncomplicated malaria (UM) patients" = "UM",
-                "Cerebral malaria (CM) patients" = "CM") %>% 
-  knitr::kable(align = rep('c', 4))
+                "Cerebral malaria (CM) patients" = "CM")
+
+  # remove lactate data for HV
+  df_clinical_res[11,]$`Healthy volunteers (HV)` <- "--"
+
+  # create kable df
+  df_clinical_res %>% knitr::kable(align = rep('c', 4))
 
 ```
-**Table 1.** A summary of subject information and hematological data. Group patient data presented as median (range). Note: sex not recorded for 3 CM patients.
+**Table 1.** A summary of subject information and hematological data.
+Group patient data presented as median (range).
+Note: Sex not recorded for 3 CM patients; Lactate data removed for HV due to inconsistent data.
 

### 2021-05-13 Functions/f_filter_signal.R
function to filter raw signal by max standard deviations from mean

@@ -1,0 +1,84 @@
+
+## A function to find the longest segment of the signal (where t >= 200s) within 0.5 standard deviations ##
+## Purpose is to filter blips out of signal
+
+## find the longest list with a standard deviation below a certain limit is to check all potential sublists
+
+## our parameters: f_filter_signal(df, max_sd = 0.5)
+
+
+f_filter_signal <- function(df, max_sd) {
+  
+  start = 1
+  end = 1
+  max_start = 1
+  max_end = 0
+  
+  df <- df %>% filter(!is.na(roll_mean))
+  
+  mean <- df$roll_mean %>% mean(na.rm = TRUE)
+  sd <- df$roll_mean %>% sd(na.rm = TRUE)
+  
+  for (i in 1:nrow(df)) {
+    
+    if ((mean - max_sd*sd) < df[i,]$roll_mean && df[i,]$roll_mean < (mean + max_sd*sd)) {
+      
+      end <- i
+      
+    } else {
+      
+      start <- i
+      
+    }
+    
+    if ((end - start) > (max_end - max_start)) {
+      
+      max_end <- end
+      max_start <- start
+      
+    }
+    
+  }
+  
+  return(df[max_start:max_end,])
+  
+}
+
+
+
+
+
+## Alternative function where you can filter with a max variance
+
+f_filter_signal_var <- function(df, min_length, max_var){
+  
+  for (length in nrow(df):min_length) {
+    
+    for (start in 1:(nrow(df) - length + 1)) {
+      
+      df_temp <- df[start:(start + length - 1),]
+      df_temp <- df_temp %>% mutate(start = start, length = length)
+      
+      var <- df_temp$roll_mean %>% var(na.rm = TRUE)
+      
+      if (var <= max_var) {
+        
+        return(df_temp)
+        
+      }
+      
+    }
+    
+  }
+  
+  return(NULL)
+  
+}
+
+
+
+
+
+
+
+

### 2021-05-13 figures.Rmd
table 1. changed range to quartiles figure 1. split alpha into fig 2

@@
 
-# Age, hematocrit, and lactate
+# other numeric variables
+
 df_clinical_res <- df_hbox %>%
@@
   group_by(Status) %>% 
-  summarise(`Age (years)` = median(Age, na.rm = TRUE), age_range = range(Age, na.rm = TRUE),
+  summarise(`Age (years)` = median(Age, na.rm = TRUE), age_q1 = quantile(Age, 0.25, na.rm = TRUE), age_q3 = quantile(Age, 0.75, na.rm = TRUE),
 
-            `Temperature (°C)` = median(Temperature, na.rm = TRUE), temp_range = range(Temperature, na.rm = TRUE),
-            `Heart rate` = median(HR, na.rm = TRUE), hr_range = range(HR, na.rm = TRUE),
-            `Systolic BP` = median(BP.Systolic, na.rm = TRUE), sys_range = range(BP.Systolic, na.rm = TRUE),
-            `Diastolic BP` = median(BP.Diastolic, na.rm = TRUE), dia_range = range(BP.Diastolic, na.rm = TRUE),
-            `Respiration rate` = median(RR, na.rm = TRUE), rr_range = range(RR, na.rm = TRUE),
-            `O~2~ saturation` = median(O2.Sat, na.rm = TRUE), o2_range = range(O2.Sat, na.rm = TRUE),
+            `Temperature (°C)` = median(Temperature, na.rm = TRUE), temp_q1 = quantile(Temperature, 0.25, na.rm = TRUE), temp_q3 = quantile(Temperature, 0.75, na.rm = TRUE),
+            `Heart rate` = median(HR, na.rm = TRUE), hr_q1 = quantile(HR, 0.25, na.rm = TRUE), hr_q3 = quantile(HR, 0.75, na.rm = TRUE),
+            `Systolic BP` = median(BP.Systolic, na.rm = TRUE), sys_q1 = quantile(BP.Systolic, 0.25, na.rm = TRUE), sys_q3 = quantile(BP.Systolic, 0.75, na.rm = TRUE),
+            `Diastolic BP` = median(BP.Diastolic, na.rm = TRUE), dia_q1 = quantile(BP.Diastolic, 0.25, na.rm = TRUE), dia_q3 = quantile(BP.Diastolic, 0.75, na.rm = TRUE),
+            `Respiration rate` = median(RR, na.rm = TRUE), rr_q1 = quantile(RR, 0.25, na.rm = TRUE), rr_q3 = quantile(RR, 0.75, na.rm = TRUE),
+            `O~2~ saturation` = median(O2.Sat, na.rm = TRUE), o2_q1 = quantile(O2.Sat, 0.25, na.rm = TRUE), o2_q3 = quantile(O2.Sat, 0.75, na.rm = TRUE),
 
-            `Hematocrit (%)` = median(Hematocrit, na.rm = TRUE), hem_range = range(Hematocrit, na.rm = TRUE),
-            `Lactate (mmol/L)` = median(Lactate, na.rm = TRUE), lactate_range = range(Lactate, na.rm = TRUE),
-            `Arginine (μmol/L)` = median(Arginine.umol.L, na.rm = TRUE), arg_range = range(Arginine.umol.L, na.rm = TRUE)) %>% 
+            `Hematocrit (%)` = median(Hematocrit, na.rm = TRUE), hem_q1 = quantile(Hematocrit, 0.25, na.rm = TRUE), hem_q3 = quantile(Hematocrit, 0.75, na.rm = TRUE),
+            `Lactate (mmol/L)` = median(Lactate, na.rm = TRUE), lac_q1 = quantile(Lactate, 0.25, na.rm = TRUE), lac_q3 = quantile(Lactate, 0.75, na.rm = TRUE),
+            `Arginine (μmol/L)` = median(Arginine.umol.L, na.rm = TRUE), arg_q1 = quantile(Arginine.umol.L, 0.25, na.rm = TRUE), arg_q3 = quantile(Arginine.umol.L, 0.75, na.rm = TRUE)) %>% 
   mutate_if(is.numeric, round, 1) %>%
 
-  mutate(`Age (years)` = paste0(`Age (years)`, " (", age_range[1], ", ", age_range[2], ")"),
+  mutate(`Age (years)` = paste0(`Age (years)`, " (", age_q1, ", ", age_q3, ")"),
 
-         `Temperature (°C)` = paste0(`Temperature (°C)`, " (", temp_range[1], ", ", temp_range[2], ")"),
-         `Heart rate` = paste0(`Heart rate`, " (", hr_range[1], ", ", hr_range[2], ")"),
-         `Systolic BP` = paste0(`Systolic BP`, " (", sys_range[1], ", ", sys_range[2], ")"),
-         `Diastolic BP` = paste0(`Diastolic BP`, " (", dia_range[1], ", ", dia_range[2], ")"),
-         `Respiration rate` = paste0(`Respiration rate`, " (", rr_range[1], ", ", rr_range[2], ")"),
-         `O~2~ saturation` = paste0(`O~2~ saturation`, " (", o2_range[1], ", ", o2_range[2], ")"),
+         `Temperature (°C)` = paste0(`Temperature (°C)`, " (", temp_q1, ", ", temp_q3, ")"),
+         `Heart rate` = paste0(`Heart rate`, " (", hr_q1, ", ", hr_q3, ")"),
+         `Systolic BP` = paste0(`Systolic BP`, " (", sys_q1, ", ", sys_q3, ")"),
+         `Diastolic BP` = paste0(`Diastolic BP`, " (", dia_q1, ", ", dia_q3, ")"),
+         `Respiration rate` = paste0(`Respiration rate`, " (", rr_q1, ", ", rr_q3, ")"),
+         `O~2~ saturation` = paste0(`O~2~ saturation`, " (", o2_q1, ", ", o2_q3, ")"),
 
-         `Hematocrit (%)` = paste0(`Hematocrit (%)`, " (", hem_range[1], ", ", hem_range[2], ")"),
-         `Lactate (mmol/L)` = paste0(`Lactate (mmol/L)`, " (", lactate_range[1], ", ", lactate_range[2], ")"),
-         `Arginine (μmol/L)` = paste0(`Arginine (μmol/L)`, " (", arg_range[1], ", ", arg_range[2], ")")) %>% 
-  dplyr::select(-c(age_range, temp_range, hr_range, sys_range, dia_range, rr_range, o2_range, hem_range, lactate_range, arg_range)) %>%
+         `Hematocrit (%)` = paste0(`Hematocrit (%)`, " (", hem_q1, ", ", hem_q3, ")"),
+         `Lactate (mmol/L)` = paste0(`Lactate (mmol/L)`, " (", lac_q1, ", ", lac_q3, ")"),
+         `Arginine (μmol/L)` = paste0(`Arginine (μmol/L)`, " (", arg_q1, ", ", arg_q3, ")")) %>% 
+  dplyr::select(-c(age_q1, age_q3, 
+                   temp_q1, temp_q3, hr_q1, hr_q3, sys_q1, sys_q3, dia_q1, dia_q3, rr_q1, rr_q3, o2_q1, o2_q3, 
+                   hem_q1, hem_q3, lac_q1, lac_q3, arg_q1, arg_q3)) %>%
 
@@
 **Table 1.** A summary of subject information and hematological data.
-Group patient data presented as median (range).
+Group patient data presented as median (quartile 1, quartile 3).
 Note: Sex not recorded for 3 CM patients; Lactate data removed for HV due to inconsistent data.
@@
 
-```{r fig1, fig.height = 12}
+```{r fig1}
 
@@
 
-df_hbox_voi_long %>%
+df_hbox_voi_long %>% dplyr::filter(variable %in% c("Oxy Hb", "Deoxy Hb", "Cerebral O2 sat", "THC")) %>%
 
@@
   stat_summary(fun = "median", geom = "crossbar", aes(color = Status), size = 0.2, width = 0.5) +
-  facet_wrap(~variable, scales = "free_y", nrow = 4, ncol = 2) +
+  facet_wrap(~variable, scales = "free_y", nrow = 2, ncol = 2) +
 
@@
 
-**Figure 1.** Differences in variables measured by Oxiplex and associated α values among patient groups.
-THC levels were different between all three groups (Kruskal-Wallis adjusted p-value = 3.25x10^-6^; Dunn Test post-hoc adjusted p-values: CM vs HV = 1.66x10^-6^, CM vs UM = .032, HV vs UM = .033).
+**Figure 1.** Shown are the variables measured by Oxiplex, plotted by patient status THC levels were different between all three groups (Kruskal-Wallis adjusted p-value = 3.25x10^-6^; Dunn Test post-hoc adjusted p-values: CM vs HV = 1.66x10^-6^, CM vs UM = .032, HV vs UM = .033).
 Levels of cerebral O~2~ saturation did not differ among groups (Kruskal-Wallis adjusted p-value = .07).
-Note: removed one outlier from Deoxy Hb α (3.52).
 
-### Figure 2: Logistic regression on CM vs UM
+### Figure 2: Visual summary of α values associated with NIRS variables
+
+```{r fig2}
+
+df_hbox_voi <- df_hbox_impute %>% 
+  dplyr::select(c(Status, `Oxy Hb`, `Oxy Hb α`, `Deoxy Hb`, `Deoxy Hb α`, `Cerebral O2 sat`, `Cerebral O2 sat α`, THC, `THCα`)) %>% 
+  dplyr::filter(`Deoxy Hb α` < 3)
+
+df_hbox_voi_long <- df_hbox_voi %>% 
+  pivot_longer(2:9, names_to = "variable", values_to = "value") %>%
+  mutate(variable = factor(variable, levels = c("Oxy Hb", "Oxy Hb α",   
+                                                 "Deoxy Hb", "Deoxy Hb α", 
+                                                "Cerebral O2 sat", "Cerebral O2 sat α", 
+                                                "THC", "THCα")))
+
+my_col <- c("CM" = "#00BFC4", "HV" = "#7CAE00", "UM" = "#F8766D")
+
+df_hbox_voi_long %>% dplyr::filter(variable %in% c("Oxy Hb α", "Deoxy Hb α", "Cerebral O2 sat α", "THCα")) %>%
+
+  ggplot(aes(x = Status, y = value)) +
+  geom_violin(aes(color = Status)) +
+  geom_jitter(position = position_jitter(0.2), shape = 1) +
+  stat_summary(fun = "median", geom = "crossbar", aes(color = Status), size = 0.2, width = 0.5) +
+  facet_wrap(~variable, scales = "free_y", nrow = 2, ncol = 2) +
+
+  scale_color_manual(values = my_col) +
+  theme_bw() +
+  theme(legend.position = "none")  +
+
+  theme(strip.text = element_text(size = 15),
+        axis.title = element_text(size = 15),
+        axis.title.y = element_blank(),
+        axis.text = element_text(size = 15))
+
+```
+
+**Figure 2** Shown are the α values associated with variables measured by the Oxiplex, plotted by patient status. Note: One outlier was removed from Deoxy Hb α (3.52).
+
+### Figure 3: Logistic regression on CM vs UM
 
@@
 
-```{r fig2a}
+```{r fig3a}
 # remove healthy volunteers, Oxy Hb, and Deoxy Hb from analysis
@@
 
-```{r fig2b_prep, eval = FALSE}
+```{r fig3b_prep, eval = FALSE}
 
@@
 
-```{r fig2b_plot}
+```{r fig3b_plot}
 
@@
 
-**Figure 2** Logistic regression model discriminating CM vs UM patients using only THCα as the predictor.
+**Figure 3** Logistic regression model discriminating CM vs UM patients using only THCα as the predictor.
 **(A)** PCA of patients with CM and UM as determined by NIRS variables.

### 2021-05-13 figures.Rmd
change table 1 title
change PC vector magnitude

@@
 
-### Table 1: Summary of hematological data by patient group
+### Table 1: Summary of clinical data by patient group
 
@@
   # plot variables
-  geom_point(data = df_vars, aes(x = PC1*10, y = PC2*10)) +
-  geom_text_repel(data = df_vars, aes(x = PC1*10, y = PC2*10, label = variable, size = 10), show.legend = FALSE) +
-  geom_segment(data = df_vars, aes(x = 0, y = 0, xend = PC1*10, yend = PC2*10), arrow = arrow()) +
+  geom_point(data = df_vars, aes(x = PC1*5, y = PC2*5)) +
+  geom_text_repel(data = df_vars, aes(x = PC1*5, y = PC2*5, label = variable, size = 10), show.legend = FALSE) +
+  geom_segment(data = df_vars, aes(x = 0, y = 0, xend = PC1*5, yend = PC2*5), arrow = arrow()) +
 

### 2021-05-14 figures.Rmd
switch suppl fig 1 and 2 order

@@
 
-### Supplementary figure 1: Spearman correlations of NIRS variables 
-```{r fig_s1, fig.show = "hold", out.width = "50%", fig.height = 12, fig.width = 15}
+### Supplementary figure 1: Logistic regression curves of NIRS variables
+```{r fig_s1, fig.height = 12}
+
+df_hbox_voi_long %>% mutate(Status = as.numeric(ifelse(Status == "CM", 1, 0))) %>%
+
+  ggplot(aes(x = value, y = Status)) +
+  geom_point() +
+  geom_smooth(method = "glm", method.args = list(family = "binomial")) +
+  facet_wrap(~variable, scales = "free_x", nrow = 4, ncol = 2) +
+
+  ylab("Status: 1 = CM, 0 = UM") +
+  xlab("") +
+
+  theme_bw() +
+  theme(axis.text = element_text(size = 20),
+        axis.title = element_text(size = 25),
+        legend.text = element_text(size = 15),
+        legend.title = element_text(size = 20),
+        strip.text = element_text(size = 20))
+
+```
+
+**Figure S1.** Plotted are the individual logistic regression curves of individual variables tested for their discriminatory power between UM and CM.
+On the y-axis, 1 represents CM while 0 represents UM.
+A steeper curve indicates higher discriminatory power
+### Supplementary figure 2: Spearman correlations of NIRS variables 
+```{r fig_s2, fig.show = "hold", out.width = "50%", fig.height = 12, fig.width = 15}
 
@@
 
-**Figure S1.** Shown are the spearman correlations among variables measured by Oxiplex and their associated alpha values. Red indicates a positive correlation; blue indicates a negative correlation.
-
-### Supplementary figure 2: Logistic regression curves of NIRS variables
-```{r fig_s2, fig.height = 12}
-
-df_hbox_voi_long %>% mutate(Status = as.numeric(ifelse(Status == "CM", 1, 0))) %>%
-
-  ggplot(aes(x = value, y = Status)) +
-  geom_point() +
-  geom_smooth(method = "glm", method.args = list(family = "binomial")) +
-  facet_wrap(~variable, scales = "free_x", nrow = 4, ncol = 2) +
-
-  ylab("Status: 1 = CM, 0 = UM") +
-  xlab("") +
-
-  theme_bw() +
-  theme(axis.text = element_text(size = 20),
-        axis.title = element_text(size = 25),
-        legend.text = element_text(size = 15),
-        legend.title = element_text(size = 20),
-        strip.text = element_text(size = 20))
-
-```
-
-**Figure S2.** Plotted are the individual logistic regression curves of individual variables tested for their discriminatory power between UM and CM.
-On the y-axis, 1 represents CM while 0 represents UM.
-A steeper curve indicates higher discriminatory power
+**Figure S2.** Shown are the spearman correlations among variables measured by Oxiplex and their associated alpha values. Red indicates a positive correlation; blue indicates a negative correlation.

### 2021-05-14 figures.Rmd
fix font in supplement

@@
 A steeper curve indicates higher discriminatory power
+
 ### Supplementary figure 2: Spearman correlations of NIRS variables 

### 2021-05-18 figures.Rmd
switch supplement figures again

--- 2021-05-13 figures.Rmd
@@
 
-  dplyr::rename("Mean across resamples" = "mean", "Standard error" = "std_err") %>% 
+  dplyr::rename("Mean" = "mean", "Standard error" = "std_err") %>% 
   mutate_if(is.numeric, round, 2) %>% 
@@
 
-**Table 2.** Evaluation of the logistic regression model across 1000 bootstrap resamples.
+**Table 2.** Evaluation of the best logistic regression model select from 1000.
 

### 2021-06-01 figures.Rmd
change logistic regression model to THC, THC alpha, and hematocrit

@@
   dplyr::rename(" " = "variable",
-                "Healthy volunteers (HV)" = "HV",
-                "Uncomplicated malaria (UM) patients" = "UM",
-                "Cerebral malaria (CM) patients" = "CM")
+                "Healthy volunteers" = "HV",
+                "Uncomplicated malaria patients" = "UM",
+                "Cerebral malaria patients" = "CM")
 
   # remove lactate data for HV
-  df_clinical_res[11,]$`Healthy volunteers (HV)` <- "--"
+  df_clinical_res[11,]$`Healthy volunteers` <- "--"
 
@@
 Group patient data presented as median (quartile 1, quartile 3).
-Note: Sex not recorded for 3 CM patients; Lactate data removed for HV due to inconsistent data.
+Note: Sex not recorded for 3 CM patients; Lactate data removed for healthy volunteers due to inconsistent data.
 
@@
 
-### Figure 3: Logistic regression on CM vs UM
+### Figure 3: Logistic regression model on CM vs UM
 
 #### A. PCA
-
 ```{r fig3a}
-# remove healthy volunteers, Oxy Hb, and Deoxy Hb from analysis
-df_hbox_log_voi <- df_hbox_voi %>% 
-  dplyr::filter(Status != "HV") %>% 
-  dplyr::select(-(2:5))
+# include variables of interest (THC, THC alpha, and hematocrit), remove healthy volunteers from analysis
+df_hbox_log_voi <- df_hbox_impute %>% 
+  dplyr::select(c(2,4:5,19:22)) %>% 
+  dplyr::filter(Status != "HV")
 
@@
 
-# convert to 0s and 1s
-df_hbox_voi_01 <- df_hbox_log_voi %>% mutate(Status = ifelse(Status == "CM", 1, 0) %>% as.factor())
+# select THC, THC alpha, and hematocrit
+  df_hbox_model <- df_hbox_impute %>% 
+    dplyr::select(c(2,4:5,19:22)) %>% 
+    mutate(`1/10 * THCα` = THCα/10, `1/10 * THC` = THC/10) %>% 
+    dplyr::select(-c(THCα, THC)) %>% 
+    mutate(Status = as.factor(ifelse(Status == "CM", 1, 0))) %>%
 
-# select only significant variables
-df_hbox_model <- df_hbox_voi_01 %>% 
-  dplyr::select(c(Status, THCα)) %>% 
-  mutate(`1/10 * THCα` = THCα/10) %>% 
-  dplyr::select(-THCα)
+    dplyr::select(-c(`Cerebral O2 sat`, `Cerebral O2 sat α`, `Lactate (mmol/L)`))
 
 # resample
-set.seed(123)
+set.seed(234)
 hbox_boot <- bootstraps(df_hbox_model, times = 1000)
@@
 
-# save objects so model doesn't rerun 
+# save objects so model doesn't rerun every time we knit
 save(glm_res, hbox_final, file = "glm_model.Rdata")
@@
 
-**Figure 3** Logistic regression model discriminating CM vs UM patients using only THCα as the predictor.
-**(A)** PCA of patients with CM and UM as determined by NIRS variables.
+**Figure 3.** Logistic regression model discriminating CM vs UM patients using THC, THCα, and hematocrit as the regressors **(A)** PCA of patients with CM and UM as determined by NIRS variables.
 Color indicates patient status, ellipses show median and 95% confidence interval on the cluster.

### 2021-06-01 Scripts/additional_models.R

@@ -339,7 +339,7 @@
     mutate(p.value = as.numeric(p.value)) %>% 
     mutate(p.value = scientific(p.value, digits = 3)) %>% 
     dplyr::rename("Term" = "term", "Coefficient" = "estimate", "Standard error" = "std.error", 
-                  "Statistic" = "statistic", "p-value" = "p.value") # THC ns
+                  "Statistic" = "statistic", "p-value" = "p.value") 
   
   
   

### 2021-06-04 Scripts/raw_data_processing.R
add plot fft function

@@ -720,9 +720,49 @@
   # save filtered signal
   save(df_thc_filt_pass, file = "filtered_signals.Rdata")
 
+  load("filtered_signals.Rdata")
+  
+
   
   
+
+# FFT ---------------------------------------------------------------------
+
+
+  f_plot_fft <- function(df, num) {
+    
+    # compute fft
+    
+    fft <- df %>% filter(number == num) %>% .$roll_mean %>% fft() 
+    
+    # plot power spectra
+    
+    freq <- 500  #sample frequency in Hz 
+    duration <- df %>% filter(number == num) %>% nrow()/freq # length of signal in seconds
+    amo <- Mod(fft)
+    freqvec <- 1:length(amo)
+    
+    freqvec <- freqvec/duration 
+    df <- tibble(freq = freqvec, power = amo)
+    df <- df[(1:as.integer(0.5*freq*duration)),]
+    
+    ggplot(df[3:150,], aes(x = freq, y = power)) + 
+      geom_line(stat = "identity") +
+      geom_vline(xintercept = 0.07, lty = 2, color = "red", alpha = 0.5) +
+      geom_vline(xintercept = 0.1, lty = 2, color = "red", alpha = 0.5) +
+      
+      scale_x_continuous(breaks = seq(0, 1.5, 0.1)) +
+      theme_bw() +
+      
+      theme(axis.text.x = element_text(angle = 45, hjust = 0.9))
+
+  }
   
+ f_plot_fft(df_thc_filt_pass, 1)   
+    
+  
+  
+
 # Detrended fluctuation analysis ------------------------------------------
 
   

### 2021-06-07 figures.Rmd
added scatterplot of regressors
update for Friday seminar - convert coefficients to odds ratios

@@
                 "Oxy Hb" = "cerebral.hbox", "Oxy Hb α" = "cerebral.hbox.alpha",  
-                "Cerebral O2 sat" = "cerebral.sat", "Cerebral O2 sat α" = "cerebral.sat.alpha", 
-                "THC" = "cerebral.thc", "THCα" = "cerebral.thc.alpha")
+                "HbO2%" = "cerebral.sat", "HbO2% α" = "cerebral.sat.alpha", 
+                "[HbT]" = "cerebral.thc", "[HbT]α" = "cerebral.thc.alpha")
 
@@
 df_hbox_voi <- df_hbox_impute %>% 
-  dplyr::select(c(Status, `Oxy Hb`, `Oxy Hb α`, `Deoxy Hb`, `Deoxy Hb α`, `Cerebral O2 sat`, `Cerebral O2 sat α`, THC, `THCα`)) %>% 
+  dplyr::select(c(Status, `Oxy Hb`, `Oxy Hb α`, `Deoxy Hb`, `Deoxy Hb α`, `HbO2%`, `HbO2% α`, `[HbT]`, `[HbT]α`)) %>% 
   dplyr::filter(`Deoxy Hb α` < 3)
@@
                                                  "Deoxy Hb", "Deoxy Hb α", 
-                                                "Cerebral O2 sat", "Cerebral O2 sat α", 
-                                                "THC", "THCα")))
+                                                "HbO2%", "HbO2% α", 
+                                                "[HbT]", "[HbT]α")))
 
@@
 
-df_hbox_voi_long %>% dplyr::filter(variable %in% c("Oxy Hb", "Deoxy Hb", "Cerebral O2 sat", "THC")) %>%
+df_hbox_voi_long %>% dplyr::filter(variable %in% c("Oxy Hb", "Deoxy Hb", "HbO2%", "[HbT]")) %>%
 
@@
 
-# THC
-kruskal.test(THC ~ Status, data = df_hbox_voi)
-dunnTest(THC ~ Status, data = df_hbox_voi, method = "bh")
+# [HbT]
+kruskal.test(`[HbT]` ~ Status, data = df_hbox_voi)
+dunnTest(`[HbT]` ~ Status, data = df_hbox_voi, method = "bh")
 
-# Cerebral O2 sat
-kruskal.test(`Cerebral O2 sat` ~ Status, data = df_hbox_voi)
+# HbO2%
+kruskal.test(`HbO2%` ~ Status, data = df_hbox_voi)
 
@@
 
-**Figure 1.** Shown are the variables measured by Oxiplex, plotted by patient status THC levels were different between all three groups (Kruskal-Wallis adjusted p-value = 3.25x10^-6^; Dunn Test post-hoc adjusted p-values: CM vs HV = 1.66x10^-6^, CM vs UM = .032, HV vs UM = .033).
+**Figure 1.** Shown are the variables measured by Oxiplex, plotted by patient status [HbT] levels were different between all three groups (Kruskal-Wallis adjusted p-value = 3.25x10^-6^; Dunn Test post-hoc adjusted p-values: CM vs HV = 1.66x10^-6^, CM vs UM = .032, HV vs UM = .033).
 Levels of cerebral O~2~ saturation did not differ among groups (Kruskal-Wallis adjusted p-value = .07).
@@
 
-df_hbox_voi_long %>% dplyr::filter(variable %in% c("Oxy Hb α", "Deoxy Hb α", "Cerebral O2 sat α", "THCα")) %>%
+df_hbox_voi_long %>% dplyr::filter(variable %in% c("Oxy Hb α", "Deoxy Hb α", "HbO2% α", "[HbT]α")) %>%
 
@@
   stat_summary(fun = "median", geom = "crossbar", aes(color = Status), size = 0.2, width = 0.5) +
+
+  geom_hline(yintercept = 0.4, alpha = 0.4, lty = 2) +
+  geom_hline(yintercept = 0.6, alpha = 0.4, lty = 2) +
+  geom_hline(yintercept = 1.0, alpha = 0.4, lty = 2) +
+  geom_hline(yintercept = 1.5, alpha = 0.4, lty = 2) +
+
   facet_wrap(~variable, scales = "free_y", nrow = 2, ncol = 2) +
 
+  ylim(0, 1.5) +
   scale_color_manual(values = my_col) +
   theme_bw() +
-  theme(legend.position = "none")  +
 
@@
         axis.title.y = element_blank(),
-        axis.text = element_text(size = 15))
+        axis.text = element_text(size = 15),
+        legend.position = "none")
 
@@
 
-**Figure 2** Shown are the α values associated with variables measured by the Oxiplex, plotted by patient status. Note: One outlier was removed from Deoxy Hb α (3.52).
+```{r fig2.pvals, eval = FALSE}
+
+# [HbT]α
+kruskal.test(`[HbT]α` ~ Status, data = df_hbox_voi)
+dunnTest(`[HbT]α` ~ Status, data = df_hbox_voi, method = "bh")
+
+```
+
+**Figure 2** Shown are the alpha values associated with variables measured by the Oxiplex, plotted by patient status. Note: One outlier was removed from Deoxy Hb α (3.52).
 
@@
 ```{r fig3a}
-# include variables of interest (THC, THC alpha, and hematocrit), remove healthy volunteers from analysis
+# include variables of interest ([HbT], [HbT] alpha, and hematocrit), remove healthy volunteers from analysis
 df_hbox_log_voi <- df_hbox_impute %>% 
@@
 
-# select THC, THC alpha, and hematocrit
+# select [HbT], [HbT] alpha, and hematocrit
   df_hbox_model <- df_hbox_impute %>% 
     dplyr::select(c(2,4:5,19:22)) %>% 
-    mutate(`1/10 * THCα` = THCα/10, `1/10 * THC` = THC/10) %>% 
-    dplyr::select(-c(THCα, THC)) %>% 
+    mutate(`1/10 * [HbT]α` = `[HbT]α`/10, `[HbT] (nM)` = `[HbT]`/1000) %>% 
+    dplyr::select(-c(`[HbT]α`, `[HbT]`)) %>% 
     mutate(Status = as.factor(ifelse(Status == "CM", 1, 0))) %>%
 
-    dplyr::select(-c(`Cerebral O2 sat`, `Cerebral O2 sat α`, `Lactate (mmol/L)`))
+    dplyr::select(-c(`HbO2%`, `HbO2% α`, `Lactate (mmol/L)`))
 
@@
 
-**Figure 3.** Logistic regression model discriminating CM vs UM patients using THC, THCα, and hematocrit as the regressors **(A)** PCA of patients with CM and UM as determined by NIRS variables.
+**Figure 3.** Logistic regression model discriminating CM vs UM patients using cerebral tissue hemoglobin concentration, cerebral tissue hemoglobin concentration alpha, and hematocrit as the regressors **(A)** PCA of patients with CM and UM as determined by NIRS variables.
 Color indicates patient status, ellipses show median and 95% confidence interval on the cluster.
@@
 hbox_final %>% 
-  tidy() %>% 
+  tidy(exponentiate = TRUE) %>% 
   mutate(p.value = as.character(p.value)) %>% 
@@
   mutate(p.value = scientific(p.value, digits = 3)) %>% 
-  dplyr::rename("Term" = "term", "Coefficient" = "estimate", "Standard error" = "std.error", 
+  dplyr::rename("Term" = "term", "Odds ratio" = "estimate", "Standard error" = "std.error", 
                 "Statistic" = "statistic", "p-value" = "p.value") %>%
@@
 
-### Supplementary figure 1: Spearman correlations of NIRS variables 
+### Figure 4: Relationships among [HbT], [HbT] alpha, and hematocrit
+```{r fig4}
+
+df_hbox_impute %>% filter(Status != "HV") %>%
+
+  ggplot(aes(x = `[HbT]`, y = `Hematocrit (%)`)) +
+  geom_point(aes(color = Status, size = `[HbT]α`), alpha = 0.7) +
+
+  xlab("[HbT] (μM)") +
+
+  theme_bw() +
+  theme(axis.text = element_text(size = 20),
+        axis.title = element_text(size = 25),
+        legend.text = element_text(size = 15),
+        legend.title = element_text(size = 20))
+
+```
+
+**Figure 4.**
+
+### Supplementary figure 1: Spearman correlations of potential regressors
 ```{r fig_s1, fig.show = "hold", out.width = "50%", fig.height = 12, fig.width = 15}
@@
 
-m_cor_hv <- df_hbox_voi %>% f_calc_cor(status = "HV")
+# create df of potential regressors
+df_hbox_reg <- df_hbox_impute %>% 
+  dplyr::select(c(Status, `Hematocrit (%)`, `Lactate (mmol/L)`, `Oxy Hb`, `Oxy Hb α`, `Deoxy Hb`, `Deoxy Hb α`, `HbO2%`, `HbO2% α`, `[HbT]`, `[HbT]α`)) %>% 
+  dplyr::filter(`Deoxy Hb α` < 3)
 
-m_cor_um <- df_hbox_voi %>% f_calc_cor(status = "UM")
+m_cor_hv <- df_hbox_reg %>% f_calc_cor(status = "HV")
 
-m_cor_cm <- df_hbox_voi %>% f_calc_cor(status = "CM")
+m_cor_um <- df_hbox_reg %>% f_calc_cor(status = "UM")
+
+m_cor_cm <- df_hbox_reg %>% f_calc_cor(status = "CM")
 
@@
 
-    mutate_if(is.factor, factor, levels = c("Oxy Hb", "Deoxy Hb", "Cerebral O2 sat", "THC", "Oxy Hb α", "Deoxy Hb α", "Cerebral O2 sat α", "THCα")) %>%
+    mutate_if(is.factor, factor, levels = c("Lactate (mmol/L)", "Hematocrit (%)", "Oxy Hb", "Deoxy Hb", "HbO2%", "[HbT]", "Oxy Hb α", "Deoxy Hb α", "HbO2% α", "[HbT]α")) %>%
 
@@
   # create plots
-  m_cor_hv %>% f_plot_cor() + ggtitle("HV variable correlations") + theme(legend.position = "none")
+  m_cor_hv %>% f_plot_cor() + ggtitle("Healthy volunteers") + theme(legend.position = "none")
 
-  m_cor_um %>% f_plot_cor() + ggtitle("UM variable correlations") + theme(legend.position = "none")
+  m_cor_um %>% f_plot_cor() + ggtitle("Uncomplicated malaria patients") + theme(legend.position = "none")
 
-  m_cor_cm %>% f_plot_cor() + ggtitle("CM variable correlations") + theme(legend.position = "none")
+  m_cor_cm %>% f_plot_cor() + ggtitle("Cerebral malaria patients") + theme(legend.position = "none")
 
@@
 
-### Supplementary figure 2: Logistic regression curves of NIRS variables
+### Supplementary figure 2: Logistic regression curves of potential regressors
 ```{r fig_s2, fig.height = 12}
 
-df_hbox_voi_long %>% mutate(Status = as.numeric(ifelse(Status == "CM", 1, 0))) %>%
+# pivot longer and change factor levels
+df_hbox_reg_long <- df_hbox_reg %>% 
+  pivot_longer(2:11, names_to = "variable", values_to = "value") %>% 
+  mutate(variable = factor(variable, levels = c("Lactate (mmol/L)", "Hematocrit (%)", "Oxy Hb", "Oxy Hb α", "Deoxy Hb", "Deoxy Hb α", "HbO2%", "HbO2% α" ,"[HbT]", "[HbT]α")))
+
+# plot as individual regression curves
+df_hbox_reg_long %>% mutate(Status = as.numeric(ifelse(Status == "CM", 1, 0))) %>%
 
@@
   geom_smooth(method = "glm", method.args = list(family = "binomial")) +
-  facet_wrap(~variable, scales = "free_x", nrow = 4, ncol = 2) +
+  facet_wrap(~variable, scales = "free_x", nrow = 5, ncol = 2) +
 

### 2021-06-29 figures.Rmd
updated figures

@@
 ---
+
+<style type="text/css">
+  body{
+  font-family: Arial;
+  font-size: 12pt;
+}
+</style>
 
@@
 library(reshape)
+library(extrafont)
+library(grDevices)
 
 load("glm_model.Rdata")
+
+font_import()
 
@@
 
-                "Deoxy Hb" = "cerebral.hbdeox", "Deoxy Hb α" = "cerebral.hbdeox.alpha", 
-                "Oxy Hb" = "cerebral.hbox", "Oxy Hb α" = "cerebral.hbox.alpha",  
+                "[Hb]" = "cerebral.hbdeox", "[Hb]α" = "cerebral.hbdeox.alpha", 
+                "[HbO2]" = "cerebral.hbox", "[HbO2]α" = "cerebral.hbox.alpha",  
                 "HbO2%" = "cerebral.sat", "HbO2% α" = "cerebral.sat.alpha", 
@@
 df_hbox_voi <- df_hbox_impute %>% 
-  dplyr::select(c(Status, `Oxy Hb`, `Oxy Hb α`, `Deoxy Hb`, `Deoxy Hb α`, `HbO2%`, `HbO2% α`, `[HbT]`, `[HbT]α`)) %>% 
-  dplyr::filter(`Deoxy Hb α` < 3)
+  dplyr::select(c(Status, `[HbO2]`, `[HbO2]α`, `[Hb]`, `[Hb]α`, `HbO2%`, `HbO2% α`, `[HbT]`, `[HbT]α`)) %>% 
+  dplyr::filter(`[Hb]α` < 3)
 
@@
   pivot_longer(2:9, names_to = "variable", values_to = "value") %>%
-  mutate(variable = factor(variable, levels = c("Oxy Hb", "Oxy Hb α",   
-                                                 "Deoxy Hb", "Deoxy Hb α", 
+  mutate(variable = factor(variable, levels = c("[HbO2]", "[HbO2]α",   
+                                                 "[Hb]", "[Hb]α", 
                                                 "HbO2%", "HbO2% α", 
@@
 
-df_hbox_voi_long %>% dplyr::filter(variable %in% c("Oxy Hb", "Deoxy Hb", "HbO2%", "[HbT]")) %>%
+df_hbox_voi_long %>% dplyr::filter(variable %in% c("[HbO2]", "[Hb]", "HbO2%", "[HbT]")) %>%
 
@@
 
-  theme(strip.text = element_text(size = 15),
+  theme(text = element_text(family = "Arial"),
+        strip.text = element_text(size = 15),
         axis.title = element_text(size = 15),
@@
 
-df_hbox_voi_long %>% dplyr::filter(variable %in% c("Oxy Hb α", "Deoxy Hb α", "HbO2% α", "[HbT]α")) %>%
+df_hbox_voi_long %>% dplyr::filter(variable %in% c("[HbO2]α", "[Hb]α", "HbO2% α", "[HbT]α")) %>%
 
@@
   theme_bw() +
-  theme(axis.text = element_text(size = 20),
+  theme(text = element_text(family = "Arial"),
+        axis.text = element_text(size = 20),
         axis.title = element_text(size = 25),
@@
   theme_bw() +
-  theme(axis.title = element_text(size = 25),
+  theme(text = element_text(family = "Arial"),
+        axis.title = element_text(size = 25),
         axis.text = element_text(size = 20),
@@
 df_hbox_reg <- df_hbox_impute %>% 
-  dplyr::select(c(Status, `Hematocrit (%)`, `Lactate (mmol/L)`, `Oxy Hb`, `Oxy Hb α`, `Deoxy Hb`, `Deoxy Hb α`, `HbO2%`, `HbO2% α`, `[HbT]`, `[HbT]α`)) %>% 
-  dplyr::filter(`Deoxy Hb α` < 3)
+  dplyr::select(c(Status, `Hematocrit (%)`, `Lactate (mmol/L)`, `[HbO2]`, `[HbO2]α`, `[Hb]`, `[Hb]α`, `HbO2%`, `HbO2% α`, `[HbT]`, `[HbT]α`)) %>% 
+  dplyr::filter(`[Hb]α` < 3)
 
@@
 
-    mutate_if(is.factor, factor, levels = c("Lactate (mmol/L)", "Hematocrit (%)", "Oxy Hb", "Deoxy Hb", "HbO2%", "[HbT]", "Oxy Hb α", "Deoxy Hb α", "HbO2% α", "[HbT]α")) %>%
+    mutate_if(is.factor, factor, levels = c("Lactate (mmol/L)", "Hematocrit (%)", "[HbO2]", "[Hb]", "HbO2%", "[HbT]", "[HbO2]α", "[Hb]α", "HbO2% α", "[HbT]α")) %>%
 
@@
     theme_bw() +
-    theme(axis.text.x = element_text(angle = 45, hjust = 0.9),
+    theme(text = element_text(family = "Arial"),
+          axis.text.x = element_text(angle = 45, hjust = 0.9),
           axis.text = element_text(size = 35),
@@
   pivot_longer(2:11, names_to = "variable", values_to = "value") %>% 
-  mutate(variable = factor(variable, levels = c("Lactate (mmol/L)", "Hematocrit (%)", "Oxy Hb", "Oxy Hb α", "Deoxy Hb", "Deoxy Hb α", "HbO2%", "HbO2% α" ,"[HbT]", "[HbT]α")))
+  mutate(variable = factor(variable, levels = c("Lactate (mmol/L)", "Hematocrit (%)", "[HbO2]", "[HbO2]α", "[Hb]", "[Hb]α", "HbO2%", "HbO2% α" ,"[HbT]", "[HbT]α")))
 

### 2021-06-29 Functions/f_filter_signal.R

@@ -1,64 +1,21 @@
 
-## A function to find the longest segment of the signal (where t >= 200s) within 0.5 standard deviations ##
-## Purpose is to filter blips out of signal
+## A function to find the longest segment of the signal (where t >= 200s) with variance under 1000 ##
 
-## find the longest list with a standard deviation below a certain limit is to check all potential sublists
+## Since variance does not strictly increase or decrease, the best way to find the longest list with a variance
+## below a certain limit is to check all potential sublists, starting with the biggest ones and then shrinking
+## until we find a smaller one. 
 
-## our parameters: f_filter_signal(df, max_sd = 0.5)
-
-
-f_filter_signal <- function(df, max_sd) {
-  
-  start = 1
-  end = 1
-  max_start = 1
-  max_end = 0
-  
-  df <- df %>% filter(!is.na(roll_mean))
-  
-  mean <- df$roll_mean %>% mean(na.rm = TRUE)
-  sd <- df$roll_mean %>% sd(na.rm = TRUE)
-  
-  for (i in 1:nrow(df)) {
-    
-    if ((mean - max_sd*sd) < df[i,]$roll_mean && df[i,]$roll_mean < (mean + max_sd*sd)) {
-      
-      end <- i
-      
-    } else {
-      
-      start <- i
-      
-    }
-    
-    if ((end - start) > (max_end - max_start)) {
-      
-      max_end <- end
-      max_start <- start
-      
-    }
-    
-  }
-  
-  return(df[max_start:max_end,])
-  
-}
+## our parameters: f_filter_signal(df, min_length = 20000, max_var = 1000)
 
 
-
-
-
-## Alternative function where you can filter with a max variance
-
-f_filter_signal_var <- function(df, min_length, max_var){
+f_filter_signal <- function(df, min_length, max_var){
   
   for (length in nrow(df):min_length) {
     
     for (start in 1:(nrow(df) - length + 1)) {
       
       df_temp <- df[start:(start + length - 1),]
-      df_temp <- df_temp %>% mutate(start = start, length = length)
-      
+
       var <- df_temp$roll_mean %>% var(na.rm = TRUE)
       
       if (var <= max_var) {
@@ -78,7 +35,35 @@ f_filter_signal_var <- function(df, min_length, max_var){
 
 
 
+doParallel::registerDoParallel()
[…]

### 2021-06-29 Scripts/raw_data_processing.R
move DFA to new script

@@ -97,19 +97,6 @@
   
   
 
-  
-  
-  
-  
-  
-
-
-
-
-
-
-
-
 
 # remove blips ------------------------------------------------------------
 
@@ -637,26 +624,6 @@
   
 
   
-
-  
-  
-  
-  
-  
-  
-  
-  
-  
-  
-  
-    
-  
-  
-  
-  
-  
-  
-  
   
      
 # calculate THC and THC sd ------------------------------------------------
@@ -694,6 +661,7 @@
 
 # band pass filter ---------------------------------------------------------
 
+  
 # filter each signal to below 0.1 Hz (physiological oscillations, including hemodynamics/cerebral autoregulation)
   
   library(seewave)
@@ -763,41 +731,10 @@
   
   
 
-# Detrended fluctuation analysis ------------------------------------------
+
 
   
-  library(DFA)
-  
-  df_thc_filt_pass1 <- df_thc_filt_pass %>% dplyr::filter(number == 1)
-  
-  dfa::.DFA(df_thc_filt_pass1$band_pass)
-  
-  file <- NYA2008
-  dfa.res <- DFA::DFA(NYA2008)
-  
-  
-library(MFDFA)  
-  a<-0.9
-  N<-1024
-  tsx<-MFsim(N,a)
-  scale=10:100
-  q<--10:10
-  m<-1
-  b<-MFDFA(tsx, scale, m, q) 
-  
-  dev.new()
-  par(mai=rep(1, 4))
-  plot(q, b$Hq, col=1, axes= F, ylab=expression('h'[q]), pch=16, cex.lab=1.8,
-       cex.axis=1.8, main="Hurst exponent",
-       ylim=c(min(b$Hq),max(b$Hq)))
-  grid(col="midnightblue")
-  axis(1)
-  axis(2)
-  
-  alpha <- b$spec %>% as_tibble()
-  alpha %>% ggplot(aes(x = 1:20, y = hq)) + geom_point()
-  
-  
+
   
   
   

### 2021-06-30 Scripts/raw_data_processing.R
move dfa to separate script

@@ -53,7 +53,7 @@
   # smooth data with 10s moving average
   df_thc <- df_thc %>% 
     group_by(number) %>% 
-    mutate(roll_mean = zoo::rollmean(THC, k = 1000, fill = NA)) %>% 
+    mutate(roll_mean = zoo::rollmean(THC, k = 50, fill = NA)) %>% 
     ungroup()
   
 
@@ -643,7 +643,7 @@
     theme(legend.position = "none") 
     
   
-  # THC sd
+  # THC sd --> not included in paper
   df_THCsd <- df_thc_filt %>% 
     filter(!is.na(roll_mean)) %>% 
     group_by(number, Status) %>% 
@@ -665,10 +665,13 @@
 # filter each signal to below 0.1 Hz (physiological oscillations, including hemodynamics/cerebral autoregulation)
   
   library(seewave)
+  library(dplR)
+  library(data.table)
+  # df_thc_filt <- df_thc_filt_pass[,1:5] # uncomment to recreate df_thc_filt
 
   l_thc_filt_pass <- unique(df_thc_filt$number) %>% 
     purrr::map(~dplyr::filter(df_thc_filt, number == .x, !is.na(roll_mean))) %>% 
-    purrr::map(~mutate(.x, band_pass = pass.filt(.x$roll_mean, W = c(0.07, .3), type = "pass", method = "Chebyshev", n = 4)))
+    purrr::map(~mutate(.x, band_pass = pass.filt(.x$roll_mean, W = c(0.01, .1), type = "pass", method = "Chebyshev", n = 4)))
     
   df_thc_filt_pass <- rbindlist(l_thc_filt_pass) %>% as_tibble()
   
@@ -691,7 +694,6 @@
   load("filtered_signals.Rdata")
   
 
-  
   
 
 # FFT ---------------------------------------------------------------------
@@ -716,7 +718,7 @@
     
     ggplot(df[3:150,], aes(x = freq, y = power)) + 
       geom_line(stat = "identity") +
-      geom_vline(xintercept = 0.07, lty = 2, color = "red", alpha = 0.5) +
+      geom_vline(xintercept = 0.01, lty = 2, color = "red", alpha = 0.5) +
       geom_vline(xintercept = 0.1, lty = 2, color = "red", alpha = 0.5) +
       
       scale_x_continuous(breaks = seq(0, 1.5, 0.1)) +
@@ -726,7 +728,7 @@
 
   }
[…]

### 2021-06-30 Scripts/DFA.R
preliminary dfa analysis

@@ -1,0 +1,316 @@
+
+
+
+
+# load in processed signal data -------------------------------------------
+
+  load("filtered_signals.Rdata")
+
+
+
+
+
+# calculate amplitude envelope using Hilbert transform --------------------
+
+  library(seewave)
+  library(zoo)
+  library(DescTools)
+
+  # add milliseconds in to Time
+  df_thc_filt_pass <- df_thc_filt_pass %>% 
+    group_by(number, Time) %>% 
+    left_join(count(.)) %>% 
+    mutate(Time = ifelse(n == 49, Time + 0.02*row_number(), Time + 0.02*(row_number() - 1)))
+  
+  # 1 second rolling mean?
+  df_thc_filt_pass <- df_thc_filt_pass %>% group_by(number) %>% 
+    mutate(roll_mean = zoo::rollmean(THC, k = 50, fill = NA)) %>% 
+    ungroup()
+
+  # or we can just smooth the amplitude envelope
+  hil_amp <- env(df_thc_filt_pass1$THC, f = 50, msmooth = c(50, 0), plot = FALSE) %>% 
+    as_tibble(.name_repair = "universal") %>% 
+    dplyr::rename("hil_amp" = "...1") %>% 
+    slice(rep(1:n(), each = 50)) # smoothed hilbert transform 1 sec
+  
+  # check with plotting
+  df_thc_filt_pass1[1:58450,] %>% bind_cols(hil_amp) %>% 
+    
+    ggplot(aes(x = Time)) +
+    geom_line(aes(y = THC)) +
+    geom_line(aes(y = hil_amp), color = "red")
+  
+  # use the hilbert amplitude envelope for all signals
+  l_hil_amp <- unique(df_thc_filt_pass$number) %>% 
+    purrr::map(~dplyr::filter(df_thc_filt_pass, number == .x, !is.na(THC))) %>% 
+    purrr::map(~env(.x$THC, f = 50, msmooth = c(50, 0), plot = FALSE) %>% 
+                         as_tibble(.name_repair = "universal") %>% 
+                         dplyr::rename("hil_amp" = "...1") %>% 
+                         slice(rep(1:n(), each = 50))) # nrow rounds down to the nearest 50
+  names(l_hil_amp) <-  unique(df_thc_filt_pass$number)
+
+  # combine with df_thc_filt_pass
+  l_signal_abbr <- unique(df_thc_filt_pass$number) %>% 
[…]

### 2021-07-01 Scripts/raw_data_processing.R
add milliseconds to time, 1s rolling mean

@@ -663,6 +663,7 @@
 
   
 # filter each signal to below 0.1 Hz (physiological oscillations, including hemodynamics/cerebral autoregulation)
+  # & above 0.01 (anything below is head displacements and motion noise)
   
   library(seewave)
   library(dplR)
@@ -675,7 +676,7 @@
     
   df_thc_filt_pass <- rbindlist(l_thc_filt_pass) %>% as_tibble()
   
-  # recalculate sd
+  # recalculate sd (not for use in paper)
   df_thc_filt_pass %>% 
     dplyr::filter(!is.na(roll_mean)) %>% 
     group_by(number, Status) %>% 
@@ -688,10 +689,6 @@
     theme_bw() +
     theme(legend.position = "none") 
   
-  # save filtered signal
-  save(df_thc_filt_pass, file = "filtered_signals.Rdata")
-
-  load("filtered_signals.Rdata")
   
 
   
@@ -740,3 +737,27 @@
   
   
   
+ 
+
+# r -----------------------------------------------------------------------
+
+ # add milliseconds in to Time
+ df_thc_filt_pass <- df_thc_filt_pass %>% 
+   group_by(number, Time) %>% 
+   left_join(count(.)) %>% 
+   mutate(Time = ifelse(n == 49, Time + 0.02*row_number(), Time + 0.02*(row_number() - 1))) %>% 
+   dplyr::select(-n)
+ 
+ # 1 second rolling mean?
+ df_thc_filt_pass <- df_thc_filt_pass %>% group_by(number) %>% 
+   mutate(roll_mean = zoo::rollmean(THC, k = 50, fill = NA)) %>% 
+   ungroup()
+ 
+ # save filtered signal
+ save(df_thc_filt_pass, file = "filtered_signals.Rdata")
+ 
+ load("filtered_signals.Rdata")
+ 
+ 
+ df_thc_filt_pass <-  df_thc_filt_pass %>% select(-n)
+ 

### 2021-07-02 Scripts/raw_data_processing.R
update preprocessing pipeline - SG filter, band pass filter (0.01-.1 Hz)

@@ -50,7 +50,7 @@
   df_thc$Status[df_thc$Status == "HC"] <- "HV"
   df_thc <- df_thc %>% dplyr::filter(Status != "-0") # unclear what group this patient was in
   
-  # smooth data with 10s moving average
+  # smooth data with 1s moving average
   df_thc <- df_thc %>% 
     group_by(number) %>% 
     mutate(roll_mean = zoo::rollmean(THC, k = 50, fill = NA)) %>% 
@@ -659,39 +659,26 @@
   
   
 
-# band pass filter ---------------------------------------------------------
+# savitzy-golay signal smoothing  ---------------------------------------------------------
 
   
-# filter each signal to below 0.1 Hz (physiological oscillations, including hemodynamics/cerebral autoregulation)
-  # & above 0.01 (anything below is head displacements and motion noise)
+  # 4th order savitzy-golay filter
+  library(signal)
   
-  library(seewave)
-  library(dplR)
-  library(data.table)
-  # df_thc_filt <- df_thc_filt_pass[,1:5] # uncomment to recreate df_thc_filt
-
-  l_thc_filt_pass <- unique(df_thc_filt$number) %>% 
-    purrr::map(~dplyr::filter(df_thc_filt, number == .x, !is.na(roll_mean))) %>% 
-    purrr::map(~mutate(.x, band_pass = pass.filt(.x$roll_mean, W = c(0.01, .1), type = "pass", method = "Chebyshev", n = 4)))
-    
-  df_thc_filt_pass <- rbindlist(l_thc_filt_pass) %>% as_tibble()
+  df_thc_filt <- df_thc_filt %>% 
+    dplyr::select(-roll_mean) %>% 
+    group_by(number) %>% 
+    mutate(sg_filt = sgolayfilt(THC, p = 3)) %>% 
+    ungroup()
   
-  # recalculate sd (not for use in paper)
-  df_thc_filt_pass %>% 
-    dplyr::filter(!is.na(roll_mean)) %>% 
-    group_by(number, Status) %>% 
-    summarise(THC_sd = sd(band_pass)) %>%  
-    
-    ggplot(aes(x = Status, y = THC_sd)) +
-    geom_violin(aes(color = Status)) +
-    geom_jitter(position = position_jitter(0.2), shape = 1) +
-    stat_summary(fun = "median", geom = "crossbar", aes(color = Status), size = 0.2, width = 0.5) +
[…]

### 2021-07-02 Scripts/DFA.R
different renditions of DFA - current = on SG filt, extract second slope

@@ -12,52 +12,41 @@
 
 # calculate amplitude envelope using Hilbert transform --------------------
 
-  library(seewave)
-  library(zoo)
-  library(DescTools)
+  # library(seewave)
+  # library(zoo)
+  # library(DescTools)
 
-  # add milliseconds in to Time
-  df_thc_filt_pass <- df_thc_filt_pass %>% 
-    group_by(number, Time) %>% 
-    left_join(count(.)) %>% 
-    mutate(Time = ifelse(n == 49, Time + 0.02*row_number(), Time + 0.02*(row_number() - 1)))
-  
-  # 1 second rolling mean?
-  df_thc_filt_pass <- df_thc_filt_pass %>% group_by(number) %>% 
-    mutate(roll_mean = zoo::rollmean(THC, k = 50, fill = NA)) %>% 
-    ungroup()
-
-  # or we can just smooth the amplitude envelope
-  hil_amp <- env(df_thc_filt_pass1$THC, f = 50, msmooth = c(50, 0), plot = FALSE) %>% 
-    as_tibble(.name_repair = "universal") %>% 
-    dplyr::rename("hil_amp" = "...1") %>% 
-    slice(rep(1:n(), each = 50)) # smoothed hilbert transform 1 sec
-  
-  # check with plotting
-  df_thc_filt_pass1[1:58450,] %>% bind_cols(hil_amp) %>% 
-    
-    ggplot(aes(x = Time)) +
-    geom_line(aes(y = THC)) +
-    geom_line(aes(y = hil_amp), color = "red")
-  
-  # use the hilbert amplitude envelope for all signals
-  l_hil_amp <- unique(df_thc_filt_pass$number) %>% 
-    purrr::map(~dplyr::filter(df_thc_filt_pass, number == .x, !is.na(THC))) %>% 
-    purrr::map(~env(.x$THC, f = 50, msmooth = c(50, 0), plot = FALSE) %>% 
-                         as_tibble(.name_repair = "universal") %>% 
-                         dplyr::rename("hil_amp" = "...1") %>% 
-                         slice(rep(1:n(), each = 50))) # nrow rounds down to the nearest 50
-  names(l_hil_amp) <-  unique(df_thc_filt_pass$number)
-
-  # combine with df_thc_filt_pass
-  l_signal_abbr <- unique(df_thc_filt_pass$number) %>% 
-    purrr::map(~dplyr::filter(df_thc_filt_pass, number == .x)) %>% 
[…]

### 2021-07-06 Scripts/DFA.R
additional plots

@@ -17,7 +17,6 @@
     l_means <- 1:length(l_thc_filt_pass) %>% purrr::map(~mean(l_thc_filt_pass[[.x]]$sg_filt, na.rm = TRUE))
     
     # subtract the mean and take the cumulative sum
-    l_thc_cumsum <- 1:length(l_thc_filt_pass) %>% purrr::map(~filter(l_thc_filt_pass[[.x]], !is.na(sg_filt)))
     l_thc_cumsum <- 1:length(l_means) %>% purrr::map(~mutate(l_thc_filt_pass[[.x]], mean = l_means[[.x]]))
     df_thc_cumsum <- rbindlist(l_thc_cumsum) %>% as_tibble()
     
@@ -27,27 +26,28 @@
       mutate(cumsum = cumsum(subtract_mean)) %>% 
       ungroup()
     
+
     # plot
-    df_thc_cumsum %>% filter(number == 1) %>% 
-      
+    df_thc_cumsum %>% dplyr::filter(number == 1) %>% 
       ggplot(aes(x = Time)) +
-      geom_line(aes(y = sg_filt)) +
-      geom_line(aes(y = cumsum), color = "red")
-    
-    
-    
-    
+      geom_line(aes(y = cumsum)) +
+      geom_line(aes(y = sg_filt), color = "red") +
+      
+      theme_bw() +
+      ggtitle("Cumulative sum of Savitzy-Golar smoothed signal")
+        
+
+        
 # split into windows --------------------------------------------------------------------
     
     
     
     # start with signal from patient 1
-    df_thc_cumsum1 <- df_thc_cumsum %>% filter(number == 1)
     df_thc_cumsum1 %>% nrow()/50 # signal is 1169 seconds long
     
     # determine window lengths
     
-      # criteria:
+      # criteria (Hardstone et al 2012):
       # lower end = at least 4 samples (linear detrending will perform poorly with less points)
       # high end = <10% of signal length (anything higher is more noisy bc less than 10 windows to average)
       # need to be equally spaced on a log10 scale (gives equal weight to time scales when fitting  a line)
@@ -122,31 +122,29 @@
       
       l_splits[[window_num]] %>% 
         ggplot(aes(x = Time)) +
-        geom_line(aes(y = THC), color = "blue") +
-        geom_line(aes(y = cumsum), color = "black") +
-        geom_smooth(aes(y = cumsum), method = "lm", color = "red") +
[…]

### 2021-07-06 Scripts/raw_data_processing.R

@@ -28,33 +28,33 @@
   library(zoo)
   
   # format to just THC in workable form
-  l_thc <- 1:length(l_raw_data) %>% 
+  l_raw_data <- 1:length(l_raw_data) %>% 
     
     purrr::map(~mutate(l_raw_data[[.x]], Status = l_names[[.x]])) %>% 
-    purrr::map(~filter(.x, !grepl("Patient|Data|[R|r]aw", V1))) %>% 
-    purrr::map(~dplyr::select(.x, c(ncol(.x), 2,4))) %>% 
-    purrr::map(~dplyr::rename(.x, "Time" = "V2", "THC" = "V4")) %>% 
+    purrr::map(~dplyr::filter(.x, !grepl("Patient|Data|[R|r]aw", V1))) %>% 
+    purrr::map(~dplyr::select(.x, c(ncol(.x), 2:4))) %>% 
+    purrr::map(~dplyr::rename(.x, "Time" = "V2", "O2_sat" = "V3", "THC" = "V4")) %>% 
     purrr::map(~mutate(.x, Time = as.numeric(Time))) %>% 
-    purrr::map(~mutate(.x, THC = as.numeric(THC)))
+    purrr::map(~mutate(.x, THC = as.numeric(THC))) %>% 
+    purrr::map(~mutate(.x, O2_sat = as.numeric(O2_sat)))
   
   # renumber patients
-  l_thc <- 1:length(l_thc) %>% 
-    purrr::map(~mutate(l_thc[[.x]], number = c(1:length(l_thc))[.x])) %>% 
+  l_raw_data <- 1:length(l_raw_data) %>% 
+    purrr::map(~mutate(l_raw_data[[.x]], number = c(1:length(l_raw_data))[.x])) %>% 
     purrr::map(~dplyr::select(.x, c(number, everything())))
   
   # condense list to df
-  df_thc <- rbindlist(l_thc) %>% 
-    as_tibble() 
+  df_raw_data <- rbindlist(l_raw_data) %>% as_tibble() 
   
   # make sure status is HV, UM, or CM
-  df_thc$Status[df_thc$Status == "HC"] <- "HV"
-  df_thc <- df_thc %>% dplyr::filter(Status != "-0") # unclear what group this patient was in
+  df_raw_data$Status[df_raw_data$Status == "HC"] <- "HV"
[…]

### 2021-07-07 Functions/f_filter_signal.R
while it would be nice to automate signal filtering, this iterative function takes too long

@@ -5,7 +5,7 @@
 ## below a certain limit is to check all potential sublists, starting with the biggest ones and then shrinking
 ## until we find a smaller one. 
 
-## our parameters: f_filter_signal(df, min_length = 20000, max_var = 1000)
+## our parameters: f_filter_signal(df, min_length = 200s*sampling rate, max_var = 1000)
 
 
 f_filter_signal <- function(df, min_length, max_var){
@@ -16,7 +16,7 @@ f_filter_signal <- function(df, min_length, max_var){
       
       df_temp <- df[start:(start + length - 1),]
 
-      var <- df_temp$roll_mean %>% var(na.rm = TRUE)
+      var <- df_temp$THC %>% var(na.rm = TRUE)
       
       if (var <= max_var) {
         
@@ -36,7 +36,9 @@ f_filter_signal <- function(df, min_length, max_var){
 
 
 doParallel::registerDoParallel()
-f1 <- f_filter_signal(df_thc1, min_length = 20000, max_var = 1000)
+sampling_rate <- 1/.02
+
+f1 <- f_filter_signal(df_raw_data %>% dplyr::filter(number == 1), min_length = 200*sampling_rate, max_var = 1000)
 
 
 

### 2021-07-07 Scripts/raw_data_processing.R
add o2 sat signal processing, add subject id to raw data to combine with clinical data

@@ -15,7 +15,7 @@
   
   # name with Status
   l_names <- 1:length(my_files) %>% 
-    purrr::map(~substr(my_files[.x], start = 22, stop = 23)) 
+    purrr::map(~substr(my_files[.x], start = 16, stop = 25)) 
   
   names(l_raw_data) <- l_names
 
@@ -23,14 +23,13 @@
   
 # format data -------------------------------------------------------------
 
-
   library(data.table)
   library(zoo)
   
-  # format to just THC in workable form
+  # format to just THC and o2 sat in workable form
   l_raw_data <- 1:length(l_raw_data) %>% 
     
-    purrr::map(~mutate(l_raw_data[[.x]], Status = l_names[[.x]])) %>% 
+    purrr::map(~mutate(l_raw_data[[.x]], subject_id = l_names[[.x]])) %>% 
     purrr::map(~dplyr::filter(.x, !grepl("Patient|Data|[R|r]aw", V1))) %>% 
     purrr::map(~dplyr::select(.x, c(ncol(.x), 2:4))) %>% 
     purrr::map(~dplyr::rename(.x, "Time" = "V2", "O2_sat" = "V3", "THC" = "V4")) %>% 
@@ -40,11 +39,15 @@
   
   # renumber patients
   l_raw_data <- 1:length(l_raw_data) %>% 
-    purrr::map(~mutate(l_raw_data[[.x]], number = c(1:length(l_raw_data))[.x])) %>% 
-    purrr::map(~dplyr::select(.x, c(number, everything())))
+    purrr::map(~mutate(l_raw_data[[.x]], number = c(1:length(l_raw_data))[.x])) 
   
   # condense list to df
   df_raw_data <- rbindlist(l_raw_data) %>% as_tibble() 
+  
+  # add status
+  df_raw_data <- df_raw_data %>% 
+    mutate(Status = substr(subject_id, start = 7, stop = 8)) %>% 
+    dplyr::select(c(number, subject_id, Status, everything()))
   
   # make sure status is HV, UM, or CM
[…]

### 2021-07-07 Scripts/raw_data_processing.R
remove patients with low hemoglobin oxygen saturation

@@ -1346,8 +1346,10 @@
       theme(legend.position = "none") ## negative values??
     
   
-  # remove patients with negative o2 sat
-    neg_o2_sat <- df_o2sat_mean %>% dplyr::filter(o2sat_mean < 0) %>% .$number
+  # remove patients with negative o2 sat - run DFA, but do not include these average values in the plots
+    df_o2sat_mean %>% ggplot(aes(x = o2sat_mean)) + geom_histogram(bins = 50)
+    
+    neg_o2_sat <- df_o2sat_mean %>% dplyr::filter(o2sat_mean < 20) %>% .$number
     df_o2sat_filt <- df_o2sat_filt %>% dplyr::filter(!(number %in% neg_o2_sat))
     df_o2sat_filt %>% count(number)
 

### 2021-07-07 Scripts/DFA.R
combine dfa results into df

@@ -2,6 +2,15 @@
 
 
 
+
+# libraries ---------------------------------------------------------------
+
+  library(tidyverse)
+  library(purrr)
+  library(data.table)
+
+
+
 # load in processed signal data -------------------------------------------
 
   load("filtered_signals.Rdata")
@@ -10,6 +19,8 @@
 
 # compute cumulative sum --------------------------------------------------
 
+
+## THC
     # create list, where each object is a patient
     l_thc_filt_pass <- df_thc_filt_pass %>% split(f = df_thc_filt_pass$number)
 
@@ -26,7 +37,6 @@
       mutate(cumsum = cumsum(subtract_mean)) %>% 
       ungroup()
     
-
     # plot
     df_thc_cumsum %>% dplyr::filter(number == 1) %>% 
       ggplot(aes(x = Time)) +
@@ -35,6 +45,34 @@
       
       theme_bw() +
       ggtitle("Cumulative sum of Savitzy-Golar smoothed signal")
+    
+    
+## O2 sat
+    # create list, where each object is a patient
+    l_o2sat_filt_pass <- df_o2sat_filt_pass %>% split(f = df_o2sat_filt_pass$number)
+    
+    # take the mean of each filtered signal
+    l_means <- 1:length(l_o2sat_filt_pass) %>% purrr::map(~mean(l_o2sat_filt_pass[[.x]]$sg_filt, na.rm = TRUE))
+    
+    # subtract the mean and take the cumulative sum
+    l_o2sat_cumsum <- 1:length(l_means) %>% purrr::map(~mutate(l_o2sat_filt_pass[[.x]], mean = l_means[[.x]]))
+    df_o2sat_cumsum <- rbindlist(l_o2sat_cumsum) %>% as_tibble()
+    
+    df_o2sat_cumsum <- df_o2sat_cumsum %>% 
+      mutate(subtract_mean = sg_filt - mean) %>% 
+      group_by(number) %>% 
+      mutate(cumsum = cumsum(subtract_mean)) %>% 
+      ungroup()
+    
+    # plot
+    df_o2sat_cumsum %>% dplyr::filter(number == 1) %>% 
+      ggplot(aes(x = Time)) +
+      geom_line(aes(y = cumsum)) +
+      geom_line(aes(y = sg_filt), color = "red") +
+      
+      theme_bw() +
+      ggtitle("Cumulative sum of Savitzy-Golar smoothed signal")
+    
         
 
         
@@ -234,6 +272,8 @@
     
[…]

### 2021-07-08 Scripts/combining_data.R
script to combine dfa results with clinical data

@@ -1,0 +1,93 @@
+
+
+
+
+# libraries ---------------------------------------------------------------
+
+
+  library(tidyverse)
+
+
+
+# load data ---------------------------------------------------------------
+
+
+  df_hbox <- read.csv("Data/98 first visit records with cerebral oxygenation stats_trimmed.csv") %>% as_tibble()
+  load("results.Rdata")
+  
+  
+
+# format data -------------------------------------------------------------
+
+  
+  # get rid of "Other"
+  df_hbox <- df_hbox %>% 
+    dplyr::filter(Status != "Other") %>% 
+    mutate(Status = factor(Status, levels = c("HV", "UM", "CM")))
+  
+  # identify duplicated rows (TM0003, TM2001)
+  # df_hbox <- df_hbox %>% dplyr::filter(duplicated(Subject.ID..NIAID.) == FALSE)
+  
+  # Convert age to numeric
+  df_hbox <- df_hbox %>%
+    mutate(Age = gsub(" Years, ", "_", Age)) %>%
+    mutate(Age = gsub(" Months ", "", Age)) %>%
+    separate(Age, into = c("Years", "Months"), sep = "_") %>%
+    mutate(Years = as.numeric(Years)) %>%
+    mutate(Months = as.numeric(Months)/12) %>%
+    mutate(Age = Years + Months) %>%
+    dplyr::select(-c(Years, Months))
+  
+  # Convert sex to factor
+  df_hbox <- df_hbox %>% mutate(Sex = as.factor(Sex))
+  
+  # Trim for variables of interest (VOI)
+  df_hbox_trim <- df_hbox %>% 
+    dplyr::select(c(Subject.ID.and.Visit, Status, Sex, Age, Glucose, Hematocrit, Lactate, Temperature,
+                    Arginine.umol.L, Haptoglobin..mg.dl., Hemoglobin..uM., 
+                    BP.Diastolic, BP.Systolic, O2.Sat, RR, HR)) %>% 
+    dplyr::rename("subject_id" = "Subject.ID.and.Visit") 
+  
+  
+
+# join clinical data with dfa results -------------------------------------
+
+
+  
+  df_master <- df_hbox_trim %>% 
+    left_join(df_results, by = c("subject_id", "Status")) %>% 
+    dplyr::select(-number)
+
+  
+  
+# filter  ---------------------------------------------------------
+
+  # remove patients with negative o2 sat - run DFA, but do not include these average values in the plots
[…]

### 2021-07-08 Scripts/raw_data_processing.R
remove duplicate files, identify negative o2 sat means

@@ -10,6 +10,9 @@
   # read in
   my_files <- list.files(path = "Data/Raw data/", pattern = "*.txt", full.names = TRUE, recursive = FALSE)
   
+  # remove duplicate patient files (CM 14, 15, 16, 24part2)
+  my_files <- my_files[!grepl("part2.txt|TM0014CM01.txt|TM0015CM01.txt|TM0016CM01.txt", my_files)]
+  
   l_raw_data <- 1:length(my_files) %>% 
     purrr::map(~read.delim2(file = my_files[.x], header = FALSE, sep = "\t") %>% as_tibble()) 
   
@@ -62,6 +65,7 @@
 
   save(df_raw_data, file = "raw_data.Rdata")
 
+  
 
 # explore --------------------------------------------------------------------
 
@@ -106,9 +110,9 @@
 
 # remove signals that aren't at least 200s  
   sampling_rate <- 1/.02
-  df_raw_data %>% group_by(number) %>% count() %>% dplyr::filter(n < 200*sampling_rate) # 24 and 30 not long enough
-  df_raw_data <- df_raw_data %>% filter(!(number %in% c(24, 30)))
-
+  df_raw_data %>% group_by(number) %>% count() %>% dplyr::filter(n < 200*sampling_rate) # 21 and 26 not long enough
+  df_raw_data <- df_raw_data %>% filter(!(number %in% c(21, 26)))
+  
 # plotting functions
   
   # THC
@@ -179,7 +183,7 @@
 # 4
   # THC
   df_raw_data %>% f_plot_thc(4) 
-  df_thc4 <- df_raw_data %>% dplyr::select(-O2_sat) %>% f_segment_signal(4, 0, 2000) # no blips!
+  df_thc4 <- df_raw_data %>% dplyr::select(-O2_sat) %>% f_segment_signal(4, 0, 2000)
   df_thc4$THC %>% var(na.rm = TRUE)
   
   # O2 sat
@@ -191,7 +195,7 @@
 # 5
   # THC
   df_raw_data %>% f_plot_thc(5) # ***same as 4???
[…]

### 2021-07-08 figures.Rmd
adjusting figures based on new data format, still waiting for final DFA results to create logistic regression model

@@
 load("glm_model.Rdata")
+load("results.Rdata")
 
@@
 # read in data
-df_hbox <- read.csv("Data/98 first visit records with cerebral oxygenation stats_trimmed.csv") %>% as_tibble()
 
-# get rid of "Other"
-df_hbox <- df_hbox %>% 
-  filter(Status != "Other") %>% 
+  df_hbox <- read.csv("Data/master_datatable.csv") %>% 
+    as_tibble() %>% 
+    dplyr::select(-1) %>% 
     mutate(Status = factor(Status, levels = c("HV", "UM", "CM")))
 
-# identify duplicated rows (TM0003, TM2001)
-df_hbox <- df_hbox %>% filter(duplicated(Subject.ID..NIAID.) == FALSE)
+# pivot long
 
-# Convert age to numeric
-df_hbox <- df_hbox %>% 
-  mutate(Age = gsub(" Years, ", "_", Age)) %>% 
-  mutate(Age = gsub(" Months ", "", Age)) %>% 
-  separate(Age, into = c("Years", "Months"), sep = "_") %>% 
-  mutate(Years = as.numeric(Years)) %>% 
-  mutate(Months = as.numeric(Months)/12) %>% 
-  mutate(Age = Years + Months) %>% 
-  dplyr::select(-c(Years, Months))
-
-# Convert sex to factor
-df_hbox <- df_hbox %>% mutate(Sex = as.factor(Sex))
-
-# Trim for variables of interest (VOI)
-df_hbox_trim <- df_hbox %>% 
-  dplyr::select(c(Subject.ID..NIAID., Status, Glucose, Hematocrit, Lactate, Temperature,
-                  Arginine.umol.L, Haptoglobin..mg.dl., Hemoglobin..uM., 
-                  BP.Diastolic, BP.Systolic, O2.Sat, RR, HR,
-
-                  cerebral.hbdeox, cerebral.hbdeox.alpha, cerebral.hbox, cerebral.hbox.alpha,
-                  cerebral.sat, cerebral.sat.alpha, cerebral.thc, cerebral.thc.alpha)) %>%
-
-  dplyr::rename("subject_id" = "Subject.ID..NIAID.", 
-                "Arginine (umol/L)" = "Arginine.umol.L", "Haptoglobin (mg/dL)" = "Haptoglobin..mg.dl.", 
-                "Hemoglobin (uM)" = "Hemoglobin..uM.", "Lactate (mmol/L)" = "Lactate", "Hematocrit (%)" = "Hematocrit",
-                "Systolic BP" = "BP.Systolic", "Diastolic BP" = "BP.Diastolic", "O2 sat" = "O2.Sat", 
-                "Respiration rate" = "RR", "Heart rate" = "HR",
-
-                "[Hb]" = "cerebral.hbdeox", "[Hb]α" = "cerebral.hbdeox.alpha", 
-                "[HbO2]" = "cerebral.hbox", "[HbO2]α" = "cerebral.hbox.alpha",  
-                "HbO2%" = "cerebral.sat", "HbO2% α" = "cerebral.sat.alpha", 
-                "[HbT]" = "cerebral.thc", "[HbT]α" = "cerebral.thc.alpha")
-
-# Remove patients labeled as "UM" that instead meet the criteria for severe malaria
-
-  # remove UM patient with lactate > 5 
-  df_hbox_trim <- df_hbox_trim %>% dplyr::filter(!(Status == "UM" & `Lactate (mmol/L)` > 5))
-
-  # All UM patients have hematocrit > 15%
-  # df_hbox_trim %>% dplyr::filter(Status == "UM"  & `Hematocrit (%)` <= 15)
-
-# impute missing data with median by group
-f_impute <- function(x, na.rm = TRUE) (replace(x, is.na(x), median(x, na.rm = na.rm)))
-
-df_hbox_impute <- df_hbox_trim %>% 
-  group_by(Status) %>% 
-  mutate_at(3:ncol(.), f_impute) %>% 
-  ungroup()
-
-# pivot long
-df_hbox_long <- df_hbox_trim %>% pivot_longer(cols = 3:ncol(.), names_to = "variable", values_to = "value")
+  df_hbox_long <- df_hbox %>% pivot_longer(cols = 4:ncol(.), names_to = "variable", values_to = "value")
 
@@
 # number of subjects 
-df_status <- df_hbox_impute %>% 
+df_status <- df_hbox %>% 
   count(Status) %>% 
@@
   count(Sex) %>% 
-  filter(Sex != "") %>% 
+  dplyr::filter(Sex != "") %>% 
   pivot_wider(names_from = "Sex", values_from = "n") %>% 
@@
 
-  dplyr::filter(Subject.ID..NIAID. != "TM1018") %>%
+  # dplyr::filter(Subject.ID..NIAID. != "TM1018") %>%
 
@@
 
-df_hbox_voi <- df_hbox_impute %>% 
-  dplyr::select(c(Status, `[HbO2]`, `[HbO2]α`, `[Hb]`, `[Hb]α`, `HbO2%`, `HbO2% α`, `[HbT]`, `[HbT]α`)) %>% 
-  dplyr::filter(`[Hb]α` < 3)
+# extract variables of interest
+voi <- df_results %>% colnames() %>% .[4:9]
 
-df_hbox_voi_long <- df_hbox_voi %>% 
-  pivot_longer(2:9, names_to = "variable", values_to = "value") %>%
-  mutate(variable = factor(variable, levels = c("[HbO2]", "[HbO2]α",   
-                                                 "[Hb]", "[Hb]α", 
-                                                "HbO2%", "HbO2% α", 
-                                                "[HbT]", "[HbT]α")))
+# select from the long df
+df_hbox_voi_long <- df_hbox_long %>% dplyr::filter(variable %in% voi)
 
+# create color vector
 my_col <- c("CM" = "#00BFC4", "HV" = "#7CAE00", "UM" = "#F8766D")
 
-df_hbox_voi_long %>% dplyr::filter(variable %in% c("[HbO2]", "[Hb]", "HbO2%", "[HbT]")) %>%
+# plot
+df_hbox_voi_long %>% dplyr::filter(!(grepl("alpha", variable))) %>%
 
@@
   stat_summary(fun = "median", geom = "crossbar", aes(color = Status), size = 0.2, width = 0.5) +
-  facet_wrap(~variable, scales = "free_y", nrow = 2, ncol = 2) +
+  facet_wrap(~variable, scales = "free_y") +
 
@@
 
-# [HbT]
-kruskal.test(`[HbT]` ~ Status, data = df_hbox_voi)
-dunnTest(`[HbT]` ~ Status, data = df_hbox_voi, method = "bh")
+# hemoglobin concentration
+kruskal.test(avg_Hb_conc ~ Status, data = df_hbox)
+dunnTest(avg_Hb_conc ~ Status, data = df_hbox, method = "bh")
 
-# HbO2%
-kruskal.test(`HbO2%` ~ Status, data = df_hbox_voi)
+# hemoglobin oxygen saturation
+kruskal.test(avg_Hb_o2sat ~ Status, data = df_hbox)
 
@@
 
-**Figure 1.** Shown are the variables measured by Oxiplex, plotted by patient status [HbT] levels were different between all three groups (Kruskal-Wallis adjusted p-value = 3.25x10^-6^; Dunn Test post-hoc adjusted p-values: CM vs HV = 1.66x10^-6^, CM vs UM = .032, HV vs UM = .033).
-Levels of cerebral O~2~ saturation did not differ among groups (Kruskal-Wallis adjusted p-value = .07).
+**Figure 1.** Shown are the variables measured by Oxiplex, plotted by patient status.
+Cerebral tissue hemoglobin concentration levels were different between cerebral malaria patients and the other two groups, but did not differ between uncomplicated malaria patients and healthy volunteers (Kruskal-Wallis adjusted p-value = 5.00x10^-7^; Dunn Test post-hoc adjusted p-values: CM vs HV = 3.22x10^-6^, CM vs UM = 2.76x10^-4^, HV vs UM = 0.87).
+Levels of cerebral hemoglobin O~2~ saturation did not differ among groups (Kruskal-Wallis adjusted p-value = 0.28).
 
@@
 
-df_hbox_voi <- df_hbox_impute %>% 
-  dplyr::select(c(Status, `[HbO2]`, `[HbO2]α`, `[Hb]`, `[Hb]α`, `HbO2%`, `HbO2% α`, `[HbT]`, `[HbT]α`)) %>% 
-  dplyr::filter(`[Hb]α` < 3)
-
-df_hbox_voi_long <- df_hbox_voi %>% 
-  pivot_longer(2:9, names_to = "variable", values_to = "value") %>%
-  mutate(variable = factor(variable, levels = c("[HbO2]", "[HbO2]α",   
-                                                 "[Hb]", "[Hb]α", 
-                                                "HbO2%", "HbO2% α", 
-                                                "[HbT]", "[HbT]α")))
-
-my_col <- c("CM" = "#00BFC4", "HV" = "#7CAE00", "UM" = "#F8766D")
-
-df_hbox_voi_long %>% dplyr::filter(variable %in% c("[HbO2]α", "[Hb]α", "HbO2% α", "[HbT]α")) %>%
+df_hbox_voi_long %>% dplyr::filter(grepl("second_alpha", variable)) %>%
 
@@
 
-  geom_hline(yintercept = 0.4, alpha = 0.4, lty = 2) +
-  geom_hline(yintercept = 0.6, alpha = 0.4, lty = 2) +
+  geom_hline(yintercept = 0.45, alpha = 0.4, lty = 2) +
+  geom_hline(yintercept = 0.55, alpha = 0.4, lty = 2) +
   geom_hline(yintercept = 1.0, alpha = 0.4, lty = 2) +
@@
 
-  facet_wrap(~variable, scales = "free_y", nrow = 2, ncol = 2) +
+  facet_wrap(~variable, scales = "free_y") +
 
@@
 
-# [HbT]α
-kruskal.test(`[HbT]α` ~ Status, data = df_hbox_voi)
-dunnTest(`[HbT]α` ~ Status, data = df_hbox_voi, method = "bh")
+# hemoglobin concentration alpha
+kruskal.test(Hb_conc_second_alpha ~ Status, data = df_hbox)
+dunnTest(Hb_conc_second_alpha ~ Status, data = df_hbox, method = "bh")
+
+# hemoglobin oxygen saturation alpha
+kruskal.test(Hb_o2sat_second_alpha ~ Status, data = df_hbox)
+dunnTest(Hb_o2sat_second_alpha ~ Status, data = df_hbox, method = "bh")
 
@@
 
-**Figure 2** Shown are the alpha values associated with variables measured by the Oxiplex, plotted by patient status. Note: One outlier was removed from Deoxy Hb α (3.52).
+**Figure 2.** Shown are the alpha values associated with the hemoglobin concentration and oxygen saturation signals measured by Oxiplex.
+Differences among cerebral tissue hemoglobin concentration alpha values were significant by Kruskal-Wallis (adjusted p-value = 1.57x10^-4^), and differences between all groups were significant by post-hoc Dunn Test (adjusted p-values: CM vs HV = 0.03, CM vs UM = 1.65x10^-4^, HV vs UM = 0.04).
+Cerebral hemoglobin O~2~ saturation alpha values differed between cerebral malaria patients and the other two patient groups (Kruskal-Wallis adjusted p-value = 2.71x10^-5^; Dunn Test post-hoc adjusted p-values: CM vs HV = 5.18x10^-4^, CM vs UM = 4.44x10^-4^, HV vs UM = 0.23).
 
@@
 ```{r fig3a}
-# include variables of interest ([HbT], [HbT] alpha, and hematocrit), remove healthy volunteers from analysis
-df_hbox_log_voi <- df_hbox_impute %>% 
-  dplyr::select(c(2,4:5,19:22)) %>% 
-  dplyr::filter(Status != "HV")
+
+# include variables of interest, remove healthy volunteers from analysis
+
+df_hbox_voi <- df_hbox %>% dplyr::select(c("subject_id", "Status", "Hematocrit", "Lactate", "avg_Hb_conc", "Hb_conc_second_alpha", "avg_Hb_o2sat", "Hb_o2sat_second_alpha"))
+
+# impute missing data with median by group
+  f_impute <- function(x, na.rm = TRUE) (replace(x, is.na(x), median(x, na.rm = na.rm)))
+
+  df_hbox_impute <- df_hbox_voi %>% 
+    group_by(Status) %>% 
+    mutate_at(3:ncol(.), f_impute) %>% 
+    ungroup()
+
+# remove healthy volunteers for logistic regression analysis
+df_hbox_log_reg <- df_hbox_impute %>% dplyr::filter(Status != "HV")
 
 # convert tibble to df
-df_hbox_voi_df <- df_hbox_log_voi %>% 
-  dplyr::select(-Status) %>% 
+df_hbox_voi_df <- df_hbox_log_reg %>% 
+  dplyr::select(-c(subject_id, Status)) %>% 
   as.data.frame()
-rownames(df_hbox_voi_df) <- df_hbox_impute %>% dplyr::filter(Status != "HV") %>% .$subject_id
+
+# add rownames
+rownames(df_hbox_voi_df) <- df_hbox_log_reg %>% .$subject_id
 
@@
 # define groups
-groups <- as.factor(df_hbox_log_voi$Status)
+groups <- as.factor(df_hbox_log_reg$Status)
 
@@
 
-### Figure 4: Relationships among [HbT], [HbT] alpha, and hematocrit
+### Figure 4: Relationships among hemoglobin concentration, hemoglobin concentration alpha, and hematocrit
 ```{r fig4}
@@
   df %>%
-    filter(Status == status) %>% 
-    dplyr::select(-Status) %>% 
+    dplyr::filter(Status == status) %>% 
+    dplyr::select(-c(subject_id, Status)) %>% 
     cor(method = "spearman")
@@
 
-# create df of potential regressors
-df_hbox_reg <- df_hbox_impute %>% 
-  dplyr::select(c(Status, `Hematocrit (%)`, `Lactate (mmol/L)`, `[HbO2]`, `[HbO2]α`, `[Hb]`, `[Hb]α`, `HbO2%`, `HbO2% α`, `[HbT]`, `[HbT]α`)) %>% 
-  dplyr::filter(`[Hb]α` < 3)
+# create run correlations of potential regressors
+m_cor_hv <- df_hbox_impute %>% f_calc_cor(status = "HV")
 
-m_cor_hv <- df_hbox_reg %>% f_calc_cor(status = "HV")
+m_cor_um <- df_hbox_impute %>% f_calc_cor(status = "UM")
 
-m_cor_um <- df_hbox_reg %>% f_calc_cor(status = "UM")
-
-m_cor_cm <- df_hbox_reg %>% f_calc_cor(status = "CM")
+m_cor_cm <- df_hbox_impute %>% f_calc_cor(status = "CM")
 
@@
 
-    mutate_if(is.factor, factor, levels = c("Lactate (mmol/L)", "Hematocrit (%)", "[HbO2]", "[Hb]", "HbO2%", "[HbT]", "[HbO2]α", "[Hb]α", "HbO2% α", "[HbT]α")) %>%
+    mutate_if(is.factor, factor, levels = c("Lactate", "Hematocrit", "avg_Hb_o2sat", "avg_Hb_conc", "Hb_o2sat_second_alpha", "Hb_conc_second_alpha")) %>%
 
@@
 
-**Figure S1.** Shown are the spearman correlations among variables measured by Oxiplex and their associated alpha values. Red indicates a positive correlation; blue indicates a negative correlation.
+**Figure S1.** Shown are the spearman correlations among hematocrit, lactate, hemoglobin concentration and its alpha, and hemoglobin oxygen saturation and its alpha.
+Red indicates a positive correlation; blue indicates a negative correlation.
 
@@
 # pivot longer and change factor levels
-df_hbox_reg_long <- df_hbox_reg %>% 
-  pivot_longer(2:11, names_to = "variable", values_to = "value") %>% 
-  mutate(variable = factor(variable, levels = c("Lactate (mmol/L)", "Hematocrit (%)", "[HbO2]", "[HbO2]α", "[Hb]", "[Hb]α", "HbO2%", "HbO2% α" ,"[HbT]", "[HbT]α")))
+df_hbox_reg_long <- df_hbox_impute %>% 
+  mutate(Status = as.numeric(ifelse(Status == "CM", 1, 0))) %>% 
+  pivot_longer(3:ncol(.), names_to = "variable", values_to = "value") %>% 
+  mutate(variable = factor(variable, levels = c("Lactate", "Hematocrit", "avg_Hb_o2sat", "avg_Hb_conc", "Hb_o2sat_second_alpha", "Hb_conc_second_alpha")))
 
 # plot as individual regression curves
-df_hbox_reg_long %>% mutate(Status = as.numeric(ifelse(Status == "CM", 1, 0))) %>%
+df_hbox_reg_long %>%
 
@@
 On the y-axis, 1 represents CM while 0 represents UM.
-A steeper curve indicates higher discriminatory power
+A steeper curve indicates higher discriminatory power.

### 2021-07-09 figures.Rmd
updated figures with new data

@@
 
-knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, comment = "", fig.height = 8, fig.width = 10)
+knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, comment = "", fig.height = 6, fig.width = 10)
 
 library(tidyverse)
+library(DescTools)
 library(janitor)
@@
 load("results.Rdata")
+load("dfa_example.Rdata")
+load("dfa_results.Rdata")
 
@@
 
-### Figure 1: Visual summary of NIRS variables
+### Figure 1: Cerebral tissue hemoglobin concentration and oxygen saturation
 
@@
 kruskal.test(avg_Hb_conc ~ Status, data = df_hbox)
-dunnTest(avg_Hb_conc ~ Status, data = df_hbox, method = "bh")
+DescTools::DunnTest(avg_Hb_conc ~ Status, data = df_hbox, method = "BH")
 
@@
 **Figure 1.** Shown are the variables measured by Oxiplex, plotted by patient status.
-Cerebral tissue hemoglobin concentration levels were different between cerebral malaria patients and the other two groups, but did not differ between uncomplicated malaria patients and healthy volunteers (Kruskal-Wallis adjusted p-value = 5.00x10^-7^; Dunn Test post-hoc adjusted p-values: CM vs HV = 3.22x10^-6^, CM vs UM = 2.76x10^-4^, HV vs UM = 0.87).
-Levels of cerebral hemoglobin O~2~ saturation did not differ among groups (Kruskal-Wallis adjusted p-value = 0.28).
+Cerebral tissue hemoglobin concentration levels were different between cerebral malaria patients and the other two groups, but did not differ between uncomplicated malaria patients and healthy volunteers (Kruskal-Wallis adjusted p-value = 5.0x10^-7^; Dunn Test post-hoc adjusted p-values: CM vs HV = 3.2x10^-6^, CM vs UM = 2.8x10^-4^, HV vs UM = 0.87).
+Levels of cerebral hemoglobin O~2~ saturation did not differ among groups (Kruskal-Wallis adjusted p-value = 0.47).
 
-### Figure 2: Visual summary of α values associated with NIRS variables
+### Figure 2: Detrended fluctuation analysis results
 
@@
 kruskal.test(Hb_conc_second_alpha ~ Status, data = df_hbox)
-dunnTest(Hb_conc_second_alpha ~ Status, data = df_hbox, method = "bh")
+DescTools::DunnTest(Hb_conc_second_alpha ~ Status, data = df_hbox, method = "BH")
 
@@
 kruskal.test(Hb_o2sat_second_alpha ~ Status, data = df_hbox)
-dunnTest(Hb_o2sat_second_alpha ~ Status, data = df_hbox, method = "bh")
+DescTools::DunnTest(Hb_o2sat_second_alpha ~ Status, data = df_hbox, method = "BH")
 
@@
 **Figure 2.** Shown are the alpha values associated with the hemoglobin concentration and oxygen saturation signals measured by Oxiplex.
-Differences among cerebral tissue hemoglobin concentration alpha values were significant by Kruskal-Wallis (adjusted p-value = 1.57x10^-4^), and differences between all groups were significant by post-hoc Dunn Test (adjusted p-values: CM vs HV = 0.03, CM vs UM = 1.65x10^-4^, HV vs UM = 0.04).
-Cerebral hemoglobin O~2~ saturation alpha values differed between cerebral malaria patients and the other two patient groups (Kruskal-Wallis adjusted p-value = 2.71x10^-5^; Dunn Test post-hoc adjusted p-values: CM vs HV = 5.18x10^-4^, CM vs UM = 4.44x10^-4^, HV vs UM = 0.23).
+Differences among cerebral tissue hemoglobin concentration alpha values were significant by Kruskal-Wallis (adjusted p-value = 7.16x10^-4^), and differences between uncomplicated malaria patients and the other two patient groups were significant by post-hoc Dunn Test (adjusted p-values: CM vs HV = 0.10, CM vs UM = 4.7x10^-4^, HV vs UM = 0.03).
+Cerebral hemoglobin O~2~ saturation alpha values differed between cerebral malaria patients and the other two patient groups (Kruskal-Wallis adjusted p-value = 4.2x10^-5^; Dunn Test post-hoc adjusted p-values: CM vs HV = 7.5x10^-4^, CM vs UM = 3.5x10^-4^, HV vs UM = 0.32).
 
@@
 #### A. PCA
-```{r fig3a}
+```{r fig3a, fig.height = 8}
 
@@
 # select [HbT], [HbT] alpha, and hematocrit
-  df_hbox_model <- df_hbox_impute %>% 
-    dplyr::select(c(2,4:5,19:22)) %>% 
-    mutate(`1/10 * [HbT]α` = `[HbT]α`/10, `[HbT] (nM)` = `[HbT]`/1000) %>% 
-    dplyr::select(-c(`[HbT]α`, `[HbT]`)) %>% 
-    mutate(Status = as.factor(ifelse(Status == "CM", 1, 0))) %>%
-
-    dplyr::select(-c(`HbO2%`, `HbO2% α`, `Lactate (mmol/L)`))
+  df_hbox_model <- df_hbox_log_reg %>% 
+    dplyr::select(c(Status, Hematocrit, avg_Hb_conc, Hb_conc_second_alpha)) %>% 
+    mutate(`1/10_Hb_conc_alpha` = Hb_conc_second_alpha*0.1, avg_Hb_conc_nm = avg_Hb_conc/1000) %>% 
+    dplyr::select(-c(Hb_conc_second_alpha, avg_Hb_conc)) %>% 
+    mutate(Status = as.factor(ifelse(Status == "CM", 1, 0)))
 
@@
 set.seed(234)
-hbox_boot <- bootstraps(df_hbox_model, times = 1000)
+hbox_boot <- bootstraps(df_hbox_model, strata = Status, times = 1000)
 
@@
 
-```{r fig3b_plot}
+```{r fig3b_plot, fig.height = 8}
 
@@
   tidy(exponentiate = TRUE) %>% 
-  mutate(p.value = as.character(p.value)) %>% 
-  mutate_if(is.numeric, round, 2) %>% 
-  mutate(p.value = as.numeric(p.value)) %>% 
-  mutate(p.value = scientific(p.value, digits = 3)) %>% 
+  mutate(estimate = ifelse(estimate < 10^-2, scientific(estimate, digits = 3), round(estimate, 2))) %>% 
+  mutate(p.value = scientific(p.value, digits = 3),
+         std.error = round(std.error, 2),
+         statistic = round(statistic, 2)) %>% 
   dplyr::rename("Term" = "term", "Odds ratio" = "estimate", "Standard error" = "std.error", 
@@
 
-### Figure 4: Relationships among hemoglobin concentration, hemoglobin concentration alpha, and hematocrit
-```{r fig4}
+### Figure 4: Relationships among hemoglobin concentration and hematocrit
+```{r fig4, fig.height = 8}
 
-df_hbox_impute %>% filter(Status != "HV") %>%
+df_hbox %>% filter(Status != "HV") %>%
 
-  ggplot(aes(x = `[HbT]`, y = `Hematocrit (%)`)) +
-  geom_point(aes(color = Status, size = `[HbT]α`), alpha = 0.7) +
+  ggplot(aes(x = avg_Hb_conc, y = Hematocrit)) +
+  geom_point(aes(color = Status), alpha = 0.7, size = 3) +
 
-  xlab("[HbT] (μM)") +
+  xlab("Hemoglobin concentration (μM)") +
+  ylab("Hematocrit (%)") +
 
@@
 A steeper curve indicates higher discriminatory power.
+
+### Supplementary figure 3: Detrended fluctuation analysis
+
+#### A. Pre-processed signal
+```{r fig_s3a}
+
+df_thc_cumsum1 %>%
+
+  ggplot(aes(x = Time, y = sg_filt)) +
+  geom_line() +
+
+  labs(y = "Hemoglobin concentration (μM)", x = "Time (s)") +
+  ggtitle("Pre-processed signal") +
+  theme_bw() +
+  theme(text = element_text(family = "Arial"),
+        axis.text = element_text(size = 20),
+        axis.title = element_text(size = 25),
+        legend.text = element_text(size = 15),
+        legend.title = element_text(size = 20))
+
+```
+
+#### B. Cumulative sum
+```{r fig_s3b}
+
+df_thc_cumsum1 %>%
+
+  ggplot(aes(x = Time, y = cumsum)) +
+  geom_line() +
+  geom_vline(xintercept = 1201, lty = 2, color = "black") +
+  geom_vline(xintercept = 1210, lty = 2, color = "black") +
+
+  labs(y = "Cumulative sum", x = "Time (s)") +
+  ggtitle("Cumulative signal") +
+  theme_bw() +
+  theme(text = element_text(family = "Arial"),
+        axis.text = element_text(size = 20),
+        axis.title = element_text(size = 25),
+        legend.text = element_text(size = 15),
+        legend.title = element_text(size = 20))
+
+```
+
+#### C. Window detrend
+```{r fig_s3c, fig.show = "hold", out.width = "50%"}
+
+fit <- lm(cumsum ~ Time, data = df_thc_cumsum1 %>% dplyr::filter(Time > 1201 & Time < 1210))
+
+# Window 
+df_thc_cumsum1 %>% dplyr::filter(Time > 1201 & Time < 1210) %>%
+
+  ggplot(aes(x = Time, y = cumsum)) +
+  geom_line(size = 2) +
+  geom_abline(intercept = coef(fit)[1], slope = coef(fit)[2], color = "black", lty = 2, size = 2) +
+
+  labs(y = "Cumulative sum", x = "Time (s)") +
+  ggtitle("Cumulative signal") +
+  theme_bw() +
+  theme(text = element_text(family = "Arial"),
+        axis.text = element_text(size = 20),
+        axis.title = element_text(size = 25),
+        legend.text = element_text(size = 15),
+        legend.title = element_text(size = 20),
+        title = element_text(size = 25))
+
+# Detrended
+df_thc_cumsum1 %>% dplyr::filter(Time > 1201 & Time < 1210) %>%
+
+  ggplot(aes(x = Time, y = fit$residuals)) +
+  geom_line(size = 2) +
+  geom_hline(yintercept = 0, color = "black", lty = 2, size = 2) +
+
+  labs(y = "", x = "Time (s)") +
+  ggtitle("Detrended cumulative signal") +
+  theme_bw() +
+  theme(text = element_text(family = "Arial"),
+        axis.text = element_text(size = 20),
+        axis.title = element_text(size = 25),
+        legend.text = element_text(size = 15),
+        legend.title = element_text(size = 20),
+        title = element_text(size = 25))
+
+```
+
+#### D. Results plot
+```{r fig_s3d}
+
+# find slope of segmented line
+library(segmented)
+df1 <- l_dfa_thc[[1]][[1]] %>% mutate(log10_avg_fluct = log10(avg_fluctuation), log10_window_size = log10(window_size))
+fit1 <- lm(log10_avg_fluct ~ log10_window_size, data = df1) 
+segfit1 <- segmented(fit1)
+slopes1 <- coef(segfit1) # the coefficients are the differences in slope in comparison to the previous slope
+
+# first line: 
+#y = b0 + b1*x; y = intercept1 + slope1 * x
+
+# second line:
+# y = c0 + c1*x; y = intercept2 + slope2 * x
+
+# At the breakpoint (break1), the segments b and c intersect: b0 + b1*x = c0 + c1*x
+b0 <- slopes1[[1]]
+b1 <- slopes1[[2]]
+
+c1 <- slopes1[[2]] + slopes1[[3]]
+break1 <- segfit1$psi[[2]]
+
+# Solve for c0 (intercept of second segment):
+c0 <- b0 + b1 * break1 - c1 * break1
+
+df1 %>%
+
+  ggplot(aes(x = log10_window_size, y = log10_avg_fluct)) +
+  geom_point(size = 3, shape = 1) +
+  geom_abline(intercept = c0, slope = c1, color = "black", lty = 2) +
+  annotate("text", x = -0.5, y = 2, label = paste0("alpha = ", round(c1, 2)), size = 8) +
+
+  labs(x = "log10(window length)", y = "log10(average fluctuation)") +
+  ggtitle("DFA results") +
+
+  theme_bw() +
+  theme(text = element_text(family = "Arial"),
+        axis.text = element_text(size = 20),
+        axis.title = element_text(size = 25),
+        legend.text = element_text(size = 15),
+        legend.title = element_text(size = 20))
+
+```

### 2021-07-09 Scripts/DFA.R
export results

@@ -170,7 +170,7 @@
     f_plot_window(l_splits[[11]], 1)
 
     # plot detrended window
-    p_plot_resids <- function(l_resids, window_num) {
+    f_plot_resids <- function(l_resids, window_num) {
       
       l_resids[[window_num]] %>% 
         ggplot(aes(x = Time, y = residuals)) +
@@ -180,7 +180,7 @@
       
     }
     
-    p_plot_resids(l_resids[[11]], 1)
+    f_plot_resids(l_resids[[11]], 1)
     
 
     
@@ -281,8 +281,8 @@
     registerDoParallel()
     
     
-    l_dfa_thc <- 1:length(l_thc_cumsum) %>% purrr::map(~f_dfa(l_thc_cumsum[[.x]], num_of_windows = 10))
-    l_dfa_o2sat <- 1:length(l_o2sat_cumsum) %>% purrr::map(~f_dfa(l_o2sat_cumsum[[.x]], num_of_windows = 10))
+    l_dfa_thc <- 1:length(l_thc_cumsum) %>% purrr::map(~f_dfa(l_thc_cumsum[[.x]], num_of_windows = 20))
+    l_dfa_o2sat <- 1:length(l_o2sat_cumsum) %>% purrr::map(~f_dfa(l_o2sat_cumsum[[.x]], num_of_windows = 20))
     
 
     # save DFA analysis results
@@ -346,6 +346,9 @@
         mutate(alpha = o2sat_alphas, 
                second_alpha = o2sat_second_alpha,
                Status = as.factor(Status))
+      
+      # no break point
+      df_o2sat_dfa[df_o2sat_dfa$second_alpha > 2,]$second_alpha <- df_o2sat_dfa[df_o2sat_dfa$second_alpha > 2,]$alpha
       
 
 
@@ -457,6 +460,7 @@ library(DescTools)
       
     # plot alpha by group
       
+
       # THC
       df_thc_dfa %>% 
         ggplot(aes(x = Status, y = second_alpha)) +
@@ -471,6 +475,7 @@ library(DescTools)
         theme(legend.position = "none")
       
       # test for differences
+      df_thc_dfa %>% group_by(Status) %>% summarise(median = median(second_alpha))
       kruskal.test(second_alpha ~ Status, data = df_thc_dfa)
       DunnTest(second_alpha ~ Status, data = df_thc_dfa)
       
@@ -482,24 +487,29 @@ library(DescTools)
[…]

### 2021-07-09 Scripts/additional_models.R
penalized logistic regression

@@ -370,3 +370,168 @@
   glm(Status ~ THC + THCα, data = df_hbox_voi_01, family = "binomial") %>% summary()
   glm(Status ~ THCα, data = df_hbox_voi_01, family = "binomial") %>% summary()
   
+
+# penalized logistic regression -------------------------------------------
+
+  install.packages("logistf")
+  logistf::logistf(Status ~ ., data = df_model, maxit = , maxstep = )
+
+  library(tidymodels)
+  library(vip)  
+  library(glmnet)
+
+  ### janitor::clean_names()
+  
+  # select variables
+  df_model <- df_hbox_log_reg %>% 
+    dplyr::select(-1) %>% 
+    mutate(Status = as.factor(ifelse(Status == "CM", 1, 0)))
+  
+  ## Train a model
+  
+  # create recipe
+  hbox_rec <- recipe(Status ~ ., data = df_model) %>% 
+    step_BoxCox(all_predictors()) %>% # or step_normalize()
+    step_nzv(all_predictors()) # step_rm() to get rid of what you don't want in model
+  
+  # prep
+  hbox_prep <- hbox_rec %>% prep()
+  
+  # model specification
+  lasso_spec <- logistic_reg(penalty = 0.1, mixture = 1) %>% 
+    set_engine("glmnet")
+  
+  # create workflow
+  lasso_wf <- workflow() %>% 
+    add_recipe(hbox_rec) # workflows know about training data so you don't have to juice
+    
+  lasso_fit <- lasso_wf %>% 
+    add_model(lasso_spec) %>% 
+    fit(data = df_model) # still a workflow
+  
+  lasso_fit %>% 
+    pull_workflow_fit() %>% 
+    tidy()
+    
+  
+  # How do we figure out how to pick penalty?? --> tune lasso parameters
+  
+  # resample
+  set.seed(234)
+  hbox_boot <- bootstraps(df_model, strata = Status)
+  
+  # new specification
+  tune_spec <- logistic_reg(penalty = tune(), mixture = 1) %>% 
+    set_engine("glmnet") # use resampling to find right value for this data set
+  
+  # set grid
+  lambda_grid <- grid_regular(penalty(),
+               levels = 50)
+  
+  # run across resamples
+  doParallel::registerDoParallel()
+
+  set.seed(2021)  
[…]

### 2021-07-09 Scripts/combining_data.R
combine and filter DFA results and clinical data

@@ -58,13 +58,10 @@
     left_join(df_results, by = c("subject_id", "Status")) %>% 
     dplyr::select(-number)
 
-  
-  
+
 # filter  ---------------------------------------------------------
 
-  # remove patients with negative o2 sat - run DFA, but do not include these average values in the plots
-  df_master$avg_Hb_o2sat[df_master$avg_Hb_o2sat < 0] <- NA
-  
+
 
   # Remove patients labeled as "UM" that instead meet the WHO criteria for severe malaria
   
@@ -74,10 +71,22 @@
     # All UM patients have hematocrit > 15%
     # df_master %>% dplyr::filter(Status == "UM"  & Hematocrit <= 15)
   
+  # remove duplicate patient ID
+  df_master <- df_master %>% unique()
+  
 
 
+# double check ------------------------------------------------------------
+
+
+  # any duplicated patients?
+  df_master %>% count(subject_id) %>% count(n)
+  
+  # any negative Hb o2 sat?
+  df_master %>% dplyr::filter(avg_Hb_o2sat < 20)
+  
+  
 # export as csv -----------------------------------------------------------
-
 
   write.csv(df_master, file = "Data/master_datatable.csv")
     

### 2021-07-19 figures.Rmd
adjust font sizes

@@
   # plot variables
-  geom_point(data = df_vars, aes(x = PC1*5, y = PC2*5)) +
-  geom_text_repel(data = df_vars, aes(x = PC1*5, y = PC2*5, label = variable, size = 10), show.legend = FALSE) +
+  # geom_point(data = df_vars, aes(x = PC1*5, y = PC2*5)) +
+  # geom_text_repel(data = df_vars, aes(x = PC1*5, y = PC2*5, label = variable, size = 10), show.legend = FALSE) +
   geom_segment(data = df_vars, aes(x = 0, y = 0, xend = PC1*5, yend = PC2*5), arrow = arrow()) +
@@
     theme(text = element_text(family = "Arial"),
-          axis.text.x = element_text(angle = 45, hjust = 0.9),
+          axis.text.x = element_text(angle = 45, hjust = 1),
           axis.text = element_text(size = 35),
@@
   theme(text = element_text(family = "Arial"),
-        axis.text = element_text(size = 20),
-        axis.title = element_text(size = 25),
+        axis.text = element_text(size = 25),
+        axis.title = element_text(size = 28),
         legend.text = element_text(size = 15),

### 2021-07-19 Scripts/additional_models.R
look at penalized logistic regression models

@@ -371,160 +371,6 @@
   glm(Status ~ THCα, data = df_hbox_voi_01, family = "binomial") %>% summary()
   
 
-# penalized logistic regression -------------------------------------------
-
-  install.packages("logistf")
-  logistf::logistf(Status ~ ., data = df_model, maxit = , maxstep = )
-
-  library(tidymodels)
-  library(vip)  
-  library(glmnet)
-
-  ### janitor::clean_names()
-  
-  # select variables
-  df_model <- df_hbox_log_reg %>% 
-    dplyr::select(-1) %>% 
-    mutate(Status = as.factor(ifelse(Status == "CM", 1, 0)))
-  
-  ## Train a model
-  
-  # create recipe
-  hbox_rec <- recipe(Status ~ ., data = df_model) %>% 
-    step_BoxCox(all_predictors()) %>% # or step_normalize()
-    step_nzv(all_predictors()) # step_rm() to get rid of what you don't want in model
-  
-  # prep
-  hbox_prep <- hbox_rec %>% prep()
-  
-  # model specification
-  lasso_spec <- logistic_reg(penalty = 0.1, mixture = 1) %>% 
-    set_engine("glmnet")
-  
-  # create workflow
-  lasso_wf <- workflow() %>% 
-    add_recipe(hbox_rec) # workflows know about training data so you don't have to juice
-    
-  lasso_fit <- lasso_wf %>% 
-    add_model(lasso_spec) %>% 
-    fit(data = df_model) # still a workflow
-  
-  lasso_fit %>% 
-    pull_workflow_fit() %>% 
-    tidy()
-    
-  
-  # How do we figure out how to pick penalty?? --> tune lasso parameters
-  
-  # resample
-  set.seed(234)
-  hbox_boot <- bootstraps(df_model, strata = Status)
-  
-  # new specification
-  tune_spec <- logistic_reg(penalty = tune(), mixture = 1) %>% 
-    set_engine("glmnet") # use resampling to find right value for this data set
-  
-  # set grid
-  lambda_grid <- grid_regular(penalty(),
-               levels = 50)
-  
-  # run across resamples
-  doParallel::registerDoParallel()
-
-  set.seed(2021)  
-  lasso_grid <- tune_grid(
-    lasso_wf %>% add_model(tune_spec),
-    resamples = hbox_boot,
[…]

### 2021-07-19 Scripts/raw_data_processing.R
finalized raw data processing (read in --> filter)

@@ -107,7 +107,7 @@
 
 # remove blips (manual) ------------------------------------------------------------
 
-
+my_files[26]
 # remove signals that aren't at least 200s  
   sampling_rate <- 1/.02
   df_raw_data %>% group_by(number) %>% count() %>% dplyr::filter(n < 200*sampling_rate) # 21 and 26 not long enough
@@ -651,7 +651,7 @@
   df_thc46$THC %>% var(na.rm = TRUE)
   
   # O2 sat
-  df_raw_data %>% f_plot_o2sat(46) ### signal too noisy
+  df_raw_data %>% f_plot_o2sat(46) ### signal very noisy
   df_o2sat46 <- df_raw_data %>% dplyr::select(-THC) %>% f_segment_signal(46, 500, 2000)
   df_o2sat46$O2_sat %>% var(na.rm = TRUE)
 

### 2021-07-19 Scripts/penalized_logistic_regression.R
explore penalized logistic regression to work with complete separation of variables

@@ -1,0 +1,149 @@
+
+
+
+# libraries ---------------------------------------------------------------
+
+
+    library(tidyverse)
+    library(tidymodels)
+    library(vip)  
+    library(glmnet)
+
+
+
+# Train model -------------------------------------------------------------
+
+
+    # create recipe
+    my_rec <- recipe(Status ~ ., data = df_model) %>% 
+      step_BoxCox(all_predictors()) %>%
+      step_nzv(all_predictors())
+    
+    # model specification
+    lasso_spec <- logistic_reg(penalty = 0.1, mixture = 1) %>% # picking specific penalty (tune in next section)
+      set_engine("glmnet")
+    
+    # create workflow
+    lasso_wf <- workflow() %>% 
+      add_recipe(my_rec)
+    
+    # fit the workflow to model + data
+    lasso_fit <- lasso_wf %>% 
+      add_model(lasso_spec) %>% 
+      fit(data = df_model)
+    
+    # extract results
+    lasso_fit %>% 
+      pull_workflow_fit() %>% 
+      tidy()
+
+
+
+    
+    
+# Tune LASSO parameters ---------------------------------------------------
+
+
+    
+    # resample
+    set.seed(234)
+    my_boot <- bootstraps(df_model, strata = group, times = 1000)
+    
+    # new specification
+    tune_spec <- logistic_reg(penalty = tune(), mixture = 1) %>% # tune model to determine optimal penalty value
+      set_engine("glmnet")
+    
+    # set grid
+    lambda_grid <- grid_regular(penalty(),
+                                levels = 50)
+    
+    # run across resamples
+    doParallel::registerDoParallel()
+    
+    set.seed(2021)  
+    lasso_grid <- tune_grid(
+      lasso_wf %>% add_model(tune_spec),
+      resamples = my_boot,
+      metrics = metric_set(roc_auc, accuracy, sensitivity, specificity),
+      control = control_grid(save_pred = TRUE),
+      grid = lambda_grid
+    )  
+    
+    # look at metrics
+    lasso_grid %>% 
+      collect_metrics() %>% 
+      
+      ggplot(aes(penalty, mean, color = .metric)) +
+      geom_errorbar(aes(ymin = mean - std_err,
+                        ymax = mean + std_err)) +
+      geom_line(size = 1.5, show.legend = FALSE) +
+      facet_wrap(~.metric, scales = "free", nrow = 2) +
[…]

### 2021-07-19 Scripts/DFA.R
adjust slightly to format for follow-up visits analyses

@@ -293,7 +293,7 @@
 
 # second order detrending --------------------------------------------------
 
-## separately compute scaling exponents for short-length vs long-length windws, for short-range and long-range correlations of the time series (Seleznov et al 2019)
+## separately compute scaling exponents for short-length vs long-length windows, for short-range and long-range correlations of the time series (Seleznov et al 2019)
     
     library(segmented)
     

### 2021-07-19 Scripts/raw_data_processing.R

@@ -7,12 +7,13 @@
   library(purrr)
   
 
-  # read in
+  # get names
   my_files <- list.files(path = "Data/Raw data/", pattern = "*.txt", full.names = TRUE, recursive = FALSE)
   
   # remove duplicate patient files (CM 14, 15, 16, 24part2)
   my_files <- my_files[!grepl("part2.txt|TM0014CM01.txt|TM0015CM01.txt|TM0016CM01.txt", my_files)]
   
+  # read in
   l_raw_data <- 1:length(my_files) %>% 
     purrr::map(~read.delim2(file = my_files[.x], header = FALSE, sep = "\t") %>% as_tibble()) 
   
@@ -107,7 +108,7 @@
 
 # remove blips (manual) ------------------------------------------------------------
 
-my_files[26]
+
 # remove signals that aren't at least 200s  
   sampling_rate <- 1/.02
   df_raw_data %>% group_by(number) %>% count() %>% dplyr::filter(n < 200*sampling_rate) # 21 and 26 not long enough
@@ -136,7 +137,6 @@ my_files[26]
   }
   
 # segment signal function
-  
   f_segment_signal <- function(df, num, start, end) {
     
     df %>% 

### 2021-07-19 Scripts/CM_follow_up_analysis.R
raw data processing and detrended fluctuation analysis for the cerebral malaria patient follow-up visits

@@ -1,0 +1,877 @@
+
+
+#### original code for this data processing can be found in raw_data_processing.R ####
+
+
+# read in data -------------------------------------------------------------------------
+
+    library(tidyverse)
+    library(purrr)
+    
+    
+    # get names
+    my_files <- list.files(path = "Data/CM_follow_up/", pattern = "*.txt", full.names = TRUE, recursive = FALSE)
+    
+    # # remove duplicate patient files (TM0001)
+    # my_files <- my_files[!grepl("part2.txt|TM0014CM01.txt|TM0015CM01.txt|TM0016CM01.txt", my_files)]
+    
+    # read in
+    l_follow_up <- 1:length(my_files) %>% 
+      purrr::map(~read.delim2(file = my_files[.x], header = FALSE, sep = "\t") %>% as_tibble()) 
+    
+    # name with Status
+    l_names <- 1:length(my_files) %>% 
+      purrr::map(~substr(my_files[.x], start = 20, stop = 25)) 
+    
+    names(l_follow_up) <- l_names
+
+
+    
+    
+    
+
+# format raw data ---------------------------------------------------------
+
+    
+    
+    library(data.table)
+    library(zoo)
+    
+    # format to just THC and o2 sat in workable form
+    l_follow_up <- 1:length(l_follow_up) %>% 
+      
+      purrr::map(~mutate(l_follow_up[[.x]], subject_id = l_names[[.x]])) %>% 
+      purrr::map(~dplyr::filter(.x, !grepl("Patient|Data|[R|r]aw", V1))) %>% 
+      purrr::map(~dplyr::select(.x, c(ncol(.x), 2:4))) %>% 
+      purrr::map(~dplyr::rename(.x, "Time" = "V2", "O2_sat" = "V3", "THC" = "V4")) %>% 
+      purrr::map(~mutate(.x, Time = as.numeric(Time))) %>% 
+      purrr::map(~mutate(.x, THC = as.numeric(THC))) %>% 
+      purrr::map(~mutate(.x, O2_sat = as.numeric(O2_sat)))
+    
+    # number patients
+    l_follow_up <- 1:length(l_follow_up) %>% 
[…]

### 2021-07-19 cm_follow_up_figures.Rmd
manuscript figure 5: CM follow-up data

---
title: "Cerebral malaria follow-up visit data"
author: "Rachel Smith"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include = FALSE}

knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, comment = "", fig.height = 6, fig.width = 10)

library(tidyverse)
library(extrafont)

load("results.Rdata")
load("follow_up_results.Rdata")

font_import()

```

```{r data}

# combine visit 1 with follow-ups
df_results_all <- df_results %>% 
  mutate(subject_id = substr(subject_id, start = 1, stop = 6)) %>% 
  left_join(df_fu_results %>% select(-c(number, Status)), by = "subject_id")

# reorganize for plotting
df_results_all_plot <- df_results_all %>% 
  dplyr::filter(subject_id %in% df_fu_results$subject_id) %>% 
  pivot_longer(cols = 4:ncol(.), names_to = "variable", values_to = "value") %>% 
  mutate(follow_up = ifelse(grepl("fu", variable), "follow-up", "initial visit")) %>% 
  mutate(follow_up = factor(follow_up, levels = c("initial visit", "follow-up")))

# descriptive statistics

# THC
df_thc_compare <- df_results_all_plot %>% 
  dplyr::filter(variable %in% c("Hb_conc_overall_alpha", "Hb_conc_fu_alpha"))

thc_fu_pval <- wilcox.test(value ~ follow_up, data = df_thc_compare) %>% .$p.value

# O2 sat
df_o2sat_compare <- df_results_all_plot %>% 
  dplyr::filter(variable %in% c("Hb_o2sat_overall_alpha", "Hb_o2sat_fu_alpha"))

o2sat_fu_pval <- wilcox.test(value ~ follow_up, data = df_o2sat_compare) %>% .$p.value

```

### Figure 5: Cerebral malaria patient follow-up visits

```{r fig5}

# plot O2 sat
df_results_all_plot %>% 
  dplyr::filter(!grepl("avg|second", variable)) %>% 
  mutate(measurement = ifelse(grepl("conc", variable), "THC", "O2_sat")) %>% 
  mutate(measurement = factor(measurement, levels = c("THC", "O2_sat"))) %>% 
  mutate(pvals = ifelse(measurement == "THC", round(thc_fu_pval, 3), round(o2sat_fu_pval, 3))) %>%

  ggplot(aes(x = follow_up, y = value)) +
  geom_violin(aes(color = follow_up)) +
  geom_point(shape = 1) +
  stat_summary(fun = "median", geom = "crossbar", aes(x = follow_up, color = follow_up), size = 0.2, width = 0.5) +
  geom_path(aes(group = subject_id), lty = 2, alpha = 0.7) +

  geom_segment(aes(x = 1, y = 1.4, xend = 2, yend = 1.4), size = 0.3) +
  geom_segment(aes(x = 1, y = 1.4, xend = 1, yend = 1.35), size = 0.3) +
  geom_segment(aes(x = 2, y = 1.4, xend = 2, yend = 1.35), size = 0.3) +
  geom_text(x = 1.5, y = 1.43, aes(label = paste0("p = ", pvals))) +

  facet_wrap(~measurement) +

  ylim(0.5, 1.5) +
  labs(y = "alpha", x = "") +

  theme_bw() +
  theme(legend.position = "none") +
  theme(text = element_text(family = "Arial"),
        strip.text = element_text(size = 15),
        axis.title = element_text(size = 15),
        axis.text = element_text(size = 15),
        legend.position = "none")

```

### 2021-07-21 figures.Rmd
updated model with up-sample

@@
 
+library(themis)
+
 # select [HbT], [HbT] alpha, and hematocrit
@@
   step_BoxCox(all_predictors()) %>% 
-  step_nzv(all_predictors())
+  step_nzv(all_predictors()) %>% 
+  step_smote(Status) # upsample to deal with unbalanced sample sizes
 
@@
 
+total_true_UM <- final_res_conf_mat %>% dplyr::filter(Truth == "UM") %>% .$Freq %>% sum()
+total_true_CM <- final_res_conf_mat %>% dplyr::filter(Truth == "CM") %>% .$Freq %>% sum()
+
 final_res_conf_mat %>% 
-  group_by(Truth) %>% 
-  summarise(total_true = sum(Freq)) %>% 
-  right_join(final_res_conf_mat) %>% 
-  mutate(`Percent occurence` = round(Freq/total_true*100, 2)) %>%
+  mutate(total_true = ifelse(Truth == "CM", Freq/total_true_CM, Freq/total_true_UM)) %>% 
+  mutate(`Percent occurrence` = round(total_true*100, 2)) %>%
 
   # plot
-  ggplot(aes(x = Truth, y = Prediction, fill = `Percent occurence`, label = `Percent occurence`)) +
+  ggplot(aes(x = Truth, y = Prediction, fill = `Percent occurrence`, label = `Percent occurrence`)) +
   geom_tile() +

### 2021-07-25 cm_follow_up_figures.Rmd
change order of variables in figures; export table to csv

@@
   mutate(measurement = ifelse(grepl("conc", variable), "THC", "O2_sat")) %>% 
-  mutate(measurement = factor(measurement, levels = c("THC", "O2_sat"))) %>% 
+  mutate(measurement = factor(measurement, levels = c("O2_sat", "THC"))) %>% 
   mutate(pvals = ifelse(measurement == "THC", round(thc_fu_pval, 3), round(o2sat_fu_pval, 3))) %>%
@@
   geom_violin(aes(color = follow_up)) +
-  geom_point(shape = 1) +
+  geom_point(shape = 1, size = 2) +
   stat_summary(fun = "median", geom = "crossbar", aes(x = follow_up, color = follow_up), size = 0.2, width = 0.5) +

### 2021-07-25 figures.Rmd
change order of variables in figures; export table to csv

@@
     dplyr::select(-1) %>% 
-    mutate(Status = factor(Status, levels = c("HV", "UM", "CM")))
+    mutate(Status = ifelse(Status == "HV", "HC", Status)) %>% 
+    mutate(Status = factor(Status, levels = c("HC", "UM", "CM")))
 
@@
   dplyr::rename(" " = "variable",
-                "Healthy volunteers" = "HV",
-                "Uncomplicated malaria patients" = "UM",
-                "Cerebral malaria patients" = "CM")
+                "Healthy children" = "HC",
+                "Uncomplicated malaria" = "UM",
+                "Cerebral malaria" = "CM")
 
-  # remove lactate data for HV
-  df_clinical_res[11,]$`Healthy volunteers` <- "--"
+  # remove lactate data for HC
+  df_clinical_res[11,]$`Healthy children` <- "--"
+
+  # export to CSV
+  # write.csv(df_clinical_res, file = "Outputs/table_1.csv")
 
@@
 # select from the long df
-df_hbox_voi_long <- df_hbox_long %>% dplyr::filter(variable %in% voi)
+df_hbox_voi_long <- df_hbox_long %>% 
+  dplyr::filter(variable %in% voi) %>% 
+  mutate(variable = factor(variable, levels = c("avg_Hb_o2sat", "avg_Hb_conc",
+                                                "Hb_o2sat_overall_alpha", "Hb_conc_overall_alpha",
+                                                "Hb_o2sat_second_alpha", "Hb_conc_second_alpha")))
 
 # create color vector
-my_col <- c("CM" = "#00BFC4", "HV" = "#7CAE00", "UM" = "#F8766D")
+my_col <- c("CM" = "#00BFC4", "HC" = "#7CAE00", "UM" = "#F8766D")
 
@@
   geom_violin(aes(color = Status)) +
-  geom_jitter(position = position_jitter(0.2), shape = 1) +
+  geom_jitter(position = position_jitter(0.2), shape = 1, size = 2) +
   stat_summary(fun = "median", geom = "crossbar", aes(color = Status), size = 0.2, width = 0.5) +
@@
 **Figure 1.** Shown are the variables measured by Oxiplex, plotted by patient status.
-Cerebral tissue hemoglobin concentration levels were different between cerebral malaria patients and the other two groups, but did not differ between uncomplicated malaria patients and healthy volunteers (Kruskal-Wallis adjusted p-value = 5.0x10^-7^; Dunn Test post-hoc adjusted p-values: CM vs HV = 3.2x10^-6^, CM vs UM = 2.8x10^-4^, HV vs UM = 0.87).
+Cerebral tissue hemoglobin concentration levels were different between cerebral malaria patients and the other two groups, but did not differ between uncomplicated malaria patients and healthy volunteers (Kruskal-Wallis adjusted p-value = 5.0x10^-7^; Dunn Test post-hoc adjusted p-values: CM vs HC = 3.2x10^-6^, CM vs UM = 2.8x10^-4^, HC vs UM = 0.87).
 Levels of cerebral hemoglobin O~2~ saturation did not differ among groups (Kruskal-Wallis adjusted p-value = 0.47).
@@
 
-# include variables of interest, remove healthy volunteers from analysis
+# include variables of interest, remove healthy children from analysis
 
-df_hbox_voi <- df_hbox %>% dplyr::select(c("subject_id", "Status", "Hematocrit", "Lactate", "avg_Hb_conc", "Hb_conc_second_alpha", "avg_Hb_o2sat", "Hb_o2sat_second_alpha"))
+df_hbox_voi <- df_hbox %>% 
+  dplyr::select(c("subject_id", "Status", "Hematocrit", "Lactate", "avg_Hb_conc", "Hb_conc_second_alpha", "avg_Hb_o2sat", "Hb_o2sat_second_alpha"))
 
@@
 # remove healthy volunteers for logistic regression analysis
-df_hbox_log_reg <- df_hbox_impute %>% dplyr::filter(Status != "HV")
+df_hbox_log_reg <- df_hbox_impute %>% dplyr::filter(Status != "HC")
 
@@
 
-collect_metrics(glm_res) %>% 
+table2a <- collect_metrics(glm_res) %>% 
   dplyr::select(.metric, mean, std_err) %>% 
@@
   mutate_if(is.numeric, round, 2) %>% 
-  as.data.frame() %>%
+  as.data.frame()
 
-  knitr::kable(align = rep('c', 4))
+table2a %>% knitr::kable(align = rep('c', 4))
 
@@
 
-hbox_final %>% 
+table2b <- hbox_final %>% 
   tidy(exponentiate = TRUE) %>% 
@@
                 "Statistic" = "statistic", "p-value" = "p.value") %>% 
+  as.data.frame()
 
-  as.data.frame() %>%
-  knitr::kable(align = rep('c', 5))
+table2b %>% knitr::kable(align = rep('c', 5))
+
+```
+
+```{r export_table2, eval = FALSE}
+
+list(table2a, table2b) %>% write.csv("Outputs/table_2.csv")
 
@@
 
-df_hbox %>% filter(Status != "HV") %>%
+df_hbox %>% filter(Status != "HC") %>%
 
@@
 # create run correlations of potential regressors
-m_cor_hv <- df_hbox_impute %>% f_calc_cor(status = "HV")
+m_cor_hv <- df_hbox_impute %>% f_calc_cor(status = "HC")
 
@@
   # create plots
-  m_cor_hv %>% f_plot_cor() + ggtitle("Healthy volunteers") + theme(legend.position = "none")
+  m_cor_hv %>% f_plot_cor() + ggtitle("Healthy chilren") + theme(legend.position = "none")
 
-  m_cor_um %>% f_plot_cor() + ggtitle("Uncomplicated malaria patients") + theme(legend.position = "none")
+  m_cor_um %>% f_plot_cor() + ggtitle("Uncomplicated malaria") + theme(legend.position = "none")
 
-  m_cor_cm %>% f_plot_cor() + ggtitle("Cerebral malaria patients") + theme(legend.position = "none")
+  m_cor_cm %>% f_plot_cor() + ggtitle("Cerebral malaria") + theme(legend.position = "none")
 

### 2021-08-19 cm_follow_up_figures.Rmd
updated follow-up analysis

@@
 df_results_all_plot <- df_results_all %>% 
+  dplyr::select(-visit) %>% 
   dplyr::filter(subject_id %in% df_fu_results$subject_id) %>% 
@@
 
+  # averages
+
     # THC
     df_thc_compare <- df_results_all_plot %>% 
-  dplyr::filter(variable %in% c("Hb_conc_overall_alpha", "Hb_conc_fu_alpha"))
+      dplyr::filter(variable %in% c("avg_Hb_conc", "avg_Hb_conc_fu"))
 
@@
     df_o2sat_compare <- df_results_all_plot %>% 
+      dplyr::filter(variable %in% c("avg_Hb_o2sat", "avg_Hb_o2sat_fu"))
+
+    o2sat_fu_pval <- wilcox.test(value ~ follow_up, data = df_o2sat_compare) %>% .$p.value
+
+  # alphas
+
+    # THC
+    df_thc_alpha_compare <- df_results_all_plot %>% 
+      dplyr::filter(variable %in% c("Hb_conc_overall_alpha", "Hb_conc_fu_alpha"))
+
+    thc_alpha_fu_pval <- wilcox.test(value ~ follow_up, data = df_thc_alpha_compare) %>% .$p.value
+
+    # O2 sat
+    df_o2sat_alpha_compare <- df_results_all_plot %>% 
       dplyr::filter(variable %in% c("Hb_o2sat_overall_alpha", "Hb_o2sat_fu_alpha"))
 
-o2sat_fu_pval <- wilcox.test(value ~ follow_up, data = df_o2sat_compare) %>% .$p.value
+    o2sat_alpha_fu_pval <- wilcox.test(value ~ follow_up, data = df_o2sat_alpha_compare) %>% .$p.value
 
@@
 
-```{r fig5}
+#### A: Cerebral hemoglobin oxygen saturation & hemolgobin concentration
+```{r fig5a}
 
-# plot O2 sat
 df_results_all_plot %>%
+
+  dplyr::filter(grepl("avg_Hb", variable)) %>% 
+  mutate(measurement = ifelse(grepl("conc", variable), "THC", "O2_sat")) %>% 
+  mutate(measurement = factor(measurement, levels = c("O2_sat", "THC"))) %>% 
+  mutate(pvals = ifelse(measurement == "THC", round(thc_fu_pval, 3), round(o2sat_fu_pval, 3))) %>%
+
+  ggplot(aes(x = follow_up, y = value)) +
+  geom_violin(aes(color = follow_up)) +
+  geom_point(shape = 1, size = 2) +
+  stat_summary(fun = "median", geom = "crossbar", aes(x = follow_up, color = follow_up), size = 0.2, width = 0.5) +
+  geom_path(aes(group = subject_id), lty = 2, alpha = 0.7) +
+
+  geom_segment(aes(x = 1, y = 1.4, xend = 2, yend = 1.4), size = 0.3) +
+  geom_segment(aes(x = 1, y = 1.4, xend = 1, yend = 1.35), size = 0.3) +
+  geom_segment(aes(x = 2, y = 1.4, xend = 2, yend = 1.35), size = 0.3) +
+  geom_text(x = 1.5, y = 1.43, aes(label = paste0("p = ", pvals))) +
+
+  facet_wrap(~measurement, scales = "free_y") +
+  labs(y = "", x = "") +
+
+  theme_bw() +
+  theme(legend.position = "none") +
+  theme(text = element_text(family = "Arial"),
+        strip.text = element_text(size = 15),
+        axis.title = element_text(size = 15),
+        axis.text = element_text(size = 15),
+        legend.position = "none")
+
+```
+
+#### B: Cerebral hemoglobin oxygen saturation & hemolgobin concentration alphas
+```{r fig5b}
+
+df_results_all_plot %>%
+
   dplyr::filter(!grepl("avg|second", variable)) %>% 
@@
   mutate(measurement = factor(measurement, levels = c("O2_sat", "THC"))) %>% 
-  mutate(pvals = ifelse(measurement == "THC", round(thc_fu_pval, 3), round(o2sat_fu_pval, 3))) %>%
+  mutate(pvals = ifelse(measurement == "THC", round(thc_alpha_fu_pval, 3), round(o2sat_alpha_fu_pval, 3))) %>%
 

### 2021-08-19 Scripts/CM_follow_up_analysis.R

@@ -27,8 +27,6 @@
 
 
     
-    
-    
 
 # format raw data ---------------------------------------------------------
 
@@ -38,15 +36,27 @@
     library(zoo)
     
     # format to just THC and o2 sat in workable form
-    l_follow_up <- 1:length(l_follow_up) %>% 
+    l_follow_up_formatted <- 3:length(l_follow_up) %>% 
       
       purrr::map(~mutate(l_follow_up[[.x]], subject_id = l_names[[.x]])) %>% 
       purrr::map(~dplyr::filter(.x, !grepl("Patient|Data|[R|r]aw", V1))) %>% 
-      purrr::map(~dplyr::select(.x, c(ncol(.x), 2:4))) %>% 
-      purrr::map(~dplyr::rename(.x, "Time" = "V2", "O2_sat" = "V3", "THC" = "V4")) %>% 
-      purrr::map(~mutate(.x, Time = as.numeric(Time))) %>% 
-      purrr::map(~mutate(.x, THC = as.numeric(THC))) %>% 
-      purrr::map(~mutate(.x, O2_sat = as.numeric(O2_sat)))
+      purrr::map(~dplyr::select(.x, c(ncol(.x), 2, 7:8))) %>% 
+      purrr::map(~dplyr::rename(.x, "Time" = "V2", "O2_sat" = "V7", "THC" = "V8")) %>% 
+      purrr::map(~mutate(.x, Time = as.numeric(as.character(Time)))) %>% 
+      purrr::map(~mutate(.x, THC = as.numeric(as.character(THC)))) %>% 
+      purrr::map(~mutate(.x, O2_sat = as.numeric(as.character(O2_sat))))
+    
+    # append data from TM0001 (data in different columns)
+    df1 <- l_follow_up[[1]] %>% 
+      mutate(subject_id = l_names[[1]]) %>% 
+      dplyr::select(c(ncol(.), 2:4)) %>% 
+      dplyr::rename("Time" = "V2", "O2_sat" = "V3", "THC" = "V4") %>% 
+      dplyr::filter(!grepl("Time", Time)) %>% 
+      mutate(Time = as.numeric(as.character(Time)), 
+             THC = as.numeric(as.character(THC)),
+             O2_sat = as.numeric(as.character(O2_sat)))
+    
[…]

### 2021-08-19 Scripts/combining_data.R

@@ -43,19 +43,20 @@
   
   # Trim for variables of interest (VOI)
   df_hbox_trim <- df_hbox %>% 
-    dplyr::select(c(Subject.ID.and.Visit, Status, Sex, Age, Glucose, Hematocrit, Lactate, Temperature,
+    dplyr::select(c(Subject.ID..NIAID., Status, Subject.ID.and.Visit, Admission.Date, DOB, Sex, Age, Glucose, Hematocrit, Lactate, Temperature,
                     Arginine.umol.L, Haptoglobin..mg.dl., Hemoglobin..uM., 
                     BP.Diastolic, BP.Systolic, O2.Sat, RR, HR)) %>% 
-    dplyr::rename("subject_id" = "Subject.ID.and.Visit") 
-  
+    dplyr::filter(!grepl("blood", Subject.ID.and.Visit)) %>% 
+    dplyr::rename("subject_id" = "Subject.ID..NIAID.", "visit" = "Subject.ID.and.Visit") %>% 
+    dplyr::mutate(visit = substr(visit, start = 9, stop = 10))
   
 
+  
 # join clinical data with dfa results -------------------------------------
 
 
-  
   df_master <- df_hbox_trim %>% 
-    left_join(df_results, by = c("subject_id", "Status")) %>% 
+    left_join(df_results %>% mutate(subject_id = substr(subject_id, start = 1, stop = 6)), by = c("subject_id", "Status")) %>% 
     dplyr::select(-number)
 
 
@@ -81,16 +82,17 @@
 
   # any duplicated patients?
   df_master %>% count(subject_id) %>% count(n)
-  
+
   # any negative Hb o2 sat?
   df_master %>% dplyr::filter(avg_Hb_o2sat < 20)
+  
+  
   
   
 # export as csv -----------------------------------------------------------
 
   write.csv(df_master, file = "Data/master_datatable.csv")
     
-  
 
   
   

### 2021-08-20 cm_follow_up_figures.Rmd
finalized figure 5 with updated follow-up analysis

@@
   mutate(pvals = ifelse(measurement == "THC", round(thc_fu_pval, 3), round(o2sat_fu_pval, 3))) %>% 
+  mutate(ycoord = ifelse(measurement == "THC", 400, 100)) %>%
 
@@
 
-  geom_segment(aes(x = 1, y = 1.4, xend = 2, yend = 1.4), size = 0.3) +
-  geom_segment(aes(x = 1, y = 1.4, xend = 1, yend = 1.35), size = 0.3) +
-  geom_segment(aes(x = 2, y = 1.4, xend = 2, yend = 1.35), size = 0.3) +
-  geom_text(x = 1.5, y = 1.43, aes(label = paste0("p = ", pvals))) +
+  geom_segment(aes(x = 1, y = ycoord, xend = 2, yend = ycoord), size = 0.3) +
+  geom_segment(aes(x = 1, y = ycoord, xend = 1, yend = ycoord - .04*ycoord), size = 0.3) +
+  geom_segment(aes(x = 2, y = ycoord, xend = 2, yend = ycoord - .04*ycoord), size = 0.3) +
+  geom_text(x = 1.5, aes(y = ycoord + .025*ycoord, label = paste0("p = ", pvals))) +
 
@@
 
+library(scales)
+
 df_results_all_plot %>%
@@
   mutate(measurement = factor(measurement, levels = c("O2_sat", "THC"))) %>% 
-  mutate(pvals = ifelse(measurement == "THC", round(thc_alpha_fu_pval, 3), round(o2sat_alpha_fu_pval, 3))) %>%
+  mutate(pvals = ifelse(measurement == "THC", scientific(thc_alpha_fu_pval, 3), scientific(o2sat_alpha_fu_pval, 3))) %>%
 
@@
   geom_segment(aes(x = 2, y = 1.4, xend = 2, yend = 1.35), size = 0.3) +
-  geom_text(x = 1.5, y = 1.43, aes(label = paste0("p = ", pvals))) +
+  geom_text(x = 1.5, y = 1.43, aes(label = paste0("p < .0001"))) +
 

### 2021-08-20 figures.Rmd
added figure 6

@@
 
-  df_hbox_long <- df_hbox %>% pivot_longer(cols = 4:ncol(.), names_to = "variable", values_to = "value")
+  df_hbox_long <- df_hbox %>% pivot_longer(cols = 7:ncol(.), names_to = "variable", values_to = "value")
 
@@
 
+### Figure 6: Hemoglobin concentration trends in extravascular edema vs vascular engorgement
+
+```{r fig6, fig.width = 12}
+
+df_fig6 <- tibble(edema = rnorm(n = 90, mean = 20, sd = 20),
+                  normal = rnorm(n = 90, mean = 70, sd = 20),
+                  engorge = rnorm(n = 90, mean = 130, sd = 20)) %>% 
+  pivot_longer(1:3, names_to = "variable", values_to = "value") %>% 
+  mutate(variable = factor(variable, levels = c("edema", "normal", "engorge")))
+
+df_fig6 %>% 
+  ggplot(aes(x = variable, y = value)) +
+  geom_point(shape = 1, size = 3, position = position_jitter(0.1)) +
+  stat_summary(fun = "median", geom = "crossbar", size = 0.2, width = 0.5) +
+  labs(x = "", y = "") +
+  theme_classic() +
+  theme(axis.text = element_blank(),
+        axis.ticks = element_blank())
+
+```
+
 ### Supplementary figure 1: Spearman correlations of potential regressors

### 2021-08-20 figures.Rmd
ROC curves

@@
 load("results.Rdata")
+load("roc_curves.Rdata")
 load("dfa_example.Rdata")
@@
 
+#### C. ROC curve of final model compared to other iterations
+
+```{r fig3c_prep, eval = FALSE}
+
+## write function to create model on given df
+
+f_log_reg_model <- function(df_model) {
+
+  # resample
+  hbox_boot <- bootstraps(df_model, strata = Status, times = 1000)
+
+  # recipe
+  hbox_rec <- recipe(Status ~ ., data = df_model) %>% 
+    step_BoxCox(all_predictors()) %>% 
+    step_nzv(all_predictors()) %>% 
+    step_smote(Status) # upsample to deal with unbalanced sample sizes
+
+  # create workflow
+  hbox_wf <- workflow() %>% add_recipe(hbox_rec)
+
+  # logistic regression model specifications
+  glm_spec <- logistic_reg() %>% set_engine("glm")
+
+  # create model on resamples
+  glm_res <- hbox_wf %>% 
+    add_model(glm_spec) %>%
+    fit_resamples(
+      resamples = hbox_boot,
+      metrics = metric_set(roc_auc, accuracy, sensitivity, specificity),
+      control = tune::control_resamples(save_pred = TRUE, verbose = TRUE)
+    )
+
+  # select best model to be used for evaluation
+  glm_best <- glm_res %>% select_best("roc_auc")
+
+  # finalize model
+  hbox_final <- hbox_wf %>% 
+    add_model(glm_spec) %>% 
+    finalize_workflow(glm_best) %>%
+    fit(df_model) %>%
+    pull_workflow_fit()
+
+  # view coefficients & p-values
+  model_coefs <- hbox_final %>% tidy()
+
+  # return list including model and a table of best model variable statistics
+  list(glm_res, model_coefs)
+
+}
+
+## write function to collect predictions for ROC plot
+
+f_collect_preds <- function(glm_res) {
+
+      glm_res %>% 
+      collect_predictions() %>% 
+      roc_curve(Status, .pred_CM)
+
+}
+
+## Perform functions on all iterations of model
+
+set.seed(123)
+
+    # select starting variables
+    df_hbox_model1 <- df_hbox_log_reg %>% 
+      dplyr::select(c(Status, Hematocrit, Lactate, avg_Hb_o2sat, Hb_o2sat_second_alpha, avg_Hb_conc, Hb_conc_second_alpha)) %>% 
+      mutate(Status = factor(Status, levels = c("CM", "UM")))
+
+    # model 1: Status ~ Hematocrit + Lactate + avg_Hb_o2sat + Hb_o2sat_second_alpha + avg_Hb_conc + Hb_conc_second_alpha
+    model1 <- f_log_reg_model(df_hbox_model1)
+    roc1 <- f_collect_preds(model1[[1]]) %>% mutate(model = "1")
+    model1[[2]] # remove Hb_o2sat_second_alpha (highest p-val)
+
+    # model 2: Status ~ Hematocrit + Lactate + avg_Hb_o2sat + avg_Hb_conc + Hb_conc_second_alpha
+    df_hbox_model2 <-  df_hbox_model1 %>% dplyr::select(-Hb_o2sat_second_alpha)
+    model2 <- f_log_reg_model(df_hbox_model2)
+    roc2 <- f_collect_preds(model2[[1]]) %>% mutate(model = "2")
+    model2[[2]] # remove Lactate
+
+    # model 3: Status ~ Hematocrit + avg_Hb_o2sat + avg_Hb_conc + Hb_conc_second_alpha
+    df_hbox_model3 <-  df_hbox_model2 %>% dplyr::select(-Lactate)
+    model3 <- f_log_reg_model(df_hbox_model3)
+    roc3 <- f_collect_preds(model3[[1]]) %>% mutate(model = "3")
+    model3[[2]] # remove avg_Hb_o2sat
+
+    # model 4: Status ~ Hematocrit + avg_Hb_conc + Hb_conc_second_alpha
+    df_hbox_model4 <-  df_hbox_model3 %>% dplyr::select(-avg_Hb_o2sat)
+    model4 <- f_log_reg_model(df_hbox_model4)
+    roc4 <- f_collect_preds(model4[[1]]) %>% mutate(model = "4")
+    model4[[2]] # none to remove, final model
+
+save(roc1, roc2, roc3, file = "roc_curves.Rdata")
+
+# calculate areas under curves
+
+library(MESS)
+roc1 <- roc1 %>% mutate(false_pos = 1 - specificity)
+auc(roc1$false_pos, roc1$sensitivity)
+
+roc2 <- roc2 %>% mutate(false_pos = 1 - specificity)
+auc(roc2$false_pos, roc2$sensitivity)
+
+roc3 <- roc3 %>% mutate(false_pos = 1 - specificity)
+auc(roc3$false_pos, roc3$sensitivity)
+
+roc_final <- glm_res %>% 
+  collect_predictions() %>% 
+  roc_curve(Status, .pred_0) %>% 
+  mutate(false_pos = 1 - specificity)
+auc(roc_final$false_pos, roc_final$sensitivity)
+
+```
+
+```{r fig3c_plot, fig.height = 8}
+
+df_roc <- glm_res %>% 
+  collect_predictions() %>% 
+  roc_curve(Status, .pred_0) %>% 
+  mutate(model = "final") %>% 
+  bind_rows(roc1, roc2, roc3)
+
+ggplot() +
+  geom_path(data = df_roc %>% filter(model == "final"), 
+            aes(x = 1 - specificity, y = sensitivity, color = model), 
+            size = 2.5) +
+  geom_path(data = df_roc %>% filter(model != "final"), 
+            aes(x = 1 - specificity, y = sensitivity, color = model), 
+            size = 1) +
+  geom_abline(lty = 3) +
+  labs(x = "False positive rate", y = "True positive rate") +
+  theme_bw() +
+
+  theme(text = element_text(family = "Arial"),
+        axis.title = element_text(size = 25),
+        axis.text = element_text(size = 20),
+        legend.title = element_text(size = 20),
+        legend.text = element_text(size = 15),
+        legend.position = "top")
+
+```
+
 **Figure 3.** Logistic regression model discriminating CM vs UM patients using cerebral tissue hemoglobin concentration, cerebral tissue hemoglobin concentration alpha, and hematocrit as the regressors **(A)** PCA of patients with CM and UM as determined by NIRS variables.

### 2021-10-05 cm_follow_up_figures.Rmd
plot all tracings

--- 2021-08-19 cm_follow_up_figures.Rmd
@@
 
-load("results.Rdata")
-load("follow_up_results.Rdata")
+load("all_dfa_results.Rdata")
 
@@
 
-# combine visit 1 with follow-ups
-df_results_all <- df_results %>% 
-  mutate(subject_id = substr(subject_id, start = 1, stop = 6)) %>% 
-  left_join(df_fu_results %>% select(-c(number, Status)), by = "subject_id")
-
 # reorganize for plotting
-df_results_all_plot <- df_results_all %>% 
-  dplyr::select(-visit) %>% 
-  dplyr::filter(subject_id %in% df_fu_results$subject_id) %>% 
-  pivot_longer(cols = 4:ncol(.), names_to = "variable", values_to = "value") %>% 
+df_results_all_plot <- df_brain_dfa_results %>% 
+  dplyr::filter(!is.na(avg_Hb_conc_fu)) %>% 
+  pivot_longer(cols = 3:ncol(.), names_to = "variable", values_to = "value") %>% 
   mutate(follow_up = ifelse(grepl("fu", variable), "follow-up", "initial visit")) %>% 
@@
 
-  geom_segment(aes(x = 1, y = 1.4, xend = 2, yend = 1.4), size = 0.3) +
-  geom_segment(aes(x = 1, y = 1.4, xend = 1, yend = 1.35), size = 0.3) +
-  geom_segment(aes(x = 2, y = 1.4, xend = 2, yend = 1.35), size = 0.3) +
-  geom_text(x = 1.5, y = 1.43, aes(label = paste0("p = ", pvals))) +
+  # geom_segment(aes(x = 1, y = ycoord, xend = 2, yend = ycoord), size = 0.3) +
+  # geom_segment(aes(x = 1, y = ycoord, xend = 1, yend = ycoord - .04*ycoord), size = 0.3) +
+  # geom_segment(aes(x = 2, y = ycoord, xend = 2, yend = ycoord - .04*ycoord), size = 0.3) +
+  # geom_text(x = 1.5, aes(y = ycoord + .025*ycoord, label = paste0("p = ", pvals))) +
 
@@
 ```{r fig5b}
+
+library(scales)
 

### 2021-10-05 Scripts/CM_follow_up_analysis.R

@@ -95,8 +95,48 @@
           dplyr::filter(Time > start & Time < end) # time is in SECONDS
       }
     
+
+# Plot all together
       
-   ## variance needs to be < 10000 ##
+      df_follow_up_long <- df_follow_up %>% 
+        pivot_longer(5:6, values_to = "value", names_to = "variable")
+      
+      f_plot_all <- function(df, num) {
+        
+        subject_id <- df %>% 
+          count(number, subject_id) %>% 
+          dplyr::filter(number == num) %>% 
+          .$subject_id 
+        
+        df %>% 
+          dplyr::filter(number == num) %>% 
+          ggplot(aes(x = Time, y = value)) +
+          geom_line() +
+          facet_wrap(~variable, scales = "free_y") +
+          ggtitle(subject_id) +
+          theme_bw()
+        
+      }
+      
+      f_plot_all(df_follow_up_long, 2)
+      
+      # save as pdf
+      l_plot <- list()
+      for (i in unique(df_follow_up_long$number)) {
+        
+        p <- f_plot_all(df_follow_up_long, i)
+        l_plot[[i]] <- p
+        
+      }
+      
+      pdf("Outputs/follow_up_brain_tracings.pdf")
+      for (i in unique(df_follow_up_long$number)) {
+        print(l_plot[[i]])    
+      }
+      dev.off()
+    
+        
+ ## variance needs to be < 10000 ##
 
   # 1: TM0005CMF
       

### 2021-10-05 Scripts/raw_data_processing.R

@@ -66,8 +66,8 @@
 
   save(df_raw_data, file = "raw_data.Rdata")
 
-  
 
+  
 # explore --------------------------------------------------------------------
 
 
@@ -109,6 +109,9 @@
 # remove blips (manual) ------------------------------------------------------------
 
 
+  load("raw_data.Rdata")
+  
+  
 # remove signals that aren't at least 200s  
   sampling_rate <- 1/.02
   df_raw_data %>% group_by(number) %>% count() %>% dplyr::filter(n < 200*sampling_rate) # 21 and 26 not long enough
@@ -122,7 +125,7 @@
     df %>% 
       dplyr::filter(number == num) %>% 
       ggplot(aes(x = Time, y = THC)) +
-      geom_line()
+      geom_line() 
     
   }
   
@@ -144,9 +147,49 @@
       dplyr::filter(Time > start & Time < end) # time is in SECONDS
   }
 
+  
+# Plot all together
+  
+  df_raw_data_long <- df_raw_data %>% 
+    pivot_longer(5:6, values_to = "value", names_to = "variable")
+  
+  f_plot_all <- function(df, num) {
+    
+    subject_id <- df %>% 
+      count(number, subject_id) %>% 
+      dplyr::filter(number == num) %>% 
+      .$subject_id 
+    
+    df %>% 
+      dplyr::filter(number == num) %>% 
+      ggplot(aes(x = Time, y = value)) +
+      geom_line() +
+      facet_wrap(~variable, scales = "free_y") +
+      ggtitle(subject_id) +
+      theme_bw()
+    
+  }
+  
+  f_plot_all(df_raw_data_long, 2)
+
+  # save as pdf
+  l_plot <- list()
+  for (i in unique(df_raw_data_long$number)) {
+    
+    p <- f_plot_all(df_raw_data_long, i)
+    l_plot[[i]] <- p
+    
+  }
+  
+  pdf("Outputs/brain_tracings.pdf")
+  for (i in unique(df_raw_data_long$number)) {
+    print(l_plot[[i]])    
+  }
+  dev.off()
+  
 # 1
   # THC
-  df_raw_data %>% f_plot_thc(1) +theme_bw()# eyeball where after blips is
+  df_raw_data %>% f_plot_thc(1) + theme_bw()# eyeball where after blips is
[…]

### 2021-10-05 figures.Rmd
plot all brain tracings

@@
 load("dfa_results.Rdata")
+load("all_dfa_results.Rdata")
 
@@
     dplyr::select(-1) %>% 
-    mutate(Status = ifelse(Status == "HV", "HC", Status)) %>% 
+    mutate(Status = replace(Status, Status == "HV", "HC")) %>% 
     mutate(Status = factor(Status, levels = c("HC", "UM", "CM")))
@@
 # extract variables of interest
-voi <- df_results %>% colnames() %>% .[4:9]
+voi <- df_results %>% colnames() %>% .[3:8]
 
@@
                                                 "Hb_o2sat_second_alpha", "Hb_conc_second_alpha")))
-
 # create color vector
@@
 
+  df_hbox_model <- df_hbox_log_reg %>% 
+    dplyr::select(c(Status, Hematocrit, avg_Hb_conc, Hb_conc_second_alpha)) %>% 
+    mutate(`1.10_Hb_conc_alpha` = Hb_conc_second_alpha*0.1) %>% 
+    mutate(`1.10_Hb_conc_alpha_z` = (Hb_conc_second_alpha - mean(Hb_conc_second_alpha)) / sd(Hb_conc_second_alpha), 
+           Hb_conc_z = (avg_Hb_conc - mean(avg_Hb_conc)) / sd(avg_Hb_conc),
+           Hematocrit_z = (Hematocrit - mean(Hematocrit)) / sd(Hematocrit)) %>% 
+    dplyr::select(-c(Hb_conc_second_alpha, `1.10_Hb_conc_alpha`, avg_Hb_conc, Hematocrit)) %>% 
+    mutate(Status = as.factor(ifelse(Status == "CM", 1, 0)))
+
 # resample
 set.seed(234)
-hbox_boot <- bootstraps(df_hbox_model, strata = Status, times = 1000)
+hbox_boot <- bootstraps(df_hbox_model, strata = Status)#, times = 1000)
 
@@
   dplyr::rename("Term" = "term", "Odds ratio" = "estimate", "Standard error" = "std.error", 
-                "Statistic" = "statistic", "p-value" = "p.value") %>% 
+                "t-statistic" = "statistic", "p-value" = "p.value") %>% 
   as.data.frame()
@@
 
-### Figure 6: Hemoglobin concentration trends in extravascular edema vs vascular engorgement
-
-```{r fig6, fig.width = 12}
+### Figure 5: Cerebral malaria patient follow-up visits
+
+```{r fig5_data}
+
+# reorganize for plotting
+df_results_all_plot <- df_brain_dfa_results %>% 
+  dplyr::filter(!is.na(avg_Hb_conc_fu)) %>% 
+  pivot_longer(cols = 3:ncol(.), names_to = "variable", values_to = "value") %>% 
+  mutate(follow_up = ifelse(grepl("fu", variable), "follow-up", "initial visit")) %>% 
+  mutate(follow_up = factor(follow_up, levels = c("initial visit", "follow-up")))
+
+# descriptive statistics
+
+  # averages
+
+    # THC
+    df_thc_compare <- df_results_all_plot %>% 
+      dplyr::filter(variable %in% c("avg_Hb_conc", "avg_Hb_conc_fu"))
+
+    thc_fu_pval <- wilcox.test(value ~ follow_up, data = df_thc_compare) %>% .$p.value
+
+    # O2 sat
+    df_o2sat_compare <- df_results_all_plot %>% 
+      dplyr::filter(variable %in% c("avg_Hb_o2sat", "avg_Hb_o2sat_fu"))
+
+    o2sat_fu_pval <- wilcox.test(value ~ follow_up, data = df_o2sat_compare) %>% .$p.value
+
+  # alphas
+
+    # THC
+    df_thc_alpha_compare <- df_results_all_plot %>% 
+      dplyr::filter(variable %in% c("Hb_conc_overall_alpha", "Hb_conc_fu_alpha"))
+
+    thc_alpha_fu_pval <- wilcox.test(value ~ follow_up, data = df_thc_alpha_compare) %>% .$p.value
+
+    # O2 sat
+    df_o2sat_alpha_compare <- df_results_all_plot %>% 
+      dplyr::filter(variable %in% c("Hb_o2sat_overall_alpha", "Hb_o2sat_fu_alpha"))
+
+    o2sat_alpha_fu_pval <- wilcox.test(value ~ follow_up, data = df_o2sat_alpha_compare) %>% .$p.value
+
+```
+
+#### A: Cerebral hemoglobin oxygen saturation & hemolgobin concentration
+```{r fig5a}
+
+# create color vector
+my_col_fu <- c("initial visit" = "#B80F0A", "follow up" = "black")
+
+# plot
+df_results_all_plot %>%
+
+  dplyr::filter(grepl("avg_Hb", variable)) %>% 
+  mutate(measurement = ifelse(grepl("conc", variable), "THC", "O2_sat")) %>% 
+  mutate(measurement = factor(measurement, levels = c("O2_sat", "THC"))) %>% 
+  mutate(pvals = ifelse(measurement == "THC", round(thc_fu_pval, 3), round(o2sat_fu_pval, 3))) %>% 
+  mutate(ycoord = ifelse(measurement == "THC", 400, 100)) %>%
+
+  ggplot(aes(x = follow_up, y = value)) +
+  geom_violin(aes(color = follow_up)) +
+  geom_point(shape = 1, size = 2) +
+  stat_summary(fun = "median", geom = "crossbar", aes(x = follow_up, color = follow_up), size = 0.2, width = 0.5) +
+  geom_path(aes(group = subject_id), lty = 2, alpha = 0.7) +
+
+  # geom_segment(aes(x = 1, y = ycoord, xend = 2, yend = ycoord), size = 0.3) +
+  # geom_segment(aes(x = 1, y = ycoord, xend = 1, yend = ycoord - .04*ycoord), size = 0.3) +
+  # geom_segment(aes(x = 2, y = ycoord, xend = 2, yend = ycoord - .04*ycoord), size = 0.3) +
+  # geom_text(x = 1.5, aes(y = ycoord + .025*ycoord, label = paste0("p = ", pvals))) +
+
+  facet_wrap(~measurement, scales = "free_y") +
+  labs(y = "", x = "") +
+  scale_color_manual(values = my_col_fu) +
+
+  theme_bw() +
+  theme(legend.position = "none") +
+  theme(text = element_text(family = "Arial"),
+        strip.text = element_text(size = 15),
+        axis.title = element_text(size = 15),
+        axis.text = element_text(size = 15),
+        legend.position = "none")
+
+```
+No difference from initial visit to follow up - Hb_oxy pval: 0.90; Hb_tot pval: 0.32
+
+#### B: Cerebral hemoglobin oxygen saturation & hemolgobin concentration alphas
+```{r fig5b}
+
+library(scales)
+
+df_results_all_plot %>%
+
+  dplyr::filter(!grepl("avg|second", variable)) %>% 
+  mutate(measurement = ifelse(grepl("conc", variable), "THC", "O2_sat")) %>% 
+  mutate(measurement = factor(measurement, levels = c("O2_sat", "THC"))) %>% 
+  mutate(pvals = ifelse(measurement == "THC", round(thc_alpha_fu_pval, 3), round(o2sat_alpha_fu_pval, 3))) %>%
+
+  ggplot(aes(x = follow_up, y = value)) +
+  geom_violin(aes(color = follow_up)) +
+  geom_point(shape = 1, size = 2) +
+  stat_summary(fun = "median", geom = "crossbar", aes(x = follow_up, color = follow_up), size = 0.2, width = 0.5) +
+  geom_path(aes(group = subject_id), lty = 2, alpha = 0.7) +
+
+  geom_hline(yintercept = 0.45, alpha = 0.4, lty = 2) +
+  geom_hline(yintercept = 0.55, alpha = 0.4, lty = 2) +
+  geom_hline(yintercept = 1.0, alpha = 0.4, lty = 2) +
+  geom_hline(yintercept = 1.5, alpha = 0.4, lty = 2) +
+
+  geom_segment(aes(x = 1, y = 0.93, xend = 2, yend = 0.93), size = 0.3) +
+  geom_segment(aes(x = 1, y = 0.93, xend = 1, yend = 0.89), size = 0.3) +
+  geom_segment(aes(x = 2, y = 0.93, xend = 2, yend = 0.89), size = 0.3) +
+  geom_text(x = 1.5, y = 0.96, aes(label = paste0("p = ", pvals))) +
+
+  facet_wrap(~measurement) +
+
+  ylim(0.45, 1.1) +
+  labs(y = "alpha", x = "") +
+  scale_color_manual(values = my_col_fu) +
+
+  theme_bw() +
+  theme(legend.position = "none") +
+  theme(text = element_text(family = "Arial"),
+        strip.text = element_text(size = 15),
+        axis.title = element_text(size = 15),
+        axis.text = element_text(size = 15),
+        legend.position = "none")
+
+```
+
+### Figure 6: Muscle hemoglobin concentration and oxygen saturation data from initial visit and follow-up
+
+#### A. Muscle hemoglobin concentration and oxygen saturation
+
+```{r fig6A}
+
+# select from the long df
+df_muscle_long <- df_muscle_dfa_results %>% 
+  dplyr::select(c(Status, avg_Hb_conc_muscle, avg_Hb_o2sat_muscle)) %>% 
+  pivot_longer(2:3, names_to = "variable") %>% 
+  mutate(Status = factor(Status, levels = c("HC", "UM", "CM")),
+         variable = factor(variable, levels = c("avg_Hb_o2sat_muscle", "avg_Hb_conc_muscle")))
+
+# create color vector
+my_col <- c("CM" = "#00BFC4", "HC" = "#7CAE00", "UM" = "#F8766D")
+
+# plot
+df_muscle_long %>% dplyr::filter(Status != "?") %>%
+
+  ggplot(aes(x = Status, y = value)) +
+  geom_violin(aes(color = Status)) +
+  geom_jitter(position = position_jitter(0.2), shape = 1, size = 2) +
+  stat_summary(fun = "median", geom = "crossbar", aes(color = Status), size = 0.2, width = 0.5) +
+  facet_wrap(~variable, scales = "free_y") +
+
+  scale_color_manual(values = my_col) +
+  theme_bw() +
+  theme(legend.position = "none")  +
+
+  theme(text = element_text(family = "Arial"),
+        strip.text = element_text(size = 15),
+        axis.title = element_text(size = 15),
+        axis.title.y = element_blank(),
+        axis.text = element_text(size = 15))
+
+```
+
+```{r fig6a.pvals, eval = FALSE}
+
+# hemoglobin concentration
+kruskal.test(avg_Hb_conc_muscle ~ Status, data = df_muscle_dfa_results)
+
+# hemoglobin oxygen saturation
+kruskal.test(avg_Hb_o2sat_muscle ~ Status, data = df_muscle_dfa_results)
+
+```
+No significant differences: Hb_oxy pval = 0.11 by KW; Hb_tot pval = 0.051 by KW
+
+#### B. Muscle DFA results
+```{r fig6B}
+
+# select from the long df
+df_muscle_dfa_long <- df_muscle_dfa_results %>% 
+  dplyr::select(c(Status, muscle_Hb_conc_alpha, muscle_Hb_o2sat_alpha)) %>% 
+  pivot_longer(2:3, names_to = "variable") %>% 
+  mutate(Status = factor(Status, levels = c("HC", "UM", "CM")),
+         variable = factor(variable, levels = c("muscle_Hb_o2sat_alpha", "muscle_Hb_conc_alpha")))
+
+# create color vector
+my_col <- c("CM" = "#00BFC4", "HC" = "#7CAE00", "UM" = "#F8766D")
+
+# plot
+df_muscle_dfa_long %>% dplyr::filter(Status != "?") %>%
+
+  ggplot(aes(x = Status, y = value)) +
+  geom_violin(aes(color = Status)) +
+  geom_jitter(position = position_jitter(0.2), shape = 1, size = 2) +
+  stat_summary(fun = "median", geom = "crossbar", aes(color = Status), size = 0.2, width = 0.5) +
+
+  geom_hline(yintercept = 0.45, alpha = 0.4, lty = 2) +
+  geom_hline(yintercept = 0.55, alpha = 0.4, lty = 2) +
+  geom_hline(yintercept = 1.0, alpha = 0.4, lty = 2) +
+  geom_hline(yintercept = 1.5, alpha = 0.4, lty = 2) +
+
+  facet_wrap(~variable) +
+  ylim(0, 1.5) +
+
+  scale_color_manual(values = my_col) +
+  theme_bw() +
+  theme(legend.position = "none")  +
+
+  theme(text = element_text(family = "Arial"),
+        strip.text = element_text(size = 15),
+        axis.title = element_text(size = 15),
+        axis.title.y = element_blank(),
+        axis.text = element_text(size = 15))
+
+```
+
+```{r fig6b.pvals, eval = FALSE}
+
+# hemoglobin concentration alpha
+kruskal.test(muscle_Hb_conc_alpha ~ Status, data = df_muscle_dfa_results)
+
+# hemoglobin oxygen saturation alpha
+kruskal.test(muscle_Hb_o2sat_alpha ~ Status, data = df_muscle_dfa_results)
+
+```
+No significant differences: Hb_oxy pval = 0.08 by KW; Hb_tot pval = 0.15 by KW
+
+#### C. CM follow-up muscle results
+```{r fig6c_data}
+
+# reorganize for plotting
+df_muscle_fu_plot <- df_muscle_dfa_results %>% 
+  dplyr::filter(!is.na(avg_Hb_o2sat_muscle_fu)) %>% 
+  pivot_longer(cols = 3:ncol(.), names_to = "variable", values_to = "value") %>% 
+  mutate(follow_up = ifelse(grepl("fu", variable), "follow-up", "initial visit")) %>% 
+  mutate(follow_up = factor(follow_up, levels = c("initial visit", "follow-up")))
+
+# descriptive statistics
+
+  # averages
+
+    # THC
+    df_thc_muscle_compare <- df_muscle_fu_plot %>% 
+      dplyr::filter(variable %in% c("avg_Hb_conc_muscle", "avg_Hb_conc_muscle_fu"))
+
+    thc_muscle_fu_pval <- wilcox.test(value ~ follow_up, data = df_thc_muscle_compare) %>% .$p.value
+
+    # O2 sat
+    df_o2sat_muscle_compare <- df_muscle_fu_plot %>% 
+      dplyr::filter(variable %in% c("avg_Hb_o2sat_muscle", "avg_Hb_o2sat_muscle_fu"))
+
+    o2sat_muscle_fu_pval <- wilcox.test(value ~ follow_up, data = df_o2sat_muscle_compare) %>% .$p.value
+
+  # alphas
+
+    # THC
+    df_thc_muscle_alpha_compare <- df_muscle_fu_plot %>% 
+      dplyr::filter(variable %in% c("muscle_Hb_conc_alpha", "muscle_Hb_conc_fu_alpha"))
+
+    thc_alpha_muscle_fu_pval <- wilcox.test(value ~ follow_up, data = df_thc_muscle_alpha_compare) %>% .$p.value
+
+    # O2 sat
+    df_o2sat_muscle_alpha_compare <- df_muscle_fu_plot %>% 
+      dplyr::filter(variable %in% c("muscle_Hb_o2sat_alpha", "muscle_Hb_o2sat_fu_alpha"))
+
+    o2sat_alpha_muscle_fu_pval <- wilcox.test(value ~ follow_up, data = df_o2sat_muscle_alpha_compare) %>% .$p.value
+
+```
+
+Muscle hemoglobin concentration and hemoglobin oxygen saturation
+```{r fig6c_avgs}
+
+df_muscle_fu_plot %>%
+
+  dplyr::filter(grepl("avg_Hb", variable)) %>% 
+  mutate(measurement = ifelse(grepl("conc", variable), "THC", "O2_sat")) %>% 
+  mutate(measurement = factor(measurement, levels = c("O2_sat", "THC"))) %>% 
+  mutate(pvals = ifelse(measurement == "THC", round(thc_muscle_fu_pval, 3), round(o2sat_muscle_fu_pval, 3))) %>% 
+  mutate(ycoord = ifelse(measurement == "THC", 400, 100)) %>%
+
+  ggplot(aes(x = follow_up, y = value)) +
+  geom_violin(aes(color = follow_up)) +
+  geom_point(shape = 1, size = 2) +
+  stat_summary(fun = "median", geom = "crossbar", aes(x = follow_up, color = follow_up), size = 0.2, width = 0.5) +
+  geom_path(aes(group = subject_id), lty = 2, alpha = 0.7) +
+
+  # geom_segment(aes(x = 1, y = ycoord, xend = 2, yend = ycoord), size = 0.3) +
+  # geom_segment(aes(x = 1, y = ycoord, xend = 1, yend = ycoord - .04*ycoord), size = 0.3) +
+  # geom_segment(aes(x = 2, y = ycoord, xend = 2, yend = ycoord - .04*ycoord), size = 0.3) +
+  # geom_text(x = 1.5, aes(y = ycoord + .025*ycoord, label = paste0("p = ", pvals))) +
+
+  facet_wrap(~measurement, scales = "free_y") +
+  labs(y = "", x = "") +
+  scale_color_manual(values = my_col_fu) +
+
+  theme_bw() +
+  theme(legend.position = "none") +
+  theme(text = element_text(family = "Arial"),
+        strip.text = element_text(size = 15),
+        axis.title = element_text(size = 15),
+        axis.text = element_text(size = 15),
+        legend.position = "none")
+
+```
+No difference in hemoglobin concentration from initial visit to follow up - Hb_tot pval: 0.72.
+However, there was a significant drop in muscle hemoglobin oxygen saturation at the follow-up compared to initial visit (p = 0.044).
+
+Muscle DFA results
+```{r fig6c_dfa}
+
+df_muscle_fu_plot %>%
+
+  dplyr::filter(!grepl("avg_Hb", variable) & !grepl("second", variable)) %>% 
+  mutate(measurement = ifelse(grepl("conc", variable), "THC", "O2_sat")) %>% 
+  mutate(measurement = factor(measurement, levels = c("O2_sat", "THC"))) %>% 
+  mutate(pvals = ifelse(measurement == "THC", round(thc_muscle_fu_pval, 3), round(o2sat_muscle_fu_pval, 3))) %>% 
+  mutate(ycoord = ifelse(measurement == "THC", 400, 100)) %>%
+
+  ggplot(aes(x = follow_up, y = value)) +
+  geom_violin(aes(color = follow_up)) +
+  geom_point(shape = 1, size = 2) +
+  stat_summary(fun = "median", geom = "crossbar", aes(x = follow_up, color = follow_up), size = 0.2, width = 0.5) +
+  geom_path(aes(group = subject_id), lty = 2, alpha = 0.7) +
+
+  geom_hline(yintercept = 0.45, alpha = 0.4, lty = 2) +
+  geom_hline(yintercept = 0.55, alpha = 0.4, lty = 2) +
+  geom_hline(yintercept = 1.0, alpha = 0.4, lty = 2) +
+  geom_hline(yintercept = 1.5, alpha = 0.4, lty = 2) +
+
+  # geom_segment(aes(x = 1, y = ycoord, xend = 2, yend = ycoord), size = 0.3) +
+  # geom_segment(aes(x = 1, y = ycoord, xend = 1, yend = ycoord - .04*ycoord), size = 0.3) +
+  # geom_segment(aes(x = 2, y = ycoord, xend = 2, yend = ycoord - .04*ycoord), size = 0.3) +
+  # geom_text(x = 1.5, aes(y = ycoord + .025*ycoord, label = paste0("p = ", pvals))) +
+
+  facet_wrap(~measurement) +
+  ylim(0.45, 1.05) +
+  labs(y = "", x = "") +
+  scale_color_manual(values = my_col_fu) +
+
+  theme_bw() +
+  theme(legend.position = "none") +
+  theme(text = element_text(family = "Arial"),
+        strip.text = element_text(size = 15),
+        axis.title = element_text(size = 15),
+        axis.text = element_text(size = 15),
+        legend.position = "none")
+
+```
+No difference from initial visit to follow up - Hb_oxy pval: 0.17; Hb_tot pval: 0.10.
+
+### Figure 7: Hemoglobin concentration trends in extravascular edema vs vascular engorgement
+
+```{r fig7, fig.width = 12, eval = FALSE }
 

### 2021-11-05 final_figures.Rmd
scripts, no edits

--- 2021-08-20 figures.Rmd
@@
 ---
-title: "Dysregulation of Cerebral Hemoglobin Content in Malawian Children with Malaria"
-author: "Rachel Smith"
+title: "Cerebral oxygenation manuscript figures"
 date: "`r Sys.Date()`"
@@
 
-```{r setup, include = FALSE}
+```{r setup, include=FALSE}
 
@@
 
+# Load libraries
 library(tidyverse)
@@
 
+# Load data
+load("psd_plot_data.Rdata")
 load("glm_model.Rdata")
@@
 
-### Table 1: Summary of clinical data by patient group
+### Table 1. Baseline clinical characteristics
 
@@
 
-### Figure 1: Cerebral tissue hemoglobin concentration and oxygen saturation
+### Fig 0. (optional) Power spectral density curves
+
+```{r fig0}
+
+# plot
+f_plot_psd <- function(df, status, upper_lim) {
+
+  df %>% 
+    dplyr::filter(Status == status) %>% 
+    dplyr::filter(freq > 0.01 & freq < upper_lim) %>% 
+    ggplot(aes(x = freq, y = med_power)) +
+    geom_line() +
+    geom_vline(xintercept = 0.1, color = "red", lty = 2) +
+    facet_grid(tissue ~ variable, scales = "free_y") +
+    labs(x = "Frequency", y = "Median power across patients") +
+    theme_bw()
+
+}
+
+## 0-1 Hz
+
+  # HC
+  f_plot_psd(df_psd, "HC", 1) + ggtitle("Healthy children")
+
+  # UM
+  f_plot_psd(df_psd, "UM", 1) + ggtitle("Uncomplicated malaria")
+
+  # CM
+  f_plot_psd(df_psd, "CM", 1) + ggtitle("Cerebral malaria")
+
+## 0-3 Hz
+
+  # HC
+  f_plot_psd(df_psd, "HC", 3) + ggtitle("Healthy children")
+
+  # UM
+  f_plot_psd(df_psd, "UM", 3) + ggtitle("Uncomplicated malaria")
+
+  # CM
+  f_plot_psd(df_psd, "CM", 3) + ggtitle("Cerebral malaria")
+
+```
+
+### Fig 1. Hb_oxy and Hb_tot in the brain and muscle
 
@@
 # extract variables of interest
-voi <- df_results %>% colnames() %>% .[4:9]
+voi <- df_all_dfa_results %>% colnames() %>% .[-(1:2)]
 
@@
 df_hbox_voi_long <- df_hbox_long %>% 
-  dplyr::filter(variable %in% voi) %>% 
-  mutate(variable = factor(variable, levels = c("avg_Hb_o2sat", "avg_Hb_conc",
-                                                "Hb_o2sat_overall_alpha", "Hb_conc_overall_alpha",
-                                                "Hb_o2sat_second_alpha", "Hb_conc_second_alpha")))
+  dplyr::filter(variable %in% voi & !grepl("fu", variable)) %>%
+
+  mutate(tissue = as.factor(ifelse(grepl("muscle", variable), "muscle", "brain")),
+         measurement = factor(ifelse(grepl("conc", variable), "Hb_tot", "Hb_oxy"), levels = c("Hb_oxy", "Hb_tot")),
+         metric = case_when(
+
+           grepl("overall_alpha|conc_alpha|o2sat_alpha", variable) ~ "alpha",
+           grepl("second_alpha", variable) ~ "second_alpha",
+           TRUE ~ "mean"
+
+         )) %>% 
+  dplyr::select(-variable)
 
@@
 # plot
-df_hbox_voi_long %>% dplyr::filter(!(grepl("alpha", variable))) %>%
+df_hbox_voi_long %>% dplyr::filter(!(grepl("alpha", metric))) %>%
 
@@
   stat_summary(fun = "median", geom = "crossbar", aes(color = Status), size = 0.2, width = 0.5) +
-  facet_wrap(~variable, scales = "free_y") +
+  facet_grid(measurement ~ tissue, scales = "free_y") +
 
   scale_color_manual(values = my_col) +
+  labs(x = "") +
   theme_bw() +
@@
 
-  theme(text = element_text(family = "Arial"),
+  theme(# text = element_text(family = "Arial"),
         strip.text = element_text(size = 15),
@@
 
-```{r fig1.pvals, eval = FALSE}
-
-# hemoglobin concentration
-kruskal.test(avg_Hb_conc ~ Status, data = df_hbox)
-DescTools::DunnTest(avg_Hb_conc ~ Status, data = df_hbox, method = "BH")
-
-# hemoglobin oxygen saturation
-kruskal.test(avg_Hb_o2sat ~ Status, data = df_hbox)
-
-```
-
-**Figure 1.** Shown are the variables measured by Oxiplex, plotted by patient status.
-Cerebral tissue hemoglobin concentration levels were different between cerebral malaria patients and the other two groups, but did not differ between uncomplicated malaria patients and healthy volunteers (Kruskal-Wallis adjusted p-value = 5.0x10^-7^; Dunn Test post-hoc adjusted p-values: CM vs HC = 3.2x10^-6^, CM vs UM = 2.8x10^-4^, HC vs UM = 0.87).
-Levels of cerebral hemoglobin O~2~ saturation did not differ among groups (Kruskal-Wallis adjusted p-value = 0.47).
-
-### Figure 2: Detrended fluctuation analysis results
+### Fig 2. Hb_oxy alpha and Hb_tot alpha in the brain and muscle
 
@@
 
-df_hbox_voi_long %>% dplyr::filter(grepl("second_alpha", variable)) %>%
+# plot
+df_hbox_voi_long %>% dplyr::filter((grepl("second_alpha", metric))) %>%
 
@@
   stat_summary(fun = "median", geom = "crossbar", aes(color = Status), size = 0.2, width = 0.5) +
+  facet_grid(measurement ~ tissue) +
 
@@
 
-  facet_wrap(~variable, scales = "free_y") +
-
   ylim(0, 1.5) +
   scale_color_manual(values = my_col) +
+  labs(x = "") +
   theme_bw() +
+  theme(legend.position = "none")  +
 
-  theme(text = element_text(family = "Arial"),
+  theme(# text = element_text(family = "Arial"),
         strip.text = element_text(size = 15),
@@
         axis.title.y = element_blank(),
-        axis.text = element_text(size = 15),
-        legend.position = "none")
+        axis.text = element_text(size = 15))
 
@@
 
-```{r fig2.pvals, eval = FALSE}
+### Fig 3. Logistic regression modeling of uncomplicated and cerebral malaria subjects
 
-# hemoglobin concentration alpha
-kruskal.test(Hb_conc_second_alpha ~ Status, data = df_hbox)
-DescTools::DunnTest(Hb_conc_second_alpha ~ Status, data = df_hbox, method = "BH")
+```{r fig3_data}
 
-# hemoglobin oxygen saturation alpha
-kruskal.test(Hb_o2sat_second_alpha ~ Status, data = df_hbox)
-DescTools::DunnTest(Hb_o2sat_second_alpha ~ Status, data = df_hbox, method = "BH")
+# include variables of interest
 
-```
-
-**Figure 2.** Shown are the alpha values associated with the hemoglobin concentration and oxygen saturation signals measured by Oxiplex.
-Differences among cerebral tissue hemoglobin concentration alpha values were significant by Kruskal-Wallis (adjusted p-value = 7.16x10^-4^), and differences between uncomplicated malaria patients and the other two patient groups were significant by post-hoc Dunn Test (adjusted p-values: CM vs HV = 0.10, CM vs UM = 4.7x10^-4^, HV vs UM = 0.03).
-Cerebral hemoglobin O~2~ saturation alpha values differed between cerebral malaria patients and the other two patient groups (Kruskal-Wallis adjusted p-value = 4.2x10^-5^; Dunn Test post-hoc adjusted p-values: CM vs HV = 7.5x10^-4^, CM vs UM = 3.5x10^-4^, HV vs UM = 0.32).
-
-### Figure 3: Logistic regression model on CM vs UM
-
-#### A. PCA
-```{r fig3a, fig.height = 8}
-
-# include variables of interest, remove healthy children from analysis
-
-df_hbox_voi <- df_hbox %>% 
+df_hbox_log_reg <- df_hbox %>% 
   dplyr::select(c("subject_id", "Status", "Hematocrit", "Lactate", "avg_Hb_conc", "Hb_conc_second_alpha", "avg_Hb_o2sat", "Hb_o2sat_second_alpha"))
@@
 
-  df_hbox_impute <- df_hbox_voi %>% 
+  df_hbox_impute <- df_hbox_log_reg %>% 
     group_by(Status) %>% 
@@
 
+```
+
+#### A. Correlations of variables within subject groups
+
+```{r fig3a, fig.show = "hold", out.width = "50%", fig.height = 12, fig.width = 15}
+
+# calculate spearman correlations in each group
+
+f_calc_cor <- function(df, status) {
+
+  m <- df %>%
+    dplyr::filter(Status == status) %>% 
+    dplyr::select(-c(subject_id, Status)) %>% 
+    cor(method = "spearman")
+
+  diag(m) <- 0
+
+  m
+}
+
+# create run correlations of potential regressors
+m_cor_um <- df_hbox_impute %>% f_calc_cor(status = "UM")
+
+m_cor_cm <- df_hbox_impute %>% f_calc_cor(status = "CM")
+
+# correlation matrix
+
+# write plot function
+f_plot_cor <- function(mat) {
+
+  melt(mat) %>%
+
+    mutate_if(is.factor, factor, levels = c("Lactate", "Hematocrit", "avg_Hb_o2sat", "avg_Hb_conc", "Hb_o2sat_second_alpha", "Hb_conc_second_alpha")) %>%
+
+    ggplot(aes(x = X1, y = X2, fill = value, label = ifelse(value != 0, round(value, 2), NA))) + 
+    geom_tile() +
+    scale_fill_gradient2(low = "#4575b4", high = "#d73027") +
+    geom_text(size = 9) +
+
+    xlab("") +
+    ylab("") +
+    theme_bw() +
+    theme(# text = element_text(family = "Arial"),
+          axis.text.x = element_text(angle = 45, hjust = 1),
+          axis.text = element_text(size = 35),
+          legend.text = element_text(size = 40),
+          legend.title = element_text(size = 50),
+          legend.key.size = unit(2, 'cm'),
+          plot.title = element_text(size = 30)
+    )
+
+}
+
+  # extract legend
+  p_leg <- m_cor_um %>% f_plot_cor() %>% get_legend()
+
+  # create plots
+  m_cor_um %>% f_plot_cor() + ggtitle("Uncomplicated malaria") + theme(legend.position = "none")
+
+  m_cor_cm %>% f_plot_cor() + ggtitle("Cerebral malaria") + theme(legend.position = "none")
+
+  as_ggplot(p_leg)
+
+```
+
+#### B. Principal component analysis
+
+```{r fig3b}
 # remove healthy volunteers for logistic regression analysis
-df_hbox_log_reg <- df_hbox_impute %>% dplyr::filter(Status != "HC")
+df_hbox_impute <- df_hbox_impute %>% dplyr::filter(Status != "HC")
 
 # convert tibble to df
-df_hbox_voi_df <- df_hbox_log_reg %>% 
+df_hbox_df <- df_hbox_impute %>% 
   dplyr::select(-c(subject_id, Status)) %>% 
@@
 # add rownames
-rownames(df_hbox_voi_df) <- df_hbox_log_reg %>% .$subject_id
+rownames(df_hbox_df) <- df_hbox_impute %>% .$subject_id
 
 # compute PCA
-my_pca <- prcomp(df_hbox_voi_df, scale = TRUE)
+my_pca <- prcomp(df_hbox_df, scale = TRUE)
 
 # define groups
-groups <- as.factor(df_hbox_log_reg$Status)
+groups <- as.factor(df_hbox_impute$Status)
 
@@
   theme_bw() +
-  theme(text = element_text(family = "Arial"),
+  theme(# text = element_text(family = "Arial"),
         axis.text = element_text(size = 20),
@@
 
-#### B. Confusion matrix
+#### C. Coefficients
+```{r table2c}
 
-```{r fig3b_prep, eval = FALSE}
+df_CI <- confint(hbox_final$fit) %>% exp() %>% as_tibble(rownames = "term")
 
-library(themis)
+table2c <- hbox_final %>%
+  tidy(exponentiate = TRUE) %>%
+  left_join(df_CI, by = "term") %>%
+  mutate(estimate = ifelse(estimate < 10^-2, scientific(estimate, digits = 3), round(estimate, 2)),
+         conf.low = ifelse(`2.5 %` < 10^-2, scientific(`2.5 %`, digits = 3), round(`2.5 %`, 2)),
+         conf.high = ifelse(`97.5 %` < 10^-2, scientific(`97.5 %`, digits = 3), round(`97.5 %`, 2))) %>%
+  mutate(p.value = scientific(p.value, digits = 3),
+         statistic = round(statistic, 2)) %>%
+  unite("95% CI", c(conf.low, conf.high), sep = ", ") %>%
+  dplyr::select(c(term, estimate, `95% CI`, statistic, p.value)) %>% 
+  dplyr::rename("Term" = "term", "Odds ratio" = "estimate",
+                "t-statistic" = "statistic", "p-value" = "p.value") %>%
+  as.data.frame()
 
-# select [HbT], [HbT] alpha, and hematocrit
-  df_hbox_model <- df_hbox_log_reg %>% 
-    dplyr::select(c(Status, Hematocrit, avg_Hb_conc, Hb_conc_second_alpha)) %>% 
-    mutate(`1/10_Hb_conc_alpha` = Hb_conc_second_alpha*0.1, avg_Hb_conc_nm = avg_Hb_conc/1000) %>% 
-    dplyr::select(-c(Hb_conc_second_alpha, avg_Hb_conc)) %>% 
-    mutate(Status = as.factor(ifelse(Status == "CM", 1, 0)))
-
-# resample
-set.seed(234)
-hbox_boot <- bootstraps(df_hbox_model, strata = Status, times = 1000)
-
-# recipe
-hbox_rec <- recipe(Status ~ ., data = df_hbox_model) %>% 
-  step_BoxCox(all_predictors()) %>% 
-  step_nzv(all_predictors()) %>% 
-  step_smote(Status) # upsample to deal with unbalanced sample sizes
-
-# create workflow
-hbox_wf <- workflow() %>% add_recipe(hbox_rec)
-
-# logistic regression model specifications
-glm_spec <- logistic_reg() %>% set_engine("glm")
-
-# create model on 1000 different resamples
-glm_res <- hbox_wf %>% 
-  add_model(glm_spec) %>%
-  fit_resamples(
-    resamples = hbox_boot,
-    metrics = metric_set(roc_auc, accuracy, sensitivity, specificity),
-    control = tune::control_resamples(save_pred = TRUE, verbose = TRUE)
-  )
-
-# select best model to be used for evaluation
-glm_best <- glm_res %>% select_best("roc_auc")
-
-# finalize model
-hbox_final <- hbox_wf %>% 
-  add_model(glm_spec) %>% 
-  finalize_workflow(glm_best) %>%
-  fit(df_hbox_model) %>%
-  pull_workflow_fit()
-
-# save objects so model doesn't rerun every time we knit
-save(glm_res, hbox_final, file = "glm_model.Rdata")
+table2c %>% knitr::kable(align = rep('c', 5))
 
@@
 
-```{r fig3b_plot, fig.height = 8}
+#### D. ROC curve of final model
 
-final_res_conf_mat <- glm_res %>% 
-  conf_mat_resampled() %>% 
-  mutate(Prediction = ifelse(Prediction == 1, "CM", "UM")) %>% 
-  mutate(Truth = ifelse(Truth == 1, "CM", "UM"))
-
-total_true_UM <- final_res_conf_mat %>% dplyr::filter(Truth == "UM") %>% .$Freq %>% sum()
-total_true_CM <- final_res_conf_mat %>% dplyr::filter(Truth == "CM") %>% .$Freq %>% sum()
-
-final_res_conf_mat %>% 
-  mutate(total_true = ifelse(Truth == "CM", Freq/total_true_CM, Freq/total_true_UM)) %>% 
-  mutate(`Percent occurrence` = round(total_true*100, 2)) %>%
-
-  # plot
-  ggplot(aes(x = Truth, y = Prediction, fill = `Percent occurrence`, label = `Percent occurrence`)) +
-  geom_tile() +
-  geom_text(size = 6) +
-  scale_fill_gradient2(low = "#4575b4", high = "#d73027") +
-
-  theme_bw() +
-  theme(text = element_text(family = "Arial"),
-        axis.title = element_text(size = 25),
-        axis.text = element_text(size = 20),
-        legend.title = element_text(size = 20),
-        legend.text = element_text(size = 15))
-
-```
-
-#### C. ROC curve of final model compared to other iterations
-
-```{r fig3c_prep, eval = FALSE}
+```{r fig3d_prep, eval = FALSE}
 
@@
 
-```{r fig3c_plot, fig.height = 8}
+```{r fig3d_plot, fig.height = 8}
 
@@
             size = 2.5) +
-  geom_path(data = df_roc %>% filter(model != "final"), 
-            aes(x = 1 - specificity, y = sensitivity, color = model), 
-            size = 1) +
   geom_abline(lty = 3) +
@@
 
-  theme(text = element_text(family = "Arial"),
+  theme(# text = element_text(family = "Arial"),
         axis.title = element_text(size = 25),
@@
         legend.text = element_text(size = 15),
-        legend.position = "top")
+        legend.position = "none")
 
@@
 
-**Figure 3.** Logistic regression model discriminating CM vs UM patients using cerebral tissue hemoglobin concentration, cerebral tissue hemoglobin concentration alpha, and hematocrit as the regressors **(A)** PCA of patients with CM and UM as determined by NIRS variables.
-Color indicates patient status, ellipses show median and 95% confidence interval on the cluster.
-**(B)** Confusion matrix representing the results of the logistic regression model averaged across 1000 bootstrap resamples.
-Percent occurrence indicates percentage times that the event (shown by the x- and y-axes) occurred.
+```{r fig3d_inset}
 
-### Table 2. Logistic regression model evaluation
-
-#### A. Model metrics
-```{r table2a}
-
-table2a <- collect_metrics(glm_res) %>% 
+fig3d_inset <- collect_metrics(glm_res) %>% 
   dplyr::select(.metric, mean, std_err) %>% 
@@
 
-table2a %>% knitr::kable(align = rep('c', 4))
+fig3d_inset %>% knitr::kable(align = rep('c', 4))
 
@@
 
-#### B. Coefficients
-```{r table2b}
+### Fig 4. Explanatory model of how to interpret hemoglobin concentration
 
-table2b <- hbox_final %>% 
-  tidy(exponentiate = TRUE) %>% 
-  mutate(estimate = ifelse(estimate < 10^-2, scientific(estimate, digits = 3), round(estimate, 2))) %>% 
-  mutate(p.value = scientific(p.value, digits = 3),
-         std.error = round(std.error, 2),
-         statistic = round(statistic, 2)) %>% 
-  dplyr::rename("Term" = "term", "Odds ratio" = "estimate", "Standard error" = "std.error", 
-                "Statistic" = "statistic", "p-value" = "p.value") %>% 
-  as.data.frame()
+See manuscript draft (figure created in PowerPoint)
 
-table2b %>% knitr::kable(align = rep('c', 5))
-
-```
-
-```{r export_table2, eval = FALSE}
-
-list(table2a, table2b) %>% write.csv("Outputs/table_2.csv")
-
-```
-
-**Table 2.** Evaluation of the best logistic regression model select from 1000.
-
-### Figure 4: Relationships among hemoglobin concentration and hematocrit
-```{r fig4, fig.height = 8}
-
-df_hbox %>% filter(Status != "HC") %>%
-
-  ggplot(aes(x = avg_Hb_conc, y = Hematocrit)) +
-  geom_point(aes(color = Status), alpha = 0.7, size = 3) +
-
-  xlab("Hemoglobin concentration (μM)") +
-  ylab("Hematocrit (%)") +
-
-  theme_bw() +
-  theme(text = element_text(family = "Arial"),
-        axis.text = element_text(size = 20),
-        axis.title = element_text(size = 25),
-        legend.text = element_text(size = 15),
-        legend.title = element_text(size = 20))
-
-```
-
-**Figure 4.**
-
-### Figure 6: Hemoglobin concentration trends in extravascular edema vs vascular engorgement
-
-```{r fig6, fig.width = 12}
-
-df_fig6 <- tibble(edema = rnorm(n = 90, mean = 20, sd = 20),
-                  normal = rnorm(n = 90, mean = 70, sd = 20),
-                  engorge = rnorm(n = 90, mean = 130, sd = 20)) %>% 
-  pivot_longer(1:3, names_to = "variable", values_to = "value") %>% 
-  mutate(variable = factor(variable, levels = c("edema", "normal", "engorge")))
-
-df_fig6 %>% 
-  ggplot(aes(x = variable, y = value)) +
-  geom_point(shape = 1, size = 3, position = position_jitter(0.1)) +
-  stat_summary(fun = "median", geom = "crossbar", size = 0.2, width = 0.5) +
-  labs(x = "", y = "") +
-  theme_classic() +
-  theme(axis.text = element_blank(),
-        axis.ticks = element_blank())
-
-```
-
-### Supplementary figure 1: Spearman correlations of potential regressors
-```{r fig_s1, fig.show = "hold", out.width = "50%", fig.height = 12, fig.width = 15}
-
-# calculate spearman correlations in each group
-
-f_calc_cor <- function(df, status) {
-
-  df %>%
-    dplyr::filter(Status == status) %>% 
-    dplyr::select(-c(subject_id, Status)) %>% 
-    cor(method = "spearman")
-
-}
-
-# create run correlations of potential regressors
-m_cor_hv <- df_hbox_impute %>% f_calc_cor(status = "HC")
-
-m_cor_um <- df_hbox_impute %>% f_calc_cor(status = "UM")
-
-m_cor_cm <- df_hbox_impute %>% f_calc_cor(status = "CM")
-
-# correlation matrix
-
-# write plot function
-f_plot_cor <- function(mat) {
-
-  melt(mat) %>%
-
-    mutate_if(is.factor, factor, levels = c("Lactate", "Hematocrit", "avg_Hb_o2sat", "avg_Hb_conc", "Hb_o2sat_second_alpha", "Hb_conc_second_alpha")) %>%
-
-    ggplot(aes(x = X1, y = X2, fill = value, label = round(value, 2))) + 
-    geom_tile() +
-    scale_fill_gradient2(low = "#4575b4", high = "#d73027") +
-    geom_text(size = 9) +
-
-    xlab("") +
-    ylab("") +
-    theme_bw() +
-    theme(text = element_text(family = "Arial"),
-          axis.text.x = element_text(angle = 45, hjust = 1),
-          axis.text = element_text(size = 35),
-          legend.text = element_text(size = 40),
-          legend.title = element_text(size = 50),
-          legend.key.size = unit(2, 'cm'),
-          plot.title = element_text(size = 30)
-    )
-
-}
-
-  # extract legend
-  p_leg <- m_cor_hv %>% f_plot_cor() %>% get_legend()
-
-  # create plots
-  m_cor_hv %>% f_plot_cor() + ggtitle("Healthy chilren") + theme(legend.position = "none")
-
-  m_cor_um %>% f_plot_cor() + ggtitle("Uncomplicated malaria") + theme(legend.position = "none")
-
-  m_cor_cm %>% f_plot_cor() + ggtitle("Cerebral malaria") + theme(legend.position = "none")
-
-  as_ggplot(p_leg)
-
-```
-
-**Figure S1.** Shown are the spearman correlations among hematocrit, lactate, hemoglobin concentration and its alpha, and hemoglobin oxygen saturation and its alpha.
-Red indicates a positive correlation; blue indicates a negative correlation.
-
-### Supplementary figure 2: Logistic regression curves of potential regressors
-```{r fig_s2, fig.height = 12}
-
-# pivot longer and change factor levels
-df_hbox_reg_long <- df_hbox_impute %>% 
-  mutate(Status = as.numeric(ifelse(Status == "CM", 1, 0))) %>% 
-  pivot_longer(3:ncol(.), names_to = "variable", values_to = "value") %>% 
-  mutate(variable = factor(variable, levels = c("Lactate", "Hematocrit", "avg_Hb_o2sat", "avg_Hb_conc", "Hb_o2sat_second_alpha", "Hb_conc_second_alpha")))
-
-# plot as individual regression curves
-df_hbox_reg_long %>%
-
-  ggplot(aes(x = value, y = Status)) +
-  geom_point() +
-  geom_smooth(method = "glm", method.args = list(family = "binomial")) +
-  facet_wrap(~variable, scales = "free_x", nrow = 5, ncol = 2) +
-
-  ylab("Status: 1 = CM, 0 = UM") +
-  xlab("") +
-
-  theme_bw() +
-  theme(text = element_text(family = "Arial"),
-        axis.text = element_text(size = 20),
-        axis.title = element_text(size = 25),
-        legend.text = element_text(size = 15),
-        legend.title = element_text(size = 20),
-        strip.text = element_text(size = 20))
-
-```
-
-**Figure S2.** Plotted are the individual logistic regression curves of individual variables tested for their discriminatory power between UM and CM.
-On the y-axis, 1 represents CM while 0 represents UM.
-A steeper curve indicates higher discriminatory power.
-
-### Supplementary figure 3: Detrended fluctuation analysis
+## Supplement
+### SFig 1. DFA analysis illustration
 
@@
   theme_bw() +
-  theme(text = element_text(family = "Arial"),
+  theme(# text = element_text(family = "Arial"),
         axis.text = element_text(size = 25),
@@
 ```
+
+### SFig 2. Logistic regression curves
+
+```{r fig_s2, fig.height = 12}
+
+# pivot longer and change factor levels
+df_hbox_reg_long <- df_hbox_impute %>% 
+  mutate(Status = as.numeric(ifelse(Status == "CM", 1, 0))) %>% 
+  pivot_longer(3:ncol(.), names_to = "variable", values_to = "value") %>% 
+  mutate(variable = factor(variable, levels = c("Lactate", "Hematocrit", "avg_Hb_o2sat", "avg_Hb_conc", "Hb_o2sat_second_alpha", "Hb_conc_second_alpha")))
+
+# plot as individual regression curves
+df_hbox_reg_long %>%
+
+  ggplot(aes(x = value, y = Status)) +
+  geom_point() +
+  geom_smooth(method = "glm", method.args = list(family = "binomial")) +
+  facet_wrap(~variable, scales = "free_x", nrow = 5, ncol = 2) +
+
+  ylab("Status: 1 = CM, 0 = UM") +
+  xlab("") +
+
+  theme_bw() +
+  theme(# text = element_text(family = "Arial"),
+        axis.text = element_text(size = 20),
+        axis.title = element_text(size = 25),
+        legend.text = element_text(size = 15),
+        legend.title = element_text(size = 20),
+        strip.text = element_text(size = 20))
+
+```
+
+**Figure S2.** Plotted are the individual logistic regression curves of individual variables tested for their discriminatory power between UM and CM.
+On the y-axis, 1 represents CM while 0 represents UM.
+A steeper curve indicates higher discriminatory power.

### 2021-11-05 Scripts/psd_plots.R

@@ -1,0 +1,155 @@
+
+
+
+# libraries ---------------------------------------------------------------
+
+
+  library(tidyverse)
+  library(reticulate)
+  library(purrr)
+  library(data.table)
+
+
+  
+# data --------------------------------------------------------------------
+
+  
+  load("filtered_signals.Rdata")
+  load("filtered_muscle_signals.Rdata")
+  
+
+# python script to calculate freq & power for each patient ---------------------------------
+
+
+  np <- import("numpy")
+  
+  sampling_rate <- 50 # Hz
+  
+  f_psd <- function(df, num){
+    
+    df_id <- df %>% count(number, Status) %>% dplyr::select(-n)
+    
+    df <- df %>% filter(number == num)
+    fourier_transform <- np$fft$rfft(df$sg_filt)  
+    abs_fourier_transform <- np$abs(fourier_transform)  
+    power_spectrum <- np$square(abs_fourier_transform)  
+    frequency <- np$linspace(0, sampling_rate/2, length(power_spectrum))  
+    df_psd <- tibble(number = num, freq = frequency, power = power_spectrum)
+    
+    df_psd <- df_psd %>% left_join(df_id, by = "number")
+    
+    df_psd <- df_psd %>% 
+      dplyr::filter(freq > 0.001) %>% 
+      mutate(freq_range = cut(freq, seq(from = 0, to = max(freq), by = 0.001)))
+    
+    df_psd
+    
+  }
+
+  
+
+# run function on all signals ---------------------------------------------
+
+
+
+  # Brain Hb_tot
+  df_brain_hbtot_psd <- unique(df_thc_filt_pass$number) %>% 
+    purrr::map(~f_psd(df_thc_filt_pass, .x)) %>% 
+    rbindlist() %>% 
+    as_tibble() %>% 
+    group_by(Status, freq_range) %>% 
+    summarise(med_power = median(power)) %>% 
+    mutate(tissue = "brain", variable = "Hb_tot")
+  
+  # Brain Hb_oxy
+  df_brain_hboxy_psd <- unique(df_o2sat_filt_pass$number) %>% 
+    purrr::map(~f_psd(df_o2sat_filt_pass, .x)) %>% 
+    rbindlist() %>% 
+    as_tibble() %>% 
+    group_by(Status, freq_range) %>% 
[…]

### 2022-01-10 final_figures.Rmd
updated final figures with final analysis data

@@
 library(tidyverse)
+library(readxl)
 library(DescTools)
@@
 # Load data
-load("psd_plot_data.Rdata")
-load("glm_model.Rdata")
-load("results.Rdata")
-load("roc_curves.Rdata")
-load("dfa_example.Rdata")
-load("dfa_results.Rdata")
-load("all_dfa_results.Rdata")
+load("objects/glm_model.Rdata")
+load("objects/bandpass_filtered_dfa_results.Rdata")
+load("objects/sims.Rdata")
 
@@
 
-  df_hbox <- read.csv("Data/master_datatable.csv") %>% 
+df_hbox <- read_csv("Data/Hans_datatable_exports/malawi key data v04Jan2022.csv") %>% 
   as_tibble() %>% 
-    dplyr::select(-1) %>% 
-    mutate(Status = replace(Status, Status == "HV", "HC")) %>% 
-    mutate(Status = factor(Status, levels = c("HC", "UM", "CM")))
+  janitor::clean_names() %>% 
+  mutate(status = replace(status, status == "HV", "HC")) %>% 
+  dplyr::rename("subject_id" = "subject_id_session") %>%
 
-# pivot long
-
-  df_hbox_long <- df_hbox %>% pivot_longer(cols = 7:ncol(.), names_to = "variable", values_to = "value")
+  dplyr::filter(status %in% c("CM", "HC", "UM") & session_no == 1 | subject_id == "TM0003CM01") %>% 
+  dplyr::filter(!grepl("blood", subject_id)) %>% 
+  mutate(status = factor(status, levels = c("HC", "UM", "CM"))) %>% 
+  mutate_at(c("bp_dias", "bp_sys", "glucose", "lactate"), as.numeric) %>% # converts remaining character variables to numerics
+  select(subject_id, status, everything()) %>% 
+  group_by(status)
 
@@
 
-```{r table1}
+#### Table 1a. All clinical data
+```{r table1a, eval = FALSE}
 
@@
 df_status <- df_hbox %>% 
-  count(Status) %>% 
-  pivot_wider(1:2, names_from = "Status", values_from = "n") %>% 
+  count() %>% 
+  pivot_wider(1:2, names_from = "status", values_from = "n") %>% 
   mutate(variable = "Number of subjects") %>% 
@@
 df_sex <- df_hbox %>% 
-  group_by(Status) %>% 
-  count(Sex) %>% 
-  dplyr::filter(Sex != "") %>% 
-  pivot_wider(names_from = "Sex", values_from = "n") %>% 
+  count(sex) %>% 
+  dplyr::filter(sex != "") %>% 
+  pivot_wider(names_from = "sex", values_from = "n") %>% 
   mutate(perc_fem = round(Female/(Female + Male)*100, 2)) %>%
 
-  pivot_wider(c(1,4), names_from = "Status", values_from = "perc_fem") %>% 
+  pivot_wider(c(1,4), names_from = "status", values_from = "perc_fem") %>% 
   mutate(variable = "Sex (% female)") %>% 
@@
 
+  # select variables for table
+  clinical_voi <- c("age_calc", "temperature", "hr", "bp_sys", "bp_dias", "resp_rate", "pulse_ox", "hct", "lactate", "glucose", "aa_arg")
+
+  # calculate median, first and third quartiles
   df_clinical_res <- df_hbox %>% 
+    select_at(clinical_voi) %>% 
+    summarise_all(funs(median = median, q1 = quantile(., probs = 0.25), q3 = quantile(., probs = 0.75)), na.rm = TRUE) %>% 
+    mutate_if(is.numeric, round, 2)
 
-  # dplyr::filter(Subject.ID..NIAID. != "TM1018") %>%
+  # write function to paste together values
+  f_unite_cols <- function(col) {
+    df_clinical_res %>% 
+      unite(!! col, starts_with(col), sep = "_") %>% 
+      dplyr::select(!! col) 
+    }
 
-  group_by(Status) %>% 
-  summarise(`Age (years)` = median(Age, na.rm = TRUE), age_q1 = quantile(Age, 0.25, na.rm = TRUE), age_q3 = quantile(Age, 0.75, na.rm = TRUE),
+  # paste together and transpose
+  df_clinical_res <- 1:length(clinical_voi) %>% 
+    purrr::map_dfc(~f_unite_cols(clinical_voi[.x])) %>% 
+    mutate_all(funs(str_replace(., "_", " ("))) %>% 
+    mutate_all(funs(str_replace(., "_", ", "))) %>% 
+    mutate_all(funs(paste0(., ")"))) %>%
 
-            `Temperature (°C)` = median(Temperature, na.rm = TRUE), temp_q1 = quantile(Temperature, 0.25, na.rm = TRUE), temp_q3 = quantile(Temperature, 0.75, na.rm = TRUE),
-            `Heart rate` = median(HR, na.rm = TRUE), hr_q1 = quantile(HR, 0.25, na.rm = TRUE), hr_q3 = quantile(HR, 0.75, na.rm = TRUE),
-            `Systolic BP` = median(BP.Systolic, na.rm = TRUE), sys_q1 = quantile(BP.Systolic, 0.25, na.rm = TRUE), sys_q3 = quantile(BP.Systolic, 0.75, na.rm = TRUE),
-            `Diastolic BP` = median(BP.Diastolic, na.rm = TRUE), dia_q1 = quantile(BP.Diastolic, 0.25, na.rm = TRUE), dia_q3 = quantile(BP.Diastolic, 0.75, na.rm = TRUE),
-            `Respiration rate` = median(RR, na.rm = TRUE), rr_q1 = quantile(RR, 0.25, na.rm = TRUE), rr_q3 = quantile(RR, 0.75, na.rm = TRUE),
-            `O~2~ saturation` = median(O2.Sat, na.rm = TRUE), o2_q1 = quantile(O2.Sat, 0.25, na.rm = TRUE), o2_q3 = quantile(O2.Sat, 0.75, na.rm = TRUE),
-
-            `Hematocrit (%)` = median(Hematocrit, na.rm = TRUE), hem_q1 = quantile(Hematocrit, 0.25, na.rm = TRUE), hem_q3 = quantile(Hematocrit, 0.75, na.rm = TRUE),
-            `Lactate (mmol/L)` = median(Lactate, na.rm = TRUE), lac_q1 = quantile(Lactate, 0.25, na.rm = TRUE), lac_q3 = quantile(Lactate, 0.75, na.rm = TRUE),
-            `Arginine (μmol/L)` = median(Arginine.umol.L, na.rm = TRUE), arg_q1 = quantile(Arginine.umol.L, 0.25, na.rm = TRUE), arg_q3 = quantile(Arginine.umol.L, 0.75, na.rm = TRUE)) %>% 
-  mutate_if(is.numeric, round, 1) %>%
-
-  mutate(`Age (years)` = paste0(`Age (years)`, " (", age_q1, ", ", age_q3, ")"),
-
-         `Temperature (°C)` = paste0(`Temperature (°C)`, " (", temp_q1, ", ", temp_q3, ")"),
-         `Heart rate` = paste0(`Heart rate`, " (", hr_q1, ", ", hr_q3, ")"),
-         `Systolic BP` = paste0(`Systolic BP`, " (", sys_q1, ", ", sys_q3, ")"),
-         `Diastolic BP` = paste0(`Diastolic BP`, " (", dia_q1, ", ", dia_q3, ")"),
-         `Respiration rate` = paste0(`Respiration rate`, " (", rr_q1, ", ", rr_q3, ")"),
-         `O~2~ saturation` = paste0(`O~2~ saturation`, " (", o2_q1, ", ", o2_q3, ")"),
-
-         `Hematocrit (%)` = paste0(`Hematocrit (%)`, " (", hem_q1, ", ", hem_q3, ")"),
-         `Lactate (mmol/L)` = paste0(`Lactate (mmol/L)`, " (", lac_q1, ", ", lac_q3, ")"),
-         `Arginine (μmol/L)` = paste0(`Arginine (μmol/L)`, " (", arg_q1, ", ", arg_q3, ")")) %>% 
-  dplyr::select(-c(age_q1, age_q3, 
-                   temp_q1, temp_q3, hr_q1, hr_q3, sys_q1, sys_q3, dia_q1, dia_q3, rr_q1, rr_q3, o2_q1, o2_q3, 
-                   hem_q1, hem_q3, lac_q1, lac_q3, arg_q1, arg_q3)) %>%
-
-  unique() %>% 
     t() %>% 
-  row_to_names(row_number = 1) %>% 
     as_tibble(rownames = "variable") %>% 
+    dplyr::rename("HC" = "V1", "UM" = "V2", "CM" = "V3") %>% 
     add_row(df_status, .before = 1) %>% 
-  add_row(df_sex, .before = 2) %>%
-
-  dplyr::rename(" " = "variable",
-                "Healthy children" = "HC",
-                "Uncomplicated malaria" = "UM",
-                "Cerebral malaria" = "CM")
-
-  # remove lactate data for HC
-  df_clinical_res[11,]$`Healthy children` <- "--"
+    add_row(df_sex, .before = 2)
 
 # export to CSV
-  # write.csv(df_clinical_res, file = "Outputs/table_1.csv")
+  write.csv(df_clinical_res, file = "Outputs/table_1.csv")
 
@@
 ```
-**Table 1.** A summary of subject information and hematological data.
-Group patient data presented as median (quartile 1, quartile 3).
-Note: Sex not recorded for 3 CM patients; Lactate data removed for healthy volunteers due to inconsistent data.
+**Table 1a.** A summary of subject information and hematological data. Group patient data presented as median (quartile 1, quartile 3).
 
-### Fig 0. (optional) Power spectral density curves
+#### Table 1b. CM-specific data
+```{r table1b, eval = FALSE}
 
-```{r fig0}
+cm_voi <- c("bcs_session", "brain_swell", "hrp2")
 
-# plot
-f_plot_psd <- function(df, status, upper_lim) {
+df_cm_voi <- df_hbox %>% 
+  ungroup %>% 
+  filter(status == "CM") %>% 
+  select_at(cm_voi) %>% 
+  mutate_at(c("bcs_session", "brain_swell"), as.factor)
 
-  df %>% 
-    dplyr::filter(Status == status) %>% 
-    dplyr::filter(freq > 0.01 & freq < upper_lim) %>% 
-    ggplot(aes(x = freq, y = med_power)) +
-    geom_line() +
-    geom_vline(xintercept = 0.1, color = "red", lty = 2) +
-    facet_grid(tissue ~ variable, scales = "free_y") +
-    labs(x = "Frequency", y = "Median power across patients") +
-    theme_bw()
+# BCS counts
+df_cm_voi %>% count(bcs_session) %>% knitr::kable()
 
-}
+# Brain swell score counts
+df_cm_voi %>% count(brain_swell) %>% knitr::kable()
 
-## 0-1 Hz
+# HRP
+df_cm_hrp2 <- df_cm_voi %>% 
+  select(cm_voi[3]) %>% 
+  summarise_all(funs(median = median, q1 = quantile(., probs = 0.25), q3 = quantile(., probs = 0.75)), na.rm = TRUE) %>% 
+  unite(hrp2, 1:2, sep = " (") %>% 
+  unite(hrp2, 1:2, sep = ", ") %>% 
+  mutate(hrp2 = paste0(hrp2, ")"))
 
-  # HC
-  f_plot_psd(df_psd, "HC", 1) + ggtitle("Healthy children")
-
-  # UM
-  f_plot_psd(df_psd, "UM", 1) + ggtitle("Uncomplicated malaria")
-
-  # CM
-  f_plot_psd(df_psd, "CM", 1) + ggtitle("Cerebral malaria")
-
-## 0-3 Hz
-
-  # HC
-  f_plot_psd(df_psd, "HC", 3) + ggtitle("Healthy children")
-
-  # UM
-  f_plot_psd(df_psd, "UM", 3) + ggtitle("Uncomplicated malaria")
-
-  # CM
-  f_plot_psd(df_psd, "CM", 3) + ggtitle("Cerebral malaria")
+df_cm_hrp2 %>% knitr::kable()
 
 ```
+**Table 1b.** CM-specific clinical characteristics
 
@@
 
-# extract variables of interest
-voi <- df_all_dfa_results %>% colnames() %>% .[-(1:2)]
-
-# select from the long df
-df_hbox_voi_long <- df_hbox_long %>% 
-  dplyr::filter(variable %in% voi & !grepl("fu", variable)) %>%
-
-  mutate(tissue = as.factor(ifelse(grepl("muscle", variable), "muscle", "brain")),
-         measurement = factor(ifelse(grepl("conc", variable), "Hb_tot", "Hb_oxy"), levels = c("Hb_oxy", "Hb_tot")),
-         metric = case_when(
-
-           grepl("overall_alpha|conc_alpha|o2sat_alpha", variable) ~ "alpha",
-           grepl("second_alpha", variable) ~ "second_alpha",
-           TRUE ~ "mean"
-
-         )) %>% 
-  dplyr::select(-variable)
+# extract hb_tot and hb_oxy for the brain and muscle
+df_fig1 <- df_hbox %>% 
+  select(contains(c("hb_tot","hb_oxy")), -contains(c("alpha", "bsl"))) %>%
+  ungroup() %>% 
+  pivot_longer(2:ncol(.), names_to = "variable", values_to = "value") %>% 
+  separate(variable, into = c("tissue", "variable"), sep = "_", extra = "merge")
 
@@
 # plot
-df_hbox_voi_long %>% dplyr::filter(!(grepl("alpha", metric))) %>%
+df_fig1 %>%
 
-  ggplot(aes(x = Status, y = value)) +
-  geom_violin(aes(color = Status)) +
+  ggplot(aes(x = status, y = value)) +
+  geom_violin(aes(color = status)) +
   geom_jitter(position = position_jitter(0.2), shape = 1, size = 2) +
-  stat_summary(fun = "median", geom = "crossbar", aes(color = Status), size = 0.2, width = 0.5) +
-  facet_grid(measurement ~ tissue, scales = "free_y") +
+  stat_summary(fun = "median", geom = "crossbar", aes(color = status), size = 0.2, width = 0.5) +
+  facet_grid(variable ~ tissue, scales = "free_y") +
 
@@
 
+# extract hb_tot and hb_oxy for the brain and muscle
+df_fig2 <- df_hbox %>% 
+  select(contains("alpha2")) %>%
+  ungroup() %>% 
+  pivot_longer(2:ncol(.), names_to = "variable", values_to = "value") %>% 
+  separate(variable, into = c("tissue", "variable"), sep = "_", extra = "merge") %>% 
+  mutate(variable = str_remove(variable, "2"))
+
 # plot
-df_hbox_voi_long %>% dplyr::filter((grepl("second_alpha", metric))) %>%
+df_fig2 %>%
 
-  ggplot(aes(x = Status, y = value)) +
-  geom_violin(aes(color = Status)) +
+  ggplot(aes(x = status, y = value)) +
+  geom_violin(aes(color = status)) +
   geom_jitter(position = position_jitter(0.2), shape = 1, size = 2) +
-  stat_summary(fun = "median", geom = "crossbar", aes(color = Status), size = 0.2, width = 0.5) +
-  facet_grid(measurement ~ tissue) +
+  stat_summary(fun = "median", geom = "crossbar", aes(color = status), size = 0.2, width = 0.5) +
+  facet_grid(variable ~ tissue, scales = "free_y") +
 
-  geom_hline(yintercept = 0.45, alpha = 0.4, lty = 2) +
-  geom_hline(yintercept = 0.55, alpha = 0.4, lty = 2) +
-  geom_hline(yintercept = 1.0, alpha = 0.4, lty = 2) +
-  geom_hline(yintercept = 1.5, alpha = 0.4, lty = 2) +
+  geom_hline(yintercept = 0.4, lty = 2, alpha = 0.7) +
+  geom_hline(yintercept = 0.6, lty = 2, alpha = 0.7) +
+  geom_hline(yintercept = 1.0, lty = 2, alpha = 0.7) +
+  geom_hline(yintercept = 1.5, lty = 2, alpha = 0.7) +
 
-  ylim(0, 1.5) +
   scale_color_manual(values = my_col) +
+  ylim(c(0, 1.5)) +
   labs(x = "") +
@@
 
-### Fig 3. Logistic regression modeling of uncomplicated and cerebral malaria subjects
+### Fig 3. Increased cerebral_hb_tot and decreased cerebral_hb_tot_alpha are distinct clinical features of cerebral malaria
 
@@
 
-# include variables of interest
+# select model variables of interest
+model_voi <- c("subject_id", "status", 
+               "hct", "lactate",
+               "cerebral_hb_tot", "cerebral_hb_oxy",
+               "cerebral_hb_tot_alpha2", "cerebral_hb_oxy_alpha2")
 
-df_hbox_log_reg <- df_hbox %>% 
-  dplyr::select(c("subject_id", "Status", "Hematocrit", "Lactate", "avg_Hb_conc", "Hb_conc_second_alpha", "avg_Hb_o2sat", "Hb_o2sat_second_alpha"))
+df_model <- df_hbox %>% 
+  select_at(model_voi) %>%
+
+  # remove HC
+  ungroup %>% 
+  filter(status != "HC")
 
@@
 
-  df_hbox_impute <- df_hbox_log_reg %>% 
-    group_by(Status) %>% 
+  df_model_impute <- df_model %>% 
+    group_by(status) %>% 
     mutate_at(3:ncol(.), f_impute) %>% 
@@
 
-f_calc_cor <- function(df, status) {
+f_calc_cor <- function(df, group) {
 
   m <- df %>%
-    dplyr::filter(Status == status) %>% 
-    dplyr::select(-c(subject_id, Status)) %>% 
+    dplyr::filter(status == group) %>% 
+    dplyr::select(-c(subject_id, status)) %>% 
     cor(method = "spearman")
@@
 # create run correlations of potential regressors
-m_cor_um <- df_hbox_impute %>% f_calc_cor(status = "UM")
+m_cor_um <- df_model_impute %>% f_calc_cor(group = "UM")
 
-m_cor_cm <- df_hbox_impute %>% f_calc_cor(status = "CM")
+m_cor_cm <- df_model_impute %>% f_calc_cor(group = "CM")
 
@@
 
-    mutate_if(is.factor, factor, levels = c("Lactate", "Hematocrit", "avg_Hb_o2sat", "avg_Hb_conc", "Hb_o2sat_second_alpha", "Hb_conc_second_alpha")) %>%
+    mutate_if(is.factor, factor, levels = c("hct", "lactate", "cerebral_hb_oxy", "cerebral_hb_tot", "cerebral_hb_oxy_alpha2", "cerebral_hb_tot_alpha2")) %>%
 
@@
 
+# plot
+
   # extract legend
@@
 
-  as_ggplot(p_leg)
+  # as_ggplot(p_leg)
 
@@
 ```{r fig3b}
-# remove healthy volunteers for logistic regression analysis
-df_hbox_impute <- df_hbox_impute %>% dplyr::filter(Status != "HC")
 
 # convert tibble to df
-df_hbox_df <- df_hbox_impute %>% 
-  dplyr::select(-c(subject_id, Status)) %>% 
+df_model_pca <- df_model_impute %>% 
+  dplyr::select(-c(subject_id, status)) %>% 
   as.data.frame()
@@
 # add rownames
-rownames(df_hbox_df) <- df_hbox_impute %>% .$subject_id
+rownames(df_model_pca) <- df_model_impute$subject_id
 
 # compute PCA
-my_pca <- prcomp(df_hbox_df, scale = TRUE)
+my_pca <- prcomp(df_model_pca, scale = TRUE)
 
 # define groups
-groups <- as.factor(df_hbox_impute$Status)
+groups <- as.factor(df_model_impute$status)
 
@@
 df_vars <- my_pca$rotation %>% as_tibble() %>% mutate(variable = rownames(my_pca$rotation))
-df_pnts <- my_pca$x %>% as_tibble() %>% mutate(group = groups)
+df_pnts <- my_pca$x %>% 
+  as_tibble() %>% 
+  mutate(status = groups, 
+         subject_id = df_model_impute$subject_id)
 
@@
 ggplot() +
-  geom_point(data = df_pnts, mapping = aes(x = PC1, y = PC2, color = group), size = 3) +
-  stat_conf_ellipse(data = df_pnts, mapping = aes(x = PC1, y = PC2, color = group, fill = group),
+  geom_point(data = df_pnts, mapping = aes(x = PC1, y = PC2, color = status), size = 3) +
+  stat_conf_ellipse(data = df_pnts, mapping = aes(x = PC1, y = PC2, color = status, fill = status),
                     level = 0.95, npoint = 100, bary = TRUE, alpha = 0.1, geom = "polygon") +
-  stat_mean(data = df_pnts, mapping = aes(x = PC1, y = PC2, color = group, shape = group),
-            na.rm = FALSE) +
+  # stat_mean(data = df_pnts, mapping = aes(x = PC1, y = PC2, color = status, shape = status),
+  #           na.rm = FALSE) +
+
+  # View patient IDs
+  # geom_text_repel(data = df_pnts, aes(x = PC1, y = PC2, label = subject_id, size = 10), show.legend = FALSE) +
 
@@
   # geom_point(data = df_vars, aes(x = PC1*5, y = PC2*5)) +
-  # geom_text_repel(data = df_vars, aes(x = PC1*5, y = PC2*5, label = variable, size = 10), show.legend = FALSE) +
+  geom_text_repel(data = df_vars, aes(x = PC1*5, y = PC2*5, label = variable, size = 10), show.legend = FALSE) +
   geom_segment(data = df_vars, aes(x = 0, y = 0, xend = PC1*5, yend = PC2*5), arrow = arrow()) +
@@
 #### C. Coefficients
-```{r table2c}
 
-df_CI <- confint(hbox_final$fit) %>% exp() %>% as_tibble(rownames = "term")
+```{r model, eval = FALSE}
 
-table2c <- hbox_final %>%
-  tidy(exponentiate = TRUE) %>%
-  left_join(df_CI, by = "term") %>%
-  mutate(estimate = ifelse(estimate < 10^-2, scientific(estimate, digits = 3), round(estimate, 2)),
-         conf.low = ifelse(`2.5 %` < 10^-2, scientific(`2.5 %`, digits = 3), round(`2.5 %`, 2)),
-         conf.high = ifelse(`97.5 %` < 10^-2, scientific(`97.5 %`, digits = 3), round(`97.5 %`, 2))) %>%
-  mutate(p.value = scientific(p.value, digits = 3),
-         statistic = round(statistic, 2)) %>%
-  unite("95% CI", c(conf.low, conf.high), sep = ", ") %>%
-  dplyr::select(c(term, estimate, `95% CI`, statistic, p.value)) %>% 
-  dplyr::rename("Term" = "term", "Odds ratio" = "estimate",
-                "t-statistic" = "statistic", "p-value" = "p.value") %>%
-  as.data.frame()
-
-table2c %>% knitr::kable(align = rep('c', 5))
-
-```
-
-#### D. ROC curve of final model
-
-```{r fig3d_prep, eval = FALSE}
-
-## write function to create model on given df
-
-f_log_reg_model <- function(df_model) {
+### shows how logistic regression model was set up and run, results saved in objects/glm_model ###
 
 # resample
-  hbox_boot <- bootstraps(df_model, strata = Status, times = 1000)
+set.seed(234)
+hbox_boot <- bootstraps(df_model, strata = Status, times = 5000)
 
@@
 hbox_rec <- recipe(Status ~ ., data = df_model) %>% 
-    step_BoxCox(all_predictors()) %>% 
-    step_nzv(all_predictors()) %>% 
-    step_smote(Status) # upsample to deal with unbalanced sample sizes
+  step_impute_knn(all_predictors()) %>% 
+  step_normalize(all_predictors()) %>%
+  step_zv(all_predictors()) %>%
+  step_smote(Status) %>% 
+  update_role(subject_id, new_role = "id")
 
@@
 
-  # create model on resamples
+# create model on 5000 different resamples
 glm_res <- hbox_wf %>% 
@@
   fit(df_model) %>%
-    pull_workflow_fit()
-
-  # view coefficients & p-values
-  model_coefs <- hbox_final %>% tidy()
-
-  # return list including model and a table of best model variable statistics
-  list(glm_res, model_coefs)
-
-}
-
-## write function to collect predictions for ROC plot
-
-f_collect_preds <- function(glm_res) {
-
-      glm_res %>% 
-      collect_predictions() %>% 
-      roc_curve(Status, .pred_CM)
-
-}
-
-## Perform functions on all iterations of model
-
-set.seed(123)
-
-    # select starting variables
-    df_hbox_model1 <- df_hbox_log_reg %>% 
-      dplyr::select(c(Status, Hematocrit, Lactate, avg_Hb_o2sat, Hb_o2sat_second_alpha, avg_Hb_conc, Hb_conc_second_alpha)) %>% 
-      mutate(Status = factor(Status, levels = c("CM", "UM")))
-
-    # model 1: Status ~ Hematocrit + Lactate + avg_Hb_o2sat + Hb_o2sat_second_alpha + avg_Hb_conc + Hb_conc_second_alpha
-    model1 <- f_log_reg_model(df_hbox_model1)
-    roc1 <- f_collect_preds(model1[[1]]) %>% mutate(model = "1")
-    model1[[2]] # remove Hb_o2sat_second_alpha (highest p-val)
-
-    # model 2: Status ~ Hematocrit + Lactate + avg_Hb_o2sat + avg_Hb_conc + Hb_conc_second_alpha
-    df_hbox_model2 <-  df_hbox_model1 %>% dplyr::select(-Hb_o2sat_second_alpha)
-    model2 <- f_log_reg_model(df_hbox_model2)
-    roc2 <- f_collect_preds(model2[[1]]) %>% mutate(model = "2")
-    model2[[2]] # remove Lactate
-
-    # model 3: Status ~ Hematocrit + avg_Hb_o2sat + avg_Hb_conc + Hb_conc_second_alpha
-    df_hbox_model3 <-  df_hbox_model2 %>% dplyr::select(-Lactate)
-    model3 <- f_log_reg_model(df_hbox_model3)
-    roc3 <- f_collect_preds(model3[[1]]) %>% mutate(model = "3")
-    model3[[2]] # remove avg_Hb_o2sat
-
-    # model 4: Status ~ Hematocrit + avg_Hb_conc + Hb_conc_second_alpha
-    df_hbox_model4 <-  df_hbox_model3 %>% dplyr::select(-avg_Hb_o2sat)
-    model4 <- f_log_reg_model(df_hbox_model4)
-    roc4 <- f_collect_preds(model4[[1]]) %>% mutate(model = "4")
-    model4[[2]] # none to remove, final model
-
-save(roc1, roc2, roc3, file = "roc_curves.Rdata")
-
-# calculate areas under curves
-
-library(MESS)
-roc1 <- roc1 %>% mutate(false_pos = 1 - specificity)
-auc(roc1$false_pos, roc1$sensitivity)
-
-roc2 <- roc2 %>% mutate(false_pos = 1 - specificity)
-auc(roc2$false_pos, roc2$sensitivity)
-
-roc3 <- roc3 %>% mutate(false_pos = 1 - specificity)
-auc(roc3$false_pos, roc3$sensitivity)
-
-roc_final <- glm_res %>% 
-  collect_predictions() %>% 
-  roc_curve(Status, .pred_0) %>% 
-  mutate(false_pos = 1 - specificity)
-auc(roc_final$false_pos, roc_final$sensitivity)
+  extract_fit_parsnip()
 
 ```
+
+```{r table3c}
+
+df_CI <- confint(hbox_final$fit) %>% exp() %>% as_tibble(rownames = "term")
+
+table3c <- hbox_final %>%
+  tidy(exponentiate = TRUE) %>%
+  filter(p.value < 0.05) %>% 
+  left_join(df_CI, by = "term") %>%
+  mutate(estimate = ifelse(estimate < 10^-2, scientific(estimate, digits = 3), round(estimate, 2)),
+         conf.low = ifelse(`2.5 %` < 10^-2, scientific(`2.5 %`, digits = 3), round(`2.5 %`, 2)),
+         conf.high = ifelse(`97.5 %` < 10^-2, scientific(`97.5 %`, digits = 3), round(`97.5 %`, 2))) %>%
+  mutate(p.value = scientific(p.value, digits = 3),
+         statistic = round(statistic, 2)) %>%
+  unite("95% CI", c(conf.low, conf.high), sep = ", ") %>%
+  dplyr::select(c(term, estimate, `95% CI`, statistic, p.value)) %>% 
+  dplyr::rename("Term" = "term", "Odds ratio" = "estimate",
+                "t-statistic" = "statistic", "p-value" = "p.value") %>%
+  as.data.frame()
+
+table3c %>% knitr::kable(align = rep('c', 5))
+
+```
+
+#### D. ROC curve of final model
 
@@
   collect_predictions() %>% 
-  roc_curve(Status, .pred_0) %>% 
-  mutate(model = "final") %>% 
-  bind_rows(roc1, roc2, roc3)
+  roc_curve(status, .pred_CM)
 
 ggplot() +
-  geom_path(data = df_roc %>% filter(model == "final"), 
-            aes(x = 1 - specificity, y = sensitivity, color = model), 
+  geom_path(data = df_roc, 
+            aes(x = 1 - specificity, y = sensitivity), 
             size = 2.5) +
@@
 
-fig3d_inset <- collect_metrics(glm_res) %>% 
+fig3d_inset <- glm_res %>% 
+  collect_metrics() %>% 
   dplyr::select(.metric, mean, std_err) %>% 
@@
 
-### Fig 4. Explanatory model of how to interpret hemoglobin concentration
+### Fig 4. An increase in cerebral_hb_tot is significantly associated with an increase in brain swelling score
+
+```{r fig4_plot}
+
+# select model voi
+brain_swell_voi <- c("brain_swell", "cerebral_hb_tot", "hct")
+df_fig4 <- df_hbox %>% select_at(brain_swell_voi)
+
+# remove cerebral_hb_tot outliers
+  df_fig4_outliers <- df_fig4 %>% 
+    mutate(Q1_hbtot = quantile(cerebral_hb_tot, 0.25, na.rm = TRUE),
+           Q3_hbtot = quantile(cerebral_hb_tot, 0.75, na.rm = TRUE),
+           IQR_hbtot = Q3_hbtot - Q1_hbtot,
+           outlier_hbtot = case_when(
+
+             cerebral_hb_tot < Q1_hbtot - IQR_hbtot ~ 1,
+             cerebral_hb_tot > Q3_hbtot + IQR_hbtot ~ 1,
+             TRUE ~ 0
+
+           )
+
+    ) %>% 
+    filter(outlier_hbtot != 1)
+
+# model
+  lm_fig4 <- lm(brain_swell ~ cerebral_hb_tot + hct, data = df_fig4_outliers)
+  f <- summary(lm_fig4)$fstatistic
+  p <- pf(f[1], f[2], f[3], lower.tail = FALSE)
+
+  model <- "model: brain_swell ~ cerebral_hb_tot + hematocrit"
+  r2 <- paste0("adjusted R^2: ", lm_fig4 %>% summary() %>% .$adj.r.squared %>% round(2))
+  model_p_val <- paste0("model: ", round(p, 3))
+  hb_tot_p_val <- paste0("cerebral_hb_tot: ", lm_fig4 %>% tidy() %>% .[2, "p.value"] %>% .$p.value %>% round(3))
+
+# plot variables
+require(ggpmisc)
+
+df_fig4_outliers %>% 
+  ggplot(aes(x = cerebral_hb_tot, y = brain_swell)) +
+  geom_point(shape = 1, aes(size = hct, color = as.factor(brain_swell))) +
+  #geom_abline(intercept = ) +
+
+  geom_smooth(method = "lm", lty = 2, color = "#696969", alpha = 0.2) +
+  annotate("text", label = model, x = 25, y = 8, hjust = 0, vjust = 1, size = 5) +
+  annotate("text", label = r2, x = 25, y = 7.5, hjust = 0, vjust = 1, size = 5) +
+  annotate("text", label = "p-values", x = 25, y = 7, hjust = 0, vjust = 1, size = 5) +
+  annotate("text", label = model_p_val, x = 35, y = 6.75, hjust = 0, vjust = 1, size = 4) +
+  annotate("text", label = hb_tot_p_val, x = 35, y = 6.5, hjust = 0, vjust = 1, size = 4) +
+
+  lims(x = c(25, 225)) +
+  theme_bw() +
+
+  theme(# text = element_text(family = "Arial"),
+        strip.text = element_text(size = 15),
+        axis.title = element_text(size = 15),
+        axis.title.y = element_text(size = 15),
+        axis.text = element_text(size = 15),
+        legend.title = element_text(size = 12),
+        legend.text = element_text(size = 10))
+
+# 3d plot?  
+# require(plotly)
+#   x <- df_fig4_outliers$cerebral_hb_tot
+#   y <- df_fig4_outliers$hematocrit
+#   z <- df_fig4_outliers$brain_swell
+#   
+#   df_fig4_outliers %>% plot_ly(x = x, y = y, z = z, type = "scatter3d", mode = "markers", color = x)
+
+```
+
+### Fig 5. Explanatory model of how to interpret hemoglobin concentration
 
@@
 
-#### A. Pre-processed signal
-```{r fig_s3a}
+#### A. Raw signal
+```{r fig_s1a}
 
-df_thc_cumsum1 %>%
+df_figs1a <- read.table("Data/signal_segments/Hb_tot/TM0001CM01.txt")
 
-  ggplot(aes(x = Time, y = sg_filt)) +
-  geom_line() +
+df_figs1a %>%
+
+  ggplot(aes(x = Time)) +
+  geom_line(aes(y = THC)) +
 
   labs(y = "Hemoglobin concentration (μM)", x = "Time (s)") +
-  ggtitle("Pre-processed signal") +
+  ggtitle("Raw signal") +
   theme_bw() +
@@
 
-#### B. Cumulative sum
-```{r fig_s3b}
+#### B. Processed signal
+```{r fig_s1b}
 
-df_thc_cumsum1 %>%
+df_figs1b <- read.table("Data/filtered_signals/0.2s_mean_0.001_1_filt/Hb_tot/TM0001CM01.txt") %>% mutate(time = row_number())
 
-  ggplot(aes(x = Time, y = cumsum)) +
-  geom_line() +
-  geom_vline(xintercept = 1201, lty = 2, color = "black") +
-  geom_vline(xintercept = 1210, lty = 2, color = "black") +
+df_figs1b %>% 
+  ggplot(aes(x = time)) +
+  geom_line(aes(y = V1)) +
 
-  labs(y = "Cumulative sum", x = "Time (s)") +
-  ggtitle("Cumulative signal") +
+  lims(y = c(-0.001, 0.001)) +
+
+  labs(y = "Bandpass filtered signal", x = "Time") +
+  ggtitle("Pre-processed signal") +
   theme_bw() +
   theme(# text = element_text(family = "Arial"),
-        axis.text = element_text(size = 20),
+        axis.text = element_blank(),
         axis.title = element_text(size = 25),
@@
 
-#### C. Window detrend
-```{r fig_s3c, fig.show = "hold", out.width = "50%"}
+#### C. Cumulative sum
+```{r fig_s1c}
 
-fit <- lm(cumsum ~ Time, data = df_thc_cumsum1 %>% dplyr::filter(Time > 1201 & Time < 1210))
+df_figs1c <- df_figs1b %>% 
+  mutate(mean = mean(V1),
+         subtract_mean = V1 - mean,
+         cumsum = cumsum(subtract_mean))
+
+df_figs1c %>%
+
+  ggplot(aes(x = time, y = cumsum)) +
+  geom_line() +
+  geom_vline(xintercept = 1, lty = 2, color = "black") +
+  geom_vline(xintercept = 1000, lty = 2, color = "black") +
+
+  labs(y = "Cumulative sum", x = "Time") +
+  ggtitle("Cumulative signal") +
+  theme_bw() +
+  theme(# text = element_text(family = "Arial"),
+        axis.text = element_blank(),
+        axis.title = element_text(size = 25),
+        legend.text = element_text(size = 15),
+        legend.title = element_text(size = 20))
+
+```
+
+#### D. Window detrend
+```{r fig_s1d, fig.show = "hold", out.width = "50%"}
+
+fit <- lm(cumsum ~ time, data = df_figs1c %>% dplyr::filter(time > 1 & time < 1000))
 
 # Window 
-df_thc_cumsum1 %>% dplyr::filter(Time > 1201 & Time < 1210) %>%
+df_figs1c %>% dplyr::filter(time > 1 & time < 1000) %>%
 
-  ggplot(aes(x = Time, y = cumsum)) +
+  ggplot(aes(x = time, y = cumsum)) +
   geom_line(size = 2) +
@@
 
-  labs(y = "Cumulative sum", x = "Time (s)") +
+  labs(y = "Cumulative sum", x = "Time") +
   ggtitle("Cumulative signal") +
@@
   theme(# text = element_text(family = "Arial"),
-        axis.text = element_text(size = 20),
+        axis.text.x = element_text(size = 20),
+        axis.text.y = element_blank(),
         axis.title = element_text(size = 25),
@@
 # Detrended
-df_thc_cumsum1 %>% dplyr::filter(Time > 1201 & Time < 1210) %>%
+df_figs1c %>% dplyr::filter(time > 1 & time < 1000) %>%
 
-  ggplot(aes(x = Time, y = fit$residuals)) +
+  ggplot(aes(x = time, y = fit$residuals)) +
   geom_line(size = 2) +
@@
 
-  labs(y = "", x = "Time (s)") +
+  labs(y = "", x = "Time") +
   ggtitle("Detrended cumulative signal") +
@@
   theme(# text = element_text(family = "Arial"),
-        axis.text = element_text(size = 25),
+        axis.text.x = element_text(size = 20),
+        axis.text.y = element_blank(),
         axis.title = element_text(size = 28),
@@
 
-#### D. Results plot
-```{r fig_s3d}
+#### E. Results plot
+```{r fig_s1e}
+
+df_figs1e <- l_dfa_hbtot[[1]][[1]]
 
@@
 library(segmented)
-df1 <- l_dfa_thc[[1]][[1]] %>% mutate(log10_avg_fluct = log10(avg_fluctuation), log10_window_size = log10(window_size))
-fit1 <- lm(log10_avg_fluct ~ log10_window_size, data = df1) 
-segfit1 <- segmented(fit1)
-slopes1 <- coef(segfit1) # the coefficients are the differences in slope in comparison to the previous slope
+df_figs1e <- df_figs1e %>% mutate(log10_avg_fluct = log10(avg_fluctuation), log10_window_size = log10(window_size))
+fit_figs1e <- lm(log10_avg_fluct ~ log10_window_size, data = df_figs1e) 
+segfit <- segmented(fit_figs1e)
+slopes <- coef(segfit) # the coefficients are the differences in slope in comparison to the previous slope
 
@@
 # At the breakpoint (break1), the segments b and c intersect: b0 + b1*x = c0 + c1*x
-b0 <- slopes1[[1]]
-b1 <- slopes1[[2]]
+b0 <- slopes[[1]]
+b1 <- slopes[[2]]
 
-c1 <- slopes1[[2]] + slopes1[[3]]
-break1 <- segfit1$psi[[2]]
+c1 <- slopes[[2]] + slopes[[3]]
+break1 <- segfit$psi[[2]]
 
@@
 
-df1 %>%
+df_figs1e %>%
 
@@
   geom_abline(intercept = c0, slope = c1, color = "black", lty = 2) +
-  annotate("text", x = -0.5, y = 2, label = paste0("alpha = ", round(c1, 2)), size = 8) +
+  annotate("text", x = -1, y = -2, label = paste0("alpha = ", round(c1, 2)), size = 8, vjust = 1, hjust = 0) +
 
@@
 
+#### F. Simulation results
+
+```{r fig_s1f}
+
+wn_alpha <- 1:length(l_wn_dfa) %>% purrr::map_df(~l_wn_dfa[[.x]][[2]]) %>% dplyr::rename("white" = "log10(window_size)")
+pn_alpha <- 1:length(l_pn_dfa) %>% purrr::map_df(~l_pn_dfa[[.x]][[2]]) %>% dplyr::rename("pink" = "log10(window_size)")
+bn_alpha <- 1:length(l_bn_dfa) %>% purrr::map_df(~l_bn_dfa[[.x]][[2]]) %>% dplyr::rename("brown" = "log10(window_size)")
+
+df_figs1f <- bind_cols(wn_alpha, pn_alpha, bn_alpha) %>%
+  pivot_longer(1:3, names_to = "noise", values_to = "alpha")
+
+# create color vector
+my_col <- c("white" = "#00BFC4", "brown" = "#7CAE00", "pink" = "#F8766D")
+
+# plot
+df_figs1f %>%
+
+  ggplot(aes(x = noise, y = alpha)) +
+  geom_violin(aes(color = noise)) +
+  geom_jitter(position = position_jitter(0.2), shape = 1, size = 2) +
+  stat_summary(fun = "median", geom = "crossbar", aes(color = noise), size = 0.2, width = 0.5) +
+
+  geom_hline(yintercept = 0.4, lty = 2, alpha = 0.7) +
+  geom_hline(yintercept = 0.6, lty = 2, alpha = 0.7) +
+  geom_hline(yintercept = 1.0, lty = 2, alpha = 0.7) +
+  geom_hline(yintercept = 1.5, lty = 2, alpha = 0.7) +
+
+  scale_color_manual(values = my_col) +
+  labs(x = "noise type") +
+  scale_y_continuous(limits = c(0, 1.7)) +
+  theme_bw() +
+  theme(legend.position = "none")  +
+
+  theme(# text = element_text(family = "Arial"),
+        strip.text = element_text(size = 15),
+        axis.title = element_text(size = 15),
+        axis.text = element_text(size = 15))
+
+```
+
 ### SFig 2. Logistic regression curves
@@
 # pivot longer and change factor levels
-df_hbox_reg_long <- df_hbox_impute %>% 
-  mutate(Status = as.numeric(ifelse(Status == "CM", 1, 0))) %>% 
+df_figs2 <- df_model %>% 
+  mutate(status = as.numeric(ifelse(status == "CM", 1, 0))) %>% 
   pivot_longer(3:ncol(.), names_to = "variable", values_to = "value") %>% 
-  mutate(variable = factor(variable, levels = c("Lactate", "Hematocrit", "avg_Hb_o2sat", "avg_Hb_conc", "Hb_o2sat_second_alpha", "Hb_conc_second_alpha")))
+  mutate(variable = factor(variable, levels = c("hct", "lactate", "cerebral_hb_oxy", "cerebral_hb_tot", "cerebral_hb_oxy_alpha2", "cerebral_hb_tot_alpha2")))
 
 # plot as individual regression curves
-df_hbox_reg_long %>%
+df_figs2 %>%
 
-  ggplot(aes(x = value, y = Status)) +
+  ggplot(aes(x = value, y = status)) +
   geom_point() +
-  geom_smooth(method = "glm", method.args = list(family = "binomial")) +
+  geom_smooth(method = "glm", method.args = list(family = "binomial"), color = "black") +
   facet_wrap(~variable, scales = "free_x", nrow = 5, ncol = 2) +
@@
 A steeper curve indicates higher discriminatory power.
+
+## Table of manuscript p-values
+
+```{r manuscript_stats, eval = FALSE}
+
+# select voi
+df_stats <- df_hbox %>% 
+  select(c("status",
+
+           # clinical
+           "age_calc", "temperature", "hr", "bp_sys", "bp_dias", "resp_rate", "pulse_ox", "hct", "lactate", "glucose", "aa_arg",
+
+           # concentrations
+           matches("_hb_tot$|_hb_oxy$"),
+
+           # alphas
+           matches("alpha2")
+
+  ))
+
+# perform kruskal-wallis test on all variables
+df_pvals <- 2:ncol(df_stats) %>% 
+  purrr::map_dfr(~tibble(
+
+    variable = names(df_stats[.x]),
+    kw_pval = kruskal.test(df_stats[[.x]] ~ df_stats$status)$p.value,
+    um_hc = DunnTest(df_stats[[.x]] ~ df_stats$status)[[1]][1, 2],
+    cm_hc = DunnTest(df_stats[[.x]] ~ df_stats$status)[[1]][2, 2],
+    cm_um = DunnTest(df_stats[[.x]] ~ df_stats$status)[[1]][3, 2]
+
+  ))
+
+df_pvals %>% knitr::kable()
+
+# write to csv
+write.csv(df_pvals, file = "Outputs/manuscript_pvals.csv")
+
+```
+
+```{r export_xlsx, eval = FALSE}
+
+require(xlsx)
+
+wb <- createWorkbook()
+
+sheet = createSheet(wb, "Table 1a")
+
+  addDataFrame(df_clinical_res %>% as.data.frame(), sheet = sheet, startColumn = 1, row.names = FALSE)
+
+sheet = createSheet(wb, "Table 1b")
+
+  addDataFrame(df_cm_voi %>% count(bcs_session) %>% as.data.frame(), sheet = sheet, startColumn = 1, row.names = FALSE)
+  addDataFrame(df_cm_voi %>% count(brain_swell) %>% as.data.frame(), sheet = sheet, startColumn = 4, row.names = FALSE)
+  addDataFrame(df_cm_hrp2 %>% as.data.frame(), sheet = sheet, startColumn = 7, row.names = FALSE)
+
+sheet = createSheet(wb, "Table 3c & d")
+
+  addDataFrame(table3c %>% as.data.frame(), sheet = sheet, startColumn = 1, row.names = FALSE)
+  addDataFrame(fig3d_inset %>% as.data.frame(), sheet = sheet, startColumn = 7, row.names = FALSE)
+
+sheet = createSheet(wb, "manuscript p-vals")
+
+  addDataFrame(df_pvals %>% as.data.frame(), sheet = sheet, startColumn = 1, row.names = FALSE)
+
+saveWorkbook(wb, "Outputs/all_manuscript_tables.xlsx")
+
+```

### 2022-01-10 Scripts/combining_data.R

@@ -73,8 +73,6 @@
         rbindlist() %>% as_tibble()
       
       
-      
-
 
 # format and combine data -------------------------------------------------------------
 
@@ -94,22 +92,22 @@
         select(-c(contains("alpha"), cerebral_hb_oxy, cerebral_hb_tot)) %>% 
         
         # add muscle alphas & medians
-        right_join(df_bp_alphas_muscle %>% 
+        full_join(df_bp_alphas_muscle %>% 
                      select(-c(X, contains("breakpoint"))) %>% 
-                     left_join(df_med_hboxy_muscle, by = "subject_id") %>% 
-                     left_join(df_med_hbtot_muscle, by = "subject_id"), 
+                    full_join(df_med_hboxy_muscle, by = "subject_id") %>% 
+                    full_join(df_med_hbtot_muscle, by = "subject_id"), 
                    by = c("subject_id", "Status")) %>% 
         
         # add brain alphas & medians
-        left_join(df_bp_alphas_brain %>% 
+        full_join(df_bp_alphas_brain %>% 
                     select(-c(X, contains("breakpoint"))) %>% 
                     rename_with(~paste0("cerebral_", .x), 3:8) %>% 
-                    left_join(df_med_hboxy_brain, by = "subject_id") %>% 
-                    left_join(df_med_hbtot_brain, by = "subject_id"), 
+                    full_join(df_med_hboxy_brain, by = "subject_id") %>% 
+                    full_join(df_med_hbtot_brain, by = "subject_id"), 
                   by = c("subject_id", "Status")) %>% 
         
         # replace missing UM values
-        left_join(df_bp_alphas_brain_missing_um, by = c("subject_id", "Status")) %>%
+        full_join(df_bp_alphas_brain_missing_um, by = c("subject_id", "Status")) %>%
         mutate(cerebral_hbtot_overall_a = ifelse(subject_id %in% subjs, cerebral_hbtot_overall_a.y, cerebral_hbtot_overall_a.x),
                cerebral_hbtot_short_a = ifelse(subject_id %in% subjs, cerebral_hbtot_short_a.y, cerebral_hbtot_short_a.x),
                cerebral_hbtot_long_a = ifelse(subject_id %in% subjs, cerebral_hbtot_long_a.y, cerebral_hbtot_long_a.x),
@@ -119,6 +117,12 @@
[…]

### 2022-01-10 Scripts/tidymodels_analysis.R

@@ -14,6 +14,7 @@
         library(usemodels)
         library(ranger)
         library(vip)
+        library(janitor)
 
 
 
@@ -222,299 +223,93 @@
         
 # Logistic regression model --------------------------------------------------------------
 
+        
+      
+        df_model <- read_csv("Data/Hans_datatable_exports/malawi key data v04Jan2022.csv") %>% 
+          as_tibble() %>% 
+          janitor::clean_names() %>% 
+          dplyr::rename("subject_id" = "subject_id_session") %>% 
+          
+          dplyr::filter(status %in% c("CM", "UM") & session_no == 1 | subject_id == "TM0003CM01") %>% 
+          dplyr::filter(!grepl("blood", subject_id)) %>% 
+          mutate(status = factor(status, levels = c("CM", "UM"))) %>% 
+          select(subject_id, status, hct, lactate, cerebral_hb_tot, cerebral_hb_tot_alpha2, cerebral_hb_oxy, cerebral_hb_oxy_alpha2) %>% 
+          mutate_at(3:8, as.numeric)
+        
+        # look at missing data
+        df_model %>% is.na %>% colSums
+        df_model %>% filter(is.na(cerebral_hb_oxy_alpha2))
+        
+        require(naniar)
+        df_model %>% select(-c(subject_id, status)) %>% gg_miss_upset() # 12 pts missing all 4 NIRS voi
+        df_model %>% count(status) 
+        
+        # remove patients with all 4 NIRS variables missing
+        df_model <- df_model %>% drop_na(contains("hb_tot"))
+        df_model %>% select(-c(subject_id, status)) %>% gg_miss_upset() # 3 patients missing hb_oxy, 3 missing hb_oxy_alpha, 3 missing both, 2 missing lactate, 1 missing hct & lactate
 
-        # Data
-        df_model <- df_master_voi %>% select(-subject_id)
+        # add 1/10 alpha as variable
+        df_model <- df_model %>% 
+          mutate(one_tenth_cerebral_hb_tot_alpha = 1/10 * cerebral_hb_tot_alpha2,
+                 one_tenth_cerebral_hb_oxy_alpha = 1/10 * cerebral_hb_oxy_alpha2) %>% 
+          select(-c(cerebral_hb_tot_alpha2, cerebral_hb_oxy_alpha2))
         
-        ## write function to create model on given df
+        # resample
+        set.seed(234)
+        hbox_boot <- bootstraps(df_model, strata = status, times = 5000)
         
[…]

### 2022-02-09 final_figures.Rmd
add NIRS median, q1, q3 to manuscript tables file

@@
 
+```{r fig_1_and_2_stats, eval = FALSE}
+
+df_fig1_fig2_stats <-
+
+  df_fig1 %>% 
+    group_by(variable, status) %>% 
+    dplyr::select(value) %>% 
+    summarise_all(funs(median = median, q1 = quantile(., probs = 0.25), q3 = quantile(., probs = 0.75)), na.rm = TRUE) %>% 
+    mutate_if(is.numeric, round, 2) %>%
+
+  bind_rows(
+
+    df_fig2 %>% 
+      group_by(variable, status) %>% 
+      dplyr::select(value) %>% 
+      summarise_all(funs(median = median, q1 = quantile(., probs = 0.25), q3 = quantile(., probs = 0.75)), na.rm = TRUE) %>% 
+      mutate_if(is.numeric, round, 2)
+
+  )
+
+```
+
 ```{r manuscript_stats, eval = FALSE}
@@
 
+sheet = createSheet(wb, "Figs 1 & 2 stats")
+
+  addDataFrame(df_fig1_fig2_stats %>% as.data.frame(), sheet = sheet, startColumn = 1, row.names = FALSE)
+
 sheet = createSheet(wb, "Table 3c & d")

### 2022-02-09 Scripts/old_analyses_versions/penalized_logistic_regression.R
commit earlier changes

@@ -1,5 +1,7 @@
 
 
+### not used in final manuscript because we want to be able to evaluate coefficients
+
 
 # libraries ---------------------------------------------------------------
 
@@ -19,78 +21,60 @@
   df_model <- read.csv("Data/master_datatable.csv") %>% 
     as_tibble() %>% 
     select(c(subject_id, Status, Hematocrit, Lactate,
-             cerebral_hb_tot, cerebral_hbtot_long_a, cerebral_hb_oxy, cerebral_hboxy_long_a)) %>% 
-    mutate(Status = factor(Status, c("HC", "UM", "CM")))
+             cerebral_hb_tot, cerebral_hbtot_long_a, cerebral_hb_oxy, cerebral_hboxy_long_a))
 
 
-# Train model -------------------------------------------------------------
+# Build model -------------------------------------------------------------
 
    
-  df_model <- df_model %>% filter(Status != "HC")
+  # Remove HC and look at missing data
+  df_model <- df_model %>% filter(Status != "HC") %>% mutate(Status = factor(Status, levels = c("CM", "UM")))
   df_model %>% is.na %>% colSums
   df_model %>% filter(is.na(cerebral_hboxy_long_a))
   
-  # look at missing data
   require(naniar)
-  df_model %>% select(-c(subject_id, Status)) %>% gg_miss_upset() 
-  df_model %>% count(Status) # 11 out of 24 UM measurements don't have any good recordings
+  df_model %>% select(-c(subject_id, Status)) %>% gg_miss_upset() # one patient is missing Hct and Lactate, 3 patients missing Hb_oxy alpha
+  df_model %>% count(Status) 
+  
   
     # create recipe
     my_rec <- recipe(Status ~ ., data = df_model) %>% 
       step_impute_knn(all_predictors()) %>% 
       step_normalize(all_predictors()) %>%
       step_zv(all_predictors()) %>%
-      step_upsample(all_outcomes()) %>% 
+      step_smote(Status) %>% 
       update_role(subject_id, new_role = "id")
     
-    # prep
+    # prep recipe
     my_prep <- my_rec %>% prep(strings_as_factors = FALSE)
     
-
     # model specification
-    lasso_spec <- logistic_reg(penalty = 0.1, mixture = 1) %>% # picking specific penalty (tune in next section) for LASSO regression
+    lasso_spec <- logistic_reg(penalty = tune(), mixture = 1) %>%
       set_engine("glmnet")
     
     # create workflow
[…]

### 2022-02-09 final_figures.Rmd
add NIRS stats to overall table
median, q1, q3, for NIRS variables in brain and tissue

--- 2022-01-10 final_figures.Rmd
@@
 
+```{r fig_1_and_2_stats, eval = FALSE}
+
+df_fig1_fig2_stats <-
+
+  df_fig1 %>% 
+    group_by(tissue, variable, status) %>% 
+    dplyr::select(value) %>% 
+    summarise_all(funs(median = median, 
+                       q1 = quantile(., probs = 0.25), 
+                       q3 = quantile(., probs = 0.75)), na.rm = TRUE) %>% 
+    mutate_if(is.numeric, round, 2) %>%
+
+  bind_rows(
+
+    df_fig2 %>% 
+      group_by(tissue, variable, status) %>% 
+      dplyr::select(value) %>% 
+      summarise_all(funs(median = median, 
+                         q1 = quantile(., probs = 0.25), 
+                         q3 = quantile(., probs = 0.75)), na.rm = TRUE) %>% 
+      mutate_if(is.numeric, round, 2)
+
+  )
+
+```
+
 ```{r manuscript_stats, eval = FALSE}

### 2022-02-14 final_figures.Rmd
export to pdf, run additional brain_swell models

@@
 date: "`r Sys.Date()`"
-output: html_document
+output: pdf_document
 ---
@@
 
-knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, comment = "", fig.height = 6, fig.width = 10)
+knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, comment = "", fig.height = 6, fig.width = 10, 
+                      fig.align = 'center')
 
@@
 ### Fig 1. Hb_oxy and Hb_tot in the brain and muscle
-
 ```{r fig1}
@@
 ```
+ \newpage
 
 ### Fig 2. Hb_oxy alpha and Hb_tot alpha in the brain and muscle
-
 ```{r fig2}
@@
 ```
+ \newpage
 
 ### Fig 3. Increased cerebral_hb_tot and decreased cerebral_hb_tot_alpha are distinct clinical features of cerebral malaria
-
 ```{r fig3_data}
@@
 #### A. Correlations of variables within subject groups
-
 ```{r fig3a, fig.show = "hold", out.width = "50%", fig.height = 12, fig.width = 15}
@@
 ```
+\newpage
 
 #### B. Principal component analysis
-
 ```{r fig3b}
@@
 ```
+\newpage
 
 #### C. Coefficients
-
 ```{r model, eval = FALSE}
@@
                 "t-statistic" = "statistic", "p-value" = "p.value") %>%
+  filter(Term != "(Intercept)") %>% 
   as.data.frame()
@@
 ```
+**note: for the OR, denominator is probability of patient having CM; numerator is p patient having UM
+
+\newpage
 
 #### D. ROC curve of final model
-
 ```{r fig3d_plot, fig.height = 8}
@@
 ```
+\newpage
 
 ### Fig 4. An increase in cerebral_hb_tot is significantly associated with an increase in brain swelling score
+```{r brainswell_model_comparison, eval = FALSE}
+
+# remove cerebral_hb_tot outliers
+  df_hbox_hbtot_outliers <- df_hbox %>% 
+    mutate(Q1_hbtot = quantile(cerebral_hb_tot, 0.25, na.rm = TRUE),
+           Q3_hbtot = quantile(cerebral_hb_tot, 0.75, na.rm = TRUE),
+           IQR_hbtot = Q3_hbtot - Q1_hbtot,
+           outlier_hbtot = case_when(
+
+             cerebral_hb_tot < Q1_hbtot - IQR_hbtot ~ 1,
+             cerebral_hb_tot > Q3_hbtot + IQR_hbtot ~ 1,
+             TRUE ~ 0
+
+           )
+
+    ) %>% 
+    filter(outlier_hbtot != 1)
+
+lm1 <- lm(brain_swell ~ cerebral_hb_tot, data = df_hbox_hbtot_outliers)
+lm2 <- lm(brain_swell ~ cerebral_hb_tot + hct, data = df_hbox_hbtot_outliers)
+lm3 <- lm(brain_swell ~ cerebral_hb_tot + cerebral_hb_tot_alpha2, data = df_hbox_hbtot_outliers)
+lm4 <- lm(brain_swell ~ cerebral_hb_tot + cerebral_hb_tot_alpha2 + hct, data = df_hbox_hbtot_outliers)
+lm5 <- lm(brain_swell ~ cerebral_hb_oxy + cerebral_hb_oxy_alpha2 + cerebral_hb_tot, data = df_hbox_hbtot_outliers)
+
+# evaluate parameters for all models
+brain_swell_models <- list(lm1, lm2, lm3, lm4, lm5)
+
+df_brain_swell_models <- tibble(formula = as.character(),
+                                term = as.character(),
+                                aic = as.numeric(),
+                                r2 = as.numeric(),
+                                adj_r2 = as.numeric(),
+                                model_p_val = as.numeric(),
+                                estimate = as.numeric(),
+                                std_error = as.numeric(),
+                                statistic = as.numeric(),
+                                term_p_val = as.numeric())
+
+for (i in 1:length(brain_swell_models)) {
+
+  formula <- brain_swell_models[[i]] %>% .$call
+  formula <- gsub(".*=", "", formula)[[2]]
+
+  aic <- brain_swell_models[[i]] %>% AIC
+  r2 <- brain_swell_models[[i]] %>% summary %>% .$r.squared
+  adj_r2 <- brain_swell_models[[i]] %>% summary %>% .$adj.r.squared
+  f <- brain_swell_models[[i]] %>% summary %>% .$fstatistic
+  p <- pf(f[1], f[2], f[3], lower.tail = FALSE)
+  coef <- brain_swell_models[[i]] %>% tidy %>% filter(term != "(Intercept)") %>% dplyr::rename("term_p_val" = "p.value")
+
+  df_tmp <- tibble(formula = formula,
+                   term = coef$term,
+                   aic = aic,
+                   r2 = r2,
+                   adj_r2 = adj_r2,
+                   model_p_val = p) %>%
+    left_join(coef) %>%
+    clean_names
+
+  df_brain_swell_models <- df_brain_swell_models %>% 
+    bind_rows(df_tmp) %>% 
+    relocate(term, .before = estimate)
+
+}
+
+df_brain_swell_models %>% write.csv(file = "outputs/brain_swell_models.csv")
+
+```
 
@@
   geom_smooth(method = "lm", lty = 2, color = "#696969", alpha = 0.2) +
-  annotate("text", label = model, x = 25, y = 8, hjust = 0, vjust = 1, size = 5) +
-  annotate("text", label = r2, x = 25, y = 7.5, hjust = 0, vjust = 1, size = 5) +
-  annotate("text", label = "p-values", x = 25, y = 7, hjust = 0, vjust = 1, size = 5) +
-  annotate("text", label = model_p_val, x = 35, y = 6.75, hjust = 0, vjust = 1, size = 4) +
-  annotate("text", label = hb_tot_p_val, x = 35, y = 6.5, hjust = 0, vjust = 1, size = 4) +
+  # annotate("text", label = model, x = 25, y = 8, hjust = 0, vjust = 1, size = 5) +
+  # annotate("text", label = r2, x = 25, y = 7.5, hjust = 0, vjust = 1, size = 5) +
+  # annotate("text", label = "p-values", x = 25, y = 7, hjust = 0, vjust = 1, size = 5) +
+  # annotate("text", label = model_p_val, x = 35, y = 6.75, hjust = 0, vjust = 1, size = 4) +
+  # annotate("text", label = hb_tot_p_val, x = 35, y = 6.5, hjust = 0, vjust = 1, size = 4) +
 
@@
 ```
+**Figure 4.** model: brain_swell ~ cerebral_hb_tot + hematocrit; adjusted R^2: 0.25; model p-value: 0.007; cerebral_hb_tot p-value: 0.003
+
+\newpage
 
@@
 ```
+ \newpage
 
@@
 ```
+ \newpage
 
 #### F. Simulation results
-
 ```{r fig_s1f}
@@
 ```
+ \newpage
 
 ### SFig 2. Logistic regression curves
-
 ```{r fig_s2, fig.height = 12}
@@
 ```
-
 **Figure S2.** Plotted are the individual logistic regression curves of individual variables tested for their discriminatory power between UM and CM.

### 2023-06-12 Scripts/R/markdown_files/response_to_reviewers.Rmd
12 June 2023 clean and organize scripts for publishing

---
title: "Cerebral oxygenation manuscript response to reviewers"
date: "last updated: `r Sys.Date()`"
output: pdf_document
---

<style type="text/css">
  body{
  font-family: Arial;
  font-size: 12pt;
}
</style>

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, comment = "", fig.height = 6, fig.width = 10, 
                      fig.align = 'center')

# Load libraries
library(tidyverse)
library(readxl)
library(DescTools)
library(janitor)
library(FSA)
library(factoextra)
library(ggrepel)
library(ggpubr)
library(tidymodels)
library(vip)
library(scales)
library(reshape)
library(extrafont)
library(grDevices)
library(psych)
library(patchwork)

# Load data
base_dir <- "/Users/ray/Library/CloudStorage/OneDrive-NationalInstitutesofHealth/Ackerman Lab 2019-2021/cerebral_oxygenation/"

# ALT LOADING: use base_dir
#load(paste0(base_dir, "objects/20221125_final_glm_models.Rdata"))
#load(paste0(base_dir, "objects/bandpass_filtered_dfa_results.Rdata"))
#load(paste0(base_dir, "objects/sims.Rdata"))

font_import()

# set plot theme
theme_set(theme_bw() +
              theme(plot.title = element_text(size = 15),
                    axis.title = element_text(size = 15),
                    axis.text = element_text(size = 15),
                    strip.text = element_text(size = 15),
                    legend.title = element_text(size = 15),
                    legend.text = element_text(size = 12)))

```

```{r data}

# read in data

df_hbox <- read_csv(paste0(base_dir, "Data/Hans_datatable_exports/malawi key data v04Jan2022.csv")) %>% 
  as_tibble() %>% 
  janitor::clean_names() %>% 
  mutate(status = replace(status, status == "HV", "HC")) %>% 
  dplyr::rename("subject_id" = "subject_id_session") %>%

  dplyr::filter(status %in% c("CM", "HC", "UM") & session_no == 1 | subject_id == "TM0003CM01") %>% 
  dplyr::filter(!grepl("blood", subject_id)) %>% 
  mutate(status = factor(status, levels = c("HC", "UM", "CM"))) %>% 
  mutate_at(c("bp_dias", "bp_sys", "glucose", "lactate"), as.numeric) %>% # converts remaining character variables to numerics
  select(subject_id, status, everything()) %>% 
  group_by(status)

```

## Reviewer 2

### Minor comment 1

```{r r2c1, fig.height = 10, fig.width = 15}

# MEDIAN (25th, 75th)
df_hbox %>% 
  filter(status == "CM") %>% 
  ungroup %>% 
  select(hrp2, parasitemia) %>% 
  pivot_longer(1:ncol(.), names_to = "variable", values_to = "value") %>% 
  group_by(variable) %>% 
  summarize(median = median(value, na.rm = TRUE),
            `25th` = quantile(value, probs = 0.25, na.rm = TRUE), 
            `75th` = quantile(value, probs = 0.75, na.rm = TRUE)
  ) %>%

  knitr::kable()

# TEST FOR NORMALITY
features <- c("hrp2", "parasitemia", "cerebral_hb_tot", "muscle_hb_tot", "cerebral_hb_tot_alpha2", "muscle_hb_tot_alpha2")

df_normality <- df_hbox %>% 
  filter(status == "CM") %>% 
  ungroup %>% 
  select_at(features) %>% 
  pivot_longer(1:ncol(.), names_to = "feature", values_to = "value") %>%
  group_by(feature) %>% 
  mutate(sw_p = shapiro.test(value)$p.value,
         sw_p = ifelse(sw_p < 0.05, scientific(sw_p, digits = 3), as.character(round(sw_p, 2))),
         sw_p = paste0("SW p = ", sw_p),

         xmin = density(value, na.rm = TRUE)$x %>% min %>% floor,
         xmax = density(value, na.rm = TRUE)$x %>% max %>% ceiling,
         ymin = density(value, na.rm = TRUE)$y %>% min %>% floor,
         ymax = density(value, na.rm = TRUE)$y %>% max %>% ceiling

  )

df_text <- df_normality %>% 
  dplyr::select(-value) %>% 
  distinct() %>% 
  mutate(x = xmax - 0.1*xmax,
         y = ymax - 0.1*ymax) %>% 
  dplyr::select(-contains("min"), -contains("max")) %>% 
  dplyr::rename("label" = "sw_p")

p_sw <- df_normality %>% 
  ggplot(aes(x = value)) +
  geom_density(trim = FALSE) +
  geom_text(
    data = df_text,
    aes(x = Inf, y = Inf, label = label),
    hjust = 1,
    vjust = 1,
    size = 5
  ) + 
  labs(y = "") +
  facet_wrap(vars(feature), scales = "free")

# PLOT CORRELATIONS BETWEEN QUANTITATIVE PARASITEMIA MEASUREMENTS IN CM AND CEREBRAL HBTOT
p_hrp2_cerebral <- df_hbox %>% 
  filter(status == "CM") %>%

  ggplot(aes(x = log10(hrp2), y = log10(cerebral_hb_tot))) +
  geom_point(size = 4, shape = 1) +
  geom_smooth(method = "lm") +
  stat_cor(size = 5) +
  labs(x = paste0("log10(HRP2) (ng/\u03bcL)"), y = "log10(cerebral Hbtot)") +
  ggtitle("NIRS-derived cerebral Hbtot and HRP2 in children with CM") +
  theme_bw() +
  theme(axis.text = element_text(size = 15),
        axis.title = element_text(size = 15))

p_parasitemia_cerebral <- df_hbox %>% 
  filter(status == "CM") %>%

  ggplot(aes(x = log10(parasitemia), y = log10(cerebral_hb_tot))) +
  geom_point(size = 4, shape = 1) +
  geom_smooth(method = "lm") +
  stat_cor(size = 5) +
  labs(x = "log10(parasitemia) (RBCs/\u03bcL)", y = "log10(cerebral Hbtot)") +
  ggtitle("NIRS-derived cerebral Hbtot and parasitemia in children with CM") +
  theme_bw() +
  theme(axis.text = element_text(size = 15),
        axis.title = element_text(size = 15))

# PLOT CORRELATIONS BETWEEN QUANTITATIVE PARASITEMIA MEASUREMENTS IN CM AND MUSCle HBTOT
p_hrp2_muscle <- df_hbox %>% 
  filter(status == "CM") %>%

  ggplot(aes(x = log10(hrp2), y = muscle_hb_tot)) +
  geom_point(size = 4, shape = 1) +
  geom_smooth(method = "lm") +
  stat_cor(size = 5) +
  xlim(c(2.5, 4.5)) +
  labs(x = paste0("log10(HRP2) (ng/\u03bcL)"), y = bquote(`Muscle Hb`[tot])) +
  ggtitle("NIRS-derived muscle Hbtot and HRP2 in children with CM") +
  theme_bw() +
  theme(axis.text = element_text(size = 15),
        axis.title = element_text(size = 15))

p_parasitemia_muscle <- df_hbox %>% 
  filter(status == "CM") %>%

  ggplot(aes(x = log10(parasitemia), y = muscle_hb_tot)) +
  geom_point(size = 4, shape = 1) +
  geom_smooth(method = "lm") +
  stat_cor(size = 5) +
  labs(x = "log10(parasitemia) (RBCs/\u03bcL)", y = bquote(`muscle Hb`[tot])) +
  ggtitle("NIRS-derived muscle Hbtot and parasitemia in children with CM") +
  theme_bw() +
  theme(axis.text = element_text(size = 15),
        axis.title = element_text(size = 15))

# PLOT CORRELATIONS BETWEEN QUANTITATIVE PARASITEMIA MEASUREMENTS IN CM AND CEREBRAL HBTOT ALPHA
p_hrp2_cerebral_alpha <- df_hbox %>% 
  filter(status == "CM") %>%

  ggplot(aes(x = log10(hrp2), y = cerebral_hb_tot_alpha2)) +
  geom_point(size = 4, shape = 1) +
  geom_smooth(method = "lm") +
  stat_cor(size = 5) +
  labs(x = paste0("log10(HRP2) (ng/\u03bcL)"), y = "Cerebral Hbtot alpha") +
  ggtitle("NIRS-derived cerebral Hbtot alpha and HRP2 in children with CM") +
  theme_bw() +
  theme(axis.text = element_text(size = 15),
        axis.title = element_text(size = 15))

p_parasitemia_cerebral_alpha <- df_hbox %>% 
  filter(status == "CM") %>%

  ggplot(aes(x = log10(parasitemia), y = cerebral_hb_tot_alpha2)) +
  geom_point(size = 4, shape = 1) +
  geom_smooth(method = "lm") +
  stat_cor(size = 5) +
  labs(x = "log10(parasitemia) (RBCs/\u03bcL)", y = "Cerebral Hbtot alpha") +
  ggtitle("NIRS-derived cerebral Hbtot alpha and parasitemia in children with CM") +
  theme_bw() +
  theme(axis.text = element_text(size = 15),
        axis.title = element_text(size = 15))

# PLOT CORRELATIONS BETWEEN QUANTITATIVE PARASITEMIA MEASUREMENTS IN CM AND MUSCLE HBTOT ALPHA
p_hrp2_muscle_alpha <- df_hbox %>% 
  filter(status == "CM") %>%

  ggplot(aes(x = log10(hrp2), y = muscle_hb_tot_alpha)) +
  geom_point(size = 4, shape = 1) +
  geom_smooth(method = "lm") +
  stat_cor(size = 5) +
  xlim(c(2.5, 4.5)) +
  labs(x = paste0("log10(HRP2) (ng/\u03bcL)"), y = "Muscle Hbtot alpha") +
  ggtitle("NIRS-derived muscle Hbtot alpha and HRP2 in children with CM") +
  theme_bw() +
  theme(axis.text = element_text(size = 15),
        axis.title = element_text(size = 15))

p_parasitemia_muscle_alpha <- df_hbox %>% 
  filter(status == "CM") %>%

  ggplot(aes(x = log10(parasitemia), y = muscle_hb_tot_alpha2)) +
  geom_point(size = 4, shape = 1) +
  geom_smooth(method = "lm") +
  stat_cor(size = 5) +
  labs(x = "log10(parasitemia) (RBCs/\u03bcL)", y = "Muscle Hbtot alpha") +
  ggtitle("NIRS-derived muscle Hbtot alpha and parasitemia in children with CM") +
  theme_bw() +
  theme(axis.text = element_text(size = 15),
        axis.title = element_text(size = 15))

# COMBINE PLOTS

p_sw

(p_hrp2_cerebral | p_parasitemia_cerebral) /
(p_hrp2_muscle | p_parasitemia_muscle)

(p_hrp2_cerebral_alpha | p_parasitemia_cerebral_alpha) /
(p_hrp2_muscle_alpha | p_parasitemia_muscle_alpha)

```

## Reviewer 3

### Recommendation 1

```{r r3r1}

# REGRESSION
# df_hbox %>% 
#   filter(status == "CM") %>% 
#   
#   ggplot(aes(x = cerebral_hb_tot, y = outcome)) +
#   geom_point(size = 4, shape = 1) +
#   geom_smooth(method = "lm") +
#   stat_cor(size = 5) +
#   labs(x = "Cerebral Hbtot", y = "Outcome") +
#   ggtitle("Correlation between NIRS-derived cerebral Hbtot and outcome in children with CM") +
#   theme_bw() +
#   theme(axis.text = element_text(size = 15),
#         axis.title = element_text(size = 15))

# KRUSKAL-WALLIS
kw_p <- kruskal.test(cerebral_hb_tot ~ outcome,
             data = df_hbox %>%
               filter(status == "CM") %>% 
               mutate(outcome = as.factor(outcome))) %>% 
  .$p.value %>% 
  round(2)

df_hbox %>%
  filter(status == "CM") %>%
  mutate(outcome =

           case_when(

             outcome == 1 ~ "survival",
             outcome == 2 ~ "survival with neurological sequelae",
             outcome == 3 ~ "non-survival"

           ) %>%

           factor(levels = c("survival", "survival with neurological sequelae", "non-survival"))

           ) %>%

  ggplot(aes(x = outcome, y = cerebral_hb_tot, color = outcome)) +
  geom_boxplot(outlier.shape = NA) +
  geom_point(size = 4, alpha = 0.5,
             position = position_jitter(width = 0.1)) +
  #annotate(geom = "text", label = paste0("KW p-value = ", kw_p), x = 3, y = 350, size = 5) +
  labs(x = "", y = bquote(`Cerebral Hb`[tot])) +
  ggtitle("Outcome and NIRS-derived cerebral Hbtot in children with CM") +
  theme_bw() +
  theme(axis.text = element_text(size = 15),
        axis.title = element_text(size = 15),
        legend.position = "none"
        )

```

### 2023-06-12 Scripts/R/markdown_files/20221125_revise_model.Rmd
12 June 2023 clean and organize scripts for publishing

---
title: "Revise LR model"
author: "Rachel Smith"
date: "`r Sys.Date()`"
output: pdf_document
---

Code to identify the best fitting model for each of cerebral Hb_tot and cerebral Hb_tot alpha

```{r setup, include = FALSE}

knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.height = 5.5, fig.width = 8)

# libraries
library(tidyverse)
library(RColorBrewer)
library(gridExtra)
library(patchwork)
library(Hmisc)
library(corrr)
library(tidymodels)
library(themis)
library(vip)
library(janitor)
library(kableExtra)

# set theme for plots
theme_set(theme_bw() +
            theme(plot.title = element_text(size = 18),
                  axis.title = element_text(size = 15),
                  axis.text = element_text(size = 12),
                  strip.text = element_text(size = 12),
                  legend.title = element_text(size = 15),
                  legend.text = element_text(size = 12)))

# create color vector
my_col <- c("CM" = "#00BFC4", "HC" = "#7CAE00", "UM" = "#F8766D")

```

```{r data, echo = FALSE}

# load
df_hbox <- read_csv("Data/Hans_datatable_exports/malawi key data v04Jan2022.csv") %>% 
  as_tibble() %>% 
  janitor::clean_names() %>% 
  mutate(status = replace(status, status == "HV", "HC")) %>% 
  dplyr::rename("subject_id" = "subject_id_session") %>%

  dplyr::filter(status %in% c("CM", "HC", "UM") & session_no == 1 | subject_id == "TM0003CM01") %>% 
  dplyr::filter(!grepl("blood", subject_id)) %>% 
  mutate(status = factor(status, levels = c("HC", "UM", "CM"))) %>% 
  mutate_at(c("bp_dias", "bp_sys", "glucose", "lactate"), as.numeric) %>% # converts remaining character variables to numerics
  select(subject_id, status, everything()) %>% 
  group_by(status)

# format
  # define clinical variables of interest
  clinical_voi <- c("age_calc", "temperature", "hr", "bp_sys", "bp_dias", "resp_rate", "pulse_ox", "hct", "lactate", "glucose", "aa_arg")
  hbox_voi <- c("cerebral_hb_tot", "cerebral_hb_tot_alpha2")

  # select clinical VOI
  df_model <- df_hbox %>% 
    dplyr::select(c(status, sex, all_of(hbox_voi), all_of(clinical_voi))) %>% 
    mutate(sex_coded = ifelse(sex == "Female", 1, 0)) %>% 
    dplyr::select(-sex)

  # impute missing data with median by group
  f_impute <- function(x, na.rm = TRUE) (replace(x, is.na(x), median(x, na.rm = na.rm)))

  df_model_impute <- df_model %>% 
    group_by(status) %>% 
    mutate_at(2:ncol(.), f_impute) %>% 
    ungroup()

```

## Chi-squared test for differences in proportion of sex by patient group

HC, UM, CM
```{r sex_chisq_all}

# HC, UM, CM
df_sex_chisq_all <- df_hbox %>% 
  dplyr::select(status, sex) %>% 
  count(status, sex) %>% 
  pivot_wider(id_cols = status, names_from = sex, values_from = n) %>% 
  as.data.frame %>% 
  column_to_rownames("status")

knitr::kable(df_sex_chisq_all)

chisq.test(df_sex_chisq_all)

```

UM vs CM
```{r sex_chisq_um_cm}

# UM vs CM
df_sex_chisq_um_cm <- df_hbox %>% 
  dplyr::select(status, sex) %>% 
  filter(status != "HC") %>% 
  count(status, sex) %>% 
  pivot_wider(id_cols = status, names_from = sex, values_from = n) %>% 
  as.data.frame %>% 
  column_to_rownames("status")

knitr::kable(df_sex_chisq_um_cm)

chisq.test(df_sex_chisq_um_cm)

```

```{r lr_model}

# write function to run LR model
f_run_lr <- function(df_model, resamples) {

  # resample
  hbox_boot <- bootstraps(df_model, strata = status, times = resamples)

  # recipe
  hbox_rec <- recipe(status ~ ., data = df_model) %>% 
    step_impute_knn(all_predictors()) %>% 
    step_normalize(all_predictors()) %>%
    step_zv(all_predictors()) %>%
    step_smote(status) %>% 
    update_role(subject_id, new_role = "id")

  # create workflow
  hbox_wf <- workflow() %>% add_recipe(hbox_rec)

  # logistic regression model specifications
  glm_spec <- logistic_reg() %>% set_engine("glm")

  # create model on 5000 different resamples
  glm_res <- hbox_wf %>% 
    add_model(glm_spec) %>%
    fit_resamples(
      resamples = hbox_boot,
      metrics = metric_set(roc_auc, accuracy, sensitivity, specificity),
      control = tune::control_resamples(save_pred = TRUE, verbose = TRUE)
    )

  # select best model to be used for evaluation
  glm_best <- glm_res %>% select_best("roc_auc")

  # finalize model
  hbox_wf %>% 
    add_model(glm_spec) %>% 
    finalize_workflow(glm_best) %>%
    fit(df_model) %>%
    extract_fit_parsnip()

}

```

## Shapiro-Wilk test for normality
```{r shapiro_wilk}

voi <- c("cerebral_hb_tot", "cerebral_hb_tot_alpha2", "sex", "resp_rate", "hct", "lactate")

df_hbox_voi <- df_hbox %>% 
  select(status, all_of(voi)) %>% 
  dplyr::select(-sex) %>% 
  pivot_longer(2:ncol(.), names_to = "variable", values_to = "value")

# plot distributions
df_hbox_voi %>%  
  ggplot(aes(x = status, y = value)) +
  geom_violin(aes(color = status)) +
  geom_jitter(position = position_jitter(0.15), shape = 1, size = 1.5) +
  stat_summary(fun = "median", geom = "crossbar", aes(color = status), size = 0.2, width = 0.5) +
  facet_wrap(~ variable, scales = "free") +
  labs(x = "", y = "") +
  ggtitle("Clinical variable distributions by patient group") +
  theme(legend.position = "none")

# Shapiro-Wilk test
df_shapiro <- df_hbox_voi %>% 
  group_by(status, variable) %>% 
  nest() %>%

  mutate(shapiro_test = map(.x = data,
                            .f = ~ shapiro.test(.x$value))) %>% 
  mutate(p_val = map(.x = shapiro_test,
                     .f = ~ .x$p.value)) %>% 
  unnest(cols = c(p_val)) %>% 
  mutate(p_adj = p.adjust(p_val, method = "BH")) %>% 
  dplyr::select(status, variable, p_adj) %>% 
  pivot_wider(id_cols = status, names_from = variable, values_from = p_adj) %>% 
  mutate_if(is.numeric, ~ round(.x, digits = 3))

df_shapiro %>% 
  knitr::kable()

```

## Cerebral Hb_tot model fitting

status ~ log10(cerebral_hb_tot) + sex + log10(resp_rate) + hct + log10(lactate)
```{r hb_tot_model1}

df_model <- df_hbox %>% 
  mutate(sex = ifelse(sex == "Female", 1, 0)) %>% 
  filter(status != "HC") %>% 
  mutate(status = factor(status, levels = c("UM", "CM"))) %>%

  # transform non-normal variables (by Shapiro-Wilk) using log-transform
  mutate(log10_cerebral_hb_tot = log10(cerebral_hb_tot),
         log10_resp_rate = log10(resp_rate),
         log10_lactate = log10(lactate))

set.seed(20221125)

# 1. status ~ log10_cerebral_hb_tot + sex + log10_resp_rate + hct + log10_lactate
hbtot_final1 <- f_run_lr(df_model %>% 
                           dplyr::select(subject_id, status, log10_cerebral_hb_tot, 
                                         sex, log10_resp_rate, hct, log10_lactate), 
                       10)

hbtot_final1 %>% 
  tidy %>% 
  filter(term != "(Intercept)") %>% 
  mutate_if(is.numeric, ~ round(.x, 3)) %>% 
  knitr::kable() # AIC 82.61

```

status ~ log10(cerebral_hb_tot) + sex + hct + log10(lactate)
```{r hb_tot_model2}

# 2. status ~ log10_cerebral_hb_tot + sex + hct + log10_lactate

set.seed(20221125)

# resample
hbtot_boot <- bootstraps(df_model, strata = status, times = 5000)

# recipe
hbtot_rec <- recipe(status ~ log10_cerebral_hb_tot + hct + log10_lactate + sex + subject_id, data = df_model) %>% 
  step_impute_knn(all_predictors()) %>% 
  step_normalize(all_predictors()) %>%
  step_zv(all_predictors()) %>%
  step_smote(status) %>% 
  update_role(subject_id, new_role = "id")

# create workflow
hbtot_wf <- workflow() %>% add_recipe(hbtot_rec)

# logistic regression model specifications
hbtot_glm_spec <- logistic_reg() %>% set_engine("glm")

# create model on 5000 different resamples
hbtot_glm_res <- hbtot_wf %>% 
  add_model(hbtot_glm_spec) %>%
  fit_resamples(
    resamples = hbtot_boot,
    metrics = metric_set(roc_auc, accuracy, sensitivity, specificity),
    control = tune::control_resamples(save_pred = TRUE, verbose = TRUE)
  )

# select best model to be used for evaluation
hbtot_glm_best <- hbtot_glm_res %>% select_best("roc_auc")

# finalize model
hbtot_final2 <- hbtot_wf %>% 
  add_model(hbtot_glm_spec) %>% 
  finalize_workflow(hbtot_glm_best) %>%
  fit(df_model) %>%
  extract_fit_parsnip()

hbtot_final2 %>% 
  tidy %>% 
  filter(term != "(Intercept)") %>% 
  mutate_if(is.numeric, ~ round(.x, 3)) %>% 
  knitr::kable() # AIC 81.92, every term is significant

```

```{r assess_hb_tot_model}

hbtot_final2$fit
# AIC 76.25

df_CI <- confint(hbtot_final2$fit) %>% exp() %>% as_tibble(rownames = "term")

hbtot_final2 %>%
  tidy(exponentiate = TRUE) %>%
  filter(p.value < 0.05) %>% 
  left_join(df_CI, by = "term") %>%
  mutate(estimate = ifelse(estimate < 10^-2, scientific(estimate, digits = 3), round(estimate, 2)),
         conf.low = ifelse(`2.5 %` < 10^-2, scientific(`2.5 %`, digits = 3), round(`2.5 %`, 2)),
         conf.high = ifelse(`97.5 %` < 10^-2, scientific(`97.5 %`, digits = 3), round(`97.5 %`, 2))) %>%
  mutate(p.value = scientific(p.value, digits = 3),
         statistic = round(statistic, 2)) %>%
  unite("95% CI", c(conf.low, conf.high), sep = ", ") %>%
  dplyr::select(c(term, estimate, `95% CI`, statistic, p.value)) %>% 
  dplyr::rename("Term" = "term", "Odds ratio" = "estimate",
                "t-statistic" = "statistic", "p-value" = "p.value") %>%
  filter(Term != "(Intercept)") %>% 
  as.data.frame() %>%

  knitr::kable(align = rep('c', 5))

hbtot_glm_res %>% 
  collect_metrics() %>% 
  dplyr::select(.metric, mean, std_err) %>% 
  mutate(mean = mean*100, std_err = std_err*100) %>% 
  mutate(Metric = c("Accuracy", "ROC AUC", "Sensitivity", "Specificity")) %>% 
  dplyr::select(c("Metric", "mean", "std_err")) %>%

  dplyr::rename("Mean" = "mean", "Standard error" = "std_err") %>% 
  mutate_if(is.numeric, round, 2) %>% 
  as.data.frame() %>%

  knitr::kable()

```

```{r hb_tot_imp, fig.height = 4}

set.seed(123)
hbtot_imp <- hbtot_wf %>% 
  add_model(hbtot_glm_spec) %>% 
  fit(df_model) %>% 
  pull_workflow_fit() %>%

  vi(
    method = "permute", nsim = 10,
    train = juice(hbtot_rec %>% prep()),
    target = "status", metric = "auc", reference_class = "CM",
    pred_wrapper = kernlab::predict

  )

hbtot_imp %>% 
  clean_names %>% 
  filter(variable != "subject_id") %>% 
  mutate(variable = fct_reorder(variable, importance)) %>%

  ggplot(aes(x = importance, y = variable, color = variable)) +
  geom_errorbar(aes(xmin = importance - st_dev, xmax = importance + st_dev),
                alpha = 0.5, size = 1.3) +
  geom_point(size = 3) +
  theme(legend.position = "none",
        plot.title = element_text(size = 18),
        axis.title = element_text(size = 15),
        axis.text = element_text(size = 12),
        strip.text = element_text(size = 12),
        legend.title = element_text(size = 15),
        legend.text = element_text(size = 12)) +
  labs(y = NULL, x = "relative importance") +
  ggtitle("Variable importance in distinguishing CM vs UM")

```

## Cerebral Hb_tot_alpha model fitting

status ~ cerebral_hb_tot_alpha2 + sex + log10(resp_rate) + hct + log10(lactate)
```{r hb_tot_alpha_model1}

# 1. status ~ cerebral_hb_tot_alpha2 + sex + log10_resp_rate + hct + log10_lactate
hbtot_alpha_final1 <- f_run_lr(df_model %>% 
                           dplyr::select(subject_id, status, cerebral_hb_tot_alpha2, 
                                         sex, log10_resp_rate, hct, log10_lactate), 
                       10)

hbtot_alpha_final1 %>% 
  tidy %>% 
  filter(term != "(Intercept)") %>% 
  mutate_if(is.numeric, ~ round(.x, 3)) %>% 
  knitr::kable() # AIC 62.41

```

status ~ cerebral_hb_tot_alpha2 + sex + hct + log10(lactate)
```{r hb_tot_alpha_model2}

# 2. status ~ cerebral_hb_tot_alpha2 + sex + hct + log10_lactate

set.seed(20221125)

# resample
hbtot_alpha_boot <- bootstraps(df_model, strata = status, times = 5000)

# recipe
hbtot_alpha_rec <- recipe(status ~ cerebral_hb_tot_alpha2 + hct + sex + log10_lactate + subject_id, data = df_model) %>% 
  step_impute_knn(all_predictors()) %>% 
  step_normalize(all_predictors()) %>%
  step_zv(all_predictors()) %>%
  step_smote(status) %>% 
  update_role(subject_id, new_role = "id")

# create workflow
hbtot_alpha_wf <- workflow() %>% add_recipe(hbtot_alpha_rec)

# logistic regression model specifications
hbtot_alpha_glm_spec <- logistic_reg() %>% set_engine("glm")

# create model on 5000 different resamples
hbtot_alpha_glm_res <- hbtot_alpha_wf %>% 
  add_model(hbtot_alpha_glm_spec) %>%
  fit_resamples(
    resamples = hbtot_alpha_boot,
    metrics = metric_set(roc_auc, accuracy, sensitivity, specificity),
    control = tune::control_resamples(save_pred = TRUE, verbose = TRUE)
  )

# select best model to be used for evaluation
hbtot_alpha_glm_best <- hbtot_alpha_glm_res %>% select_best("roc_auc")

# finalize model
hbtot_alpha_final2 <- hbtot_alpha_wf %>% 
  add_model(hbtot_alpha_glm_spec) %>% 
  finalize_workflow(hbtot_alpha_glm_best) %>%
  fit(df_model) %>%
  extract_fit_parsnip()

hbtot_alpha_final2 %>% 
  tidy %>% 
  filter(term != "(Intercept)") %>% 
  mutate_if(is.numeric, ~ round(.x, 3)) %>% 
  knitr::kable() # AIC 65.29

## save final models!
save(hbtot_final2, hbtot_alpha_final2,
     file = "objects/20221125_final_glm_models.Rdata")

```

```{r assess_hb_tot_alpha_model}

hbtot_alpha_final2$fit
# AIC 61.05

df_CI <- confint(hbtot_alpha_final2$fit) %>% exp() %>% as_tibble(rownames = "term")

hbtot_alpha_final2 %>%
  tidy(exponentiate = TRUE) %>%
  filter(p.value < 0.05) %>% 
  left_join(df_CI, by = "term") %>%
  mutate(estimate = ifelse(estimate < 10^-2, scientific(estimate, digits = 3), round(estimate, 2)),
         conf.low = ifelse(`2.5 %` < 10^-2, scientific(`2.5 %`, digits = 3), round(`2.5 %`, 2)),
         conf.high = ifelse(`97.5 %` < 10^-2, scientific(`97.5 %`, digits = 3), round(`97.5 %`, 2))) %>%
  mutate(p.value = scientific(p.value, digits = 3),
         statistic = round(statistic, 2)) %>%
  unite("95% CI", c(conf.low, conf.high), sep = ", ") %>%
  dplyr::select(c(term, estimate, `95% CI`, statistic, p.value)) %>% 
  dplyr::rename("Term" = "term", "Odds ratio" = "estimate",
                "t-statistic" = "statistic", "p-value" = "p.value") %>%
  filter(Term != "(Intercept)") %>% 
  as.data.frame() %>%

  knitr::kable(align = rep('c', 5))

hbtot_alpha_glm_res %>% 
  collect_metrics() %>% 
  dplyr::select(.metric, mean, std_err) %>% 
  mutate(mean = mean*100, std_err = std_err*100) %>% 
  mutate(Metric = c("Accuracy", "ROC AUC", "Sensitivity", "Specificity")) %>% 
  dplyr::select(c("Metric", "mean", "std_err")) %>%

  dplyr::rename("Mean" = "mean", "Standard error" = "std_err") %>% 
  mutate_if(is.numeric, round, 2) %>% 
  as.data.frame() %>%

  knitr::kable()

```

```{r hb_tot_alpha_imp, fig.height = 4}

set.seed(123)
hbtot_alpha_imp <- hbtot_alpha_wf %>% 
  add_model(hbtot_alpha_glm_spec) %>% 
  fit(df_model) %>% 
  pull_workflow_fit() %>%

  vi(
    method = "permute", nsim = 10,
    train = juice(hbtot_alpha_rec %>% prep()),
    target = "status", metric = "auc", reference_class = "CM",
    pred_wrapper = kernlab::predict

  )

hbtot_alpha_imp %>% 
  clean_names %>% 
  filter(variable != "subject_id") %>% 
  mutate(variable = fct_reorder(variable, importance)) %>%

  ggplot(aes(x = importance, y = variable, color = variable)) +
  geom_errorbar(aes(xmin = importance - st_dev, xmax = importance + st_dev),
                alpha = 0.5, size = 1.3) +
  geom_point(size = 3) +
  theme(legend.position = "none",
        plot.title = element_text(size = 18),
        axis.title = element_text(size = 15),
        axis.text = element_text(size = 12),
        strip.text = element_text(size = 12),
        legend.title = element_text(size = 15),
        legend.text = element_text(size = 12)) +
  labs(y = NULL, x = "relative importance") +
  ggtitle("Variable importance in distinguishing CM vs UM")

```

### 2023-06-12 Scripts/R/markdown_files/20221205_brain_swell_model.Rmd
12 June 2023 clean and organize scripts for publishing

---
output: pdf_document
---

Characterize the relationship between brain swelling and cerebral Hb_tot

```{r setup, include = FALSE}

knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.height = 5.5, fig.width = 8)

# libraries
library(tidyverse)
library(RColorBrewer)
library(gridExtra)
library(patchwork)
library(Hmisc)
library(corrr)
library(tidymodels)
library(themis)
library(vip)
library(janitor)
library(kableExtra)

# set theme for plots
theme_set(theme_bw() +
            theme(plot.title = element_text(size = 18),
                  axis.title = element_text(size = 15),
                  axis.text = element_text(size = 12),
                  strip.text = element_text(size = 12),
                  legend.title = element_text(size = 15),
                  legend.text = element_text(size = 12)))

# create color vector
my_col <- c("CM" = "#00BFC4", "HC" = "#7CAE00", "UM" = "#F8766D")

```

```{r data, echo = FALSE}

# load
df_hbox <- read_csv("Data/Hans_datatable_exports/malawi key data v04Jan2022.csv") %>% 
  as_tibble() %>% 
  janitor::clean_names() %>% 
  mutate(status = replace(status, status == "HV", "HC")) %>% 
  dplyr::rename("subject_id" = "subject_id_session") %>%

  dplyr::filter(status %in% c("CM", "HC", "UM") & session_no == 1 | subject_id == "TM0003CM01") %>% 
  dplyr::filter(!grepl("blood", subject_id)) %>% 
  mutate(status = factor(status, levels = c("HC", "UM", "CM"))) %>% 
  mutate_at(c("bp_dias", "bp_sys", "glucose", "lactate"), as.numeric) %>% # converts remaining character variables to numerics
  select(subject_id, status, everything()) %>% 
  group_by(status) %>% 
  mutate(log10_cerebral_hb_tot = log10(cerebral_hb_tot))

```

```{r fig1}

# extract hb_tot and hb_oxy for the brain and muscle
df_fig1 <- df_hbox %>% 
  select(contains(c("hb_tot","hb_oxy")), -contains(c("alpha", "bsl"))) %>%
  ungroup() %>% 
  pivot_longer(2:ncol(.), names_to = "variable", values_to = "value") %>% 
  separate(variable, into = c("tissue", "variable"), sep = "_", extra = "merge")

# create color vector
my_col <- c("CM" = "#00BFC4", "HC" = "#7CAE00", "UM" = "#F8766D")

# plot
df_fig1 %>%

  ggplot(aes(x = status, y = value)) +
  geom_violin(aes(color = status)) +
  geom_jitter(position = position_jitter(0.2), shape = 1, size = 2) +
  stat_summary(fun = "median", geom = "crossbar", aes(color = status), size = 0.2, width = 0.5) +
  facet_grid(variable ~ tissue, scales = "free_y") +

  scale_color_manual(values = my_col) +
  labs(x = "") +
  theme_bw() +
  theme(legend.position = "none")  +

  theme(# text = element_text(family = "Arial"),
        strip.text = element_text(size = 15),
        axis.title = element_text(size = 15),
        axis.title.y = element_blank(),
        axis.text = element_text(size = 15))

```

## Brain swell model fitting

1. Log10-transformed Hb_tot vs brain swelling – with hematocrit indicated by plotting character size (with all data points represented)
2. A simple plot of hct vs brain swelling for my own understanding of that relationship.

```{r plot}

df_hbox %>%
  ggplot(aes(x = log10_cerebral_hb_tot, y = brain_swell)) +
  geom_point(aes(size = hct), shape = 1) +

  geom_smooth(method = "lm", lty = 2, color = "#696969", alpha = 0.2, se = FALSE) +

  theme_bw() +
  ggtitle("1. Relationship between brain swell score and log10-transformed cerebral Hb_tot") +
  theme(strip.text = element_text(size = 15),
        axis.title = element_text(size = 15),
        axis.title.y = element_text(size = 15),
        axis.text = element_text(size = 15),
        legend.title = element_text(size = 12),
        legend.text = element_text(size = 10))

df_hbox %>%
  ggplot(aes(x = hct, y = brain_swell)) +
  geom_point(shape = 1, size = 2) +

  geom_smooth(method = "lm", lty = 2, color = "#696969", alpha = 0.2, se = FALSE) +
  theme_bw() +
  ggtitle("2. Relationship between hematocrit & brain swell score") +
  theme(strip.text = element_text(size = 15),
        axis.title = element_text(size = 15),
        axis.title.y = element_text(size = 15),
        axis.text = element_text(size = 15),
        legend.title = element_text(size = 12),
        legend.text = element_text(size = 10))

```

<!-- ```{r brain_swell} -->

<!-- cor.test( ~ brain_swell + log10_cerebral_hb_tot,  -->
<!--           data = df_hbox, -->
<!--           method = "spearman") -->

<!-- lm(brain_swell ~ log10_cerebral_hb_tot, data = df_hbox) %>% summary -->

<!-- lm(brain_swell ~ log10_cerebral_hb_tot + hct, data = df_hbox) %>% summary -->

<!-- lm(brain_swell ~ log10_cerebral_hb_tot + cerebral_hb_tot_alpha2, data = df_hbox) %>% summary -->

<!-- lm(brain_swell ~ log10_cerebral_hb_tot + cerebral_hb_tot_alpha2 + hct, data = df_hbox) %>% summary -->

<!-- lm(brain_swell ~ log10_cerebral_hb_tot + cerebral_hb_tot_alpha2, data = df_hbox) %>% summary -->

<!-- lm(brain_swell ~ cerebral_hb_oxy + cerebral_hb_oxy_alpha2 + log10_cerebral_hb_tot, data = df_hbox) %>% summary -->

<!-- ``` -->

### 2023-06-12 Scripts/R/markdown_files/20221217_parasitemia.Rmd
12 June 2023 clean and organize scripts for publishing

---
title: "parasitemia"
date: "`r Sys.Date()`"
output: pdf_document
---

Identify differences in parastiemia counts between patient groups

```{r setup, include = FALSE}

knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.height = 5.5, fig.width = 8)

# libraries
library(tidyverse)
library(RColorBrewer)
library(patchwork)
library(Hmisc)
library(corrr)
library(tidymodels)
library(themis)
library(vip)
library(janitor)
library(kableExtra)
library(MASS)

# set theme for plots
theme_set(theme_bw() +
            theme(plot.title = element_text(size = 18),
                  axis.title = element_text(size = 15),
                  axis.text = element_text(size = 12),
                  strip.text = element_text(size = 12),
                  legend.title = element_text(size = 15),
                  legend.text = element_text(size = 12)))

# create color vector
my_col <- c("CM" = "#00BFC4", "HC" = "#7CAE00", "UM" = "#F8766D")

```

```{r data, echo = FALSE}

# load
df_hbox <- read_csv("Data/Hans_datatable_exports/malawi key data v04Jan2022.csv") %>% 
  as_tibble() %>% 
  janitor::clean_names() %>% 
  mutate(status = replace(status, status == "HV", "HC")) %>% 
  dplyr::rename("subject_id" = "subject_id_session") %>%

  dplyr::filter(status %in% c("CM", "HC", "UM") & session_no == 1 | subject_id == "TM0003CM01") %>% 
  dplyr::filter(!grepl("blood", subject_id)) %>% 
  mutate(status = factor(status, levels = c("HC", "UM", "CM"))) %>% 
  mutate_at(c("bp_dias", "bp_sys", "glucose", "lactate"), as.numeric) %>% # converts remaining character variables to numerics
  dplyr::select(subject_id, status, everything()) %>% 
  group_by(status)

# format
  # define clinical variables of interest
  clinical_voi <- c("age_calc", "temperature", "hr", "bp_sys", "bp_dias", "resp_rate", "pulse_ox", "hct", "lactate", "glucose", "aa_arg")
  hbox_voi <- c("cerebral_hb_tot", "cerebral_hb_tot_alpha2")

  # select clinical VOI
  df_model <- df_hbox %>% 
    dplyr::select(c(status, sex, all_of(hbox_voi), all_of(clinical_voi))) %>% 
    mutate(sex_coded = ifelse(sex == "Female", 1, 0)) %>% 
    dplyr::select(-sex)

  # impute missing data with median by group
  f_impute <- function(x, na.rm = TRUE) (replace(x, is.na(x), median(x, na.rm = na.rm)))

  df_model_impute <- df_model %>% 
    group_by(status) %>% 
    mutate_at(2:ncol(.), f_impute) %>% 
    ungroup()

```

## Parasite counts in each patient group
```{r parasite_counts}

df_parasite_counts <- df_hbox %>% 
  dplyr::select(parasites) %>% 
  dplyr::count(parasites) %>% 
  pivot_wider(id_cols = parasites, names_from = status, values_from = n) %>% 
  dplyr::filter(!is.na(parasites)) %>% # 1 CM patient had no value for parasites
  dplyr::mutate_if(is.numeric, ~ ifelse(is.na(.x), 0, .x))

df_parasite_counts %>% knitr::kable(align = rep('c', 4))

# plot
df_parasite_counts %>% 
  pivot_longer(2:ncol(.), names_to = "status", values_to = "count") %>% 
  dplyr::mutate(status = factor(status, levels = c("HC", "UM", "CM")),
         parasites = factor(parasites, levels = c("Neg", "1+", "2+", "3+", "4+", "5+")),
         count = ifelse(count == 0, NA_real_, count)) %>%

  ggplot(aes(x = parasites, y = count, fill = status)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = my_col) +
  ylab("") +
  ggtitle("Parasite counts in each patient group")

```

## Statistical tests to compare parasites in each patient group

### Chi-squared
```{r chisq}

# Chi-squared of UM vs CM
chisq.test(df_parasite_counts %>% 
             dplyr::select(-HC) %>% 
             as.data.frame %>% 
             column_to_rownames("parasites")
)

```

### Ordinal regression (?)
```{r ord}

df_parasites <- df_hbox %>% 
  dplyr::select(parasites) %>% 
  mutate(parasites = factor(parasites, levels = c("Neg", "1+", "2+", "3+", "4+", "5+")))

ord <- polr(status ~ parasites, data = df_parasites, Hess = TRUE)
tidy(ord) %>% 
  mutate(p_val = pnorm(abs(statistic), lower.tail = FALSE)*2) %>% 
  clean_names %>% 
  dplyr::rename("coefficient" = "estimate",
                "t_stat" = "statistic")

```

### Chi-squared (binarize parasites variable)
```{r chisq_bin}

df_parasite_counts_binned <- df_parasite_counts %>% 
  mutate(parasites_bin = ifelse(parasites %in% c("Neg", "1+", "2+", "3+"),
                                "low", "high") %>% 
           factor(levels = c("low", "high"))) %>% 
  dplyr::select(parasites_bin, UM, CM) %>% 
  group_by(parasites_bin) %>% 
  summarise(UM = sum(UM), CM = sum(CM))

df_parasite_counts_binned %>% knitr::kable(align = rep('c', 3))

# Chi-squared of UM vs CM
chisq.test(df_parasite_counts_binned %>% 
             as.data.frame %>% 
             column_to_rownames("parasites_bin")
)

```

### Logistic regression (binarize parasites variable)
```{r logreg}

df_parasites_binned <- df_parasites %>% 
  mutate(parasites_bin = ifelse(parasites %in% c("Neg", "1+", "2+", "3+"),
                                "low", "high") %>% 
           factor(levels = c("low", "high")))

# Chi-squared of UM vs CM
logreg <- glm(status ~ parasites_bin, data = df_parasites_binned, family = "binomial")
tidy(logreg) %>% 
  clean_names %>% 
  dplyr::rename("coefficient" = "estimate",
                "t_stat" = "statistic")

```

### HRP & parasitemia in CM patients
```{r parasitemia}

cm_voi <- c("hrp2", "parasitemia")

df_cm_voi <- df_hbox %>% 
  ungroup %>% 
  filter(status == "CM") %>% 
  dplyr::select_at(cm_voi)

# HRP
df_cm_hrp2_parasitemia <- df_cm_voi %>% 
  summarise_all(funs(median = median, 
                     q1 = quantile(., probs = 0.25), 
                     q3 = quantile(., probs = 0.75)), 
                na.rm = TRUE) %>% 
  mutate(hrp2 = paste0(hrp2_median, " (", hrp2_q1, ", ", hrp2_q3, ")"),
         parasitemia = paste0(parasitemia_median, " (", parasitemia_q1, ", ", parasitemia_q3, ")")) %>% 
  dplyr::select(hrp2, parasitemia)

df_cm_hrp2_parasitemia %>% knitr::kable(align = rep('c', 2))

```

### 2023-06-12 Scripts/R/markdown_files/20230109_fig1_fig3_reformat.Rmd
12 June 2023 clean and organize scripts for publishing

--- 2022-02-09 final_figures.Rmd
@@
 ---
-title: "Cerebral oxygenation manuscript figures"
 date: "`r Sys.Date()`"
-output: html_document
+output: pdf_document
 ---
@@
 library(readxl)
-library(DescTools)
 library(janitor)
-library(FSA)
-library(factoextra)
-library(ggrepel)
-library(ggpubr)
-library(tidymodels)
-library(vip)
-library(scales)
-library(reshape)
 library(extrafont)
-library(grDevices)
-
-# Load data
-load("objects/glm_model.Rdata")
-load("objects/bandpass_filtered_dfa_results.Rdata")
-load("objects/sims.Rdata")
 
@@
 
-### Table 1. Baseline clinical characteristics
-
-#### Table 1a. All clinical data
-```{r table1a, eval = FALSE}
-
-# number of subjects 
-df_status <- df_hbox %>% 
-  count() %>% 
-  pivot_wider(1:2, names_from = "status", values_from = "n") %>% 
-  mutate(variable = "Number of subjects") %>% 
-  dplyr::select(variable, everything()) %>% 
-  mutate_if(is.numeric, as.character)
-
-# Sex
-df_sex <- df_hbox %>% 
-  count(sex) %>% 
-  dplyr::filter(sex != "") %>% 
-  pivot_wider(names_from = "sex", values_from = "n") %>% 
-  mutate(perc_fem = round(Female/(Female + Male)*100, 2)) %>%
-
-  pivot_wider(c(1,4), names_from = "status", values_from = "perc_fem") %>% 
-  mutate(variable = "Sex (% female)") %>% 
-  dplyr::select(variable, everything()) %>% 
-  mutate_if(is.numeric, as.character)
-
-# other numeric variables
-
-  # select variables for table
-  clinical_voi <- c("age_calc", "temperature", "hr", "bp_sys", "bp_dias", "resp_rate", "pulse_ox", "hct", "lactate", "glucose", "aa_arg")
-
-  # calculate median, first and third quartiles
-  df_clinical_res <- df_hbox %>% 
-    select_at(clinical_voi) %>% 
-    summarise_all(funs(median = median, q1 = quantile(., probs = 0.25), q3 = quantile(., probs = 0.75)), na.rm = TRUE) %>% 
-    mutate_if(is.numeric, round, 2)
-
-  # write function to paste together values
-  f_unite_cols <- function(col) {
-    df_clinical_res %>% 
-      unite(!! col, starts_with(col), sep = "_") %>% 
-      dplyr::select(!! col) 
-    }
-
-  # paste together and transpose
-  df_clinical_res <- 1:length(clinical_voi) %>% 
-    purrr::map_dfc(~f_unite_cols(clinical_voi[.x])) %>% 
-    mutate_all(funs(str_replace(., "_", " ("))) %>% 
-    mutate_all(funs(str_replace(., "_", ", "))) %>% 
-    mutate_all(funs(paste0(., ")"))) %>%
-
-    t() %>% 
-    as_tibble(rownames = "variable") %>% 
-    dplyr::rename("HC" = "V1", "UM" = "V2", "CM" = "V3") %>% 
-    add_row(df_status, .before = 1) %>% 
-    add_row(df_sex, .before = 2)
-
-# export to CSV
-  write.csv(df_clinical_res, file = "Outputs/table_1.csv")
-
-# create kable df
-  df_clinical_res %>% knitr::kable(align = rep('c', 4))
-
-```
-**Table 1a.** A summary of subject information and hematological data. Group patient data presented as median (quartile 1, quartile 3).
-
-#### Table 1b. CM-specific data
-```{r table1b, eval = FALSE}
-
-cm_voi <- c("bcs_session", "brain_swell", "hrp2")
-
-df_cm_voi <- df_hbox %>% 
-  ungroup %>% 
-  filter(status == "CM") %>% 
-  select_at(cm_voi) %>% 
-  mutate_at(c("bcs_session", "brain_swell"), as.factor)
-
-# BCS counts
-df_cm_voi %>% count(bcs_session) %>% knitr::kable()
-
-# Brain swell score counts
-df_cm_voi %>% count(brain_swell) %>% knitr::kable()
-
-# HRP
-df_cm_hrp2 <- df_cm_voi %>% 
-  select(cm_voi[3]) %>% 
-  summarise_all(funs(median = median, q1 = quantile(., probs = 0.25), q3 = quantile(., probs = 0.75)), na.rm = TRUE) %>% 
-  unite(hrp2, 1:2, sep = " (") %>% 
-  unite(hrp2, 1:2, sep = ", ") %>% 
-  mutate(hrp2 = paste0(hrp2, ")"))
-
-df_cm_hrp2 %>% knitr::kable()
-
-```
-**Table 1b.** CM-specific clinical characteristics
-
 ### Fig 1. Hb_oxy and Hb_tot in the brain and muscle
-
 ```{r fig1}
@@
   pivot_longer(2:ncol(.), names_to = "variable", values_to = "value") %>% 
-  separate(variable, into = c("tissue", "variable"), sep = "_", extra = "merge")
+  separate(variable, into = c("tissue", "variable"), sep = "_", extra = "merge") %>% 
+  mutate(variable = factor(variable, levels = c("hb_tot", "hb_oxy")))
 
@@
 
-### Fig 2. Hb_oxy alpha and Hb_tot alpha in the brain and muscle
-
-```{r fig2}
+### Fig 3. Hb_oxy alpha and Hb_tot alpha in the brain and muscle
+```{r fig3}
 
 # extract hb_tot and hb_oxy for the brain and muscle
-df_fig2 <- df_hbox %>% 
+df_fig3 <- df_hbox %>% 
   select(contains("alpha2")) %>%
@@
   separate(variable, into = c("tissue", "variable"), sep = "_", extra = "merge") %>% 
-  mutate(variable = str_remove(variable, "2"))
+  mutate(variable = str_remove(variable, "2")) %>% 
+  mutate(variable = factor(variable, levels = c("hb_tot_alpha", "hb_oxy_alpha")))
 
 # plot
-df_fig2 %>%
+df_fig3 %>%
 
   ggplot(aes(x = status, y = value)) +
-  geom_violin(aes(color = status)) +
+  geom_violin(aes(color = status), fill = "white") +
   geom_jitter(position = position_jitter(0.2), shape = 1, size = 2) +
@@
   facet_grid(variable ~ tissue, scales = "free_y") +
-
-  geom_hline(yintercept = 0.4, lty = 2, alpha = 0.7) +
-  geom_hline(yintercept = 0.6, lty = 2, alpha = 0.7) +
-  geom_hline(yintercept = 1.0, lty = 2, alpha = 0.7) +
-  geom_hline(yintercept = 1.5, lty = 2, alpha = 0.7) +
 
@@
 ```
-
-### Fig 3. Increased cerebral_hb_tot and decreased cerebral_hb_tot_alpha are distinct clinical features of cerebral malaria
-
-```{r fig3_data}
-
-# select model variables of interest
-model_voi <- c("subject_id", "status", 
-               "hct", "lactate",
-               "cerebral_hb_tot", "cerebral_hb_oxy",
-               "cerebral_hb_tot_alpha2", "cerebral_hb_oxy_alpha2")
-
-df_model <- df_hbox %>% 
-  select_at(model_voi) %>%
-
-  # remove HC
-  ungroup %>% 
-  filter(status != "HC")
-
-# impute missing data with median by group
-  f_impute <- function(x, na.rm = TRUE) (replace(x, is.na(x), median(x, na.rm = na.rm)))
-
-  df_model_impute <- df_model %>% 
-    group_by(status) %>% 
-    mutate_at(3:ncol(.), f_impute) %>% 
-    ungroup()
-
-```
-
-#### A. Correlations of variables within subject groups
-
-```{r fig3a, fig.show = "hold", out.width = "50%", fig.height = 12, fig.width = 15}
-
-# calculate spearman correlations in each group
-
-f_calc_cor <- function(df, group) {
-
-  m <- df %>%
-    dplyr::filter(status == group) %>% 
-    dplyr::select(-c(subject_id, status)) %>% 
-    cor(method = "spearman")
-
-  diag(m) <- 0
-
-  m
-}
-
-# create run correlations of potential regressors
-m_cor_um <- df_model_impute %>% f_calc_cor(group = "UM")
-
-m_cor_cm <- df_model_impute %>% f_calc_cor(group = "CM")
-
-# correlation matrix
-
-# write plot function
-  f_plot_cor <- function(mat) {
-
-  melt(mat) %>%
-
-    mutate_if(is.factor, factor, levels = c("hct", "lactate", "cerebral_hb_oxy", "cerebral_hb_tot", "cerebral_hb_oxy_alpha2", "cerebral_hb_tot_alpha2")) %>%
-
-    ggplot(aes(x = X1, y = X2, fill = value, label = ifelse(value != 0, round(value, 2), NA))) + 
-    geom_tile() +
-    scale_fill_gradient2(low = "#4575b4", high = "#d73027") +
-    geom_text(size = 9) +
-
-    xlab("") +
-    ylab("") +
-    theme_bw() +
-    theme(# text = element_text(family = "Arial"),
-          axis.text.x = element_text(angle = 45, hjust = 1),
-          axis.text = element_text(size = 35),
-          legend.text = element_text(size = 40),
-          legend.title = element_text(size = 50),
-          legend.key.size = unit(2, 'cm'),
-          plot.title = element_text(size = 30)
-    )
-
-}
-
-# plot
-
-  # extract legend
-  p_leg <- m_cor_um %>% f_plot_cor() %>% get_legend()
-
-  # create plots
-  m_cor_um %>% f_plot_cor() + ggtitle("Uncomplicated malaria") + theme(legend.position = "none")
-
-  m_cor_cm %>% f_plot_cor() + ggtitle("Cerebral malaria") + theme(legend.position = "none")
-
-  # as_ggplot(p_leg)
-
-```
-
-#### B. Principal component analysis
-
-```{r fig3b}
-
-# convert tibble to df
-df_model_pca <- df_model_impute %>% 
-  dplyr::select(-c(subject_id, status)) %>% 
-  as.data.frame()
-
-# add rownames
-rownames(df_model_pca) <- df_model_impute$subject_id
-
-# compute PCA
-my_pca <- prcomp(df_model_pca, scale = TRUE)
-
-# define groups
-groups <- as.factor(df_model_impute$status)
-
-# Eigenvalues
-pca_eig_val <- get_eigenvalue(my_pca)
-
-# two data frames for plotting
-df_vars <- my_pca$rotation %>% as_tibble() %>% mutate(variable = rownames(my_pca$rotation))
-df_pnts <- my_pca$x %>% 
-  as_tibble() %>% 
-  mutate(status = groups, 
-         subject_id = df_model_impute$subject_id)
-
-# plot
-
-  # plot patients
-ggplot() +
-  geom_point(data = df_pnts, mapping = aes(x = PC1, y = PC2, color = status), size = 3) +
-  stat_conf_ellipse(data = df_pnts, mapping = aes(x = PC1, y = PC2, color = status, fill = status),
-                    level = 0.95, npoint = 100, bary = TRUE, alpha = 0.1, geom = "polygon") +
-  # stat_mean(data = df_pnts, mapping = aes(x = PC1, y = PC2, color = status, shape = status),
-  #           na.rm = FALSE) +
-
-  # View patient IDs
-  # geom_text_repel(data = df_pnts, aes(x = PC1, y = PC2, label = subject_id, size = 10), show.legend = FALSE) +
-
-  # plot variables
-  # geom_point(data = df_vars, aes(x = PC1*5, y = PC2*5)) +
-  geom_text_repel(data = df_vars, aes(x = PC1*5, y = PC2*5, label = variable, size = 10), show.legend = FALSE) +
-  geom_segment(data = df_vars, aes(x = 0, y = 0, xend = PC1*5, yend = PC2*5), arrow = arrow()) +
-
-  # draw lines
-  geom_vline(xintercept = 0, lty = 2, color = "black") +
-  geom_hline(yintercept = 0, lty = 2, color = "black") +
-
-  # design
-  xlab(paste0("PC1: ", round(pca_eig_val$variance.percent[1], 2), "% of variance")) +
-  ylab(paste0("PC2: ", round(pca_eig_val$variance.percent[2], 2), "% of variance")) +
-
-  theme_bw() +
-  theme(# text = element_text(family = "Arial"),
-        axis.text = element_text(size = 20),
-        axis.title = element_text(size = 25),
-        legend.text = element_text(size = 15),
-        legend.title = element_text(size = 20))
-
-```
-
-#### C. Coefficients
-
-```{r model, eval = FALSE}
-
-### shows how logistic regression model was set up and run, results saved in objects/glm_model ###
-
-# resample
-set.seed(234)
-hbox_boot <- bootstraps(df_model, strata = Status, times = 5000)
-
-# recipe
-hbox_rec <- recipe(Status ~ ., data = df_model) %>% 
-  step_impute_knn(all_predictors()) %>% 
-  step_normalize(all_predictors()) %>%
-  step_zv(all_predictors()) %>%
-  step_smote(Status) %>% 
-  update_role(subject_id, new_role = "id")
-
-# create workflow
-hbox_wf <- workflow() %>% add_recipe(hbox_rec)
-
-# logistic regression model specifications
-glm_spec <- logistic_reg() %>% set_engine("glm")
-
-# create model on 5000 different resamples
-glm_res <- hbox_wf %>% 
-  add_model(glm_spec) %>%
-  fit_resamples(
-    resamples = hbox_boot,
-    metrics = metric_set(roc_auc, accuracy, sensitivity, specificity),
-    control = tune::control_resamples(save_pred = TRUE, verbose = TRUE)
-  )
-
-# select best model to be used for evaluation
-glm_best <- glm_res %>% select_best("roc_auc")
-
-# finalize model
-hbox_final <- hbox_wf %>% 
-  add_model(glm_spec) %>% 
-  finalize_workflow(glm_best) %>%
-  fit(df_model) %>%
-  extract_fit_parsnip()
-
-```
-
-```{r table3c}
-
-df_CI <- confint(hbox_final$fit) %>% exp() %>% as_tibble(rownames = "term")
-
-table3c <- hbox_final %>%
-  tidy(exponentiate = TRUE) %>%
-  filter(p.value < 0.05) %>% 
-  left_join(df_CI, by = "term") %>%
-  mutate(estimate = ifelse(estimate < 10^-2, scientific(estimate, digits = 3), round(estimate, 2)),
-         conf.low = ifelse(`2.5 %` < 10^-2, scientific(`2.5 %`, digits = 3), round(`2.5 %`, 2)),
-         conf.high = ifelse(`97.5 %` < 10^-2, scientific(`97.5 %`, digits = 3), round(`97.5 %`, 2))) %>%
-  mutate(p.value = scientific(p.value, digits = 3),
-         statistic = round(statistic, 2)) %>%
-  unite("95% CI", c(conf.low, conf.high), sep = ", ") %>%
-  dplyr::select(c(term, estimate, `95% CI`, statistic, p.value)) %>% 
-  dplyr::rename("Term" = "term", "Odds ratio" = "estimate",
-                "t-statistic" = "statistic", "p-value" = "p.value") %>%
-  as.data.frame()
-
-table3c %>% knitr::kable(align = rep('c', 5))
-
-```
-
-#### D. ROC curve of final model
-
-```{r fig3d_plot, fig.height = 8}
-
-df_roc <- glm_res %>% 
-  collect_predictions() %>% 
-  roc_curve(status, .pred_CM)
-
-ggplot() +
-  geom_path(data = df_roc, 
-            aes(x = 1 - specificity, y = sensitivity), 
-            size = 2.5) +
-  geom_abline(lty = 3) +
-  labs(x = "False positive rate", y = "True positive rate") +
-  theme_bw() +
-
-  theme(# text = element_text(family = "Arial"),
-        axis.title = element_text(size = 25),
-        axis.text = element_text(size = 20),
-        legend.title = element_text(size = 20),
-        legend.text = element_text(size = 15),
-        legend.position = "none")
-
-```
-
-```{r fig3d_inset}
-
-fig3d_inset <- glm_res %>% 
-  collect_metrics() %>% 
-  dplyr::select(.metric, mean, std_err) %>% 
-  mutate(mean = mean*100, std_err = std_err*100) %>% 
-  mutate(Metric = c("Accuracy", "ROC AUC", "Sensitivity", "Specificity")) %>% 
-  dplyr::select(c("Metric", "mean", "std_err")) %>%
-
-  dplyr::rename("Mean" = "mean", "Standard error" = "std_err") %>% 
-  mutate_if(is.numeric, round, 2) %>% 
-  as.data.frame()
-
-fig3d_inset %>% knitr::kable(align = rep('c', 4))
-
-```
-
-### Fig 4. An increase in cerebral_hb_tot is significantly associated with an increase in brain swelling score
-
-```{r fig4_plot}
-
-# select model voi
-brain_swell_voi <- c("brain_swell", "cerebral_hb_tot", "hct")
-df_fig4 <- df_hbox %>% select_at(brain_swell_voi)
-
-# remove cerebral_hb_tot outliers
-  df_fig4_outliers <- df_fig4 %>% 
-    mutate(Q1_hbtot = quantile(cerebral_hb_tot, 0.25, na.rm = TRUE),
-           Q3_hbtot = quantile(cerebral_hb_tot, 0.75, na.rm = TRUE),
-           IQR_hbtot = Q3_hbtot - Q1_hbtot,
-           outlier_hbtot = case_when(
-
-             cerebral_hb_tot < Q1_hbtot - IQR_hbtot ~ 1,
-             cerebral_hb_tot > Q3_hbtot + IQR_hbtot ~ 1,
-             TRUE ~ 0
-
-           )
-
-    ) %>% 
-    filter(outlier_hbtot != 1)
-
-# model
-  lm_fig4 <- lm(brain_swell ~ cerebral_hb_tot + hct, data = df_fig4_outliers)
-  f <- summary(lm_fig4)$fstatistic
-  p <- pf(f[1], f[2], f[3], lower.tail = FALSE)
-
-  model <- "model: brain_swell ~ cerebral_hb_tot + hematocrit"
-  r2 <- paste0("adjusted R^2: ", lm_fig4 %>% summary() %>% .$adj.r.squared %>% round(2))
-  model_p_val <- paste0("model: ", round(p, 3))
-  hb_tot_p_val <- paste0("cerebral_hb_tot: ", lm_fig4 %>% tidy() %>% .[2, "p.value"] %>% .$p.value %>% round(3))
-
-# plot variables
-require(ggpmisc)
-
-df_fig4_outliers %>% 
-  ggplot(aes(x = cerebral_hb_tot, y = brain_swell)) +
-  geom_point(shape = 1, aes(size = hct, color = as.factor(brain_swell))) +
-  #geom_abline(intercept = ) +
-
-  geom_smooth(method = "lm", lty = 2, color = "#696969", alpha = 0.2) +
-  annotate("text", label = model, x = 25, y = 8, hjust = 0, vjust = 1, size = 5) +
-  annotate("text", label = r2, x = 25, y = 7.5, hjust = 0, vjust = 1, size = 5) +
-  annotate("text", label = "p-values", x = 25, y = 7, hjust = 0, vjust = 1, size = 5) +
-  annotate("text", label = model_p_val, x = 35, y = 6.75, hjust = 0, vjust = 1, size = 4) +
-  annotate("text", label = hb_tot_p_val, x = 35, y = 6.5, hjust = 0, vjust = 1, size = 4) +
-
-  lims(x = c(25, 225)) +
-  theme_bw() +
-
-  theme(# text = element_text(family = "Arial"),
-        strip.text = element_text(size = 15),
-        axis.title = element_text(size = 15),
-        axis.title.y = element_text(size = 15),
-        axis.text = element_text(size = 15),
-        legend.title = element_text(size = 12),
-        legend.text = element_text(size = 10))
-
-# 3d plot?  
-# require(plotly)
-#   x <- df_fig4_outliers$cerebral_hb_tot
-#   y <- df_fig4_outliers$hematocrit
-#   z <- df_fig4_outliers$brain_swell
-#   
-#   df_fig4_outliers %>% plot_ly(x = x, y = y, z = z, type = "scatter3d", mode = "markers", color = x)
-
-```
-
-### Fig 5. Explanatory model of how to interpret hemoglobin concentration
-
-See manuscript draft (figure created in PowerPoint)
-
-## Supplement
-### SFig 1. DFA analysis illustration
-
-#### A. Raw signal
-```{r fig_s1a}
-
-df_figs1a <- read.table("Data/signal_segments/Hb_tot/TM0001CM01.txt")
-
-df_figs1a %>%
-
-  ggplot(aes(x = Time)) +
-  geom_line(aes(y = THC)) +
-
-  labs(y = "Hemoglobin concentration (μM)", x = "Time (s)") +
-  ggtitle("Raw signal") +
-  theme_bw() +
-  theme(# text = element_text(family = "Arial"),
-        axis.text = element_text(size = 20),
-        axis.title = element_text(size = 25),
-        legend.text = element_text(size = 15),
-        legend.title = element_text(size = 20))
-
-```
-
-#### B. Processed signal
-```{r fig_s1b}
-
-df_figs1b <- read.table("Data/filtered_signals/0.2s_mean_0.001_1_filt/Hb_tot/TM0001CM01.txt") %>% mutate(time = row_number())
-
-df_figs1b %>% 
-  ggplot(aes(x = time)) +
-  geom_line(aes(y = V1)) +
-
-  lims(y = c(-0.001, 0.001)) +
-
-  labs(y = "Bandpass filtered signal", x = "Time") +
-  ggtitle("Pre-processed signal") +
-  theme_bw() +
-  theme(# text = element_text(family = "Arial"),
-        axis.text = element_blank(),
-        axis.title = element_text(size = 25),
-        legend.text = element_text(size = 15),
-        legend.title = element_text(size = 20))
-
-```
-
-#### C. Cumulative sum
-```{r fig_s1c}
-
-df_figs1c <- df_figs1b %>% 
-  mutate(mean = mean(V1),
-         subtract_mean = V1 - mean,
-         cumsum = cumsum(subtract_mean))
-
-df_figs1c %>%
-
-  ggplot(aes(x = time, y = cumsum)) +
-  geom_line() +
-  geom_vline(xintercept = 1, lty = 2, color = "black") +
-  geom_vline(xintercept = 1000, lty = 2, color = "black") +
-
-  labs(y = "Cumulative sum", x = "Time") +
-  ggtitle("Cumulative signal") +
-  theme_bw() +
-  theme(# text = element_text(family = "Arial"),
-        axis.text = element_blank(),
-        axis.title = element_text(size = 25),
-        legend.text = element_text(size = 15),
-        legend.title = element_text(size = 20))
-
-```
-
-#### D. Window detrend
-```{r fig_s1d, fig.show = "hold", out.width = "50%"}
-
-fit <- lm(cumsum ~ time, data = df_figs1c %>% dplyr::filter(time > 1 & time < 1000))
-
-# Window 
-df_figs1c %>% dplyr::filter(time > 1 & time < 1000) %>%
-
-  ggplot(aes(x = time, y = cumsum)) +
-  geom_line(size = 2) +
-  geom_abline(intercept = coef(fit)[1], slope = coef(fit)[2], color = "black", lty = 2, size = 2) +
-
-  labs(y = "Cumulative sum", x = "Time") +
-  ggtitle("Cumulative signal") +
-  theme_bw() +
-  theme(# text = element_text(family = "Arial"),
-        axis.text.x = element_text(size = 20),
-        axis.text.y = element_blank(),
-        axis.title = element_text(size = 25),
-        legend.text = element_text(size = 15),
-        legend.title = element_text(size = 20),
-        title = element_text(size = 25))
-
-# Detrended
-df_figs1c %>% dplyr::filter(time > 1 & time < 1000) %>%
-
-  ggplot(aes(x = time, y = fit$residuals)) +
-  geom_line(size = 2) +
-  geom_hline(yintercept = 0, color = "black", lty = 2, size = 2) +
-
-  labs(y = "", x = "Time") +
-  ggtitle("Detrended cumulative signal") +
-  theme_bw() +
-  theme(# text = element_text(family = "Arial"),
-        axis.text.x = element_text(size = 20),
-        axis.text.y = element_blank(),
-        axis.title = element_text(size = 28),
-        legend.text = element_text(size = 15),
-        legend.title = element_text(size = 20),
-        title = element_text(size = 25))
-
-```
-
-#### E. Results plot
-```{r fig_s1e}
-
-df_figs1e <- l_dfa_hbtot[[1]][[1]]
-
-# find slope of segmented line
-library(segmented)
-df_figs1e <- df_figs1e %>% mutate(log10_avg_fluct = log10(avg_fluctuation), log10_window_size = log10(window_size))
-fit_figs1e <- lm(log10_avg_fluct ~ log10_window_size, data = df_figs1e) 
-segfit <- segmented(fit_figs1e)
-slopes <- coef(segfit) # the coefficients are the differences in slope in comparison to the previous slope
-
-# first line: 
-#y = b0 + b1*x; y = intercept1 + slope1 * x
-
-# second line:
-# y = c0 + c1*x; y = intercept2 + slope2 * x
-
-# At the breakpoint (break1), the segments b and c intersect: b0 + b1*x = c0 + c1*x
-b0 <- slopes[[1]]
-b1 <- slopes[[2]]
-
-c1 <- slopes[[2]] + slopes[[3]]
-break1 <- segfit$psi[[2]]
-
-# Solve for c0 (intercept of second segment):
-c0 <- b0 + b1 * break1 - c1 * break1
-
-df_figs1e %>%
-
-  ggplot(aes(x = log10_window_size, y = log10_avg_fluct)) +
-  geom_point(size = 3, shape = 1) +
-  geom_abline(intercept = c0, slope = c1, color = "black", lty = 2) +
-  annotate("text", x = -1, y = -2, label = paste0("alpha = ", round(c1, 2)), size = 8, vjust = 1, hjust = 0) +
-
-  labs(x = "log10(window length)", y = "log10(average fluctuation)") +
-  ggtitle("DFA results") +
-
-  theme_bw() +
-  theme(# text = element_text(family = "Arial"),
-        axis.text = element_text(size = 20),
-        axis.title = element_text(size = 25),
-        legend.text = element_text(size = 15),
-        legend.title = element_text(size = 20))
-
-```
-
-#### F. Simulation results
-
-```{r fig_s1f}
-
-wn_alpha <- 1:length(l_wn_dfa) %>% purrr::map_df(~l_wn_dfa[[.x]][[2]]) %>% dplyr::rename("white" = "log10(window_size)")
-pn_alpha <- 1:length(l_pn_dfa) %>% purrr::map_df(~l_pn_dfa[[.x]][[2]]) %>% dplyr::rename("pink" = "log10(window_size)")
-bn_alpha <- 1:length(l_bn_dfa) %>% purrr::map_df(~l_bn_dfa[[.x]][[2]]) %>% dplyr::rename("brown" = "log10(window_size)")
-
-df_figs1f <- bind_cols(wn_alpha, pn_alpha, bn_alpha) %>%
-  pivot_longer(1:3, names_to = "noise", values_to = "alpha")
-
-# create color vector
-my_col <- c("white" = "#00BFC4", "brown" = "#7CAE00", "pink" = "#F8766D")
-
-# plot
-df_figs1f %>%
-
-  ggplot(aes(x = noise, y = alpha)) +
-  geom_violin(aes(color = noise)) +
-  geom_jitter(position = position_jitter(0.2), shape = 1, size = 2) +
-  stat_summary(fun = "median", geom = "crossbar", aes(color = noise), size = 0.2, width = 0.5) +
-
-  geom_hline(yintercept = 0.4, lty = 2, alpha = 0.7) +
-  geom_hline(yintercept = 0.6, lty = 2, alpha = 0.7) +
-  geom_hline(yintercept = 1.0, lty = 2, alpha = 0.7) +
-  geom_hline(yintercept = 1.5, lty = 2, alpha = 0.7) +
-
-  scale_color_manual(values = my_col) +
-  labs(x = "noise type") +
-  scale_y_continuous(limits = c(0, 1.7)) +
-  theme_bw() +
-  theme(legend.position = "none")  +
-
-  theme(# text = element_text(family = "Arial"),
-        strip.text = element_text(size = 15),
-        axis.title = element_text(size = 15),
-        axis.text = element_text(size = 15))
-
-```
-
-### SFig 2. Logistic regression curves
-
-```{r fig_s2, fig.height = 12}
-
-# pivot longer and change factor levels
-df_figs2 <- df_model %>% 
-  mutate(status = as.numeric(ifelse(status == "CM", 1, 0))) %>% 
-  pivot_longer(3:ncol(.), names_to = "variable", values_to = "value") %>% 
-  mutate(variable = factor(variable, levels = c("hct", "lactate", "cerebral_hb_oxy", "cerebral_hb_tot", "cerebral_hb_oxy_alpha2", "cerebral_hb_tot_alpha2")))
-
-# plot as individual regression curves
-df_figs2 %>%
-
-  ggplot(aes(x = value, y = status)) +
-  geom_point() +
-  geom_smooth(method = "glm", method.args = list(family = "binomial"), color = "black") +
-  facet_wrap(~variable, scales = "free_x", nrow = 5, ncol = 2) +
-
-  ylab("Status: 1 = CM, 0 = UM") +
-  xlab("") +
-
-  theme_bw() +
-  theme(# text = element_text(family = "Arial"),
-        axis.text = element_text(size = 20),
-        axis.title = element_text(size = 25),
-        legend.text = element_text(size = 15),
-        legend.title = element_text(size = 20),
-        strip.text = element_text(size = 20))
-
-```
-
-**Figure S2.** Plotted are the individual logistic regression curves of individual variables tested for their discriminatory power between UM and CM.
-On the y-axis, 1 represents CM while 0 represents UM.
-A steeper curve indicates higher discriminatory power.
-
-## Table of manuscript p-values
-
-```{r fig_1_and_2_stats, eval = FALSE}
-
-df_fig1_fig2_stats <-
-
-  df_fig1 %>% 
-    group_by(variable, status) %>% 
-    dplyr::select(value) %>% 
-    summarise_all(funs(median = median, q1 = quantile(., probs = 0.25), q3 = quantile(., probs = 0.75)), na.rm = TRUE) %>% 
-    mutate_if(is.numeric, round, 2) %>%
-
-  bind_rows(
-
-    df_fig2 %>% 
-      group_by(variable, status) %>% 
-      dplyr::select(value) %>% 
-      summarise_all(funs(median = median, q1 = quantile(., probs = 0.25), q3 = quantile(., probs = 0.75)), na.rm = TRUE) %>% 
-      mutate_if(is.numeric, round, 2)
-
-  )
-
-```
-
-```{r manuscript_stats, eval = FALSE}
-
-# select voi
-df_stats <- df_hbox %>% 
-  select(c("status",
-
-           # clinical
-           "age_calc", "temperature", "hr", "bp_sys", "bp_dias", "resp_rate", "pulse_ox", "hct", "lactate", "glucose", "aa_arg",
-
-           # concentrations
-           matches("_hb_tot$|_hb_oxy$"),
-
-           # alphas
-           matches("alpha2")
-
-  ))
-
-# perform kruskal-wallis test on all variables
-df_pvals <- 2:ncol(df_stats) %>% 
-  purrr::map_dfr(~tibble(
-
-    variable = names(df_stats[.x]),
-    kw_pval = kruskal.test(df_stats[[.x]] ~ df_stats$status)$p.value,
-    um_hc = DunnTest(df_stats[[.x]] ~ df_stats$status)[[1]][1, 2],
-    cm_hc = DunnTest(df_stats[[.x]] ~ df_stats$status)[[1]][2, 2],
-    cm_um = DunnTest(df_stats[[.x]] ~ df_stats$status)[[1]][3, 2]
-
-  ))
-
-df_pvals %>% knitr::kable()
-
-# write to csv
-write.csv(df_pvals, file = "Outputs/manuscript_pvals.csv")
-
-```
-
-```{r export_xlsx, eval = FALSE}
-
-require(xlsx)
-
-wb <- createWorkbook()
-
-sheet = createSheet(wb, "Table 1a")
-
-  addDataFrame(df_clinical_res %>% as.data.frame(), sheet = sheet, startColumn = 1, row.names = FALSE)
-
-sheet = createSheet(wb, "Table 1b")
-
-  addDataFrame(df_cm_voi %>% count(bcs_session) %>% as.data.frame(), sheet = sheet, startColumn = 1, row.names = FALSE)
-  addDataFrame(df_cm_voi %>% count(brain_swell) %>% as.data.frame(), sheet = sheet, startColumn = 4, row.names = FALSE)
-  addDataFrame(df_cm_hrp2 %>% as.data.frame(), sheet = sheet, startColumn = 7, row.names = FALSE)
-
-sheet = createSheet(wb, "Figs 1 & 2 stats")
-
-  addDataFrame(df_fig1_fig2_stats %>% as.data.frame(), sheet = sheet, startColumn = 1, row.names = FALSE)
-
-sheet = createSheet(wb, "Table 3c & d")
-
-  addDataFrame(table3c %>% as.data.frame(), sheet = sheet, startColumn = 1, row.names = FALSE)
-  addDataFrame(fig3d_inset %>% as.data.frame(), sheet = sheet, startColumn = 7, row.names = FALSE)
-
-sheet = createSheet(wb, "manuscript p-vals")
-
-  addDataFrame(df_pvals %>% as.data.frame(), sheet = sheet, startColumn = 1, row.names = FALSE)
-
-saveWorkbook(wb, "Outputs/all_manuscript_tables.xlsx")
-
-```

### 2023-06-12 Scripts/R/markdown_files/archive/cm_follow_up_figures.Rmd
12 June 2023 clean and organize scripts for publishing

--- 2021-10-05 cm_follow_up_figures.Rmd
@@
 
-knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, comment = "", fig.height = 6, fig.width = 10)
+knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, comment = "", fig.height = 6, fig.width = 10, fig.align = 'center')
 
@@
 
-load("all_dfa_results.Rdata")
+base_dir <- "/Users/work/Library/CloudStorage/OneDrive-NationalInstitutesofHealth/Ackerman Lab 2019-2021/cerebral_oxygenation/"
+
+load(paste0(base_dir, "objects/all_dfa_results.Rdata"))
 
@@
   dplyr::filter(grepl("avg_Hb", variable)) %>% 
-  mutate(measurement = ifelse(grepl("conc", variable), "THC", "O2_sat")) %>% 
-  mutate(measurement = factor(measurement, levels = c("O2_sat", "THC"))) %>% 
-  mutate(pvals = ifelse(measurement == "THC", round(thc_fu_pval, 3), round(o2sat_fu_pval, 3))) %>% 
-  mutate(ycoord = ifelse(measurement == "THC", 400, 100)) %>%
+  mutate(measurement = ifelse(grepl("conc", variable), "Hbtot", "Hboxy")) %>% 
+  mutate(measurement = factor(measurement, levels = c("Hboxy", "Hbtot"))) %>% 
+  mutate(pvals = ifelse(measurement == "Hbtot", round(thc_fu_pval, 3), round(o2sat_fu_pval, 3))) %>% 
+  mutate(ycoord = ifelse(measurement == "Hbtot", 400, 100)) %>%
 
@@
   dplyr::filter(!grepl("avg|second", variable)) %>% 
-  mutate(measurement = ifelse(grepl("conc", variable), "THC", "O2_sat")) %>% 
-  mutate(measurement = factor(measurement, levels = c("O2_sat", "THC"))) %>% 
-  mutate(pvals = ifelse(measurement == "THC", round(thc_alpha_fu_pval, 3), round(o2sat_alpha_fu_pval, 3))) %>%
+  mutate(measurement = ifelse(grepl("conc", variable), "Hbtot", "Hboxy")) %>% 
+  mutate(measurement = factor(measurement, levels = c("Hboxy", "Hbtot"))) %>% 
+  mutate(pvals = ifelse(measurement == "Hbtot", round(thc_alpha_fu_pval, 3), round(o2sat_alpha_fu_pval, 3))) %>%
 

### 2023-06-12 Scripts/R/markdown_files/archive/fig3a_ideas.Rmd
12 June 2023 clean and organize scripts for publishing

---
title: "fig3a"
output: pdf_document
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, comment = "", fig.height = 6, fig.width = 10)

# libraries
library(tidyverse)
library(psych)
library(stringr)

```

```{r data}

# read in data

df_hbox <- read_csv("Data/Hans_datatable_exports/malawi key data v04Jan2022.csv") %>% 
  as_tibble() %>% 
  janitor::clean_names() %>% 
  mutate(status = replace(status, status == "HV", "HC")) %>% 
  dplyr::rename("subject_id" = "subject_id_session") %>%

  dplyr::filter(status %in% c("CM", "HC", "UM") & session_no == 1 | subject_id == "TM0003CM01") %>% 
  dplyr::filter(!grepl("blood", subject_id)) %>% 
  mutate(status = factor(status, levels = c("HC", "UM", "CM"))) %>% 
  mutate_at(c("bp_dias", "bp_sys", "glucose", "lactate"), as.numeric) %>% # converts remaining character variables to numerics
  select(subject_id, status, everything()) %>% 
  group_by(status)

```

```{r fig3_data}

# select model variables of interest
model_voi <- c("subject_id", "status", 
               "hct", "lactate",
               "cerebral_hb_tot", "cerebral_hb_oxy",
               "cerebral_hb_tot_alpha2", "cerebral_hb_oxy_alpha2")

df_model <- df_hbox %>% 
  select_at(model_voi) %>%

  # remove HC
  ungroup %>% 
  filter(status != "HC")

# impute missing data with median by group
  f_impute <- function(x, na.rm = TRUE) (replace(x, is.na(x), median(x, na.rm = na.rm)))

  df_model_impute <- df_model %>% 
    group_by(status) %>% 
    mutate_at(3:ncol(.), f_impute) %>% 
    ungroup()

```

```{r fig3a_cors}

# find pearson correlation across all variables of interest, including p-values and adjusted p-values
cm_cor <- df_model %>% 
  dplyr::filter(status == "CM") %>% 
  dplyr::select(-c(subject_id, status)) %>% 
  corr.test(method = "pearson", adjust = "none")

um_cor <- df_model %>% 
  dplyr::filter(status == "UM") %>% 
  dplyr::select(-c(subject_id, status)) %>% 
  corr.test(method = "pearson", adjust = "none")

# combine into tibble
df_fig3a <- cm_cor$r %>% 
  as_tibble %>% 
  mutate(var1 = colnames(.), .before = hct) %>% 
  pivot_longer(2:ncol(.), names_to = "var2", values_to = "pearsons_r") %>% 
  mutate(status = "CM", .before = var1) %>%

  left_join(

    cm_cor$p %>% 
      as_tibble %>% 
      mutate(var1 = colnames(.), .before = hct) %>% 
      pivot_longer(2:ncol(.), names_to = "var2", values_to = "p_value"),

    by = c("var1", "var2")

  ) %>%

  # adjust p-values
  mutate(adj_p_val = p.adjust(p_value, method = "BH")) %>%

  bind_rows(

    um_cor$r %>% 
      as_tibble %>% 
      mutate(var1 = colnames(.), .before = hct) %>% 
      pivot_longer(2:ncol(.), names_to = "var2", values_to = "pearsons_r") %>% 
      mutate(status = "UM", .before = var1) %>%

      left_join(

        um_cor$p %>% 
          as_tibble %>% 
          mutate(var1 = colnames(.), .before = hct) %>% 
          pivot_longer(2:ncol(.), names_to = "var2", values_to = "p_value"),

        by = c("var1", "var2")

      ) %>% 
      mutate(adj_p_val = p.adjust(p_value, method = "BH"))

  ) %>%

  # set factor levels for variables
  mutate_at(2:3, factor, levels = c("hct", "lactate", "cerebral_hb_oxy", "cerebral_hb_tot", "cerebral_hb_oxy_alpha2", "cerebral_hb_tot_alpha2")) %>%

  # set diagonal equal to zero
  mutate(pearsons_r = ifelse(var1 == var2, 0, pearsons_r))

```

### idea 1: put box around values significant by p-value
```{r fig3a_v1}

df_fig3a %>% 
  mutate(is_sig = ifelse(p_value < 0.05 & var1 != var2, "yes", "no")) %>%

  ggplot(aes(x = var1, y = var2, fill = pearsons_r, label = ifelse(pearsons_r != 0, round(pearsons_r, 2), NA))) + 
  geom_tile(aes(width = 0.9, height = 0.9)) +
  geom_tile(aes(color = is_sig, width = 0.9, height = 0.9), size = 1, show.legend = FALSE) +

  facet_wrap(~ status) +

  scale_color_manual(values = c("white", "black")) +
  scale_fill_gradient2(low = "#4575b4", high = "#d73027") +
  geom_text(size = 4) +

  xlab("") +
  ylab("") +
  theme_classic() +
  theme(axis.text = element_text(size = 12),
        axis.text.x = element_text(angle = 45, hjust = 0.95))

```

### idea 2: put box around values significant by BH-adjusted p-value
```{r fig3a_v2}

df_fig3a %>% 
  mutate(is_sig = ifelse(adj_p_val < 0.05 & var1 != var2, "yes", "no")) %>%

  ggplot(aes(x = var1, y = var2, fill = pearsons_r, label = ifelse(pearsons_r != 0, round(pearsons_r, 2), NA))) + 
  geom_tile(aes(width = 0.9, height = 0.9)) +
  geom_tile(aes(color = is_sig, width = 0.9, height = 0.9), size = 1, show.legend = FALSE) +

  facet_wrap(~ status) +

  scale_color_manual(values = c("white", "black")) +
  scale_fill_gradient2(low = "#4575b4", high = "#d73027") +
  geom_text(size = 4) +

  xlab("") +
  ylab("") +
  theme_classic() +
  theme(axis.text = element_text(size = 12),
        axis.text.x = element_text(angle = 45, hjust = 0.95))

```

### idea 3: add p-value to box label
```{r fig3a_v3}

df_fig3a %>% 
  mutate(label = paste0(round(pearsons_r, 2), " ", round(p_value, 2))) %>% 
  mutate(label = ifelse(var1 == var2, NA, label)) %>% 
  mutate(label = str_wrap(label, width = 2)) %>%

  ggplot(aes(x = var1, y = var2, fill = pearsons_r, label = label)) + 
  geom_tile(aes(width = 0.95, height = 0.95)) +

  facet_wrap(~ status) +

  scale_fill_gradient2(low = "#4575b4", high = "#d73027") +
  geom_text(size = 3) +

  xlab("") +
  ylab("") +
  theme_classic() +
  theme(axis.text = element_text(size = 12),
        axis.text.x = element_text(angle = 45, hjust = 0.95))

```

### idea 3: add BH-adjusted p-value to box label
```{r fig3a_v4}

df_fig3a %>% 
  mutate(label = paste0(round(pearsons_r, 2), " ", round(adj_p_val, 2))) %>% 
  mutate(label = ifelse(var1 == var2, NA, label)) %>% 
  mutate(label = str_wrap(label, width = 2)) %>%

  ggplot(aes(x = var1, y = var2, fill = pearsons_r, label = label)) + 
  geom_tile(aes(width = 0.95, height = 0.95)) +

  facet_wrap(~ status) +

  scale_fill_gradient2(low = "#4575b4", high = "#d73027") +
  geom_text(size = 3) +

  xlab("") +
  ylab("") +
  theme_classic() +
  theme(axis.text = element_text(size = 12),
        axis.text.x = element_text(angle = 45, hjust = 0.95))

```

### 2023-06-12 Scripts/R/markdown_files/archive/final_figures.Rmd
12 June 2023 clean and organize scripts for publishing

--- 2022-02-14 final_figures.Rmd
@@
 library(grDevices)
+library(psych)
 
 # Load data
-load("objects/glm_model.Rdata")
-load("objects/bandpass_filtered_dfa_results.Rdata")
-load("objects/sims.Rdata")
+base_dir <- "/Users/ray/Library/CloudStorage/OneDrive-NationalInstitutesofHealth/Ackerman Lab 2019-2021/cerebral_oxygenation/"
+
+load(paste0(base_dir, "objects/20221125_final_glm_models.Rdata"))
+load(paste0(base_dir, "objects/glm_model.Rdata"))
+load(paste0(base_dir, "objects/brain_bandpass_filtered_dfa_results.Rdata"))
+load(paste0(base_dir, "objects/sims.Rdata"))
+
+# ALT LOADING: these objects are located in a directory 3 levels above, i.e.:
+load("../../../objects/20221125_final_glm_models.Rdata")
 
@@
 
-df_hbox <- read_csv("Data/Hans_datatable_exports/malawi key data v04Jan2022.csv") %>% 
+df_hbox <- read_csv(paste0(base_dir, "Data/Hans_datatable_exports/malawi key data v04Jan2022.csv")) %>% 
   as_tibble() %>% 
@@
   count() %>% 
-  pivot_wider(1:2, names_from = "status", values_from = "n") %>% 
+  pivot_wider(names_from = "status", values_from = "n") %>% 
   mutate(variable = "Number of subjects") %>% 
@@
   count(sex) %>% 
-  dplyr::filter(sex != "") %>% 
-  pivot_wider(names_from = "sex", values_from = "n") %>% 
+  pivot_wider(id_cols = "status", names_from = "sex", values_from = "n") %>% 
   mutate(perc_fem = round(Female/(Female + Male)*100, 2)) %>% 
-
-  pivot_wider(c(1,4), names_from = "status", values_from = "perc_fem") %>% 
+  dplyr::select(-c(Female, Male)) %>%
+
+  pivot_wider(names_from = "status", values_from = "perc_fem") %>% 
   mutate(variable = "Sex (% female)") %>% 
@@
   select(cm_voi[3]) %>% 
-  summarise_all(funs(median = median, q1 = quantile(., probs = 0.25), q3 = quantile(., probs = 0.75)), na.rm = TRUE) %>% 
+  summarise_all(funs(median = median, 
+                     q1 = quantile(., probs = 0.25), 
+                     q3 = quantile(., probs = 0.75)), 
+                na.rm = TRUE) %>% 
   unite(hrp2, 1:2, sep = " (") %>% 
@@
 #### A. Correlations of variables within subject groups
-```{r fig3a, fig.show = "hold", out.width = "50%", fig.height = 12, fig.width = 15}
-
-# calculate spearman correlations in each group
-
-f_calc_cor <- function(df, group) {
-
-  m <- df %>%
-    dplyr::filter(status == group) %>% 
+```{r fig3a_cors}
+
+# find pearson correlation across all variables of interest, including p-values and adjusted p-values
+
+cm_cor <- df_model %>% 
+  dplyr::filter(status == "CM") %>% 
   dplyr::select(-c(subject_id, status)) %>% 
-    cor(method = "spearman")
-
-  diag(m) <- 0
-
-  m
-}
-
-# create run correlations of potential regressors
-m_cor_um <- df_model_impute %>% f_calc_cor(group = "UM")
-
-m_cor_cm <- df_model_impute %>% f_calc_cor(group = "CM")
-
-# correlation matrix
-
-# write plot function
-  f_plot_cor <- function(mat) {
-
-  melt(mat) %>%
-
-    mutate_if(is.factor, factor, levels = c("hct", "lactate", "cerebral_hb_oxy", "cerebral_hb_tot", "cerebral_hb_oxy_alpha2", "cerebral_hb_tot_alpha2")) %>%
-
-    ggplot(aes(x = X1, y = X2, fill = value, label = ifelse(value != 0, round(value, 2), NA))) + 
-    geom_tile() +
+  corr.test(method = "pearson", adjust = "none")
+
+um_cor <- df_model %>% 
+  dplyr::filter(status == "UM") %>% 
+  dplyr::select(-c(subject_id, status)) %>% 
+  corr.test(method = "pearson", adjust = "none")
+
+# combine into tibble
+df_fig3a <- cm_cor$r %>% 
+  as_tibble %>% 
+  mutate(var1 = colnames(.), .before = hct) %>% 
+  pivot_longer(2:ncol(.), names_to = "var2", values_to = "pearsons_r") %>% 
+  mutate(status = "CM", .before = var1) %>%
+
+  left_join(
+
+    cm_cor$p %>% 
+      as_tibble %>% 
+      mutate(var1 = colnames(.), .before = hct) %>% 
+      pivot_longer(2:ncol(.), names_to = "var2", values_to = "p_value"),
+
+    by = c("var1", "var2")
+
+  ) %>%
+
+  # adjust p-values
+  mutate(adj_p_val = p.adjust(p_value, method = "BH")) %>%
+
+  bind_rows(
+
+    um_cor$r %>% 
+      as_tibble %>% 
+      mutate(var1 = colnames(.), .before = hct) %>% 
+      pivot_longer(2:ncol(.), names_to = "var2", values_to = "pearsons_r") %>% 
+      mutate(status = "UM", .before = var1) %>%
+
+      left_join(
+
+        um_cor$p %>% 
+          as_tibble %>% 
+          mutate(var1 = colnames(.), .before = hct) %>% 
+          pivot_longer(2:ncol(.), names_to = "var2", values_to = "p_value"),
+
+        by = c("var1", "var2")
+
+      ) %>% 
+      mutate(adj_p_val = p.adjust(p_value, method = "BH"))
+
+  ) %>%
+
+  # set factor levels for variables
+  mutate_at(2:3, factor, levels = c("hct", "lactate", "cerebral_hb_oxy", "cerebral_hb_tot", "cerebral_hb_oxy_alpha2", "cerebral_hb_tot_alpha2")) %>%
+
+  # set diagonal equal to zero
+  mutate(pearsons_r = ifelse(var1 == var2, 0, pearsons_r))
+
+```
+
+```{r fig3a_sig_cors}
+
+### find lowest |r| value that corresponds to a p < 0.05
+
+t_stat_cor <- function(r, n){(r*sqrt(n - 2)) / (sqrt(1 - r^2))} # equation to calculate t-stat given correlation and n
+
+## for CM:
+
+# combine pearson's r and sample size
+df_t_cor_cm <- cm_cor$r %>% 
+  as_tibble %>% 
+  mutate(var1 = colnames(.), .before = hct) %>% 
+  pivot_longer(2:ncol(.), names_to = "var2", values_to = "pearsons_r") %>%
+
+  left_join(
+
+    cm_cor$n %>% 
+      as_tibble %>% 
+      mutate(var1 = colnames(.), .before = hct) %>% 
+      pivot_longer(2:ncol(.), names_to = "var2", values_to = "n")
+
+  ) %>%
+
+  # calculate t-stat of each
+  mutate(t = ifelse(pearsons_r != 1, t_stat_cor(pearsons_r, n), NA))
+
+# compare with t-distribution table
+lowest_sig_r_cm <- read_excel(paste0(base_dir, "Data/t_distribution_table.xlsx")) %>% 
+  dplyr::rename("n" = "One Tail") %>% 
+  filter(!(n %in% c("Two Tails", "df"))) %>% 
+  dplyr::select(1:2) %>% 
+  mutate(n = as.numeric(n)) %>%
+
+  left_join(df_t_cor_cm, by = "n") %>% 
+  filter(!is.na(t)) %>%
+
+  # find significance by t-value
+  mutate(is_sig = ifelse(abs(t) > `0.05`, "yes", "no")) %>%
+
+  # pull only significant correlations
+  filter(is_sig == "yes") %>%
+
+  # find lowest r value associated with significant correlation
+  slice(which.min(abs(pearsons_r))) %>% pull(pearsons_r)
+
+## repeat for UM:
+
+# combine pearson's r and sample size
+df_t_cor_um <- um_cor$r %>% 
+  as_tibble %>% 
+  mutate(var1 = colnames(.), .before = hct) %>% 
+  pivot_longer(2:ncol(.), names_to = "var2", values_to = "pearsons_r") %>%
+
+  left_join(
+
+    um_cor$n %>% 
+      as_tibble %>% 
+      mutate(var1 = colnames(.), .before = hct) %>% 
+      pivot_longer(2:ncol(.), names_to = "var2", values_to = "n")
+
+  ) %>%
+
+  # calculate t-stat of each
+  mutate(t = ifelse(pearsons_r != 1, t_stat_cor(pearsons_r, n), NA))
+
+# compare with t-distribution table
+lowest_sig_r_um <-  read_excel(paste0(base_dir, "Data/t_distribution_table.xlsx")) %>% 
+  dplyr::rename("n" = "One Tail") %>% 
+  filter(!(n %in% c("Two Tails", "df"))) %>% 
+  dplyr::select(1:2) %>% 
+  mutate(n = as.numeric(n)) %>%
+
+  left_join(df_t_cor_um, by = "n") %>% 
+  filter(!is.na(t)) %>%
+
+  # find significance by t-value
+  mutate(is_sig = ifelse(abs(t) > `0.05`, "yes", "no")) %>%
+
+  # pull only significant correlations
+  filter(is_sig == "yes") %>%
+
+  # find lowest r value associated with significant correlation
+  slice(which.min(abs(pearsons_r))) %>% pull(pearsons_r)
+
+```
+
+```{r fig3a}
+
+df_fig3a %>%
+
+  ggplot(aes(x = var1, y = var2, fill = pearsons_r, label = ifelse(pearsons_r != 0, round(pearsons_r, 2), NA))) + 
+  geom_tile(aes(width = 0.9, height = 0.9)) +
+
+  facet_wrap(~ status) +
+
   scale_fill_gradient2(low = "#4575b4", high = "#d73027") +
-    geom_text(size = 9) +
+  geom_text(size = 4) +
 
@@
   ylab("") +
-    theme_bw() +
-    theme(# text = element_text(family = "Arial"),
-          axis.text.x = element_text(angle = 45, hjust = 1),
-          axis.text = element_text(size = 35),
-          legend.text = element_text(size = 40),
-          legend.title = element_text(size = 50),
-          legend.key.size = unit(2, 'cm'),
-          plot.title = element_text(size = 30)
-    )
-
-}
-
-# plot
-
-  # extract legend
-  p_leg <- m_cor_um %>% f_plot_cor() %>% get_legend()
-
-  # create plots
-  m_cor_um %>% f_plot_cor() + ggtitle("Uncomplicated malaria") + theme(legend.position = "none")
-
-  m_cor_cm %>% f_plot_cor() + ggtitle("Cerebral malaria") + theme(legend.position = "none")
-
-  # as_ggplot(p_leg)
-
-```
+  theme_classic() +
+  theme(axis.text = element_text(size = 12),
+        axis.text.x = element_text(angle = 45, hjust = 0.95))
+
+```
+In the CM group (n = 46), absolute correlation values greater 0.32 are statistically significant at a threshold of p < 0.05.
+In the UM group (n = 36), absolute correlation values greater than the absolute value of 0.43 are statistically significant.
 \newpage
@@
   geom_point(data = df_pnts, mapping = aes(x = PC1, y = PC2, color = status), size = 3) +
-  stat_conf_ellipse(data = df_pnts, mapping = aes(x = PC1, y = PC2, color = status, fill = status),
-                    level = 0.95, npoint = 100, bary = TRUE, alpha = 0.1, geom = "polygon") +
+  # stat_conf_ellipse(data = df_pnts, mapping = aes(x = PC1, y = PC2, color = status, fill = status),
+  #                   level = 0.95, npoint = 100, bary = TRUE, alpha = 0.1, geom = "polygon") +
   # stat_mean(data = df_pnts, mapping = aes(x = PC1, y = PC2, color = status, shape = status),
@@
 
-df_brain_swell_models %>% write.csv(file = "outputs/brain_swell_models.csv")
+df_brain_swell_models %>% write.csv(file = paste0("outputs/brain_swell_models.csv"))
 
@@
 
-df_figs1a <- read.table("Data/signal_segments/Hb_tot/TM0001CM01.txt")
+df_figs1a <- read.table(paste0(base_dir, "Data/signal_segments/Hb_tot/TM0001CM01.txt"))
 
@@
 
-df_figs1b <- read.table("Data/filtered_signals/0.2s_mean_0.001_1_filt/Hb_tot/TM0001CM01.txt") %>% mutate(time = row_number())
+df_figs1b <- read.table(paste0(base_dir, "Data/filtered_signals/0.2s_mean_0.001_1_filt/Hb_tot/TM0001CM01.txt")) %>% 
+  mutate(time = row_number())
 
@@
 
-  lims(y = c(-0.001, 0.001)) +
+  #lims(y = c(-0.001, 0.001)) +
 
@@
   theme(# text = element_text(family = "Arial"),
-        axis.text = element_blank(),
+        #axis.text = element_blank(),
         axis.title = element_text(size = 25),
@@
 # create color vector
-my_col <- c("white" = "#00BFC4", "brown" = "#7CAE00", "pink" = "#F8766D")
+my_col <- c("white" = "white", "brown" = "sienna2", "pink" = "pink")
 
@@
 
-  ggplot(aes(x = noise, y = alpha)) +
-  geom_violin(aes(color = noise)) +
-  geom_jitter(position = position_jitter(0.2), shape = 1, size = 2) +
+  ggplot(aes(x = noise, y = alpha, color = noise)) +
+  geom_violin(fill = "transparent") +
+  geom_jitter(position = position_jitter(width = 0.1), shape = 1, size = 2) +
   stat_summary(fun = "median", geom = "crossbar", aes(color = noise), size = 0.2, width = 0.5) +
 
-  geom_hline(yintercept = 0.4, lty = 2, alpha = 0.7) +
-  geom_hline(yintercept = 0.6, lty = 2, alpha = 0.7) +
-  geom_hline(yintercept = 1.0, lty = 2, alpha = 0.7) +
-  geom_hline(yintercept = 1.5, lty = 2, alpha = 0.7) +
-
   scale_color_manual(values = my_col) +
-  labs(x = "noise type") +
+  labs(x = "noise type", y = "alpha (\u03b1)") +
   scale_y_continuous(limits = c(0, 1.7)) +
-  theme_bw() +
+  theme_dark() +
   theme(legend.position = "none")  +

### 2023-06-12 Scripts/R/markdown_files/archive/tidymodels_analysis.Rmd
12 June 2023 clean and organize scripts for publishing

--- 2021-05-04 tidymodels_analysis.Rmd
@@
 #### Biplot
-```{r pca_biplot, fig.width = 12, fig.height = 12}
+```{r pca_biplot, fig.width = 10, fig.height = 10}
 
@@
   xlab(paste0("PC1: ", round(pca_eig_val$variance.percent[1], 2), "% of variance")) +
-  ylab(paste0("PC1: ", round(pca_eig_val$variance.percent[2], 2), "% of variance")) +
+  ylab(paste0("PC2: ", round(pca_eig_val$variance.percent[2], 2), "% of variance")) +
 

### 2023-06-12 Scripts/MatLab/plot_bandpass.m

@@ -1,7 +1,12 @@
-function bp = bandpass_filter(data, mean_length, cut, low_thresh, high_thresh)
+signal_path = "Data/signal_segments/Hb_tot/TM0001CM01.txt";
 
-    %% read in data
-    data = importdata(data);
+mean_length = 0.2;
+cut = 120;
+low_thresh = 0.001;
+high_thresh = 1;
+
+ %% read in data
+    data = importdata(signal_path);
     variable = data.data(:,2);
     
     %% average and cut
@@ -13,7 +18,4 @@ function bp = bandpass_filter(data, mean_length, cut, low_thresh, high_thresh)
     variable_movmean_cut = variable_movmean((cut/mean_length):(numel(variable_movmean) - (cut/mean_length))); % take s seconds off each side
 
     %% filter
-    bp = bpfilt(variable_movmean_cut, low_thresh, high_thresh, Fs, false);
-    
-
-
+    bp = bpfilt(variable_movmean_cut, low_thresh, high_thresh, Fs);

### 2023-06-12 Scripts/R/20221115_revise_model.R

@@ -1,0 +1,290 @@
+
+
+#### Plot correlations of clinical variables across each patient group to determine what variables to include in the LR model
+
+
+# libraries -----------------------------------------------------------------
+
+  
+  library(tidyverse)
+  library(RColorBrewer)
+  library(gridExtra)
+  library(patchwork)
+  library(Hmisc)
+  library(corrr)
+
+
+# set theme for plots -----------------------------------------------------
+
+  theme_set(theme_bw() +
+              theme(plot.title = element_text(size = 18),
+                    axis.title = element_text(size = 15),
+                    axis.text = element_text(size = 12),
+                    strip.text = element_text(size = 15),
+                    legend.title = element_text(size = 15),
+                    legend.text = element_text(size = 12)))
+  
+
+
+
+# data --------------------------------------------------------------------
+
+
+  df_hbox <- read_csv("Data/Hans_datatable_exports/malawi key data v04Jan2022.csv") %>% 
+    as_tibble() %>% 
+    janitor::clean_names() %>% 
+    mutate(status = replace(status, status == "HV", "HC")) %>% 
+    dplyr::rename("subject_id" = "subject_id_session") %>% 
+    
+    dplyr::filter(status %in% c("CM", "HC", "UM") & session_no == 1 | subject_id == "TM0003CM01") %>% 
+    dplyr::filter(!grepl("blood", subject_id)) %>% 
+    mutate(status = factor(status, levels = c("HC", "UM", "CM"))) %>% 
+    mutate_at(c("bp_dias", "bp_sys", "glucose", "lactate"), as.numeric) %>% # converts remaining character variables to numerics
+    select(subject_id, status, everything()) %>% 
+    group_by(status)
+
+
+# format data -------------------------------------------------------------
+
+
+# define clinical variables of interest
+  clinical_voi <- c("age_calc", "temperature", "hr", "bp_sys", "bp_dias", "resp_rate", "pulse_ox", "hct", "lactate", "glucose", "aa_arg")
+  hbox_voi <- c("cerebral_hb_tot", "cerebral_hb_tot_alpha2")
+
+# select clinical VOI
+  df_model <- df_hbox %>% 
[…]

### 2023-06-12 Scripts/R/archive/tidymodels_analysis.R

@@ -1,5 +1,7 @@
 
 
+### Run the LR model on variables of interest to identify variables that drive discrimination between UM and CM
+
 
 # Libraries ---------------------------------------------------------------
 

### 2023-06-12 Scripts/R/brain.swell_and_dev.cog.R

@@ -1,5 +1,7 @@
 
 
+### Correlate brain swelling score with different metrics of cognitive development
+
 
 # libraries ---------------------------------------------------------------
 
@@ -12,7 +14,7 @@
 # data --------------------------------------------------------------------
 
 
-  df_all_data <- read_xlsx("Data/malawi key data v13Dec2021.xlsx") %>% as_tibble() 
+  df_all_data <- read_xlsx("Data/Hans/datatable_exports/malawi key data v13Dec2021.xlsx") %>% as_tibble() 
   df_alpha_data <- read.csv("Data/master_datatable.csv") %>% as_tibble() %>% select(-X)
   
   

### 2023-06-12 Scripts/R/combining_data.R

@@ -1,5 +1,6 @@
 
 
+### Generate master datatable by combining data across spreadsheets
 
 
 # libraries ---------------------------------------------------------------
@@ -164,13 +165,4 @@
   write.csv(df_master, file = "Data/master_datatable.csv")
 
 
-  
-  
-
-
-
-
-
-  
-  
   

### 2023-06-12 Scripts/R/filtered_dfa.R

@@ -1,5 +1,9 @@
 
 
+
+### Run DFA on signals that have been filtered using a bandpass filter of 0.001 - 1 Hz
+
+
 # libraries ---------------------------------------------------------------
 
 

### 2023-06-12 Scripts/R/final_figures/fig1.R

@@ -1,0 +1,56 @@
+
+### Code to generate Figure 1: cerebral and muscle Hb_tot and Hb_oxy in each patient group
+
+
+# libraries ---------------------------------------------------------------
+
+library(tidyverse)
+
+# data load & format --------------------------------------------------------------------
+
+df_hbox <- read_csv("Data/Hans_datatable_exports/malawi key data v04Jan2022.csv") %>% 
+  as_tibble() %>% 
+  janitor::clean_names() %>% 
+  mutate(status = replace(status, status == "HV", "HC")) %>% 
+  dplyr::rename("subject_id" = "subject_id_session") %>% 
+  
+  dplyr::filter(status %in% c("CM", "HC", "UM") & session_no == 1 | subject_id == "TM0003CM01") %>% 
+  dplyr::filter(!grepl("blood", subject_id)) %>% 
+  mutate(status = factor(status, levels = c("HC", "UM", "CM"))) %>% 
+  mutate_at(c("bp_dias", "bp_sys", "glucose", "lactate"), as.numeric) %>% # converts remaining character variables to numerics
+  select(subject_id, status, everything()) %>% 
+  group_by(status)
+
+
+# plot --------------------------------------------------------------------
+
+# EXTRACT HB_TOT AND HB_OXY FOR THE BRAIN AND MUSCLE
+df_fig1 <- df_hbox %>% 
+  select(contains(c("hb_tot","hb_oxy")), -contains(c("alpha", "bsl"))) %>%
+  ungroup() %>% 
+  pivot_longer(2:ncol(.), names_to = "variable", values_to = "value") %>% 
+  separate(variable, into = c("tissue", "variable"), sep = "_", extra = "merge")
+
+# CREATE COLOR VECTOR
+my_col <- c("CM" = "#00BFC4", "HC" = "#7CAE00", "UM" = "#F8766D")
+
+# PLOT
+df_fig1 %>% 
+  
+  ggplot(aes(x = status, y = value)) +
+  geom_violin(aes(color = status)) +
+  geom_jitter(position = position_jitter(0.2), shape = 1, size = 2) +
+  stat_summary(fun = "median", geom = "crossbar", aes(color = status), size = 0.2, width = 0.5) +
+  facet_grid(variable ~ tissue, scales = "free_y") +
+  
+  scale_color_manual(values = my_col) +
+  labs(x = "") +
+  theme_bw() +
+  theme(legend.position = "none")  +
+  
[…]

### 2023-06-12 Scripts/R/final_figures/fig2.R

@@ -1,0 +1,45 @@
+
+### Code to generate Figure 2: correlation between brain swell score and Hb_tot
+
+
+# libraries ---------------------------------------------------------------
+
+library(tidyverse)
+
+# data load & format --------------------------------------------------------------------
+
+df_hbox <- read_csv("Data/Hans_datatable_exports/malawi key data v04Jan2022.csv") %>% 
+  as_tibble() %>% 
+  janitor::clean_names() %>% 
+  mutate(status = replace(status, status == "HV", "HC")) %>% 
+  dplyr::rename("subject_id" = "subject_id_session") %>% 
+  
+  dplyr::filter(status %in% c("CM", "HC", "UM") & session_no == 1 | subject_id == "TM0003CM01") %>% 
+  dplyr::filter(!grepl("blood", subject_id)) %>% 
+  mutate(status = factor(status, levels = c("HC", "UM", "CM"))) %>% 
+  mutate_at(c("bp_dias", "bp_sys", "glucose", "lactate"), as.numeric) %>% # converts remaining character variables to numerics
+  select(subject_id, status, everything()) %>% 
+  group_by(status)
+
+
+# plot --------------------------------------------------------------------
+
+df_hbox %>%
+  
+  # log10 transform data that are not normal
+  mutate(log10_cerebral_hb_tot = log10(cerebral_hb_tot)) %>% 
+  
+  # plot
+  ggplot(aes(x = log10_cerebral_hb_tot, y = brain_swell)) +
+  geom_point(aes(size = hct), shape = 1) +
+  
+  geom_smooth(method = "lm", lty = 2, color = "#696969", alpha = 0.2, se = FALSE) +
+  theme_bw() +
+  ggtitle("1. Relationship between brain swell score and log10-transformed cerebral Hb_tot") +
+  theme(strip.text = element_text(size = 15),
+        axis.title = element_text(size = 15),
+        axis.title.y = element_text(size = 15),
+        axis.text = element_text(size = 15),
+        legend.title = element_text(size = 12),
+        legend.text = element_text(size = 10))
+

### 2023-06-12 Scripts/R/final_figures/fig3.R

@@ -1,0 +1,58 @@
+
+### Code to generate Figure 1: cerebral and muscle Hb_tot alpha and Hb_oxy alpha in each patient group
+
+
+# libraries ---------------------------------------------------------------
+
+library(tidyverse)
+
+# data load & format --------------------------------------------------------------------
+
+df_hbox <- read_csv("Data/Hans_datatable_exports/malawi key data v04Jan2022.csv") %>% 
+  as_tibble() %>% 
+  janitor::clean_names() %>% 
+  mutate(status = replace(status, status == "HV", "HC")) %>% 
+  dplyr::rename("subject_id" = "subject_id_session") %>% 
+  
+  dplyr::filter(status %in% c("CM", "HC", "UM") & session_no == 1 | subject_id == "TM0003CM01") %>% 
+  dplyr::filter(!grepl("blood", subject_id)) %>% 
+  mutate(status = factor(status, levels = c("HC", "UM", "CM"))) %>% 
+  mutate_at(c("bp_dias", "bp_sys", "glucose", "lactate"), as.numeric) %>% # converts remaining character variables to numerics
+  select(subject_id, status, everything()) %>% 
+  group_by(status)
+
+
+# plot --------------------------------------------------------------------
+
+# EXTRACT HB_TOT ALPHA AND HB_OXY ALPHA FOR THE BRAIN AND MUSCLE
+df_fig2 <- df_hbox %>% 
+  select(contains("alpha2")) %>%
+  ungroup() %>% 
+  pivot_longer(2:ncol(.), names_to = "variable", values_to = "value") %>% 
+  separate(variable, into = c("tissue", "variable"), sep = "_", extra = "merge") %>% 
+  mutate(variable = str_remove(variable, "2"))
+
+# CREATE COLOR VECTOR
+my_col <- c("CM" = "#00BFC4", "HC" = "#7CAE00", "UM" = "#F8766D")
+
+# PLOT
+df_fig2 %>% 
+  
+  ggplot(aes(x = status, y = value)) +
+  geom_violin(aes(color = status)) +
+  geom_jitter(position = position_jitter(0.2), shape = 1, size = 2) +
+  stat_summary(fun = "median", geom = "crossbar", aes(color = status), size = 0.2, width = 0.5) +
+  facet_grid(variable ~ tissue, scales = "free_y") +
+  
+  scale_color_manual(values = my_col) +
+  labs(x = "") +
+  theme_bw() +
[…]

### 2023-06-12 Scripts/R/final_figures/figS1.R

@@ -1,0 +1,44 @@
+
+
+### Code to generate fig S1: DFA proof-of-concept on simulated signals
+
+
+# libraries ---------------------------------------------------------------
+
+library(tidyverse)
+
+# data load & format --------------------------------------------------------------------
+
+load("objects/sims.Rdata")
+
+wn_alpha <- 1:length(l_wn_dfa) %>% purrr::map_df(~l_wn_dfa[[.x]][[2]]) %>% dplyr::rename("white" = "log10(window_size)")
+pn_alpha <- 1:length(l_pn_dfa) %>% purrr::map_df(~l_pn_dfa[[.x]][[2]]) %>% dplyr::rename("pink" = "log10(window_size)")
+bn_alpha <- 1:length(l_bn_dfa) %>% purrr::map_df(~l_bn_dfa[[.x]][[2]]) %>% dplyr::rename("brown" = "log10(window_size)")
+
+
+# plot --------------------------------------------------------------------
+
+df_figs1 <- bind_cols(wn_alpha, pn_alpha, bn_alpha) %>%
+  pivot_longer(1:3, names_to = "noise", values_to = "alpha")
+
+# create color vector
+my_col <- c("white" = "white", "brown" = "sienna2", "pink" = "pink")
+
+# plot
+df_figs1 %>% 
+  
+  ggplot(aes(x = noise, y = alpha, color = noise)) +
+  geom_violin(fill = "transparent") +
+  geom_jitter(position = position_jitter(width = 0.1), shape = 1, size = 2) +
+  stat_summary(fun = "median", geom = "crossbar", aes(color = noise), size = 0.2, width = 0.5) +
+  
+  scale_color_manual(values = my_col) +
+  labs(x = "noise type", y = "alpha (\u03b1)") +
+  scale_y_continuous(limits = c(0, 1.7)) +
+  theme_dark() +
+  theme(legend.position = "none")  +
+  
+  theme(# text = element_text(family = "Arial"),
+    strip.text = element_text(size = 15),
+    axis.title = element_text(size = 15),
+    axis.text = element_text(size = 15))

### 2023-06-12 Scripts/R/final_figures/figS2.R

@@ -1,0 +1,172 @@
+
+
+### Code to generate fig S2: walk-through detrended fluctuation analysis
+
+
+
+# libraries ---------------------------------------------------------------
+
+library(tidyverse)
+library(segmented)
+
+
+# data load (brain DFA results) -------------------------------------------
+
+load("objects/brain_bandpass_filtered_dfa_results.Rdata")
+
+# fig S2a -----------------------------------------------------------------
+
+df_figs2a <- read.table("Data/signal_segments/Hb_tot/TM0001CM01.txt")
+
+df_figs2a %>% 
+  
+  ggplot(aes(x = Time)) +
+  geom_line(aes(y = THC)) +
+  
+  labs(y = "Hemoglobin concentration (μM)", x = "Time (s)") +
+  ggtitle("Raw signal") +
+  theme_bw() +
+  theme(# text = element_text(family = "Arial"),
+    axis.text = element_text(size = 20),
+    axis.title = element_text(size = 25),
+    legend.text = element_text(size = 15),
+    legend.title = element_text(size = 20))
+
+
+# fig S2b -----------------------------------------------------------------
+
+df_figs2b <- read.table("Data/filtered_signals/0.2s_mean_0.001_1_filt/Hb_tot/TM0001CM01.txt") %>% 
+  mutate(time = row_number())
+
+df_figs2b %>% 
+  ggplot(aes(x = time)) +
+  geom_line(aes(y = V1)) +
+  
+  #lims(y = c(-0.001, 0.001)) +
+  
+  labs(y = "Bandpass filtered signal", x = "Time") +
+  ggtitle("Pre-processed signal") +
+  theme_bw() +
+  theme(# text = element_text(family = "Arial"),
+    #axis.text = element_blank(),
+    axis.title = element_text(size = 25),
+    legend.text = element_text(size = 15),
+    legend.title = element_text(size = 20))
+
+
+
+# fig S2c -----------------------------------------------------------------
+
+
+df_figs2c <- df_figs2b %>% 
+  mutate(mean = mean(V1),
+         subtract_mean = V1 - mean,
+         cumsum = cumsum(subtract_mean))
+
+df_figs2c %>% 
+  
+  ggplot(aes(x = time, y = cumsum)) +
+  geom_line() +
+  geom_vline(xintercept = 1, lty = 2, color = "black") +
[…]

### 2023-06-12 Scripts/R/final_figures/table1.R

@@ -1,0 +1,112 @@
+
+### Code to generate Table 1 Clinical characteristics of the study population from raw data
+
+
+# libraries ---------------------------------------------------------------
+
+library(tidyverse)
+
+# data load & format --------------------------------------------------------------------
+
+df_hbox <- read_csv("Data/Hans_datatable_exports/malawi key data v04Jan2022.csv") %>% 
+  as_tibble() %>% 
+  janitor::clean_names() %>% 
+  mutate(status = replace(status, status == "HV", "HC")) %>% 
+  dplyr::rename("subject_id" = "subject_id_session") %>% 
+  
+  dplyr::filter(status %in% c("CM", "HC", "UM") & session_no == 1 | subject_id == "TM0003CM01") %>% 
+  dplyr::filter(!grepl("blood", subject_id)) %>% 
+  mutate(status = factor(status, levels = c("HC", "UM", "CM"))) %>% 
+  mutate_at(c("bp_dias", "bp_sys", "glucose", "lactate"), as.numeric) %>% # converts remaining character variables to numerics
+  select(subject_id, status, everything()) %>% 
+  group_by(status)
+
+# data format -------------------------------------------------------------
+
+### TABLE FOR NUMBER OF SUBJECTS IN EACH PATIENT GROUP
+df_status <- df_hbox %>% 
+  count() %>% 
+  pivot_wider(names_from = "status", values_from = "n") %>% 
+  mutate(variable = "Number of subjects") %>% 
+  dplyr::select(variable, everything()) %>% 
+  mutate_if(is.numeric, as.character)
+
+### TABLE FOR % FEMALE IN EACH PATIENT GROUP
+df_sex <- df_hbox %>% 
+  count(sex) %>% 
+  pivot_wider(id_cols = "status", names_from = "sex", values_from = "n") %>% 
+  mutate(perc_fem = round(Female/(Female + Male)*100, 2)) %>% 
+  dplyr::select(-c(Female, Male)) %>% 
+  
+  pivot_wider(names_from = "status", values_from = "perc_fem") %>% 
+  mutate(variable = "Sex (% female)") %>% 
+  dplyr::select(variable, everything()) %>% 
+  mutate_if(is.numeric, as.character)
+
+### NUMERIC VARIABLES
+
[…]

### 2023-06-12 Scripts/R/final_figures/tables2_3.R

@@ -1,0 +1,53 @@
+
+### Code to generate Tables 2 & 3 Logistic regression model results
+
+
+# libraries ---------------------------------------------------------------
+
+library(tidyverse)
+library(tidymodels)
+
+# data --------------------------------------------------------------------
+
+
+load("objects/20221125_final_glm_models.Rdata") # models compared and results generated in in scripts/markdown_files/20221115_revise_model.R
+
+# table 2 -----------------------------------------------------------------
+
+df_CI <- confint(hbtot_final2$fit) %>% exp() %>% as_tibble(rownames = "term")
+
+hbtot_final2 %>%
+  tidy(exponentiate = TRUE) %>%
+  filter(p.value < 0.05) %>% 
+  left_join(df_CI, by = "term") %>%
+  mutate(estimate = ifelse(estimate < 10^-2, scientific(estimate, digits = 3), round(estimate, 2)),
+         conf.low = ifelse(`2.5 %` < 10^-2, scientific(`2.5 %`, digits = 3), round(`2.5 %`, 2)),
+         conf.high = ifelse(`97.5 %` < 10^-2, scientific(`97.5 %`, digits = 3), round(`97.5 %`, 2))) %>%
+  mutate(p.value = scientific(p.value, digits = 3),
+         statistic = round(statistic, 2)) %>%
+  unite("95% CI", c(conf.low, conf.high), sep = ", ") %>%
+  dplyr::select(c(term, estimate, `95% CI`, statistic, p.value)) %>% 
+  dplyr::rename("Term" = "term", "Odds ratio" = "estimate",
+                "t-statistic" = "statistic", "p-value" = "p.value") %>%
+  filter(Term != "(Intercept)")
+
+
+# table 3 -----------------------------------------------------------------
+
+df_CI <- confint(hbtot_alpha_final2$fit) %>% exp() %>% as_tibble(rownames = "term")
+
+hbtot_alpha_final2 %>%
+  tidy(exponentiate = TRUE) %>%
+  filter(p.value < 0.05) %>% 
+  left_join(df_CI, by = "term") %>%
+  mutate(estimate = ifelse(estimate < 10^-2, scientific(estimate, digits = 3), round(estimate, 2)),
+         conf.low = ifelse(`2.5 %` < 10^-2, scientific(`2.5 %`, digits = 3), round(`2.5 %`, 2)),
[…]

### 2023-06-12 Scripts/R/full_analysis_pipeline.R

@@ -74,11 +74,6 @@
     
     # may be useful to save df_raw_data at this point using the save() function so you don't have to read everything in again
 
-  
-
-    
-    
-
 
 # signal preprocessing ------------------------------------------------------
 
@@ -620,22 +615,5 @@
                 
                 
                 
-                
-                
-                
-                
-            
-            
-            
-            
-            
-            
-            
-  
-  
-  
-  
-  
-  
-  
+
   

### 2023-06-12 Scripts/R/missing_UM_analysis.R

@@ -1,5 +1,7 @@
 
 
+### Run the full analysis pipeline on UM patient signals that had been missing from the original datatable
+
 
 # libraries ---------------------------------------------------------------
 
@@ -516,63 +518,5 @@
     
     
     
-    
-    
-    
-    
-    
-    
-    
-    
-    
-    
-    
-    
-    
-    
-    
-    
-    
-    
-    
-    
-    
-    
-    
-    
-    
-    
-    
-    
-    
-    
-    
-    
-    
-    
-    
-    
-    
-    
-    
-    
-        
-    
-    
-    
-    
-       
-    
-    
-    
-    
-    
-    
-    
-    
-       
-    
-    
-    
-    
+
     

### 2023-06-12 Scripts/R/raw_data_processing.R

@@ -1,6 +1,9 @@
 
 
 
+### Select signal segments for each NIRS signal that doesn't have 'blips'
+
+
 # read in data -------------------------------------------------------------------------
 
   library(tidyverse)

### 2023-06-12 Scripts/R/markdown_files/archive/20221117_revise_model.Rmd
add objects to run scripts, other data will be available in data repository

---
title: "Revise LR model"
author: "Rachel Smith"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include = FALSE}

knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.height = 5.5, fig.width = 8)

# libraries
library(tidyverse)
library(RColorBrewer)
library(gridExtra)
library(patchwork)
library(Hmisc)
library(corrr)
library(tidymodels)
library(themis)
library(vip)
library(janitor)

# set theme for plots
theme_set(theme_bw() +
            theme(plot.title = element_text(size = 18),
                  axis.title = element_text(size = 15),
                  axis.text = element_text(size = 12),
                  strip.text = element_text(size = 12),
                  legend.title = element_text(size = 15),
                  legend.text = element_text(size = 12)))

# create color vector
my_col <- c("CM" = "#00BFC4", "HC" = "#7CAE00", "UM" = "#F8766D")

```

```{r data, echo = FALSE}

# load
df_hbox <- read_csv("Data/Hans_datatable_exports/malawi key data v04Jan2022.csv") %>% 
  as_tibble() %>% 
  janitor::clean_names() %>% 
  mutate(status = replace(status, status == "HV", "HC")) %>% 
  dplyr::rename("subject_id" = "subject_id_session") %>%

  dplyr::filter(status %in% c("CM", "HC", "UM") & session_no == 1 | subject_id == "TM0003CM01") %>% 
  dplyr::filter(!grepl("blood", subject_id)) %>% 
  mutate(status = factor(status, levels = c("HC", "UM", "CM"))) %>% 
  mutate_at(c("bp_dias", "bp_sys", "glucose", "lactate"), as.numeric) %>% # converts remaining character variables to numerics
  select(subject_id, status, everything()) %>% 
  group_by(status)

# format
  # define clinical variables of interest
  clinical_voi <- c("age_calc", "temperature", "hr", "bp_sys", "bp_dias", "resp_rate", "pulse_ox", "hct", "lactate", "glucose", "aa_arg")
  hbox_voi <- c("cerebral_hb_tot", "cerebral_hb_tot_alpha2")

  # select clinical VOI
  df_model <- df_hbox %>% 
    dplyr::select(c(status, sex, all_of(hbox_voi), all_of(clinical_voi))) %>% 
    mutate(sex_coded = ifelse(sex == "Female", 1, 0)) %>% 
    dplyr::select(-sex)

  # impute missing data with median by group
  f_impute <- function(x, na.rm = TRUE) (replace(x, is.na(x), median(x, na.rm = na.rm)))

  df_model_impute <- df_model %>% 
    group_by(status) %>% 
    mutate_at(2:ncol(.), f_impute) %>% 
    ungroup()

```

```{r p_distributions, fig.height = 8, fig.width = 10}

df_model %>% 
  pivot_longer(2:ncol(.), names_to = "variable", values_to = "value") %>% 
  mutate(variable = factor(variable, levels = c(clinical_voi, "sex_coded", hbox_voi))) %>% 
  filter(!(variable == "age_calc" & value > 20000)) %>%

  ggplot(aes(x = value)) +
  geom_density(aes(fill = status), alpha = 0.5) +
  facet_wrap(~ variable, scales = "free") +
  labs(x = "") +
  ggtitle("Distributions of variables of interest")

```

```{r correlate}

# correlation
df_cor_r <- df_model_impute %>% 
  group_by(status) %>% 
  nest() %>% 
  mutate(corr = map(.x = data, 
                    .f = ~ .x %>% 
                      correlate %>% 
                      dplyr::select(term, contains("cerebral_hb_tot")) %>% 
                      filter(term %in% clinical_voi) 
  )
  ) %>% 
  unnest(cols = c(corr)) %>% 
  pivot_longer(4:5, names_to = "voi", values_to = "r") %>% 
  mutate(voi = ifelse(voi == "cerebral_hb_tot", "Hb_tot", "Hb_tot_a") %>% factor()) %>% 
  mutate(term = factor(term, levels = clinical_voi))

# p-values
df_cor_p <- df_model_impute %>% 
  group_by(status) %>% 
  nest() %>% 
  mutate(corr = map(.x = data, 
                    .f = ~ .x %>% 
                      as.data.frame %>% 
                      as.matrix %>% 
                      rcorr %>% 
                      .$P %>% 
                      corrr::as_cordf() %>% 
                      dplyr::select(term, contains("cerebral_hb_tot")) %>% 
                      filter(term %in% clinical_voi) 
  )
  ) %>% 
  unnest(cols = c(corr)) %>% 
  pivot_longer(4:5, names_to = "voi", values_to = "p") %>% 
  mutate(voi = ifelse(voi == "cerebral_hb_tot", "Hb_tot", "Hb_tot_a") %>% factor()) %>% 
  mutate(term = factor(term, levels = clinical_voi))

# combine for plot
df_plot_cor <- df_cor_r %>% 
  left_join(df_cor_p) %>% #filter(status == "HC")
  mutate(label = as.character(round(r, 2)),
         label = ifelse(p < 0.05, paste0(label, "*"), label))

```

```{r p_correlations}

df_plot_cor %>% 
  ggplot() +
  geom_tile(aes(x = voi, y = term, fill = r)) +
  geom_text(aes(x = voi,y = term, label = label)) +
  facet_wrap( ~ status) +
  labs(x = "", y = "") +
  scale_fill_gradient2(high = "#B2182B", mid = "white", low = "#2166AC",
                       limits = c(-1, 1)) +
  ylim(rev(levels(df_cor_p$term))) +
  ggtitle("Correlations of clinical variables with cerebral Hb_tot") +
  theme_classic() +
  theme(plot.title = element_text(size = 18),
        axis.title = element_text(size = 15),
        axis.text = element_text(size = 12),
        strip.text = element_text(size = 15),
        legend.title = element_text(size = 15),
        legend.text = element_text(size = 12))

```
*denotes significant correlation (p < 0.05)

```{r p_sig_cor}

p_um1 <- df_hbox %>% 
  filter(status == "UM") %>% 
  dplyr::select(cerebral_hb_tot, hr) %>%

  ggplot(aes(x = cerebral_hb_tot, y = hr, color = status)) +
  geom_point(size = 2, show.legend = FALSE) +
  geom_smooth(method = "lm", color = "black", lty = 2, se = FALSE) +
  annotate(geom = "text", 
           label = "r = -0.44; p < 0.01",
           x = 300, y = 160,
           size = 5) + 
  scale_color_manual(values = my_col) +
  ggtitle("UM HR & Hb_tot")

p_um2 <- df_hbox %>% 
  filter(status == "UM") %>% 
  dplyr::select(cerebral_hb_tot, hct) %>%

  ggplot(aes(x = cerebral_hb_tot, y = hct, color = status)) +
  geom_point(size = 2, show.legend = FALSE) +
  geom_smooth(method = "lm", color = "black", lty = 2, se = FALSE) +
  annotate(geom = "text",
           label = "r = 0.39; p = 0.03",
           x = 300, y = 27,
           size = 5) +
  scale_color_manual(values = my_col) +
  ggtitle("UM Hematocrit & Hb_tot")

p_hc1 <- df_hbox %>% 
  filter(status == "HC") %>% 
  dplyr::select(cerebral_hb_tot, aa_arg) %>%

  ggplot(aes(x = cerebral_hb_tot, y = aa_arg, color = status)) +
  geom_point(size = 2, show.legend = FALSE) +
  geom_smooth(method = "lm", color = "black", lty = 2, se = FALSE) +
  annotate(geom = "text",
           label = "r = 0.51; p < 0.01",
           x = 95, y = 45,
           size = 5) +
  scale_color_manual(values = my_col) +
  ggtitle("HC arg & Hb_tot")

p_hc2 <- df_hbox %>% 
  filter(status == "HC") %>% 
  dplyr::select(cerebral_hb_tot_alpha, aa_arg) %>%

  ggplot(aes(x = cerebral_hb_tot_alpha, y = aa_arg, color = status)) +
  geom_point(size = 2, show.legend = FALSE) +
  geom_smooth(method = "lm", color = "black", lty = 2, se = FALSE) +
  annotate(geom = "text",
           label = "r = -0.43; p = 0.02",
           x = 1.02, y = 45,
           size = 5) +
  scale_color_manual(values = my_col) +
  ggtitle("HC arg & Hb_tot_a")

(p_um1 + p_um2) / (p_hc1 + p_hc2)

```

```{r lr_model}

# write function to run LR model
f_run_lr <- function(df_model, resamples) {

  # resample
  hbox_boot <- bootstraps(df_model, strata = status, times = resamples)

  # recipe
  hbox_rec <- recipe(status ~ ., data = df_model) %>% 
    step_impute_knn(all_predictors()) %>% 
    step_normalize(all_predictors()) %>%
    step_zv(all_predictors()) %>%
    step_smote(status) %>% 
    update_role(subject_id, new_role = "id")

  # create workflow
  hbox_wf <- workflow() %>% add_recipe(hbox_rec)

  # logistic regression model specifications
  glm_spec <- logistic_reg() %>% set_engine("glm")

  # create model on 5000 different resamples
  glm_res <- hbox_wf %>% 
    add_model(glm_spec) %>%
    fit_resamples(
      resamples = hbox_boot,
      metrics = metric_set(roc_auc, accuracy, sensitivity, specificity),
      control = tune::control_resamples(save_pred = TRUE, verbose = TRUE)
    )

  # select best model to be used for evaluation
  glm_best <- glm_res %>% select_best("roc_auc")

  # finalize model
  hbox_wf %>% 
    add_model(glm_spec) %>% 
    finalize_workflow(glm_best) %>%
    fit(df_model) %>%
    extract_fit_parsnip()

}

```

### Cerebral Hb_tot model fitting

status ~ Hb_tot + age_calc + bp_sys + bp_dias + hr + hct + lactate + aa_arg + sex
```{r hb_tot_model1}

# VOI: age_calc, bp_sys, bp_dias, hr, hct, lactate, and aa_arg plus either Hb_tot or Hb_tot_a. I would probably include sex as well

df_model <- df_hbox %>% 
  mutate(sex = ifelse(sex == "Female", 1, 0)) %>% 
  filter(status != "HC") %>% 
  mutate(status = factor(status, levels = c("UM", "CM")))

set.seed(20221117)

# 1. status ~ Hb_tot + age_calc + bp_sys + bp_dias + hr + hct + lactate + aa_arg + sex
hbtot_final1 <- f_run_lr(df_model %>% 
                           dplyr::select(subject_id, status, cerebral_hb_tot, 
                                         aa_arg, age_calc, bp_sys, bp_dias, hct, hr, lactate, sex), 
                       10)

hbtot_final1 %>% 
  tidy %>% 
  knitr::kable() # AIC 86.9

```

status ~ Hb_tot + age_calc + bp_sys + bp_dias + hr + hct + lactate + sex
```{r hb_tot_model2}

# 2. status ~ Hb_tot + age_calc + bp_sys + bp_dias + hr + hct + lactate + sex
hbtot_final2 <- f_run_lr(df_model %>% 
                           dplyr::select(subject_id, status, cerebral_hb_tot, 
                                         age_calc, bp_sys, bp_dias, hct, hr, lactate, sex), 
                       10)

hbtot_final2 %>% 
  tidy %>% 
  knitr::kable() # AIC 85.6

```

status ~ Hb_tot + age_calc + bp_dias + hr + hct + lactate + sex
```{r hb_tot_model3}

# 3. status ~ Hb_tot + age_calc + bp_dias + hr + hct + lactate + sex
hbtot_final3 <- f_run_lr(df_model %>% 
                           dplyr::select(subject_id, status, cerebral_hb_tot, 
                                         age_calc, bp_dias, hct, hr, lactate, sex), 
                       10)

hbtot_final3 %>% 
  tidy %>% 
  knitr::kable() # AIC 83.52

```

status ~ Hb_tot + bp_dias + hr + hct + lactate + sex
```{r hb_tot_model4}

# 4. status ~ Hb_tot + bp_dias + hr + hct + lactate + sex
hbtot_final4 <- f_run_lr(df_model %>% 
                           dplyr::select(subject_id, status, cerebral_hb_tot, 
                                         bp_dias, hct, hr, lactate, sex), 
                       10)

hbtot_final4 %>% 
  tidy %>% 
  knitr::kable() # AIC 82.83

```

status ~ Hb_tot + hr + hct + lactate + sex
```{r hb_tot_model5}

# 5. status ~ Hb_tot + hr + hct + lactate + sex
hbtot_final5 <- f_run_lr(df_model %>% 
                           dplyr::select(subject_id, status, cerebral_hb_tot, 
                                         hct, hr, lactate, sex), 
                       10)

hbtot_final5 %>% 
  tidy %>% 
  knitr::kable() # AIC 81.36

```

status ~ Hb_tot + hct + lactate + sex
```{r hb_tot_model6}

# 6. status ~ Hb_tot + hct + lactate + sex

set.seed(20221118)

# resample
hbtot_boot <- bootstraps(df_model, strata = status, times = 5000)

# recipe
hbtot_rec <- recipe(status ~ cerebral_hb_tot + hct + lactate + sex + subject_id, data = df_model) %>% 
  step_impute_knn(all_predictors()) %>% 
  step_normalize(all_predictors()) %>%
  step_zv(all_predictors()) %>%
  step_smote(status) %>% 
  update_role(subject_id, new_role = "id")

# create workflow
hbtot_wf <- workflow() %>% add_recipe(hbtot_rec)

# logistic regression model specifications
hbtot_glm_spec <- logistic_reg() %>% set_engine("glm")

# create model on 5000 different resamples
hbtot_glm_res <- hbtot_wf %>% 
  add_model(hbtot_glm_spec) %>%
  fit_resamples(
    resamples = hbtot_boot,
    metrics = metric_set(roc_auc, accuracy, sensitivity, specificity),
    control = tune::control_resamples(save_pred = TRUE, verbose = TRUE)
  )

# select best model to be used for evaluation
hbtot_glm_best <- hbtot_glm_res %>% select_best("roc_auc")

# finalize model
hbtot_final6 <- hbtot_wf %>% 
  add_model(hbtot_glm_spec) %>% 
  finalize_workflow(hbtot_glm_best) %>%
  fit(df_model) %>%
  extract_fit_parsnip()

hbtot_final6 %>% 
  tidy %>% 
  knitr::kable() # AIC 82.17, every term is significant

```

```{r assess_hb_tot_model}

df_CI <- confint(hbtot_final6$fit) %>% exp() %>% as_tibble(rownames = "term")

hbtot_final6 %>%
  tidy(exponentiate = TRUE) %>%
  filter(p.value < 0.05) %>% 
  left_join(df_CI, by = "term") %>%
  mutate(estimate = ifelse(estimate < 10^-2, scientific(estimate, digits = 3), round(estimate, 2)),
         conf.low = ifelse(`2.5 %` < 10^-2, scientific(`2.5 %`, digits = 3), round(`2.5 %`, 2)),
         conf.high = ifelse(`97.5 %` < 10^-2, scientific(`97.5 %`, digits = 3), round(`97.5 %`, 2))) %>%
  mutate(p.value = scientific(p.value, digits = 3),
         statistic = round(statistic, 2)) %>%
  unite("95% CI", c(conf.low, conf.high), sep = ", ") %>%
  dplyr::select(c(term, estimate, `95% CI`, statistic, p.value)) %>% 
  dplyr::rename("Term" = "term", "Odds ratio" = "estimate",
                "t-statistic" = "statistic", "p-value" = "p.value") %>%
  filter(Term != "(Intercept)") %>% 
  as.data.frame() %>%

  knitr::kable(align = rep('c', 5))

hbtot_glm_res %>% 
  collect_metrics() %>% 
  dplyr::select(.metric, mean, std_err) %>% 
  mutate(mean = mean*100, std_err = std_err*100) %>% 
  mutate(Metric = c("Accuracy", "ROC AUC", "Sensitivity", "Specificity")) %>% 
  dplyr::select(c("Metric", "mean", "std_err")) %>%

  dplyr::rename("Mean" = "mean", "Standard error" = "std_err") %>% 
  mutate_if(is.numeric, round, 2) %>% 
  as.data.frame() %>%

  knitr::kable()

```

```{r hb_tot_imp, fig.height = 4}

set.seed(123)
hbtot_imp <- hbtot_wf %>% 
  add_model(hbtot_glm_spec) %>% 
  fit(df_model) %>% 
  pull_workflow_fit() %>%

  vi(
    method = "permute", nsim = 10,
    train = juice(hbtot_rec %>% prep()),
    target = "status", metric = "auc", reference_class = "CM",
    pred_wrapper = kernlab::predict

  )

hbtot_imp %>% 
  clean_names %>% 
  filter(variable != "subject_id") %>% 
  mutate(variable = fct_reorder(variable, importance)) %>%

  ggplot(aes(x = importance, y = variable, color = variable)) +
  geom_errorbar(aes(xmin = importance - st_dev, xmax = importance + st_dev),
                alpha = 0.5, size = 1.3) +
  geom_point(size = 3) +
  theme(legend.position = "none",
        plot.title = element_text(size = 18),
        axis.title = element_text(size = 15),
        axis.text = element_text(size = 12),
        strip.text = element_text(size = 12),
        legend.title = element_text(size = 15),
        legend.text = element_text(size = 12)) +
  labs(y = NULL, x = "relative importance") +
  ggtitle("Variable importance in distinguishing CM vs UM")

```

### Cerebral Hb_tot_alpha model fitting

status ~ Hb_tot_alpha + age_calc + bp_sys + bp_dias + hr + hct + lactate + aa_arg + sex
```{r hb_tot_alpha_model1}

# 1. status ~ Hb_tot_alpha + age_calc + bp_sys + bp_dias + hr + hct + lactate + aa_arg + sex
hbtot_alpha_final1 <- f_run_lr(df_model %>% 
                           dplyr::select(subject_id, status, cerebral_hb_tot_alpha2, 
                                         aa_arg, age_calc, bp_sys, bp_dias, hct, hr, lactate, sex), 
                       10)

hbtot_alpha_final1 %>% 
  tidy %>% 
  knitr::kable() # AIC 64.6

```

status ~ Hb_tot_alpha + age_calc + bp_sys + bp_dias + hct + lactate + aa_arg + sex
```{r hb_tot_alpha_model2}

# 2. status ~ Hb_tot_alpha + age_calc + bp_sys + bp_dias + hct + lactate + aa_arg + sex
hbtot_alpha_final2 <- f_run_lr(df_model %>% 
                           dplyr::select(subject_id, status, cerebral_hb_tot_alpha2, 
                                         aa_arg, age_calc, bp_sys, bp_dias, hct, lactate, sex), 
                       10)

hbtot_alpha_final2 %>% 
  tidy %>% 
  knitr::kable() # AIC 67.99

```

status ~ Hb_tot_alpha + age_calc + bp_sys + bp_dias + hct + lactate + sex
```{r hb_tot_alpha_model3}

# 3. status ~ Hb_tot_alpha + age_calc + bp_sys + bp_dias + hct + lactate + sex
hbtot_alpha_final3 <- f_run_lr(df_model %>% 
                           dplyr::select(subject_id, status, cerebral_hb_tot_alpha2, 
                                         age_calc, bp_sys, bp_dias, hct, lactate, sex), 
                       10)

hbtot_alpha_final3 %>% 
  tidy %>% 
  knitr::kable() # AIC 61.78

```

status ~ Hb_tot_alpha + bp_sys + bp_dias + hct + lactate + sex
```{r hb_tot_alpha_model4}

# 4. status ~ Hb_tot_alpha + bp_sys + bp_dias + hct + lactate + sex
hbtot_alpha_final4 <- f_run_lr(df_model %>% 
                           dplyr::select(subject_id, status, cerebral_hb_tot_alpha2, 
                                         bp_sys, bp_dias, hct, lactate, sex), 
                       10)

hbtot_alpha_final4 %>% 
  tidy %>% 
  knitr::kable() # AIC 60.53

```

status ~ Hb_tot_alpha + bp_dias + hct + lactate + sex
```{r hb_tot_alpha_model5}

# 5. status ~ Hb_tot_alpha + bp_dias + hct + lactate + sex
hbtot_alpha_final5 <- f_run_lr(df_model %>% 
                           dplyr::select(subject_id, status, cerebral_hb_tot_alpha2, 
                                         bp_dias, hct, lactate, sex), 
                       10)

hbtot_alpha_final5 %>% 
  tidy %>% 
  knitr::kable() # AIC 61.65

```

status ~ Hb_tot_alpha + hct + lactate + sex
```{r hb_tot_alpha_model6}

# 6. status ~ Hb_tot_alpha + hct + lactate + sex
hbtot_alpha_final6 <- f_run_lr(df_model %>% 
                           dplyr::select(subject_id, status, cerebral_hb_tot_alpha2, 
                                         hct, lactate, sex), 
                       10)

hbtot_alpha_final6 %>% 
  tidy %>% 
  knitr::kable() # AIC 67.5

```

status ~ Hb_tot_alpha + hct + sex
```{r hb_tot_alpha_model7}

# 7. status ~ Hb_tot_alpha + hct + sex

set.seed(20221118)

# resample
hbtot_alpha_boot <- bootstraps(df_model, strata = status, times = 5000)

# recipe
hbtot_alpha_rec <- recipe(status ~ cerebral_hb_tot_alpha2 + hct + sex + subject_id, data = df_model) %>% 
  step_impute_knn(all_predictors()) %>% 
  step_normalize(all_predictors()) %>%
  step_zv(all_predictors()) %>%
  step_smote(status) %>% 
  update_role(subject_id, new_role = "id")

# create workflow
hbtot_alpha_wf <- workflow() %>% add_recipe(hbtot_alpha_rec)

# logistic regression model specifications
hbtot_alpha_glm_spec <- logistic_reg() %>% set_engine("glm")

# create model on 5000 different resamples
hbtot_alpha_glm_res <- hbtot_alpha_wf %>% 
  add_model(hbtot_alpha_glm_spec) %>%
  fit_resamples(
    resamples = hbtot_alpha_boot,
    metrics = metric_set(roc_auc, accuracy, sensitivity, specificity),
    control = tune::control_resamples(save_pred = TRUE, verbose = TRUE)
  )

# select best model to be used for evaluation
hbtot_alpha_glm_best <- hbtot_alpha_glm_res %>% select_best("roc_auc")

# finalize model
hbtot_alpha_final7 <- hbtot_alpha_wf %>% 
  add_model(hbtot_alpha_glm_spec) %>% 
  finalize_workflow(hbtot_alpha_glm_best) %>%
  fit(df_model) %>%
  extract_fit_parsnip()

hbtot_alpha_final7 %>% 
  tidy %>% 
  knitr::kable() # AIC 65.23

```

```{r assess_hb_tot_alpha_model}

df_CI <- confint(hbtot_alpha_final7$fit) %>% exp() %>% as_tibble(rownames = "term")

hbtot_alpha_final7 %>%
  tidy(exponentiate = TRUE) %>%
  filter(p.value < 0.05) %>% 
  left_join(df_CI, by = "term") %>%
  mutate(estimate = ifelse(estimate < 10^-2, scientific(estimate, digits = 3), round(estimate, 2)),
         conf.low = ifelse(`2.5 %` < 10^-2, scientific(`2.5 %`, digits = 3), round(`2.5 %`, 2)),
         conf.high = ifelse(`97.5 %` < 10^-2, scientific(`97.5 %`, digits = 3), round(`97.5 %`, 2))) %>%
  mutate(p.value = scientific(p.value, digits = 3),
         statistic = round(statistic, 2)) %>%
  unite("95% CI", c(conf.low, conf.high), sep = ", ") %>%
  dplyr::select(c(term, estimate, `95% CI`, statistic, p.value)) %>% 
  dplyr::rename("Term" = "term", "Odds ratio" = "estimate",
                "t-statistic" = "statistic", "p-value" = "p.value") %>%
  filter(Term != "(Intercept)") %>% 
  as.data.frame() %>%

  knitr::kable(align = rep('c', 5))

hbtot_alpha_glm_res %>% 
  collect_metrics() %>% 
  dplyr::select(.metric, mean, std_err) %>% 
  mutate(mean = mean*100, std_err = std_err*100) %>% 
  mutate(Metric = c("Accuracy", "ROC AUC", "Sensitivity", "Specificity")) %>% 
  dplyr::select(c("Metric", "mean", "std_err")) %>%

  dplyr::rename("Mean" = "mean", "Standard error" = "std_err") %>% 
  mutate_if(is.numeric, round, 2) %>% 
  as.data.frame() %>%

  knitr::kable()

```

```{r hb_tot_alpha_imp, fig.height = 4}

set.seed(123)
hbtot_alpha_imp <- hbtot_alpha_wf %>% 
  add_model(hbtot_alpha_glm_spec) %>% 
  fit(df_model) %>% 
  pull_workflow_fit() %>%

  vi(
    method = "permute", nsim = 10,
    train = juice(hbtot_alpha_rec %>% prep()),
    target = "status", metric = "auc", reference_class = "CM",
    pred_wrapper = kernlab::predict

  )

hbtot_alpha_imp %>% 
  clean_names %>% 
  filter(variable != "subject_id") %>% 
  mutate(variable = fct_reorder(variable, importance)) %>%

  ggplot(aes(x = importance, y = variable, color = variable)) +
  geom_errorbar(aes(xmin = importance - st_dev, xmax = importance + st_dev),
                alpha = 0.5, size = 1.3) +
  geom_point(size = 3) +
  theme(legend.position = "none",
        plot.title = element_text(size = 18),
        axis.title = element_text(size = 15),
        axis.text = element_text(size = 12),
        strip.text = element_text(size = 12),
        legend.title = element_text(size = 15),
        legend.text = element_text(size = 12)) +
  labs(y = NULL, x = "relative importance") +
  ggtitle("Variable importance in distinguishing CM vs UM")

```

### Distributions of NIRS variables by sex

```{r sex_associations}

df_hbox %>% 
  dplyr::select(subject_id, status, sex, cerebral_hb_tot, cerebral_hb_tot_alpha2) %>% 
  pivot_longer(4:5, names_to = "nirs_variable", values_to = "value") %>%

  ggplot(aes(x = status, y = value)) +
  geom_violin(aes(color = status)) +
  geom_jitter(position = position_jitter(0.15), shape = 1, size = 2) +
  stat_summary(fun = "median", geom = "crossbar", aes(color = status), size = 0.2, width = 0.5) +
  facet_grid(nirs_variable ~ sex, scales = "free")

# check for significant differences between M-F for each variable
df_hbox %>% 
  dplyr::select(subject_id, status, sex, cerebral_hb_tot, cerebral_hb_tot_alpha2) %>% 
  pivot_longer(4:5, names_to = "nirs_variable", values_to = "value") %>%

  group_by(status, nirs_variable) %>% 
  nest() %>%

  mutate(wilcox = map(.x = data,
                      .f = ~ wilcox.test(.x %>% filter(sex == "Female") %>% pull(value),
                                         .x %>% filter(sex == "Male") %>% pull(value)))) %>% 
  mutate(p_val = map(.x = wilcox, .f = ~.x$p.value)) %>% 
  unnest(cols = c(p_val)) %>%

  knitr::kable()

```

**no sex differences in distributions of NIRS variables in each patient group**

### 2023-06-12 Scripts/R/markdown_files/response_to_reviewers.Rmd
master data table updates

@@
 
-font_import()
+#font_import()
 
@@
 
-df_hbox <- read_csv(paste0(base_dir, "Data/Hans_datatable_exports/malawi key data v04Jan2022.csv")) %>% 
+df_hbox <- read_csv(paste0(base_dir, "Data/master_datatable.csv")) %>% 
   as_tibble() %>% 
@@
   mutate(status = replace(status, status == "HV", "HC")) %>%
-  dplyr::rename("subject_id" = "subject_id_session") %>%
 
@@
   mutate(status = factor(status, levels = c("HC", "UM", "CM"))) %>% 
-  mutate_at(c("bp_dias", "bp_sys", "glucose", "lactate"), as.numeric) %>% # converts remaining character variables to numerics
+  mutate_at(c("bp_diastolic", "bp_systolic", "glucose", "lactate"), as.numeric) %>% # converts remaining character variables to numerics
   select(subject_id, status, everything()) %>% 
@@
 ```
+
+## Editor
+
+```{r ed}
+
+# final terms used in model
+final_voi <- c("cerebral_hb_tot", "cerebral_hb_tot_alpha2", "hematocrit", "lactate")
+
+# create df from just these variables
+df_model <- df_hbox %>% 
+  clean_names %>% 
+  dplyr::select(c(subject_id, status, sex, all_of(final_voi))) %>% 
+  filter(status != "HC")
+df_model
+
+l_hbtot_filt_brain[[10]] %>% mean(THC, na.rm = TRUE)
+
+```

### 2023-06-12 Scripts/R/combining_data.R

@@ -6,10 +6,9 @@
 # libraries ---------------------------------------------------------------
 
 
-  library(tidyverse)
-  library(readxl)
-
-
+library(tidyverse)
+library(readxl)
+library(data.table)
 
 # load data ---------------------------------------------------------------
 
@@ -80,7 +79,7 @@
   df_hbox %>% colnames()
       # fix Status and subject_id to match other dfs
       df_hbox <- df_hbox %>% 
-        rename("subject_id" = "Subject ID and Visit") %>% 
+        dplyr::rename("subject_id" = "Subject ID and Visit") %>% 
         dplyr::filter(Status %in% c("CM", "HC", "UM") & session.no == 1 | subject_id == "TM0003CM01") %>% 
         dplyr::filter(!grepl("blood", subject_id)) %>% 
         mutate(Status = factor(Status, levels = c("HC", "UM", "CM"))) %>% 
@@ -131,7 +130,6 @@
 
 
 
-
 # filter  ---------------------------------------------------------
 
 
@@ -158,11 +156,29 @@
   # in final analysis, remove any Hb_oxy < 20
   # df_master <- df_master %>% dplyr::filter(cerebral_hb_oxy > 20)
   
-  
+
+# 12 June 2023 adjustments ------------------------------------------------
+
+
+    # calculate age and remove DOB and datae of admission
+      #read_csv("Data/Hans_datatable_exports/malawi key data v04Jan2022.csv")
+      df_master <- df_master %>% 
+        mutate(age = `Study Date` - DOB) %>% 
+        dplyr::select(-c(`Calculated Age`, adm.date, DOB, `Study Date`)) %>% 
+
+        # convert to years
+        mutate(`age (days)` = str_remove(age, " days") %>% as.numeric(),
+               `age (years)` = `age (days)`/365 %>% round(2)) %>% 
+        dplyr::select(-age) %>% 
+        
+        # reorganize
+        select(c(subject_id, Status, session.no, order(colnames(.))))
+        
+      
   
 # export as csv -----------------------------------------------------------
 
-  write.csv(df_master, file = "Data/master_datatable.csv")
+  write.csv(df_master, file = "Data/master_datatable.csv", row.names = FALSE)
 
 
   

### 2023-06-12 Scripts/R/create_final_data_table.R
update creation of master data table

@@ -1,0 +1,63 @@
+
+
+library(tidyverse)
+
+# LOAD DATA
+df_hbox <- read_csv("Data/Hans_datatable_exports/malawi key data v04Jan2022.csv") %>% 
+  as_tibble() %>% 
+  janitor::clean_names() %>% 
+  mutate(status = replace(status, status == "HV", "HC")) %>% 
+  dplyr::rename("subject_id" = "subject_id_session") %>% 
+  
+  dplyr::filter(status %in% c("CM", "HC", "UM") & session_no == 1 | subject_id == "TM0003CM01") %>% 
+  dplyr::filter(!grepl("blood", subject_id)) %>% 
+  mutate(status = factor(status, levels = c("HC", "UM", "CM"))) %>% 
+  mutate_at(c("bp_dias", "bp_sys", "glucose", "lactate"), as.numeric) %>% # converts remaining character variables to numerics
+  select(subject_id, status, everything()) %>% 
+  group_by(status)
+
+# SUBSET FOR VARIABLES USED IN ANALYSIS
+voi <- c(
+  "subject_id", "status", "sex",
+  "age_calc", "temperature", "hr", "bp_sys", "bp_dias", "resp_rate", "pulse_ox", "hct", "lactate", "parasites", "glucose", "aa_arg",
+  "brain_swell",
+  "cerebral_hb_tot", "cerebral_hb_tot_alpha2", "cerebral_hb_oxy", "cerebral_hb_oxy_alpha2",
+  "muscle_hb_tot", "muscle_hb_tot_alpha2", "muscle_hb_oxy", "muscle_hb_oxy_alpha2"
+)
+
+df_final <- df_hbox %>% 
+  dplyr::select_at(voi)
+
+# EXPORT
+write.csv(df_final, file = "data_for_upload/12June2023_key_data.csv", row.names = FALSE)
+
+# QUANTIFY MISSING DATA
+
+df_missing_vals <- tibble(
+  
+  status = c("HC", "UM", "CM")
+  
+) %>% 
+  
+  mutate(NAs = 
+           
+           map(
+             
+             .x = status,
+             .f = ~ df_final %>% 
+               ungroup %>% 
+               filter(status == .x) %>% 
+               dplyr::select(-subject_id, -status) %>% 
+               is.na %>% 
+               colSums() %>% 
+               enframe
+             
+           )
+         
+  ) %>% 
+  unnest(cols = c(NAs)) %>% 
[…]

### 2023-06-12 Scripts/R/final_figures/table1.R

@@ -78,7 +78,7 @@ df_clinical_res <- 1:length(clinical_voi) %>%
   add_row(df_sex, .before = 2) 
 
 # export to CSV
-write.csv(df_clinical_res, file = "Outputs/table_1.csv")
+write.csv(df_clinical_res, file = "Outputs/table_1.csv", row.names = FALSE)
 
 
 

### 2023-06-13 Scripts/R/test_imputation_methods.R
test median vs KNN vs NA omit for data imputation in the LR model

@@ -1,0 +1,417 @@
+
+
+
+# libraries ---------------------------------------------------------------
+
+library(tidyverse)
+library(RColorBrewer)
+library(gridExtra)
+library(patchwork)
+library(Hmisc)
+library(corrr)
+library(tidymodels)
+library(themis)
+library(vip)
+library(janitor)
+library(kableExtra)
+
+
+
+# data --------------------------------------------------------------------
+
+
+df_hbox <- read_csv("Data/Hans_datatable_exports/malawi key data v04Jan2022.csv") %>% 
+  as_tibble() %>% 
+  janitor::clean_names() %>% 
+  mutate(status = replace(status, status == "HV", "HC")) %>% 
+  dplyr::rename("subject_id" = "subject_id_session") %>% 
+  
+  dplyr::filter(status %in% c("CM", "HC", "UM") & session_no == 1 | subject_id == "TM0003CM01") %>% 
+  dplyr::filter(!grepl("blood", subject_id)) %>% 
+  mutate(status = factor(status, levels = c("HC", "UM", "CM"))) %>% 
+  mutate_at(c("bp_dias", "bp_sys", "glucose", "lactate"), as.numeric) %>% # converts remaining character variables to numerics
+  select(subject_id, status, everything()) %>% 
+  group_by(status)
+
+
+
+# data format -------------------------------------------------------------
+
+
+df_model <- df_hbox %>% 
+  mutate(sex = ifelse(sex == "Female", 1, 0)) %>% 
+  filter(status != "HC") %>% 
+  mutate(status = factor(status, levels = c("UM", "CM"))) %>% 
+  
+  # transform non-normal variables (by Shapiro-Wilk) using log-transform
+  mutate(log10_cerebral_hb_tot = log10(cerebral_hb_tot),
+         log10_resp_rate = log10(resp_rate),
+         log10_lactate = log10(lactate)) %>% 
+  
+  # select variables of interest
+  dplyr::select(subject_id, status, log10_cerebral_hb_tot, cerebral_hb_tot_alpha2, hct, log10_lactate, sex)
+
+
+
+# impute with median ------------------------------------------------------
+
+set.seed(20221125)
+
+# resample
[…]
